---
title: "BioLog"
output: html_document
date: "2025-01-14"
---

# Load libraries

```{r, message=FALSE, warning=FALSE}
library(purrr)
library(dplyr)
library(ComplexHeatmap)
library(circlize)
library(matrixStats)
```

# Helpers

```{r}
source("../panAsp-GEM/code/Rutilities.R")
```

# Load data

```{r}
# Load GEM (the File must be closed)
panAsp_v2 <- read_excel_allsheets("../panAsp-GEM/model/panAsp_v2.xlsx")

# Load BioLog assay results
PM_01_wide <- read.csv("./Results_PM01.csv")
PM_02_wide <- read.csv("./Results_PM02.csv")
PM_03_wide <- read.csv("./Results_PM03.csv")
```

# Define a core and intermediate set of metabolites

Metabolic activity should be higher than threshold *t* in at least *k* 
replicates of *n* strains.

Define a 

- Core set: t = 100, k = 2, and n = 6
- Intermediate set: t = 75, k = 2, and n = 1

t could be define d from a confidence interval on the negative control, or on 
the say 10 lowest metabolic activities on a Plate (to have more data).

```{r}
csv_Files <- list.files(path = "./Raw", 
                        pattern = "\\.csv$",
                        full.names = TRUE)

raw_data <- lapply(csv_Files, function(x){
    data = read.csv(x)
    tail(data, n=1) # select the last time point
})

raw_data <- do.call(rbind, raw_data)

# Remove file location
raw_data$Plate.File <- NULL

# Update colname
colnames(raw_data)[which(colnames(raw_data) == "Plate.Type")] <- "Plate"

# Clean isolate names
raw_data$Field.2 <- gsub(" ", "_", raw_data$Field.2)

# Establish a baseline
raw_data$baseline <- NA
cols_to_consider <- grep("^([A-H][0-9]{2})$", names(raw_data), value = TRUE)

for (i in 1:nrow(raw_data)) {
    raw_data$baseline[i] <- mean(sort(unlist(raw_data[i,cols_to_consider]))[1:10])
}

raw_data$baseline
```

```
t = 150 # must be 100 units higher than the baseline
k = 2 # in at least 2 replicates
n = 2 # in at least 6 samples

raw_data$baseline <- NA
cols_to_consider <- grep("^([A-H][0-9]{2})$", names(raw_data), value = TRUE)

for (i in 1:nrow(raw_data)) {
    raw_data$baseline[i] <- mean(sort(unlist(raw_data[i,cols_to_consider]))[1:10])
}

raw_data$threshold <- raw_data$baseline + t
# Then, for each Plate, select the metabolites for which there are at least 6
# isolates with at least 2 replicates with a measurement above the threshold.

coreMetabolites <- vector(mode = "list", length = 3)
names(coreMetabolites) <- unique(raw_data$Plate)
for (i in 1:3) {
    PlateIDRes <- lapply(cols_to_consider, function(metabolite){
        
        dataSub <- raw_data[raw_data$Plate == unique(raw_data$Plate)[i],c("Field.2", metabolite,"threshold")]
        colnames(dataSub)[2] <- "measurement" 
        
        # Step 1: Identify Files that meet the condition for each measurement
        qualified_Files_measurement <- dataSub %>%
            group_by(Field.2) %>%
            summarise(
              count_above_threshold = sum(measurement > threshold),
              .groups = "drop"
            ) %>%
            filter(count_above_threshold >= k)
        
        # Step 2: Count the number of qualifying Files
        qualified_File_count <- nrow(qualified_Files_measurement)
        
        # Step 3: Store the name of the measurement column if condition is met
        if (qualified_File_count >= n) {
          result <- metabolite
        } else {
          result <- NULL
        }
        return(result)
    })
    
    PlateIDRes[sapply(PlateIDRes, is.null)] <- NULL
    PlateIDRes <- unlist(PlateIDRes)
    coreMetabolites[[i]] <- PlateIDRes
}
coreMetabolites
lengths(coreMetabolites)
```

Maybe best to rank metabolites and resolve 1 by 1 (or in batches of 10)? 
Then I know more certain that the effort I do is not wasted when we make the 
thresholds more or less strict.

For each metabolite, identify how many isolates have an observed metabolic 
activity that is (a) 100 or (b) 150 units higher than the baseline in at least
2 replicates. Than rank based on those numbers. Then walk down that ranked list.

## Assign wells to compounds

```{r}
# Function to update column names
update_colnames <- function(input_string) {
  
  # Step 1: Remove the first two dots and everything before them
  modified_string <- sub("^[^.]*\\.{2}", "", input_string)
  
  # Step 2: Remove the last dot
  modified_string <- sub("\\.$", "", modified_string)
  
  # Step 3: Replace remaining dots with underscores
  modified_string <- gsub("\\.", "_", modified_string)
  
  return(modified_string)
}

wellDf1 <- data.frame(Medium = unname(sapply(colnames(PM_01_wide), update_colnames)),
                      Plate = "PM01",
                      Well = substr(colnames(PM_01_wide),1,3))
wellDf2 <- data.frame(Medium = unname(sapply(colnames(PM_02_wide), update_colnames)),
                      Plate = "PM02A",
                      Well = substr(colnames(PM_02_wide),1,3))
wellDf3 <- data.frame(Medium = unname(sapply(colnames(PM_03_wide), update_colnames)),
                      Plate = "PM03B",
                      Well = substr(colnames(PM_03_wide),1,3))
wellDfall <- rbind(wellDf1,wellDf2,wellDf3)
wellDfall <- wellDfall[-which(wellDfall$Well == "X"),]
  
# Filtering the dataframe
# wellDfCore <- wellDfall[sapply(1:nrow(wellDfall), function(i) {
#   Plate <- wellDfall$Plate[i]
#   well <- wellDfall$Well[i]
#   well %in% coreMetabolites[[Plate]]
# }), ]
# 
# wellDfCore$Condition <- paste0(wellDfCore$Medium, "_", wellDfCore$Plate)
# wellDfCore$Medium 
```

```{r}
metaboliteDf <- data.frame(Plate = rep(c("PM01", "PM02A", "PM03B"), each = 96),
                           Well = rep(grep("^([A-H][0-9]{2})$", names(raw_data), value = TRUE),3),
                           Medium = wellDfall$Medium)
metaboliteDf$Condition <- paste0(metaboliteDf$Medium, "_", metaboliteDf$Plate)
metaboliteDf$basePlus150 <- metaboliteDf$basePlus100 <- 0

for (i in seq(nrow(metaboliteDf))) {
    
    data_i <- data.frame(isolates = raw_data$Field.2[which(raw_data$Plate == metaboliteDf$Plate[i])],
                         baselineSignal = raw_data$baseline[which(raw_data$Plate == metaboliteDf$Plate[i])],
                         metaboliteSignal = raw_data[which(raw_data$Plate == metaboliteDf$Plate[i]),
                                                      which(colnames(raw_data) == metaboliteDf$Well[i])])
    data_i$normalizedSignal <- data_i$metaboliteSignal - data_i$baselineSignal
    
    for (isolate in unique(data_i$isolates)) {
        index <- which(metaboliteDf$Plate == metaboliteDf$Plate[i] & metaboliteDf$Well == metaboliteDf$Well[i])
        if(sum(data_i$normalizedSignal[which(data_i$isolates == isolate)] > 150) >= 2){
          metaboliteDf$basePlus150[index] <- metaboliteDf$basePlus150[index] + 1
          metaboliteDf$basePlus100[index] <- metaboliteDf$basePlus100[index] + 1
        } else if(sum(data_i$normalizedSignal[which(data_i$isolates == isolate)] > 100) >= 2) {
          metaboliteDf$basePlus100[index] <- metaboliteDf$basePlus100[index] + 1
        }
    }
}
metaboliteDf
```

```{r}
# Link biolog media to oryzae metabolites
media_link = list("2__Deoxy_Adenosine" = "C05198[c]",
                  "L_Tryptophan" = "C00078[e]",
                  "Ammonia" = "C00014[e]",
                  "Capric_Acid" = "C01571[c]", # called decanoate in GEM, same according to kegg
                  "Dulcitol" = "C01697[e]", # called galactitol in GEM, same according to kegg
                  "Glyoxylic_Acid" = "C00048[c]",
                  "L_Glutamic_Acid" = "C00025[e]",
                  "Itaconic_Acid" = "C00490[e]",
                  "Acetoacetic_Acid" = "C00164[c]",
                  "L_Citrulline" = "C00327[c]",
                  "L_Homoserine" = "C00263[c]",
                  "Propionic_Acid" = "C00163[e]",
                  "L_Ornithine" = "C00077[c]",
                  "Fumaric_Acid" = "C00122[e]", 
                  "D_Mannitol" = "C00392[c]",
                  "Malonic_Acid" = "C00383[c]",
                  "D_Tagatose" = "C00795[e]",
                  "D_Raffinose" = "C00492[e]", # raffinose in GEM, not D-
                  "D_Alanine" = "C00133[c]",
                  "Histamine" = "C00388[c]",
                  "Stachyose" = "C01613[c]",
                  "Allantoin" = "C01551[c]",
                  "Xanthosine" = "C01762[c]",
                  "Xanthine" = "C00385[c]",
                  "Thymine" = "C00178[c]",
                  "Guanosine" = "C00387[c]",
                  "Cytidine" = "C00475[c]",
                  "Formamide" = "C00488[c]",
                  "Methylamine" = "C00218[c]",
                  "Biuret" = "C06555[c]",
                  "Amygdalin" = "C08325[c]",
                  "Ethanolamine" = "C00189[c]",
                  "Tyramine" = "C00483[c]",
                  "Inosine" = "C00294[c]", 
                  "Adenosine" = "C00212[c]",
                  "Maltotriose" = "C01835[c]",
                  "Uridine" = "C00299[c]",
                  "Thymidine" = "C00214[c]",
                  "D_Glucose" = "C00031[e]",
                  "D_Galactose" = "C00124[e]",
                  "D_Mannose" = "C00159[e]",
                  "L_Sorbose" = "C00247[e]",
                  "D_Arabinose" = "C00216[e]",
                  "L_Arabinose" = "C00259[e]",
                  "D_Fructose" = "C00095[e]",
                  "D_Ribose" = "C00121[e]",
                  "D_Xylose" = "C00181[e]",
                  "D_Maltose" = "C00208[e]",  # in GEM: Maltose
                  "Pectin" = "C00470[e]", # in GEM: Pectate
                  "D_Cellobiose" = "C00185[e]", # in GEM: Cellobiose
                  "Xylitol" = "C00379[e]",
                  "Citric_Acid" = "C00158[e]", # in GEM: Citrate
                  "Acetic_Acid" = "C00033[e]", # in GEM: Acetate
                  "L_Alanine" = "C00041[e]",
                  "L_Arginine" = "C00062[e]",
                  "L_Asparagine" = "C00152[e]",
                  "L_Cysteine" = "C00097[e]",
                  "L_Glutamine" = "C00064[e]",
                  "L_Histidine" = "C00135[e]",
                  "L_Isoleucine" = "C00407[e]",
                  "L_Leucine" = "C00123[e]",
                  "L_Lysine" = "C00047[e]",
                  "L_Methionine" = "C00073[e]",
                  "L_Phenylalanine" = "C00079[e]",
                  "L_Proline" = "C00148[e]",
                  "L_Serine" = "C00065[e]",
                  "L_Threonine" = "C00188[e]",
                  "L_Tyrosine" = "C00082[e]",
                  "L_Valine" = "C00183[e]",
                  "D_Trehalose" = "C01083[e]", # in GEM: alpha,alpha-Trehalose
                  "D_Sorbitol" = "C00794[e]",
                  "Glycogen" = "C00182[e]",
                  "Oxalic_Acid" = "C00209[e]", # in GEM: oxalate
                  "Adenine" = "C00147[e]",
                  "Cytosine" = "C00380[e]",
                  "Guanine" = "C00242[e]",
                  "Uracil" = "C00106[e]",
                  "Glycerol" = "C00116[e]",
                  "Glycine" = "C00037[e]",
                  "Putrescine" = "C00134[e]",
                  "Pyruvic_Acid" = "C00022[e]",
                  "Mannan" = "C00464[e]", # in GEM: Mannans
                  "Sucrose" = "C00089[e]",
                  "D_Melibiose" = "C05402[e]", # in GEM: melibiose
                  "Succinic_Acid" = "C00042[e]",
                  "Urea" = "C00086[e]",
                  "Agmatine" = "C00179[e]",
                  "Glycolic_Acid" = "C00160[c]",
                  "D_Glucosamine" = "C00329[e]",
                  "Dihydroxy_Acetone" = "C00184[c]")
```

```{r}
KEGG <- unname(media_link)[match(metaboliteDf$Medium, names(media_link))]
KEGG[sapply(KEGG, is.null)] <- NA
metaboliteDf$KEGG <- unlist(KEGG)
```

```{r}
write.csv2(metaboliteDf, "metaboliteDf.csv", row.names = FALSE)
```

```
KEGG <- unname(media_link)[match(wellDfCore$Medium, names(media_link))]
KEGG[sapply(KEGG, is.null)] <- NA
wellDfCore$KEGG <- unlist(KEGG)
wellDfCore[is.na(wellDfCore$KEGG),]
```

- L_Lyxose

- Rhamnose: not in GEM. Also no D-Rhamnose (C01684) or alpha-L-Rhamnose (C02476).
But we do have or beta-L-Rhamnose (C02338) in one reaction added from KEGG that is in the peroxisome...

- p_Hydroxy_Phenylacetic_Acid: same as 4-hydroxyphenylacetic acid. Not in GEM.
But we do have 2-hydroxyphenylacetate, and 4-hydroxyphenylacetaldehyde.

- p_Hydroxy_Phenylacetic_Acid: same as 3-hydroxyphenylacetic acid. Not in GEM.
But we do have 2-hydroxyphenylacetate, and 4-hydroxyphenylacetaldehyde.

```{r}
media_link = list("N_Acetyl_D_Glucosamine" = "C03878", # not in GEM
                  "L-Rhamnose" = "C00507", # not in GEM, see above
                  "Glycolic_Acid" = "C00160",
                  "p_Hydroxy_Phenylacetic_Acid" = "C00642", # not in GEM, see above
                  "m_Hydroxy_Phenylacetic_Acid" = "C05593", # not in GEM, see above
                  "L_Lyxose" = "C01508", # not in GEM, also 6-carbon sugar. Can go to xylulose L-Xylulose <=> L-Lyxose
                  "b_Cyclodextrin" = "C13183", # not in GEM
                  "g_Cyclodextrin" = NA,  # not in GEM
                  "Arbutin" = NA, # not in GEM
                  "2_Deoxy_D_Ribose" = "C01801", # not in GEM, we do have 2-deoxy-D-ribose 5-phosphate and 2-deoxyribose 1-phosphate
                  "b_Gentiobiose" = "C08240", # not in GEM, a C6+C6 sugar
                  "Palatinose" = "C01742", # not in GEM, a C6+C5 sugar
                  "Turanose" = "C19636", # not in GEM
                  "D_Glucosamine" = "C00329",
                  "Sorbic_Acid" = NA, # not in GEM, not in KEGG except as drug, and as cyclic parasorbic acid 	
                  "Dihydroxy_Acetone" = "C00184", # glycerone in GEM
                  "D_L_a_Amino_Caprylic_Acid" = NA, # not in GEM 
                  "Ala_Asp" = NA, # how to handle dipeptides? If there is growth, check if it grows on the individual AAs?
                  "Ala_Thr" = NA, # how to handle dipeptides? If there is growth, check if it grows on the individual AAs?
                  "Gly_Gln" = NA, # how to handle dipeptides? If there is growth, check if it grows on the individual AAs?
                  "Gly_Glu" = NA, # how to handle dipeptides? If there is growth, check if it grows on the individual AAs?
                  "Met_Ala" = NA) # how to handle dipeptides? If there is growth, check if it grows on the individual AAs?

```


# Functions for wrangling data

```{r}
PM_01_wide <- read.csv("./Results_PM01.csv")
PM_02_wide <- read.csv("./Results_PM02.csv")
PM_03_wide <- read.csv("./Results_PM03.csv")

# Function to update column names
update_colnames <- function(input_string) {
  
  # Step 1: Remove the first two dots and everything before them
  modified_string <- sub("^[^.]*\\.{2}", "", input_string)
  
  # Step 2: Remove the last dot
  modified_string <- sub("\\.$", "", modified_string)
  
  # Step 3: Replace remaining dots with underscores
  modified_string <- gsub("\\.", "_", modified_string)
  
  return(modified_string)
}

# Function to wrangle results, i.e., move from wide to long format
wrangleResults <- function(data_wide){
    
    wellID <- substr(colnames(PM_01_wide), start = 1, stop = 3)

    # split first column into Isolate and Replicate identifier
    data_wide$X <- gsub(" ", "_", data_wide$X)
    data_wide <- data_wide %>%
        separate(X, into = c("Isolate", "Replicate"), sep = "_replicate_", remove = FALSE)
    data_wide$X <- NULL

    data_long <- pivot_longer(data = data_wide, cols = c(3:98))
    return(data_long)
}
```

# Wrangle data

```{r}
# update column names
colnames(PM_01_wide) <- sapply(colnames(PM_01_wide), update_colnames)
colnames(PM_02_wide) <- sapply(colnames(PM_02_wide), update_colnames)
colnames(PM_03_wide) <- sapply(colnames(PM_03_wide), update_colnames)

PM_01_long <- wrangleResults(PM_01_wide)
PM_02_long <- wrangleResults(PM_02_wide)
PM_03_long <- wrangleResults(PM_03_wide)

PM_01_long$PlateID <- "PM_01"
PM_02_long$PlateID <- "PM_02"
PM_03_long$PlateID <- "PM_03"

PM_all <- rbind(PM_01_long, PM_02_long, PM_03_long)
colnames(PM_all)[c(3,4)] <- c("Medium", "Observed")
PM_all
```

```{r}
sum(is.na(PM_all$Observed)/length(PM_all$Observed))
```

# Add information

Add column that links the medium component to the corresponding uptake 
reaction in the GEM (`media_link` list object).

```
hlp <- unique(PM_all$Medium)  
hlp <- hlp[!hlp %in% names(media_link)]
sort(hlp)
```

Another option would be to query kegg and match the kegg ids

```
library(KEGGREST)
name_query <- tolower(hlp)
name_query <- gsub("_", "", name_query) # it works to replace commas for spaces
addKEGG <- lapply(name_query, function(x){
  query <- keggFind(database = "compound", query = x)
  return(cbind(x,
               paste0(gsub("cpd:", "", names(query)),collapse="/"),
               paste0(unname(query),collapse="/")))
})
KEGG <- sapply(addKEGG, function(x){x[2]})
for (i in seq_along(KEGG)) {
  if(length(unlist(strsplit(KEGG[[i]], "/"))) > 1){
    KEGG[[i]] <- ""
  }
}
KEGG[KEGG==""] <- NA
cbind(hlp, KEGG)

KEGG[!is.na(KEGG)][KEGG[!is.na(KEGG)] %in% gsub("\\[.\\]","",panAsp_v2$METS$`REPLACEMENT ID`)]
hlp[!is.na(KEGG)][which(KEGG[!is.na(KEGG)] %in% gsub("\\[.\\]","",panAsp_v2$METS$`REPLACEMENT ID`))]
```

```
hlp[grep("deoxy",tolower(hlp), ignore.case = TRUE)]
panAsp_v2$METS$NAME[grep("deoxyad",tolower(panAsp_v2$METS$NAME), ignore.case = TRUE)]
panAsp_v2$METS[grepl("fumara", panAsp_v2$METS$NAME, ignore.case = TRUE),
               c("ID","REPLACEMENT ID")]
```


```{r}
# Link biolog media to oryzae metabolites
media_link = list("2__Deoxy_Adenosine" = "C05198[c]",
                  "L_Tryptophan" = "C00078[e]",
                  "Ammonia" = "C00014[e]",
                  "Capric_Acid" = "C01571[c]", # called decanoate in GEM, same according to kegg
                  "Dulcitol" = "C01697[e]", # called galactitol in GEM, same according to kegg
                  "Glyoxylic_Acid" = "C00048[c]",
                  "L_Glutamic_Acid" = "C00025[e]",
                  "Itaconic_Acid" = "C00490[e]",
                  "Acetoacetic_Acid" = "C00164[c]",
                  "L_Citrulline" = "C00327[c]",
                  "L_Homoserine" = "C00263[c]",
                  "Propionic_Acid" = "C00163[e]",
                  "L_Ornithine" = "C00077[c]",
                  "Fumaric_Acid" = "C00122[e]", 
                  "D_Mannitol" = "C00392[c]",
                  "Malonic_Acid" = "C00383[c]",
                  "D_Tagatose" = "C00795[e]",
                  "D_Raffinose" = "C00492[e]", # raffinose in GEM, not D-
                  "D_Alanine" = "C00133[c]",
                  "Histamine" = "C00388[c]",
                  "Stachyose" = "C01613[c]",
                  "Allantoin" = "C01551[c]",
                  "Xanthosine" = "C01762[c]",
                  "Xanthine" = "C00385[c]",
                  "Thymine" = "C00178[c]",
                  "Guanosine" = "C00387[c]",
                  "Cytidine" = "C00475[c]",
                  "Formamide" = "C00488[c]",
                  "Methylamine" = "C00218[c]",
                  "Biuret" = "C06555[c]",
                  "Amygdalin" = "C08325[c]",
                  "Ethanolamine" = "C00189[c]",
                  "Tyramine" = "C00483[c]",
                  "Inosine" = "C00294[c]", 
                  "Adenosine" = "C00212[c]",
                  "Maltotriose" = "C01835[c]",
                  "Uridine" = "C00299[c]",
                  "Thymidine" = "C00214[c]",
                  "D_Glucose" = "C00031[e]",
                  "D_Galactose" = "C00124[e]",
                  "D_Mannose" = "C00159[e]",
                  "L_Sorbose" = "C00247[e]",
                  "D_Arabinose" = "C00216[e]",
                  "L_Arabinose" = "C00259[e]",
                  "D_Fructose" = "C00095[e]",
                  "D_Ribose" = "C00121[e]",
                  "D_Xylose" = "C00181[e]",
                  "D_Maltose" = "C00208[e]",  # in GEM: Maltose
                  "Pectin" = "C00470[e]", # in GEM: Pectate
                  "D_Cellobiose" = "C00185[e]", # in GEM: Cellobiose
                  "Xylitol" = "C00379[e]",
                  "Citric_Acid" = "C00158[e]", # in GEM: Citrate
                  "Acetic_Acid" = "C00033[e]", # in GEM: Acetate
                  "L_Alanine" = "C00041[e]",
                  "L_Arginine" = "C00062[e]",
                  "L_Asparagine" = "C00152[e]",
                  "L_Cysteine" = "C00097[e]",
                  "L_Glutamine" = "C00064[e]",
                  "L_Histidine" = "C00135[e]",
                  "L_Isoleucine" = "C00407[e]",
                  "L_Leucine" = "C00123[e]",
                  "L_Lysine" = "C00047[e]",
                  "L_Methionine" = "C00073[e]",
                  "L_Phenylalanine" = "C00079[e]",
                  "L_Proline" = "C00148[e]",
                  "L_Serine" = "C00065[e]",
                  "L_Threonine" = "C00188[e]",
                  "L_Tyrosine" = "C00082[e]",
                  "L_Valine" = "C00183[e]",
                  "D_Trehalose" = "C01083[e]", # in GEM: alpha,alpha-Trehalose
                  "D_Sorbitol" = "C00794[e]",
                  "Glycogen" = "C00182[e]",
                  "Oxalic_Acid" = "C00209[e]", # in GEM: oxalate
                  "Adenine" = "C00147[e]",
                  "Cytosine" = "C00380[e]",
                  "Guanine" = "C00242[e]",
                  "Uracil" = "C00106[e]",
                  "Glycerol" = "C00116[e]",
                  "Glycine" = "C00037[e]",
                  "Putrescine" = "C00134[e]",
                  "Pyruvic_Acid" = "C00022[e]",
                  "Mannan" = "C00464[e]", # in GEM: Mannans
                  "Sucrose" = "C00089[e]",
                  "D_Melibiose" = "C05402[e]", # in GEM: melibiose
                  "Succinic_Acid" = "C00042[e]",
                  "Urea" = "C00086[e]",
                  "Agmatine" = "C00179[e]")
```

```{r}
PM_all$Metabolite <- unname(unlist(media_link))[match(PM_all$Medium,names(media_link))]
```

# Export data

```{r}
write.csv2(PM_all, "PM_all.csv", row.names = FALSE)
```


```{r}
table(colnames(PM_01_wide)[3:97] %in% colnames(PM_02_wide)[3:97])
table(colnames(PM_01_wide)[3:97] %in% colnames(PM_03_wide)[3:97])
table(colnames(PM_02_wide)[3:97] %in% colnames(PM_03_wide)[3:97])
```


# Visualize results

Would be cool to have a visualization like a heatmap to cluster the BioLog Plate
results.

```{r}
plotData <- cbind(PM_01_wide,
                  PM_02_wide[,c(3,97)],
                  PM_03_wide[,c(3,97)])

# update column names
colnames(plotData) <- sapply(colnames(plotData), update_colnames)


# split first column into Isolate and Replicate identifier
plotData$X <- gsub(" ", "_", plotData$X)
plotData <- plotData %>%
    separate(X, into = c("Isolate", "Replicate"), sep = "_replicate_", remove = FALSE)
plotData$X <- NULL

plotData <- t(plotData)

colnames(plotData) <- paste0(plotData[1,], "_", plotData[2,])

plotData <- plotData[-1:-2,]

rn_save <- rownames(plotData)
plotData <- sapply(as.data.frame(plotData), as.logical)
plotData <- sapply(as.data.frame(plotData), as.numeric)
rownames(plotData) <- rn_save
```

```{r}
# Remove rows where all values are either 0 or NA
plotData <- plotData[!apply(plotData, 1, function(row) all(is.na(row) | row == 0)),]

# Remove rows where all values are either 1 or NA
plotData <- plotData[!apply(plotData, 1, function(row) all(is.na(row) | row == 1)),]

plotData <- plotData[!rowSums(is.na(plotData)) > 20,]
```


```{r}
col_fun <- colorRamp2(c(TRUE, FALSE), c("red", "white"))

vis <- Heatmap(plotData, col = col_fun, na_col="black",
        row_title = "Medium", column_title = "Strain",
        show_heatmap_legend = FALSE)

png("BioLog_heatmap_all.png", width=10, height=10, units = "in", res = 480)
vis
dev.off()
```















