---
title: "BioLog"
output: html_document
date: "2025-01-14"
---

# Load libraries

```{r, message=FALSE, warning=FALSE}
library(purrr)
library(dplyr)
library(ComplexHeatmap)
library(circlize)
library(matrixStats)
library(tidyr)
```

# Helpers

```{r}
# source("Rutilities.R")

# Function to update column names
update_colnames <- function(input_string) {
  
  # Step 1: Remove the first two dots and everything before them
  modified_string <- sub("^[^.]*\\.{2}", "", input_string)
  
  # Step 2: Remove the last dot
  modified_string <- sub("\\.$", "", modified_string)
  
  # Step 3: Replace remaining dots with underscores
  modified_string <- gsub("\\.", "_", modified_string)
  
  return(modified_string)
}

# Function to wrangle results, i.e., move from wide to long format
wrangleResults <- function(data_wide){
    
    wellID <- substr(colnames(PM_01_wide), start = 1, stop = 3)

    # split first column into Isolate and Replicate identifier
    data_wide$X <- gsub(" ", "_", data_wide$X)
    data_wide <- data_wide %>%
        separate(X, into = c("Isolate", "Replicate"), sep = "_replicate_", remove = FALSE)
    data_wide$X <- NULL

    data_long <- pivot_longer(data = data_wide, cols = c(3:98))
    return(data_long)
}
```

# Add information to the Biolog results file

## Load Biolog results

```{r}
# Load BioLog assay results (for each plate)
PM_01_wide <- read.csv("../data/biolog/Results_PM01.csv")
PM_02_wide <- read.csv("../data/biolog/Results_PM02.csv")
PM_03_wide <- read.csv("../data/biolog/Results_PM03.csv")
```

## Wrangle data

```{r}
# update column names
colnames(PM_01_wide) <- sapply(colnames(PM_01_wide), update_colnames)
colnames(PM_02_wide) <- sapply(colnames(PM_02_wide), update_colnames)
colnames(PM_03_wide) <- sapply(colnames(PM_03_wide), update_colnames)

PM_01_long <- wrangleResults(PM_01_wide)
PM_02_long <- wrangleResults(PM_02_wide)
PM_03_long <- wrangleResults(PM_03_wide)

PM_01_long$PlateID <- "PM_01"
PM_02_long$PlateID <- "PM_02"
PM_03_long$PlateID <- "PM_03"

PM_all <- rbind(PM_01_long, PM_02_long, PM_03_long)
colnames(PM_all)[c(3,4)] <- c("Medium", "Observed")
```

## Manually obtain KEGG IDs for Biolog metabolites

```{r}
# Link biolog media to oryzae metabolites
media_link = list("2__Deoxy_Adenosine" = "C05198[c]",
                  "L_Tryptophan" = "C00078[e]",
                  "Ammonia" = "C00014[e]",
                  "Capric_Acid" = "C01571[c]", # called decanoate in GEM, same according to kegg
                  "Dulcitol" = "C01697[e]", # called galactitol in GEM, same according to kegg
                  "Glyoxylic_Acid" = "C00048[c]",
                  "L_Glutamic_Acid" = "C00025[e]",
                  "Itaconic_Acid" = "C00490[e]",
                  "Acetoacetic_Acid" = "C00164[c]",
                  "L_Citrulline" = "C00327[c]",
                  "L_Homoserine" = "C00263[c]",
                  "Propionic_Acid" = "C00163[e]",
                  "L_Ornithine" = "C00077[c]",
                  "Fumaric_Acid" = "C00122[e]", 
                  "D_Mannitol" = "C00392[c]",
                  "Malonic_Acid" = "C00383[c]",
                  "D_Tagatose" = "C00795[e]",
                  "D_Raffinose" = "C00492[e]", # raffinose in GEM, not D-
                  "D_Alanine" = "C00133[c]",
                  "Histamine" = "C00388[c]",
                  "Stachyose" = "C01613[c]",
                  "Allantoin" = "C01551[c]",
                  "Xanthosine" = "C01762[c]",
                  "Xanthine" = "C00385[c]",
                  "Thymine" = "C00178[c]",
                  "Guanosine" = "C00387[c]",
                  "Cytidine" = "C00475[c]",
                  "Formamide" = "C00488[c]",
                  "Methylamine" = "C00218[c]",
                  "Biuret" = "C06555[c]",
                  "Amygdalin" = "C08325[c]",
                  "Ethanolamine" = "C00189[c]",
                  "Tyramine" = "C00483[c]",
                  "Inosine" = "C00294[c]", 
                  "Adenosine" = "C00212[c]",
                  "Maltotriose" = "C01835[c]",
                  "Uridine" = "C00299[c]",
                  "Thymidine" = "C00214[c]",
                  "D_Glucose" = "C00031[e]",
                  "D_Galactose" = "C00124[e]",
                  "D_Mannose" = "C00159[e]",
                  "L_Sorbose" = "C00247[e]",
                  "D_Arabinose" = "C00216[e]",
                  "L_Arabinose" = "C00259[e]",
                  "D_Fructose" = "C00095[e]",
                  "D_Ribose" = "C00121[e]",
                  "D_Xylose" = "C00181[e]",
                  "D_Maltose" = "C00208[e]",  # in GEM: Maltose
                  "Pectin" = "C00470[e]", # in GEM: Pectate
                  "D_Cellobiose" = "C00185[e]", # in GEM: Cellobiose
                  "Xylitol" = "C00379[e]",
                  "Citric_Acid" = "C00158[e]", # in GEM: Citrate
                  "Acetic_Acid" = "C00033[e]", # in GEM: Acetate
                  "L_Alanine" = "C00041[e]",
                  "L_Arginine" = "C00062[e]",
                  "L_Asparagine" = "C00152[e]",
                  "L_Cysteine" = "C00097[e]",
                  "L_Glutamine" = "C00064[e]",
                  "L_Histidine" = "C00135[e]",
                  "L_Isoleucine" = "C00407[e]",
                  "L_Leucine" = "C00123[e]",
                  "L_Lysine" = "C00047[e]",
                  "L_Methionine" = "C00073[e]",
                  "L_Phenylalanine" = "C00079[e]",
                  "L_Proline" = "C00148[e]",
                  "L_Serine" = "C00065[e]",
                  "L_Threonine" = "C00188[e]",
                  "L_Tyrosine" = "C00082[e]",
                  "L_Valine" = "C00183[e]",
                  "D_Trehalose" = "C01083[e]", # in GEM: alpha,alpha-Trehalose
                  "D_Sorbitol" = "C00794[e]",
                  "Glycogen" = "C00182[e]",
                  "Oxalic_Acid" = "C00209[e]", # in GEM: oxalate
                  "Adenine" = "C00147[e]",
                  "Cytosine" = "C00380[e]",
                  "Guanine" = "C00242[e]",
                  "Uracil" = "C00106[e]",
                  "Glycerol" = "C00116[e]",
                  "Glycine" = "C00037[e]",
                  "Putrescine" = "C00134[e]",
                  "Pyruvic_Acid" = "C00022[e]",
                  "Mannan" = "C00464[e]", # in GEM: Mannans
                  "Sucrose" = "C00089[e]",
                  "D_Melibiose" = "C05402[e]", # in GEM: melibiose
                  "Succinic_Acid" = "C00042[e]",
                  "Urea" = "C00086[e]",
                  "Agmatine" = "C00179[e]",
                  "Glycolic_Acid" = "C00160[c]",
                  "D_Glucosamine" = "C00329[e]",
                  "Dihydroxy_Acetone" = "C00184[c]")

PM_all$Metabolite <- unname(unlist(media_link))[match(PM_all$Medium,names(media_link))]
```

## Export data

```{r}
write.csv2(PM_all, "../data/biolog/PM_all.csv", row.names = FALSE)
```

# Define a core set of metabolites

For each metabolite, identify how many isolates have an observed metabolic 
activity that is (a) 100 or (b) 150 units higher than the baseline in at least
2 replicates. Than rank based on those numbers. Then walk down that ranked list.
Currently, I identify a core metabolite as a metabolite where there are at least
two isolates that have two replicates with a metabolic activity that is at least 
150 units higher than their baselines, or where where there are at least
four isolates that have two replicates with a metabolic activity that is at least 
100 units higher than their baaselines

## Read and wrangle Biolog raw data

```{r}
csv_Files <- list.files(path = "../data/biolog/Raw", 
                        pattern = "\\.csv$",
                        full.names = TRUE)

raw_data <- lapply(csv_Files, function(x){
    data = read.csv(x)
    tail(data, n=1) # select the last time point
})
raw_data <- do.call(rbind, raw_data)

# Remove file location
raw_data$Plate.File <- NULL

# Update colname
colnames(raw_data)[which(colnames(raw_data) == "Plate.Type")] <- "Plate"

# Clean isolate names
raw_data$Field.2 <- gsub(" ", "_", raw_data$Field.2)

# Establish a baseline
raw_data$baseline <- NA
cols_to_consider <- grep("^([A-H][0-9]{2})$", names(raw_data), value = TRUE)
for (i in 1:nrow(raw_data)) {
    raw_data$baseline[i] <- mean(sort(unlist(raw_data[i,cols_to_consider]))[1:10])
}
```

## Assign wells to compounds

```{r}
wellDf1 <- data.frame(Medium = unname(sapply(colnames(PM_01_wide), update_colnames)),
                      Plate = "PM01",
                      Well = substr(colnames(PM_01_wide),1,3))
wellDf2 <- data.frame(Medium = unname(sapply(colnames(PM_02_wide), update_colnames)),
                      Plate = "PM02A",
                      Well = substr(colnames(PM_02_wide),1,3))
wellDf3 <- data.frame(Medium = unname(sapply(colnames(PM_03_wide), update_colnames)),
                      Plate = "PM03B",
                      Well = substr(colnames(PM_03_wide),1,3))
wellDfall <- rbind(wellDf1,wellDf2,wellDf3)
wellDfall <- wellDfall[-which(wellDfall$Well == "X"),]
```

## Create a dataframe ranking metabolites according to metabolic activity

```{r}
metaboliteDf <- data.frame(Plate = rep(c("PM01", "PM02A", "PM03B"), each = 96),
                           Well = rep(grep("^([A-H][0-9]{2})$", names(raw_data), value = TRUE),3),
                           Medium = wellDfall$Medium)
metaboliteDf$Condition <- paste0(metaboliteDf$Medium, "_", metaboliteDf$Plate)
metaboliteDf$basePlus150 <- metaboliteDf$basePlus100 <- 0

for (i in seq(nrow(metaboliteDf))) {
    
    data_i <- data.frame(isolates = raw_data$Field.2[which(raw_data$Plate == metaboliteDf$Plate[i])],
                         baselineSignal = raw_data$baseline[which(raw_data$Plate == metaboliteDf$Plate[i])],
                         metaboliteSignal = raw_data[which(raw_data$Plate == metaboliteDf$Plate[i]),
                                                      which(colnames(raw_data) == metaboliteDf$Well[i])])
    data_i$normalizedSignal <- data_i$metaboliteSignal - data_i$baselineSignal
    
    for (isolate in unique(data_i$isolates)) {
        index <- which(metaboliteDf$Plate == metaboliteDf$Plate[i] & metaboliteDf$Well == metaboliteDf$Well[i])
        if(sum(data_i$normalizedSignal[which(data_i$isolates == isolate)] > 150) >= 2){
          metaboliteDf$basePlus150[index] <- metaboliteDf$basePlus150[index] + 1
          metaboliteDf$basePlus100[index] <- metaboliteDf$basePlus100[index] + 1
        } else if(sum(data_i$normalizedSignal[which(data_i$isolates == isolate)] > 100) >= 2) {
          metaboliteDf$basePlus100[index] <- metaboliteDf$basePlus100[index] + 1
        }
    }
}
head(metaboliteDf)
```


```{r}
KEGG <- unname(media_link)[match(metaboliteDf$Medium, names(media_link))]
KEGG[sapply(KEGG, is.null)] <- NA
metaboliteDf$KEGG <- unlist(KEGG)
```

```{r}
write.csv2(metaboliteDf, "../data/biolog/metaboliteDf.csv", row.names = FALSE)
```

- L_Lyxose

- Rhamnose: not in GEM. Also no D-Rhamnose (C01684) or alpha-L-Rhamnose (C02476).
But we do have or beta-L-Rhamnose (C02338) in one reaction added from KEGG that is in the peroxisome...

- p_Hydroxy_Phenylacetic_Acid: same as 4-hydroxyphenylacetic acid. Not in GEM.
But we do have 2-hydroxyphenylacetate, and 4-hydroxyphenylacetaldehyde.

- p_Hydroxy_Phenylacetic_Acid: same as 3-hydroxyphenylacetic acid. Not in GEM.
But we do have 2-hydroxyphenylacetate, and 4-hydroxyphenylacetaldehyde.

```{r}
media_link = list("N_Acetyl_D_Glucosamine" = "C03878", # not in GEM
                  "L-Rhamnose" = "C00507", # not in GEM, see above
                  "Glycolic_Acid" = "C00160",
                  "p_Hydroxy_Phenylacetic_Acid" = "C00642", # not in GEM, see above
                  "m_Hydroxy_Phenylacetic_Acid" = "C05593", # not in GEM, see above
                  "L_Lyxose" = "C01508", # not in GEM, also 6-carbon sugar. Can go to xylulose L-Xylulose <=> L-Lyxose
                  "b_Cyclodextrin" = "C13183", # not in GEM
                  "g_Cyclodextrin" = NA,  # not in GEM
                  "Arbutin" = NA, # not in GEM
                  "2_Deoxy_D_Ribose" = "C01801", # not in GEM, we do have 2-deoxy-D-ribose 5-phosphate and 2-deoxyribose 1-phosphate
                  "b_Gentiobiose" = "C08240", # not in GEM, a C6+C6 sugar
                  "Palatinose" = "C01742", # not in GEM, a C6+C5 sugar
                  "Turanose" = "C19636", # not in GEM
                  "D_Glucosamine" = "C00329",
                  "Sorbic_Acid" = NA, # not in GEM, not in KEGG except as drug, and as cyclic parasorbic acid 	
                  "Dihydroxy_Acetone" = "C00184", # glycerone in GEM
                  "D_L_a_Amino_Caprylic_Acid" = NA, # not in GEM 
                  "Ala_Asp" = NA, # how to handle dipeptides? If there is growth, check if it grows on the individual AAs?
                  "Ala_Thr" = NA, # how to handle dipeptides? If there is growth, check if it grows on the individual AAs?
                  "Gly_Gln" = NA, # how to handle dipeptides? If there is growth, check if it grows on the individual AAs?
                  "Gly_Glu" = NA, # how to handle dipeptides? If there is growth, check if it grows on the individual AAs?
                  "Met_Ala" = NA) # how to handle dipeptides? If there is growth, check if it grows on the individual AAs?

```

# Visualize Biolog results

Would be cool to have a visualization like a heatmap to cluster the BioLog Plate
results.

```{r}
plotData <- cbind(PM_01_wide,
                  PM_02_wide[,c(3,97)],
                  PM_03_wide[,c(3,97)])

# update column names
colnames(plotData) <- sapply(colnames(plotData), update_colnames)

# split first column into Isolate and Replicate identifier
plotData$X <- gsub(" ", "_", plotData$X)
plotData <- plotData %>%
    separate(X, into = c("Isolate", "Replicate"), sep = "_replicate_", remove = FALSE)
plotData$X <- NULL

plotData <- t(plotData)

colnames(plotData) <- paste0(plotData[1,], "_", plotData[2,])

plotData <- plotData[-1:-2,]

rn_save <- rownames(plotData)
plotData <- sapply(as.data.frame(plotData), as.logical)
plotData <- sapply(as.data.frame(plotData), as.numeric)
rownames(plotData) <- rn_save
```

```{r}
# Remove rows where all values are either 0 or NA
plotData <- plotData[!apply(plotData, 1, function(row) all(is.na(row) | row == 0)),]

# Remove rows where all values are either 1 or NA
plotData <- plotData[!apply(plotData, 1, function(row) all(is.na(row) | row == 1)),]

plotData <- plotData[!rowSums(is.na(plotData)) > 20,]
```

```{r}
col_fun <- colorRamp2(c(TRUE, FALSE), c("red", "white"))

vis <- Heatmap(plotData, col = col_fun, na_col="black",
        row_title = "Medium", column_title = "Strain",
        show_heatmap_legend = FALSE)

png("../data/biolog/BioLog_heatmap_sub.png", width=10, height=10, units = "in", res = 480)
vis
dev.off()
```















