---
title: "Biolog_target"
output: html_document
date: "2025-02-13"
---

Script to compare the BioLog results with our FBA predictions

# Load libraries

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
```

# Load data

```{r}
PM_all <- read.csv2("PM_all.csv")
PM_all$plateID[which(PM_all$plateID == "PM_01")] <- "PM01"
PM_all$plateID[which(PM_all$plateID == "PM_02")] <- "PM02A"
PM_all$plateID[which(PM_all$plateID == "PM_03")] <- "PM03B"


metaboliteDf <- read.csv2("metaboliteDf.csv")

#FBA_res <- read.csv2("../../medusa/growth_df_1000_sink_cofactor_noATPbound.csv", sep=",")
#FBA_res_oxy <- read.csv2("../../medusa/oxygen_df_1000_sink_cofactor_noATPbound.csv", sep=",")

# FBA_res <- read.csv2("../../medusa/growth_target.csv", sep=",")
# FBA_res_oxy <- read.csv2("../../medusa/oxygen_target.csv", sep=",")
# sink <- read.csv2("../../medusa/sink_target.csv", sep=",")
# sink$X <- NULL
```

```{r}
FBA_res <- read.csv2("../medusa/simulation_results.csv", sep=",")
FBA_res$X <- NULL
FBA_res
```

# Wrangle data

Add results of FBA analysis (growth) to Biolog plate data

```{r}
FBA_res$Isolate <- gsub("Aspergillus_oryzae_", "", FBA_res$Isolate)
FBA_res$Isolate <- gsub("_CAoGD", "", FBA_res$Isolate)

res <- PM_all %>% inner_join(FBA_res, 
        by=c('Isolate'='Isolate', 'Medium'='Medium', 'plateID'='plateID'))
res$Growth <- as.numeric(res$Growth)
res$Growth_bool <- ifelse(res$Growth > 1e-3, TRUE, FALSE)
res$ATP <- as.numeric(res$ATP)
res$ATP_bool <- ifelse(res$ATP > 1e-3, TRUE, FALSE)
res$Oxygen_growth <- as.numeric(res$Oxygen_growth)
res$Oxygen_growth_bool <- ifelse(res$Oxygen_growth > 1e-3, TRUE, FALSE)
res$Oxygen_ATP <- as.numeric(res$Oxygen_ATP)
res$Oxygen_ATP_bool <- ifelse(res$Oxygen_ATP > 1e-3, TRUE, FALSE)

# res$Condition <- paste0(res$Medium, "_", res$plateID)
# res$Sink <- sink$Location[match(res$Medium, sink$Medium)]
res$Well <- metaboliteDf$Well[match(res$Medium, metaboliteDf$Medium)]

res <- res[,c("Isolate", "Replicate", "Metabolite", "plateID", "Well", "Medium", 
              "Condition", "Location_growth", "Location_ATP", "Oxygen_growth", "Oxygen_growth_bool", 
              "Oxygen_ATP", "Oxygen_ATP_bool", "Growth","ATP", "Growth_bool", 
              "ATP_bool", "Observed")]
```

```
library(openxlsx)
write.xlsx(res[which(res$Isolate == "NRRL_2217" & res$Replicate == 1),
               c("Medium", "plateID", "Well", "Metabolite")],
           "metabolites_supporting_growth.xlsx")
```

Some minor issues

- issue: Sorbic acid


```{r}
res[which(res$Medium == "Sorbic_Acid"),
    c("Isolate", "Replicate", "Location_growth", "Oxygen_growth", "Oxygen_ATP", "Growth", "ATP", 
      "Observed")]
```

# Identify metabolites with prediction differences

And see if they match the plate data.

```{r}
for (med in unique(res$Medium)){
  print(res[which(res$Medium == med & res$Replicate == 1),
            c("Isolate", "Medium", "plateID", "Well", "Growth", "ATP", "Observed")])
}
```

Note: sorbic acid but also xantine (!) failed
 
- RIB40 predictions are consistently lower.
--> not corresponding


- p_Hydroxy_Phenylacetic_Acid, m_Hydroxy_Phenylacetic_Acid, Tyramine: higher in NRRL_3483 
--> not corresponding, NRRL_3483 is actually quite low. 
Also, NRRL_5589 shouldnt grow on it, adapt if possible!


- Arbutin, L_Norvaline, L_Glutamic_Acid, and L_Tryptophan: higher in NRRL_5592
--> not corresponding
Also, L_Glutamic_Acid should be Well A12, not B12. Where does this mistake come from?


- L-proline: only in NRRL_2217 and NRRL_3483
--> not corresponding, there seems to be growth everywhere, and growth is highest in RIB40, NRRL_35890


- myo_Inositol: only in NRRL_2217, NRRL_35890, and NRRL_5589
--> not corresponding, the worst growers on the plate are NRRL_5592 and NRRL_5589


- b_Phenylethylamine: only in NRRL_2217, NRRL_3488, NRRL_35890, and NRRL_5589
--> Worst are NRRL_3483, RIB40, and NRRL_5589


- L_Ornithine: only in NRRL_2217, NRRL_3483, NRRL_471, NRRL_35890, NRRL_5589, and RIB40
--> Corresponding, worst are NRRL_5589 and NRRL_3488. However, NRRL_3483 is pretty bad too.











