---
title: "Compartments"
output: html_document
date: "2024-10-03"
---

```{r, message=FALSE, warning=FALSE}
library(stringr)
library(openxlsx)
library("kableExtra")
```

```{r}
read_excel_allsheets <- function(filename) {
    sheets <- openxlsx::getSheetNames(filename)
    x <- lapply(sheets, function(X) openxlsx::read.xlsx(filename, sheet = X))
    names(x) <- sheets
    x
}
```

# Load gems

```{r}
# Note: excels should be closed when reading them in R
niger_gem <- read_excel_allsheets("../GEMs_orig/AspNiger/40694_2018_60_MOESM2_ESM.xlsx")
fumigatus_gem <- read_excel_allsheets("../GEMs_orig/AspFumigatus/Pan_Aspergillus_fumigatus.xlsx")
oryzae_gem <- read_excel_allsheets("../GEMs_orig/AspOryzae/iWV1314.xlsx")
```

# Establishing a link between compartments

```{r}
fumigatus_gem$COMPS
```

```{r}
oryzae_gem$COMPS
```

```{r}
niger_gem$COMPS
```

All GEMs have:

1. Extracellular
2. Cytoplasm (=cytosol)
3. Mitochonfria
4. Peroxisome

We can remove boundary from Oryzae GEM.

Finally, some GEMs additionally have:

1. Vacuole
2. Nucleus
3. Golgi
4. Lipid particle (from Golgi)
5. Endoplasmic reticulum

Perhaps it is best to lump these into cytoplasms, because if we'd keep them,
then this implies that the GEMs lacking these compartments have no reactions
there.



