---
title: "Consolidating compartments across GEMs"
author: "Jeroen Gilis"
date: "19 October 2024"
output: 
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

TODO: I tried to do this with the RAVEN toolbox in MATLAB first, but I did not
find a functionality that allows for changing compartment names, e.g., changing
cytoplasm to c. I tried doing this with the `copyToComps`

```{r, message=FALSE, warning=FALSE}
library(stringr)
library(openxlsx)
library("kableExtra")
```

```{r}
source("./Rutilities.R")
```

# Load gems

```{r}
# Note: excels should be closed when reading them in R
niger_gem <- read_excel_allsheets("../model/Ani_draft.xlsx")
fumigatus_gem <- read_excel_allsheets("../model/Afu_draft.xlsx")
oryzae_gem <- read_excel_allsheets("../model/Aor_draft.xlsx")
```

# Establishing a link between compartments

```{r}
fumigatus_gem$COMPS
```

```{r}
oryzae_gem$COMPS
```

```{r}
niger_gem$COMPS
```

All GEMs have:

1. Extracellular
2. Cytoplasm (=cytosol)
3. Mitochonfria
4. Peroxisome

We can remove boundary from Oryzae GEM.

Finally, some GEMs additionally have:

1. Vacuole
2. Nucleus
3. Golgi
4. Lipid particle (from Golgi)
5. Endoplasmic reticulum

Perhaps it is best to lump these into cytoplasms, because if we'd keep them,
then this implies that the GEMs lacking these compartments have no reactions
there.

Also, the different GEMs are very inconsistent in how they report metabolites,
including how and where they include compartment info in the metabolites. This
will need to be resolved as well to properly integrate the GEMs.

To some extent, this should be done in the script where we consolidate the
metabolites. But here, I will already take care of the compartimentalization
bit. I propose the following:

1. ID is the name of the metabolite including compartment as [c]
2. NAME is the name of the metabolite without compartment
3. REPLACEMENT.ID is the final link between the different GEMs. This will 
probably turn out to be a KEGG or a CHEBI or a name that is consisent across 
GEMs. For now, I just keep the current REPLACEMENT.ID, but with a compartment
tag [c].

# Niger

For Aspergillus niger, we need combine vacuole, golgi, and nucleus into
cytoplasm.

## METS

```{r}
# replacement.id and compartment
niger_gem$METS$`REPLACEMENT ID`[!niger_gem$METS$COMPARTMENT%in%c("c","n")] <- substr(niger_gem$METS$`REPLACEMENT ID`[!niger_gem$METS$COMPARTMENT%in%c("c","n")],1,nchar(niger_gem$METS$`REPLACEMENT ID`[!niger_gem$METS$COMPARTMENT%in%c("c","n")])-1)

niger_gem$METS[niger_gem$METS$COMPARTMENT%in%c("v","n","g"),"COMPARTMENT"] <- "c"
niger_gem$METS$`REPLACEMENT ID` <- paste0(niger_gem$METS$`REPLACEMENT ID`, "[",
                                        niger_gem$METS$COMPARTMENT, "]")

# name
niger_gem$METS$NAME <- gsub(" \\(extracellular\\)","",niger_gem$METS$NAME)
niger_gem$METS$NAME <- gsub(" \\(mitochondrial\\)","",niger_gem$METS$NAME)

# id
niger_gem$METS$ID <- gsub(" \\(extracellular\\)","",niger_gem$METS$ID)
niger_gem$METS$ID <- gsub(" \\(mitochondrial\\)","",niger_gem$METS$ID)

niger_gem$METS$ID <- gsub("\\[n\\]","\\[c\\]",niger_gem$METS$ID)
niger_gem$METS$ID <- gsub("\\[g\\]","\\[c\\]",niger_gem$METS$ID)
niger_gem$METS$ID <- gsub("\\[v\\]","\\[c\\]",niger_gem$METS$ID)
```

## RXNS

```{r}
niger_gem$RXNS$EQUATION <- gsub(" \\(extracellular\\)","",niger_gem$RXNS$EQUATION)
niger_gem$RXNS$EQUATION <- gsub(" \\(mitochondrial\\)","",niger_gem$RXNS$EQUATION)

niger_gem$RXNS$EQUATION <- gsub("\\[g\\]","\\[c\\]",niger_gem$RXNS$EQUATION)
niger_gem$RXNS$EQUATION <- gsub("\\[n\\]","\\[c\\]",niger_gem$RXNS$EQUATION)
niger_gem$RXNS$EQUATION <- gsub("\\[v\\]","\\[c\\]",niger_gem$RXNS$EQUATION)
```

It looks like I mismatched RXNS$EQUATION and METS$ID. Not all mets from the
equations are in my metabolite list, apparently.

# COMPS

```{r}
niger_gem$COMPS <- niger_gem$COMPS[niger_gem$COMPS$ABBREVIATION %in% c("p","c","m","e"),]
rownames(niger_gem$COMPS) <- 1:4
niger_gem$COMPS
```

# Fumigatus

## METS

```{r}
fumigatus_gem$METS$ID <- gsub(" \\[cytoplasm\\]\\[cytoplasm\\]","[c]",fumigatus_gem$METS$ID)
fumigatus_gem$METS$ID <- gsub(" \\[mitochondrion\\]\\[mitochondrion\\]","[m]",fumigatus_gem$METS$ID)
fumigatus_gem$METS$ID <- gsub(" \\[extracellular\\]\\[extracellular\\]","[e]",fumigatus_gem$METS$ID)
fumigatus_gem$METS$ID <- gsub(" \\[peroxisome\\]\\[peroxisome\\]","[p]",fumigatus_gem$METS$ID)
fumigatus_gem$METS$ID <- gsub(" \\[nucleus\\]\\[nucleus\\]","[c]",fumigatus_gem$METS$ID)
fumigatus_gem$METS$ID <- gsub(" \\[vacuole\\]\\[vacuole\\]","[c]",fumigatus_gem$METS$ID)
fumigatus_gem$METS$ID <- gsub(" \\[endoplasmic_reticulum\\]\\[endoplasmic_reticulum\\]","[c]",fumigatus_gem$METS$ID)
fumigatus_gem$METS$ID <- gsub(" \\[golgi\\]\\[golgi\\]","[c]",fumigatus_gem$METS$ID)
fumigatus_gem$METS$ID <- gsub(" \\[lipid_particle\\]\\[lipid_particle\\]","[c]",fumigatus_gem$METS$ID)

fumigatus_gem$METS$NAME <- gsub(" \\[cytoplasm\\]","",fumigatus_gem$METS$NAME)
fumigatus_gem$METS$NAME <- gsub(" \\[mitochondrion\\]","",fumigatus_gem$METS$NAME)
fumigatus_gem$METS$NAME <- gsub(" \\[extracellular\\]","",fumigatus_gem$METS$NAME)
fumigatus_gem$METS$NAME <- gsub(" \\[peroxisome\\]","",fumigatus_gem$METS$NAME)
fumigatus_gem$METS$NAME <- gsub(" \\[nucleus\\]","",fumigatus_gem$METS$NAME)
fumigatus_gem$METS$NAME <- gsub(" \\[vacuole\\]","",fumigatus_gem$METS$NAME)
fumigatus_gem$METS$NAME <- gsub(" \\[endoplasmic_reticulum\\]","",fumigatus_gem$METS$NAME)
fumigatus_gem$METS$NAME <- gsub(" \\[golgi\\]","",fumigatus_gem$METS$NAME)
fumigatus_gem$METS$NAME <- gsub(" \\[lipid_particle\\]","",fumigatus_gem$METS$NAME)

hlp <- cbind(c("cytoplasm", "mitochondrion", "extracellular", 
               "endoplasmic_reticulum", "nucleus", "peroxisome", "vacuole",
               "lipid_particle", "golgi"),
             c("c","m","e","c","c","p","c","c","c"))
fumigatus_gem$METS$COMPARTMENT <- hlp[match(fumigatus_gem$METS$COMPARTMENT,hlp[,1]),2]
```

## RXNS

```{r}
fumigatus_gem$RXNS$EQUATION <- gsub(" \\[cytoplasm\\]\\[cytoplasm\\]","[c]",fumigatus_gem$RXNS$EQUATION)
fumigatus_gem$RXNS$EQUATION <- gsub(" \\[mitochondrion\\]\\[mitochondrion\\]","[m]",fumigatus_gem$RXNS$EQUATION)
fumigatus_gem$RXNS$EQUATION <- gsub(" \\[extracellular\\]\\[extracellular\\]","[e]",fumigatus_gem$RXNS$EQUATION)
fumigatus_gem$RXNS$EQUATION <- gsub(" \\[peroxisome\\]\\[peroxisome\\]","[p]",fumigatus_gem$RXNS$EQUATION)
fumigatus_gem$RXNS$EQUATION <- gsub(" \\[nucleus\\]\\[nucleus\\]","[c]",fumigatus_gem$RXNS$EQUATION)
fumigatus_gem$RXNS$EQUATION <- gsub(" \\[vacuole\\]\\[vacuole\\]","[c]",fumigatus_gem$RXNS$EQUATION)
fumigatus_gem$RXNS$EQUATION <- gsub(" \\[endoplasmic_reticulum\\]\\[endoplasmic_reticulum\\]","[c]",fumigatus_gem$RXNS$EQUATION)
fumigatus_gem$RXNS$EQUATION <- gsub(" \\[golgi\\]\\[golgi\\]","[c]",fumigatus_gem$RXNS$EQUATION)
fumigatus_gem$RXNS$EQUATION <- gsub(" \\[lipid_particle\\]\\[lipid_particle\\]","[c]",fumigatus_gem$RXNS$EQUATION)
```

## COMPS

```{r}
fumigatus_gem$COMPS <- fumigatus_gem$COMPS[c(1,2,3,6),]
rownames(fumigatus_gem$COMPS) <- 1:4
fumigatus_gem$COMPS$ABBREVIATION <- c("c","m","e","p")
fumigatus_gem$COMPS$NAME <- c("Cytoplasm","Mitochondrion","Extracellular","Peroxisome")
```

# Oryzae

## METS

```{r}
oryzae_gem$METS$`REPLACEMENT ID` <- paste0(oryzae_gem$METS$`REPLACEMENT ID`, "[",
                                         oryzae_gem$METS$COMPARTMENT, "]")
```

## RXNS

```{r}
head(oryzae_gem$RXNS[oryzae_gem$RXNS$COMPARTMENT=="b",])
```

I will set these to "e"

```{r}
oryzae_gem$RXNS$COMPARTMENT[oryzae_gem$RXNS$COMPARTMENT=="b"] <- "e"
```

## COMPS

```{r}
oryzae_gem$COMPS <- oryzae_gem$COMPS[1:4,]
oryzae_gem$COMPS$NAME <- c("Peroxisome","Mitochondrion","Cytoplasm","Extracellular")
```

# Export

```{r}
# Note: excels should be closed when reading them in R
write.xlsx(niger_gem, "../model/Ani_draft.xlsx",sep.names = " ")
write.xlsx(fumigatus_gem, "../model/Afu_draft.xlsx",sep.names = " ")
write.xlsx(oryzae_gem, "../model/Aor_draft.xlsx",sep.names = " ")
```





















