---
title: "Curating panAsp GEM V1"
author: "Jeroen Gilis"
date: "6 January 2025"
output: 
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

# Libraries

```{r, message=FALSE, warning=FALSE}
library(openxlsx)
```

# Helpers

```{r}
source("../code/Rutilities.R")
```

# Load data

```{r}
# Note: excel files should be closed when reading them in R
Ani_idtokegg <- read.xlsx("../scrap/niger_metabolite_table.xlsx")

Ani_draft <- read_excel_allsheets("../data/intermediate/Ani_draft_comp.xlsx")
Aor_draft <- read_excel_allsheets("../data/intermediate/Aor_draft_comp.xlsx")

Ani_reduced <- read_excel_allsheets("..//data/intermediate/Ani_reduced.xlsx")
Afu_reduced <- read_excel_allsheets("..//data/intermediate/Afu_reduced.xlsx")

OtherAsp <- read_excel_allsheets("../data/intermediate/OtherAsp_10_50_loc.xlsx")
panAsp_v1 <- read_excel_allsheets("../data/intermediate/panAsp_v1.xlsx")
panAsp_v2 <- read_excel_allsheets("../model/panAsp_v2.xlsx")
```

# OtherAsp metabolites

I am surprised to see that when the OtherAsp reactions are added to the oryzae
GEM, 892 novel metabolites (with localization) are created: this seems like a 
lot.

```{r}
# Step 1:
Aor_kegg <- read.xlsx("../data/intermediate/oryzae_kegg.xlsx"); # obtained using R script metabolites_oryzae.Rmd
all(Aor_draft$METS$NAME == Aor_kegg$NAME) # same order

Aor_draft$METS$`REPLACEMENT ID`[!is.na(Aor_kegg$KEGG)] = Aor_kegg$KEGG[!is.na(Aor_kegg$KEGG)]
# if a kegg ID was found, use it. Otherwise, retain original ID
# paste compartment to it
Aor_draft$METS$`REPLACEMENT ID`[!is.na(Aor_kegg$KEGG)] <- paste0(Aor_draft$METS$`REPLACEMENT ID`[!is.na(Aor_kegg$KEGG)], 
                                                                 '[', 
                                                                 Aor_draft$METS$COMPARTMENT[!is.na(Aor_kegg$KEGG)], 
                                                                 ']');

# Step 2:
OtherAsp$METS$`REPLACEMENT ID` <- gsub("_Extracellular|_Mitochondrion|_Peroxisome",
                                       "", OtherAsp$METS$`REPLACEMENT ID`)
OtherAsp$METS$COMPARTMENT <- gsub("Peroxisome", "p", OtherAsp$METS$COMPARTMENT)
OtherAsp$METS$COMPARTMENT <- gsub("Extracellular", "e", OtherAsp$METS$COMPARTMENT)
OtherAsp$METS$COMPARTMENT <- gsub("Mitochondrion", "m", OtherAsp$METS$COMPARTMENT)

OtherAsp$METS$`REPLACEMENT ID` <- paste0(OtherAsp$METS$`REPLACEMENT ID`, "[",
                                         OtherAsp$METS$COMPARTMENT, "]")
```

```{r}
table(OtherAsp$METS$`REPLACEMENT ID` %in% Aor_draft$METS$`REPLACEMENT ID`)

OtherAsp$METS$RepName <- gsub("\\[.\\]", "", OtherAsp$METS$`REPLACEMENT ID`)
Aor_draft$METS$RepName <- gsub("\\[.\\]", "", Aor_draft$METS$`REPLACEMENT ID`)

table(OtherAsp$METS$RepName %in% Aor_draft$METS$RepName) 
# so 430-307 = 123 compounds that are in OtherAsp also exist in oryzae,
# but in a different compartment
```

```{r}
OtherAsp$METS[!OtherAsp$METS$RepName %in% Aor_draft$METS$RepName,]
```

So at least some of these compounds probably do exist already in oryzae, but
e.g. different resolution (peptide) or different stereochemistry 
(D-Glucose 6-phosphate).

# Niger metabolites

```{r}
# Step 1:
Aor_kegg <- read.xlsx("../data/intermediate/oryzae_kegg.xlsx"); # obtained using R script metabolites_oryzae.Rmd
all(Aor_draft$METS$NAME == Aor_kegg$NAME) # same order

Aor_draft$METS$`REPLACEMENT ID`[!is.na(Aor_kegg$KEGG)] = Aor_kegg$KEGG[!is.na(Aor_kegg$KEGG)]
# if a kegg ID was found, use it. Otherwise, retain original ID
# paste compartment to it
Aor_draft$METS$`REPLACEMENT ID`[!is.na(Aor_kegg$KEGG)] <- paste0(Aor_draft$METS$`REPLACEMENT ID`[!is.na(Aor_kegg$KEGG)], 
                                                                 '[', 
                                                                 Aor_draft$METS$COMPARTMENT[!is.na(Aor_kegg$KEGG)], 
                                                                 ']');

# Step 2:
Ani_kegg <- read.xlsx("../data/intermediate/niger_kegg.xlsx"); # obtained using R script metabolites_Ani_Afu.Rmd
all(Ani_reduced$METS$NAME == Ani_kegg$NAME) # same order

Ani_reduced$METS$`REPLACEMENT ID`[!is.na(Ani_kegg$KEGG)] = Ani_kegg$KEGG[!is.na(Ani_kegg$KEGG)]
# if a kegg ID was found, use it. Otherwise, retain original ID
# paste compartment to it
Ani_reduced$METS$`REPLACEMENT ID`[!is.na(Ani_kegg$KEGG)] <- paste0(Ani_reduced$METS$`REPLACEMENT ID`[!is.na(Ani_kegg$KEGG)], 
                                                                  '[', 
                                                                  Ani_reduced$METS$COMPARTMENT[!is.na(Ani_kegg$KEGG)], 
                                                                  ']');


table(Ani_reduced$METS$`REPLACEMENT ID` %in% Aor_draft$METS$`REPLACEMENT ID`)

Ani_reduced$METS$RepName <- gsub("\\[.\\]", "", Ani_reduced$METS$`REPLACEMENT ID`)
Aor_draft$METS$RepName <- gsub("\\[.\\]", "", Aor_draft$METS$`REPLACEMENT ID`)

table(Ani_reduced$METS$RepName %in% Aor_draft$METS$RepName) 
# so 430-307 = 123 compounds that are in Ani_reduced also exist in oryzae,
# but in a different compartment
```

```{r}
hlp <- Ani_reduced$METS$ID[!Ani_reduced$METS$RepName %in% Aor_draft$METS$RepName]
```

```{r}
panAsp_v2$METS[panAsp_v2$METS$ID %in% hlp,]
```


# Sugars

Sugars: sucrose, maltose, isomaltose, trehalose, and starch

```{r}
hlp <- panAsp_v2$RXNS[grep("sucrose",panAsp_v2$RXNS$EQUATION, ignore.case = TRUE),]
panAsp_v2$RXNS[grep("fructose",panAsp_v2$RXNS$EQUATION, ignore.case = TRUE),]

hlp
hlp[!hlp$ID %in% c("r272", "OtherAsp_R00801_EXP_1_Extracellular", 
                   "OtherAsp_R00802_EXP_1_Extracellular",
                   "OtherAsp_R00801_EXP_4", "OtherAsp_R06088_EXP_2",
                   "OtherAsp_R06088_EXP_1_Extracellular", "OtherAsp_R00802_EXP_2"),]

panAsp_v2$RXNS[panAsp_v2$RXNS$ID %in% c("r272", "OtherAsp_R00801_EXP_1_Extracellular", 
                                        "OtherAsp_R00802_EXP_1_Extracellular","OtherAsp_R06088_EXP_1_Extracellular"),c(2:6,14)]
panAsp_v2$RXNS[panAsp_v2$RXNS$ID %in% c("OtherAsp_R00801_EXP_4", "OtherAsp_R06088_EXP_2",
                                        "OtherAsp_R00802_EXP_2"),c(2:6,14)]

panAsp_v2$RXNS[grep("cluster_6858", panAsp_v2$RXNS$`GENE ASSOCIATION`),c(2:6,14)]


panAsp_v2$RXNS[grep("sucrose",panAsp_v2$RXN$EQUATION, ignore.case = TRUE),]
panAsp_v2$RXNS[grep("D-Glucose 6-phosphate",panAsp_v2$RXN$EQUATION, ignore.case = TRUE),]



panAsp_v2$RXNS[grep("G10545",panAsp_v2$RXN$EQUATION, ignore.case = TRUE),]


Aor_draft$RXNS[grep("3.2.1.3",Aor_draft$RXN$`EC-NUMBER`, ignore.case = TRUE),]
# so the bottomline is, these genes allow Aspergillus to grow on 
# maltose, isomaltose, sucrose and starch, by converting it into glucose and fructose
```

# Reactions from OtherAsp

TODO: remove hydrogen
OtherAsp_R04533: 3-hydroxybutyryl-[acp][c] + NADP(+)[c] <=> hydrogen[c] + NADPH[c] + Acetoacetyl-[acp][c]

# Reactions from niger

TODO 
Ani_r239 vs. Ani_r240: same reactions, same genes, but one has "or" and the
other has "and" relationship

```{r}
panAsp_v2$RXNS[grep("Ani_", panAsp_v2$RXNS$ID),]
```

# Reactions from fumigatus

```{r}
panAsp_v2$RXNS[grep("Afu_", panAsp_v2$RXNS$ID),]
```

```{r}
Afu_reduced$METS$RepName <- gsub("\\[.*\\]","",Afu_reduced$METS$`REPLACEMENT ID`)

table(!Afu_reduced$METS$RepName %in% Aor_draft$METS$RepName)
# only +- half of the metabolites already existed in oryzae
```


```{r}
hlp <- Afu_reduced$METS$ID[!Afu_reduced$METS$RepName %in% Aor_draft$METS$RepName]
hlp <- hlp[!grepl("glutathion", hlp, ignore.case = TRUE)]
hlp
```

## Glutathione

```{r}
panAsp_v2$RXNS[grep("glutathione", panAsp_v2$RXNS$EQUATION, ignore.case = TRUE),]
```

r1008
glutathione[c] + RX[c] => HX[c] + oxidized glutathione[c]
r864
glutathione[m] + RX[m] <=> HX[m] + R-S-Glutathione[m]

```{r}
View(panAsp_v2$RXNS[grep("Diacylglycerol", panAsp_v2$RXNS$NAME, ignore.case = TRUE),])
```

```{r}
Afu_reduced$RXNS[grep("fatty acid", Afu_reduced$RXNS$EQUATION, ignore.case = TRUE),]

Afu_reduced$RXNS[Afu_reduced$RXNS$ID %in% c("R963", "R965", "R966", "R968"),]
```

Is there any difference between R963 and R965? And between R966 and R968

```{r}
panAsp_v2$RXNS[panAsp_v2$RXNS$ID %in% c("Afu_R963", "Afu_R965", "Afu_R966", "Afu_R968"),]
```



```{r}
panAsp_v2$METS[panAsp_v2$METS$ID %in% c("H2O[c]", "1,2-Diacyl-sn-glycerol[c]",
                                        "free fatty acids[c]", "2-Acylglycerol[c]"),]
```

```{r}
View(panAsp_v2$RXNS[panAsp_v2$RXNS$`EC-NUMBER` %in% c("3.1.1.3"),])
panAsp_v2$RXNS[panAsp_v2$RXNS$`EC-NUMBER` %in% c("3.1.1.3"),][c(1,2),4]
```

```{r}
panAsp_v2$RXNS[grep("acyl-carrier protein\\[e\\]", panAsp_v2$RXNS$EQUATION),]
panAsp_v2$RXNS[grep("oleoyl-\\[acp\\]\\[e\\]", panAsp_v2$RXNS$EQUATION),]
panAsp_v2$RXNS[grep("linolenoyl-\\[acp\\]\\[e\\]", panAsp_v2$RXNS$EQUATION),]
panAsp_v2$RXNS[grep("stearoyl-\\[acp\\]\\[e\\]", panAsp_v2$RXNS$EQUATION),]

panAsp_v2$RXNS[grep("palmitoyl-\\[acp\\]\\[m\\]", panAsp_v2$RXNS$EQUATION),]

panAsp_v2$RXNS[grep("hexadecenoyl-\\[acp\\]\\[m\\]", panAsp_v2$RXNS$EQUATION),]
```

```{r}
panAsp_v2$METS[grep("hexadecenoyl-\\[acp\\]", panAsp_v2$METS$ID),"REPLACEMENT ID"]
panAsp_v2$METS[grep("oleoyl-\\[acp\\]", panAsp_v2$METS$ID),"REPLACEMENT ID"]
panAsp_v2$METS[grep("linolenoyl-\\[acp\\]", panAsp_v2$METS$ID),"REPLACEMENT ID"]
panAsp_v2$METS[grep("\\bstearoyl-\\[acp\\]", panAsp_v2$METS$ID),"REPLACEMENT ID"]

panAsp_v2$METS[grep("\\bpalmitoyl-\\[acp\\]", panAsp_v2$METS$ID),"REPLACEMENT ID"]
```



```{r}
panAsp_v2$RXNS[grep("free fatty acid", panAsp_v2$RXNS$EQUATION),]
panAsp_v2$RXNS[grep("long-chain acyl-CoAs", panAsp_v2$RXNS$EQUATION),]

panAsp_v2$RXNS[grep("OtherAsp_R05190",panAsp_v2$RXNS$ID),]
```

```{r}
panAsp_v2$RXNS[panAsp_v2$RXNS$`EC-NUMBER` %in% c("3.1.1.3"),4]
```


H2O[c] + triglycerides[c] => 0.471 palmitoyl-[acp][c] + 0.0242 hexadecenoyl-[acp][c] + 0.1535 stearoyl-[acp][c] + 0.5328 oleoyl-[acp][c] + 2.4861 linolenoyl-[acp][c] + diglycerides[c]

diglycerides[c] + H2O[c] => 0.471 palmitoyl-[acp][c] + 0.0242 hexadecenoyl-[acp][c] + 0.1535 stearoyl-[acp][c] + 0.5328 oleoyl-[acp][c] + 2.4861 linolenoyl-[acp][c] + 1-monoglycerides[c]

## B00001/B00002

```{r}
panAsp_v2$RXNS[grep("B00002", panAsp_v2$RXNS$EQUATION),]
```


# Find metabolites

```{r}
panAsp_v2$METS[grep("H2O", panAsp_v2$METS$NAME),]
panAsp_v2$METS[grep("H\\(\\+\\)", panAsp_v2$METS$NAME),]
panAsp_v2$METS[grep("hydrogen$", panAsp_v2$METS$NAME,ignore.case = TRUE),]
```

```{r}
panAsp_v2$RXNS[grep("hydrogen|H2O|H\\(\\+\\)", panAsp_v2$RXNS$EQUATION),]
```

```{r}

```


---

# OLD CODE BELOW THIS

# Libraries

```{r, message=FALSE, warning=FALSE}
library(openxlsx)
```

# Helpers

```{r}
source("../code/Rutilities.R")
```

# Load data

```{r}
panAsp_v2 <- read_excel_allsheets("../model/panAsp_v2.xlsx")
Ani_idtokegg <- read.xlsx("../scrap/niger_metabolite_table.xlsx")
```

# Wrangle data

## panAsp_v2

```{r}
# Get RepName column
panAsp_v2$METS$RepName <- gsub("\\[.\\]","",panAsp_v2$METS$`REPLACEMENT ID`)

# Get KEGG column
panAsp_v2$METS$KEGG <- panAsp_v2$METS$MIRIAM
panAsp_v2$METS$KEGG[!grepl("kegg",panAsp_v2$METS$KEGG)] <- NA
panAsp_v2$METS$KEGG <- gsub("\\;.*","",gsub(".*kegg.compound/(.+)", "\\1", panAsp_v2$METS$KEGG))

# Get chebi column
panAsp_v2$METS$CHEBI <- panAsp_v2$METS$MIRIAM
panAsp_v2$METS$CHEBI[!grepl("chebi",panAsp_v2$METS$CHEBI)] <- NA
panAsp_v2$METS$CHEBI <- gsub("\\;.*","",gsub(".*chebi/(.+)", "\\1", panAsp_v2$METS$CHEBI))
panAsp_v2$METS$CHEBI <- gsub("chebi:CHEBI/", "CHEBI:", panAsp_v2$METS$CHEBI)

# Get KEGG_self column
panAsp_v2$METS$KEGG_self <- Ani_idtokegg$KEGG_self[match(panAsp_v2$METS$NAME,Ani_idtokegg$NAME)]
```

```{r}
curationTable <- panAsp_v2$METS[,c("NAME","RepName","COMPARTMENT")]
colnames(curationTable) <- c("NAME_orig","RepName_orig","COMPARTMENT")
curationTable$NAME_curated <- curationTable$NAME_orig
curationTable$RepName_curated <- curationTable$RepName_orig
```

# 2-Oxoglutarate

Looking at the reactions involved, collapsing 
Decarboxylated 2-Oxoglutarate bound to alpha-ketoglutarate dehydrogenase and	
2-Oxoglutarate bound to dihydrolipoyl transsuccinylase to 2-Oxoglutarate
doesnt look like a good idea, so I'll keep them separated.

With regard to 4-Hydroxy-2-oxoglutarate and (4R)-4-Hydroxy-2-oxoglutarate, we
could probably collapse both to 4-Hydroxy-2-oxoglutarate.
However, looking at the reactions there may still be an issue:

2-Oxoglutarate[c] + L-erythro-4-Hydroxyglutamate[c] <=> L-Glutamate[c] + (4R)-4-Hydroxy-2-oxoglutarate[c]
4-Hydroxy-2-oxoglutarate[e] <=> Pyruvate[e] + Glyoxylate[e]
(4R)-4-Hydroxy-2-oxoglutarate[e] <=> Pyruvate[e] + Glyoxylate[e]

I am not sure if this a problem: ask Ed.

```{r}
panAsp_v2$METS[grep("Oxoglut", 
                    panAsp_v2$METS$NAME, 
                    ignore.case = TRUE),
               c("NAME", "KEGG","CHEBI","REPLACEMENT ID")]
```

```{r}
# Collapse:
# 1. (4R)-4-Hydroxy-2-oxoglutarate
# 2. 4-Hydroxy-2-oxoglutarate
# to 4-Hydroxy-2-oxoglutarate

curationTable[grep("4-Hydroxy-2-oxoglutarate", 
                   curationTable$NAME_orig, 
                   ignore.case = TRUE),]
any(curationTable$RepName_curated == "4HAKG")
curationTable$NAME_curated[grep("4-Hydroxy-2-oxoglutarate", 
                                curationTable$NAME_orig, 
                                ignore.case = TRUE)] <- "4-Hydroxy-2-oxoglutarate"
curationTable$RepName_curated[grep("4-Hydroxy-2-oxoglutarate", 
                                   curationTable$NAME_orig, 
                                   ignore.case = TRUE)] <- "4HAKG"
```

# Novel metabolites from OtherAsp

We have 424 metabolites that are at this point still considered to be unique to
the OtherAsp (fumigatus and oryzae isolates). Some of them are very likely not.
In stead, they could be:
  - failed to link, but for the new ones we have all kegg and for the original
   one we also have many kegg
  - different resolution
  
```{r}
(hlp <- panAsp_v2$METS[grep("Added_",panAsp_v2$METS$RepName,ignore.case = TRUE),])
```

```{r}
idx <- unique(unlist(lapply(hlp$ID, function(element) {
  element <- gsub("\\[","\\\\[",element)
  element <- gsub("\\]","\\\\]",element)
  element
  grep(element, panAsp_v2$RXNS$EQUATION)
})))
```


## Sugars

### Glucose

```{r}
hlp[grep("Glucose",hlp$NAME,ignore.case = TRUE),]
```

```
Ani_draft <- read_excel_allsheets("../data/intermediate/Ani_draft_comp_mets.xlsx")
Ani_draft$METS[grep("glucose", Ani_draft$METS$NAME, ignore.case = TRUE),]
Ani_draft$RXNS[grep("glucose", Ani_draft$RXNS$EQUATION, ignore.case = TRUE),]
```

```{r}
# D-Glucose 6-phosphate
panAsp_v2$RXNS[grep(" D-Glucose 6-phosphate",panAsp_v2$RXNS$EQUATION),c(2:6)]

# 1. R00299
panAsp_v2$RXNS[panAsp_v2$RXNS$ID %in% c("r5a","r6","R00299","R01600","R01786"),c(3:6,14)]
# R00299 has lower resolution -> can be safely removed

panAsp_v2$RXNS[panAsp_v2$RXNS$ID %in% c("r5a","r6","R00299","R01600","R01786"),c(6,14)]

# 2. R00837_Peroxisome
panAsp_v2$RXNS[panAsp_v2$RXNS$ID %in% c("r199","r200","r1943","R00837_Peroxisome"),c(3:6,14)]

panAsp_v2$RXNS[panAsp_v2$RXNS$ID %in% c("r199","r200","r1943","R00837_Peroxisome"),c(4:5)]
panAsp_v2$RXNS[panAsp_v2$RXNS$ID %in% c("r199","r200","r1943","R00837_Peroxisome"),c(6,14)]

# here it looks like:
# r200 should become: Water[c] + alpha,alpha-Trehalose[c] => 2 alpha-D-Glucose[c]
# R00837_Peroxisome should become: Water[p] + Trehalose 6-phosphate[p] <=> alpha-D-Glucose[p] + alpha-D-Glucose 6-phosphate[p]
```

D-Gal alpha 1->6D-Gal alpha 1->6D-Glucose is a truly new compound, leave as is.

```{r}
newCompound <- c("D-Gal alpha 1->6D-Gal alpha 1->6D-Glucose")
```

Create a list linking each cluster to a set of kegg reactions
getKeggmodelfororganism on all clusters

```{r}
View(panAsp_v2$RXNS[grep("acceptor",panAsp_v2$RXNS$EQUATION, ignore.case = TRUE),c(2:6)])
```


### Fructose

```{r}
hlp[grep("Fructose",hlp$NAME,ignore.case = TRUE),]
```

Probably these three all need to have beta- in front.

```{r}
panAsp_v2$RXNS[grep("alpha-D-Fructose",panAsp_v2$RXNS$EQUATION, ignore.case = TRUE),c(2:6)]
panAsp_v2$RXNS[grep("beta-D-Fructose",panAsp_v2$RXNS$EQUATION, ignore.case = TRUE),c(2:6)]
```

```{r}
panAsp_v2$RXNS[grep("Fructose 1-",panAsp_v2$RXNS$EQUATION, ignore.case = TRUE),c(2:6)]
```


```{r}
# D-Fructose 6-phosphate
panAsp_v2$RXNS[grep(" D-Fructose 6-phosphate",panAsp_v2$RXNS$EQUATION),c(2:6)]
panAsp_v2$RXNS[grep("D-Fructose",panAsp_v2$RXNS$EQUATION, ignore.case = TRUE),c(2:6)]

# 1. R00299
panAsp_v2$RXNS[panAsp_v2$RXNS$ID %in% c("r5a","r6","R00299","R01600","R01786"),c(3:6,14)]
# R00299 has lower resolution -> can be safely removed

# 2. R00837_Peroxisome
panAsp_v2$RXNS[panAsp_v2$RXNS$ID %in% c("r199","r200","r1943","R00837_Peroxisome"),c(3:6,14)]
# here it looks like:
# r200 should become: Water[c] + alpha,alpha-Trehalose[c] => 2 alpha-D-Glucose[c]
# R00837_Peroxisome should become: Water[p] + Trehalose 6-phosphate[p] <=> alpha-D-Glucose[p] + alpha-D-Glucose 6-phosphate[p]
```



```{r}
removeRxns <- c("R00299")
```



# Reversibility

```{r}
panAsp_v2$RXNS[panAsp_v2$RXNS$ID %in% c("r5a","r6","R01600","R01786"),c(5,6,14)]

# r5a and R01600, and r6 and R01786 are the same, respectively. 
# But the one is reversible while the other is not 
# But then how to know if we want the reversible reaction? 
# In KEGG reaction, it's reversible.
# In kegg metabolic pathway of the three Aspergillus species, its irreversible.
# It is quite likely that reaction should be irreversible.
# But it could also be that in one of the isolates it has become reversible, no?
# I mean in theory. And that info would be interesting to keep.
```



# contractmodel unexpected

r799	Phosphatidylserine synthase	CDP-diacylglycerol[c] + L-Serine[c] <=> CMP[c] + Phosphatidylserine[c]	2.7.8.8	cluster_6556 or cluster_8345
r805	Phosphatidylserine synthase	CDP-diacylglycerol[c] + L-Serine[c] <=> CMP[c] + Phosphatidylserine[c]	2.7.8.8	cluster_6556
r812	Phosphatidylserine synthase	CDP-diacylglycerol[c] + L-Serine[c] <=> CMP[c] + Phosphatidylserine[c]	2.7.8.8	cluster_6556
R01800	CDP-diacylglycerol:L-serine 3-phosphatidyltransferase	CDP-diacylglycerol[c] + L-Serine[c] <=> CMP[c] + Phosphatidylserine[c]	2.7.8.8	cluster_15562

# Fatty acid metabolism

Once this is curated, it should help downstream with e.g. the ceramides.

```{r}
View(panAsp_v2$RXNS[grep("^4.2.1.17",panAsp_v2$RXNS$`EC-NUMBER`),])

#3-oxodecanoyl-CoA[m] + Coenzyme A[m] => Acetyl coenzyme A[m] + Octanoyl-CoA[m]
#Coenzyme A[p] + 3-oxodecanoyl-CoA[p] => Acetyl coenzyme A[p] + Octanoyl-CoA[p]
```

```{r}
panAsp_v2$RXNS[panAsp_v2$RXNS$ID%in%c("r748", "r1356"),]
# same reaction, same nomenclature, but reversible in oryzae
```

```{r}
panAsp_v2$RXNS[panAsp_v2$RXNS$ID%in%c("r783", "r1386", "R04746"),]
# same reaction, same nomenclature, but reversible in oryzae
# same reaction, different nomenclature, and reversible in fumigatus (different compartment though)
```

Oct-2-enoyl-CoA[m] + Water[m] => 3-hydroxyoctanoyl-CoA[m]	4.2.1.17	
Oct-2-enoyl-CoA[m] + Water[m] <=> 3-hydroxyoctanoyl-CoA[m]	4.2.1.17	
(S)-3-Hydroxyoctanoyl-CoA[c] <=> Water[c] + trans-Oct-2-enoyl-CoA[c]	4.2.1.17;4.2.1.74

-> should  change to

Oct-2-enoyl-CoA[m] + Water[m] => 3-hydroxyoctanoyl-CoA[m]	4.2.1.17	
Oct-2-enoyl-CoA[m] + Water[m] <=> 3-hydroxyoctanoyl-CoA[m]	4.2.1.17	
3-hydroxyoctanoyl-CoA[c] <=> Water[c] + Oct-2-enoyl-CoA[m] 4.2.1.17;4.2.1.74

Look into the metabolites coming from OtherAsp, for fatty acid metabolism they
often start with trans- or (3S)-/(S)-

```{r}
panAsp_v2$RXNS[grep("trans-",panAsp_v2$RXNS$EQUATION),]
```

```{r}
panAsp_v2$RXNS[grep("\\(S\\)",panAsp_v2$RXNS$EQUATION),]
```

```{r}
panAsp_v2$RXNS[grep("Mandelate",panAsp_v2$RXNS$EQUATION,ignore.case = TRUE),]
```

```{r}
panAsp_v2$METS[grep("Oct-2-enoyl-CoA",panAsp_v2$METS$NAME,ignore.case = TRUE),]
```

```{r}
panAsp_v2$METS[grep("3-Hydroxyoctanoyl-CoA",panAsp_v2$METS$NAME,ignore.case = TRUE),]
```

```{r}
target <- names(which(table(panAsp_v2$METS$RepName)>1))
keep <- sapply(seq_along(1:length(target)), 
       function(x) length(unique(panAsp_v2$METS$NAME[which(panAsp_v2$METS$RepName == target[x])]))) > 1
target <- target[keep]

for(i in 1:length(target)){
  print(panAsp_v2$METS[which(panAsp_v2$METS$RepName == target[i]),])
}

```

```{r}
panAsp_v2$METS[grep("Malonate",panAsp_v2$METS$NAME,ignore.case = TRUE),]
```




```{r}
panAsp_v2$RXNS[grep("D-Glucose 6-phosphate",panAsp_v2$RXNS$EQUATION),]
```

```{r}
panAsp_v2$RXNS[grep("Androstenedione",panAsp_v2$RXNS$EQUATION),]
```


# 1-Acyl-sn-glycero-3-phosphocholine

Is it the same as Lysophosphatidylcholine?

```{r}
panAsp_v2$METS[grep("1-Acyl-sn-glycero-3-phosphocholine|Lysophosphatidylcholine", 
                    panAsp_v2$METS$NAME, 
                    ignore.case = TRUE),
               c("NAME", "KEGG","CHEBI","KEGG_self")]

```

```{r}
panAsp_v2$METS[grep("CHEBI:183975", 
                    panAsp_v2$METS$CHEBI),
               c("NAME", "KEGG","CHEBI","KEGG_self")]
```


# Ceramides and related compounds

This a very complicated case of different resolutions for different compunds,
including:

- Cerebroside 
- Glucocerebroside (=Glucosylceramide)
- Cerebrin
- phytosphingosin
- sphinganine
- Octadecanoyl-CoA

All of this is crazy complex, we need to simplify this somehow.
Perhaps best is to rely on KEGG. For instance, we can look into 
Sphingolipid metabolism for A. niger, fumigatus, and oryzae:
- https://www.kegg.jp/pathway/ang00600
- https://www.kegg.jp/pathway/afm00600
- https://www.kegg.jp/pathway/aor00600

Then, we can look a specific enzymes/reactions, like
- https://www.kegg.jp/entry/2.4.1.80
- https://www.kegg.jp/entry/R01497

```{r}
panAsp_v2$RXNS[grep("Glucosylceramide",panAsp_v2$RXNS$EQUATION,ignore.case = TRUE),c(4,5)]
panAsp_v2$RXNS[grep("Glucocerebroside",panAsp_v2$RXNS$EQUATION,ignore.case = TRUE),4]

panAsp_v2$RXNS[grep("2.4.1.80",panAsp_v2$RXNS$`EC-NUMBER`,ignore.case = TRUE),c(6,14)]
```

```{r}
panAsp_v2$RXNS[grep("2.4.1.80",panAsp_v2$RXNS$`EC-NUMBER`,ignore.case = TRUE),c(4,5)]
panAsp_v2$RXNS[grep("2.4.1.80",panAsp_v2$RXNS$`EC-NUMBER`,ignore.case = TRUE),c(6,14)]
```

```{r}
panAsp_v2$METS[grep("Cerebroside 1|Cerebroside 2|N-Acylsphingosine|Glucocerebroside|Glucosylceramide",panAsp_v2$METS$NAME,ignore.case = TRUE),
               c(2,6,9,12,13)]
```

Note that N-Acylsphingosine is exactly 18C shorter than Cerebroside 2, and
Glucosylceramide is exactly 18C shorter than Glucocerebroside.
So in KEGG the N-acyl group, which happens to be 18C, is not counted for
the composition -> it actually is, it is the "R" at the end! 

Then Cerebroside 1 comes from Octadecanoyl, while Cerebroside 2 comes from 
Octadecenoyl.

Downstream:

```{r}
panAsp_v2$RXNS[grep("Glucocerebroside",panAsp_v2$RXNS$EQUATION,ignore.case = FALSE),4]
```

```{r}
panAsp_v2$RXNS[grep("Glucosylceramide",panAsp_v2$RXNS$EQUATION,ignore.case = FALSE),4]
```

```{r}
panAsp_v2$RXNS[grep("Glucocerebroside",panAsp_v2$RXNS$EQUATION,ignore.case = FALSE),4]
```

So, doublecheck, but i think it could be safe to:
- Throw Cerebroside 1 and Cerebroside 2 in N-Acylsphingosine
- Throw Glucocerebroside and Glucocerebroside 2 in Glucosylceramide

But perhaps we can just keep the better resolution. Because the main reason we
are doing all this is because of:

UDPglucose[c] + N-Acylsphingosine[c] <=> UDP[c] + Glucosylceramide[c]	(2.4.1.80,	cluster_6857)

Perhaps we can just remove that reaction, and the gene to both

[1] "Cerebroside 1[c] + UDPglucose[c] => Glucocerebroside[c] + UDP[c]"
[2] "Cerebroside 2[c] + UDPglucose[c] => Glucocerebroside 2[c] + UDP[c]"     

```
panAsp_v2$RXNS[grep("Octadecenoyl",panAsp_v2$RXNS$EQUATION,ignore.case = FALSE),4]
```




---

Let's start with the fact that
1. Glucocerebroside and Glucosylceramide have the same KEGG
2. Ani differentiates between Glucocerebroside and Glucocerebroside2, while
OtherAsp does not


The suggestion that N-Acylsphingosine and Cerebroside1/2 are synonyms
Then N-Acylsphingosine also seems to be a synonym for ceramide...

So, ceramides and N-Acylsphingosine are synonyms.


These three reactions:
1. Cerebroside 1[c] + UDPglucose[c] => Glucocerebroside[c] + UDP[c]
2. Cerebroside 2[c] + UDPglucose[c] => Glucocerebroside 2[c] + UDP[c]
3. UDPglucose[c] + N-Acylsphingosine[c] <=> UDP[c] + Glucosylceramide[c]
suggest that N-Acylsphingosine and Cerebroside1/2 are synonyms, 

Their composition suggests otherwise

```{r}
panAsp_v2$METS[grep("ceramide|N-Acylsphingosine",panAsp_v2$METS$NAME,ignore.case = TRUE),]
```


```{r}
panAsp_v2$METS[grep("ceramides",panAsp_v2$METS$NAME,ignore.case = FALSE),]
panAsp_v2$METS[grep("Cerebroside 1|Cerebroside 2|N-Acylsphingosine",panAsp_v2$METS$NAME,ignore.case = FALSE),]
```


```{r}
panAsp_v2$RXNS[grep("ceramides",panAsp_v2$RXNS$EQUATION,ignore.case = TRUE),]

panAsp_v2$RXNS[grep("2.4.1.80",panAsp_v2$RXNS$`EC-NUMBER`,ignore.case = TRUE),]
```



## Cerebrin

In Ani
1. Octadecanoyl-CoA[c] + Phytosphingosine(C18)[c] => Cerebrin 1 (C18 phytosphingosin and C18:0 fatty acid)[c] + Coenzyme A[c]
2. Octadecenoyl-CoA[c] + Phytosphingosine(C18)[c] => Cerebrin 1 (C18 phytosphingosin and C18:1 fatty acid)[c] + Coenzyme A[c]
3. Octadecanoyl-CoA[c] + Phytosphingosine(C20)[c] => Cerebrin 1 (C20 phytosphingosin and C18:0 fatty acid)[c] + Coenzyme A[c]
4. Octadecenoyl-CoA[c] + Phytosphingosine(C20)[c] => Cerebrin 1 (C20 phytosphingosin and C18:1 fatty acid)[c] + Coenzyme A[c]

5. Octadecanoyl-CoA[c] + Sphinganine(C18)[c] => Cerebrin 2 (C18 sphinganine and C18:0 fatty acid)[c] + Coenzyme A[c]
6. Octadecenoyl-CoA[c] + Sphinganine(C18)[c] => Cerebrin 2 (C18 sphinganine and C18:1 fatty acid)[c] + Coenzyme A[c]
7. Octadecanoyl-CoA[c] + Sphinganine(C20)[c] => Cerebrin 2 (C20 sphinganine and C18:0 fatty acid)[c] + Coenzyme A[c]
8. Octadecenoyl-CoA[c] + Sphinganine(C20)[c] => Cerebrin 2 (C20 sphinganine and C18:1 fatty acid)[c] + Coenzyme A[c]

9. Octadecanoyl-CoA[c] + Sphinganine(C19:1)[c] => Cerebroside 1[c] + Coenzyme A[c]

In Afu ==>
1. stearoyl-CoA[c] +	phytosphingosine[c] => Cerebrin 1[c] + coenzyme A[c]
2. oleoyl-CoA[c] + phytosphingosine[c] => Cerebrin 1[c] + coenzyme A[c]
3. stearoyl-CoA[c] + Phytosphingosine(C20)[c] => Cerebrin 1[c] + coenzyme A[c]
4. oleoyl-CoA[c] + Phytosphingosine(C20)[c] => Cerebrin 1[c] + coenzyme A[c]

5. stearoyl-CoA[c] + sphinganine[c] => Cerebrin 2[c] + coenzyme A[c]
6. oleoyl-CoA[c] + sphinganine[c] => Cerebrin 2[c] + coenzyme A[c]
7. stearoyl-CoA[c] + sphinganine (C18)[c] => Cerebrin 2[c] + coenzyme A[c]
8. oleoyl-CoA[c] + sphinganine (C18)[c] => Cerebrin 2[c] + coenzyme A[c]

9. stearoyl-CoA[c] + Sphinganine[c]  => Cerebroside 1[c] + coenzyme A[c]

################################################################################

################################################################################


Sphinganine(C18), Sphinganine(C19:1) and Sphinganine(C20) -> Sphinganine
Phytosphingosine(C18) and Phytosphingosine(C20) -> Phytosphingosine
UDP-galactose (pyranose) -> UDP-galactose

---  Cerebrin 1 after Octadecanoyl or stearoyl -> Cerebrin 1a
Cerebrin 1 (C18 phytosphingosin and C18:0 fatty acid) and Cerebrin 1 (C20 phytosphingosin and C18:0 fatty acid) -> Cerebrin 1a

---  Cerebrin 1 after Octadecenoyl or oleoyl -> Cerebrin 1b
Cerebrin 1 (C18 phytosphingosin and C18:1 fatty acid) and Cerebrin 1 (C20 phytosphingosin and C18:1 fatty acid) -> Cerebrin 1b

Cerebrin 1 (C18 phytosphingosin and C14:0 fatty acid) and Cerebrin 1 (C20 phytosphingosin and C14:0 fatty acid) -> Cerebrin 1c


Cerebrin 2 (C18 sphinganine and C18:0 fatty acid) and Cerebrin 2 (C20 sphinganine and C18:0 fatty acid) -> Cerebrin 2a
Cerebrin 2 (C18 sphinganine and C18:1 fatty acid) and Cerebrin 2 (C20 sphinganine and C18:1 fatty acid) -> Cerebrin 2b
Cerebrin 2 (C18 sphinganine and C14:0 fatty acid) and Cerebrin 2 (C20 sphinganine and C14:0 fatty acid) -> Cerebrin 2c
Cerebrin 2 (C18 sphinganine and a 2-hydroxy C18:1 fatty acid) -> Cerebrin 2d

Cerebrin 3 (C18 sphingosine and C18:0 fatty acid) and Cerebrin 3 (C20 sphingosine and C18:0 fatty acid) -> Cerebrin 3a
Cerebrin 3 (C18 sphingosine and C18:1 fatty acid) and Cerebrin 3 (C20 sphingosine and C18:1 fatty acid) -> Cerebrin 3b
Cerebrin 3 (C18 sphingosine and C14:0 fatty acid) and Cerebrin 3 (C20 sphingosine and C14:0 fatty acid) -> Cerebrin 3c
Cerebrin 3 (C18 sphingosine and a 2-hydroxy C18:1 fatty acid) -> Cerebrin 3d

Cerebrin 1 (Mixture of C18 and C20 phytosphingosin) -> Cerebrin 1ab
Cerebrin 2 (Mixture of C18 and C20 sphingosin) -> Cerebrin 2ab

Galactosylceramide(Comp. A) -> Galactosylceramide A
Galactosylceramide(Comp. B) -> Galactosylceramide B
Galactosylceramide (Mixture between GALCERA and GALCERB) -> Galactosylceramide AB

Galactosylceramide (C18 sphinganine and 2-hydroxy C18:1 fatty acid) -> Galactosylceramide C
Galactosylceramide (C18 sphingosine and 2-hydroxy C18:1 fatty acid) -> Galactosylceramide D

################################################################################

Octadecanoyl-CoA[c] + Phytosphingosine[c] => Cerebrin 1a[c] + Coenzyme A[c]
Octadecenoyl-CoA[c] + Phytosphingosine[c] => Cerebrin 1b[c] + Coenzyme A[c]

0.25 Cerebrin 1a[c] + 0.25 Cerebrin 1b[c] + 0.25 Cerebrin 1a[c] + 0.25 Cerebrin 1b[c] => Cerebrin 1ab[c]

Octadecanoyl-CoA[c] + Sphinganine[c] => Cerebrin 2a[c] + Coenzyme A[c]
Octadecenoyl-CoA[c] + Sphinganine[c] => Cerebrin 2b[c] + Coenzyme A[c]

0.25 Cerebrin 2a[c] + 0.25 Cerebrin 2b[c] + 0.25 Cerebrin 2a[c] + 0.25 Cerebrin 2b[c] => Cerebrin 2ab[c]

Cerebrin 2a[c] + UDP-galactose[c] => Galactosylceramide A[c] + UDP[c]
Cerebrin 2b[c] + UDP-galactose[c] => Galactosylceramide B[c] + UDP[c]

Octadecanoyl-CoA[c] + Sphinganine[c] => Cerebroside 1[c] + Coenzyme A[c]
Cerebroside 1[c] + UDPglucose[c] => Glucocerebroside[c] + UDP[c]
Octadecenoyl-CoA[c] + Sphinganine[c] => Cerebroside 2[c] + Coenzyme A[c]
Cerebroside 2[c] + UDPglucose[c] => Glucocerebroside 2[c] + UDP[c]

Cerebrin 2ab[c] + 1-Phosphatidyl-D-myo-inositol[c] => Water[c] + Inositol phosphorylceramide[c]

0.5 Galactosylceramide A[c] + 0.5 Galactosylceramide B[c] => Galactosylceramide AB[c]

0.000223 Myristic acid[c] + 0.00148 Palmitate[c] + 0.000245 Stearate[c] + 0.001678 Octadecenoate; Oleic acid[c] + 0.001664 Octadecadienoate; Linoleic acid[c] + 4.7e-05 Octadecatrienoate; α-linolenic acid[c] + 3e-05 Cerebrin 1ab[c] + 3.1e-05 Cerebrin 2ab[c] + 0.001746 Cardiolipin[m] + 0.001009 Diacylglycerol[c] + 0.007859 Digalactosyl-diacylglycerol[c] + 0.010038 Ergosterolester[c] + 0.034062 Ergosterol[c] + 2.5e-05 Galactosylceramide AB[c] + 2.4e-05 Glucocerebroside[c] + 2.4e-05 Glucocerebroside 2[c] + 0.008912 Monoacylglycerol[c] + 0.012374 Monoglucosyloxyoctadecenoic acid[c] + 0.030053 Monogalactosyl-diacylglycerol[c] + 0.015312 Phosphatidylcholine[c] + 0.034807 Phosphatidylethanolamine[c] + 0.000359 Phosphatidylserine[c] + 0.010083 Triacylglycerol[c] + 5e-06 Trigalactosyldimannosylinositol-P-ceramide[c] => Artificial metabolite. The metabolites found in the biomass. Measured in grams[c]

Tetradecanoyl-CoA[c] + Phytosphingosine[c] => Coenzyme A[c] + Cerebrin 1c[c]
Tetradecanoyl-CoA[c] + Sphinganine[c] => Coenzyme A[c] + Cerebrin 2c[c]

UDP-galactose[c] + Cerebrin 2d[c] <=> UDP[c] + Galactosylceramide C[c]
Sphinganine[c] + 2-hydroxy octadecenoyl-CoA[c] <=> Coenzyme A[c] + Cerebrin 2d[c]

Cerebrin 2a[c] + 2 Intracellular protons[c] + Oxygen[c] => 2 Water[c] + Cerebrin 3a[c]
Cerebrin 2b[c] + 2 Intracellular protons[c] + Oxygen[c] => 2 Water[c] + Cerebrin 3b[c]
2 Intracellular protons[c] + Oxygen[c] + Cerebrin 2c[c] => 2 Water[c] + Cerebrin 3c[c]
2 Intracellular protons[c] + Oxygen[c] + Cerebrin 2d[c] => 2 Water[c] + Cerebrin 3d[c]

UDP-galactose[c] + Cerebrin 3d[c] <=> UDP[c] + Galactosylceramide D[c]

Cerebrin 1a[c] + Water[c] => Stearate[c] + Phytosphingosine[c]
Sphinganine[c] + NADPH[c] + Oxygen[c] => Phytosphingosine[c] + Water[c] + NADP+[c]


################################## Afu

sphinganine[c] + stearoyl-CoA[c] => coenzyme A[c] + Cerebrin 2[c]
phytosphingosine[c] + oleoyl-CoA[c] => coenzyme A[c] + Cerebrin 1[c]
stearoyl-CoA[c] + sphinganine (C18)[c] => coenzyme A[c] + Cerebrin 2[c]
oleoyl-CoA[c] + Phytosphingosine(C20)[c] => coenzyme A[c] + Cerebrin 1[c]
Sphinganine[c] + stearoyl-CoA[c] => coenzyme A[c] + Cerebroside 1[c]
phytosphingosine[c] + stearoyl-CoA[c] => coenzyme A[c] + Cerebrin 1[c]
sphinganine[c] + oleoyl-CoA[c] => coenzyme A[c] + Cerebrin 2[c]
stearoyl-CoA[c] + Phytosphingosine(C20)[c] => coenzyme A[c] + Cerebrin 1[c]
oleoyl-CoA[c] + sphinganine (C18)[c] => coenzyme A[c] + Cerebrin 2[c]


################################


```
target[grepl("Cerebrin",target$NAME,ignore.case = TRUE),]
```

```
target[grepl("sphing",target$NAME,ignore.case = TRUE),]
```

```
niger_gem$RXNS[grepl("ceramid",niger_gem$RXNS$NAME,ignore.case = TRUE),]

target[grepl("sphing",target$NAME,ignore.case = TRUE),]
```

Octadecanoyl-CoA[c] + Sphinganine(C19:1)[c] => Cerebroside 1[c] + Coenzyme A[c]
Cerebroside 1[c] + UDPglucose[c] => Glucocerebroside[c] + UDP[c]
Octadecenoyl-CoA[c] + Sphinganine(C19:1)[c] => Cerebroside 2[c] + Coenzyme A[c]
Cerebroside 2[c] + UDPglucose[c] => Glucocerebroside 2[c] + UDP[c]

0.000223 Myristic acid[c] + 0.00148 Palmitate[c] + 0.000245 Stearate[c] + 0.001678 Octadecenoate; Oleic acid[c] + 0.001664 Octadecadienoate; Linoleic acid[c] + 4.7e-05 Octadecatrienoate; α-linolenic acid[c] + 3e-05 Cerebrin 1 (Mixture of C18 and C20 phytosphingosin)[c] + 3.1e-05 Cerebrin 2 (Mixture of C18 and C20 sphingosin)[c] + 0.001746 Cardiolipin[m] + 0.001009 Diacylglycerol[c] + 0.007859 Digalactosyl-diacylglycerol[c] + 0.010038 Ergosterolester[c] + 0.034062 Ergosterol[c] + 2.5e-05 Galactosylceramide (Mixture between GALCERA and GALCERB)[c] + 2.4e-05 Glucocerebroside[c] + 2.4e-05 Glucocerebroside 2[c] + 0.008912 Monoacylglycerol[c] + 0.012374 Monoglucosyloxyoctadecenoic acid[c] + 0.030053 Monogalactosyl-diacylglycerol[c] + 0.015312 Phosphatidylcholine[c] + 0.034807 Phosphatidylethanolamine[c] + 0.000359 Phosphatidylserine[c] + 0.010083 Triacylglycerol[c] + 5e-06 Trigalactosyldimannosylinositol-P-ceramide[c] => Artificial metabolite. The metabolites found in the biomass. Measured in grams[c]


```{r}
panAsp_v2$METS[grep("C01190", 
                    panAsp_v2$METS$KEGG, 
                    ignore.case = TRUE),
               c("NAME", "KEGG","CHEBI","REPLACEMENT ID")]
```


```{r}
panAsp_v2$METS[grep("cere", 
                    panAsp_v2$METS$NAME, 
                    ignore.case = TRUE),
               c("NAME", "KEGG","CHEBI","REPLACEMENT ID")]
```

```{r}
curationTable[grep("cerebros", 
                   curationTable$NAME_orig, 
                   ignore.case = TRUE),]
any(curationTable$RepName_curated == "4HAKG")
curationTable$NAME_curated[grep("4-Hydroxy-2-oxoglutarate", 
                                curationTable$NAME_orig, 
                                ignore.case = TRUE)] <- "4-Hydroxy-2-oxoglutarate"
curationTable$RepName_curated[grep("4-Hydroxy-2-oxoglutarate", 
                                   curationTable$NAME_orig, 
                                   ignore.case = TRUE)] <- "4HAKG"
```

# Write output table

```{r}
curationTable$RepID_curated <- paste0(curationTable$RepName_curated, "[",
                                      curationTable$COMPARTMENT, "]")
write.xlsx(curationTable, "../data/intermediate/panAsp_curation_table.xlsx")
```


# Potential downstream issues

TODO: I might still need to remove this reaction:
0.25 Cdp-diacylglycerol used for cardiolipin[m] + 
0.25 Cdp-diacylglycerol used for phosphatidylcholine[c] + 
0.25 Cdp-diacylglycerol used for phosphatidylethanolamine[c] + 
0.25 Cdp-diacylglycerol used for phosphatidylserine[c] => CDP-diacylglycerol[c]
because I killed it's metabolites. Not sure if this will be handled 
automatically.








---

# TODO

Different resolution:

- Glucocerebroside

For Glucocerebroside vs Glucocerebroside2 it is just a difference between
double bond or not.

---

Riboflavin metabolism

KEGG ====================
 	
*3.5.4.25)* GTP + 4 H2O <=> Formate + 2,5-Diamino-6-(5-phospho-D-ribosylamino)pyrimidin-4(3H)-one + 2 Orthophosphate

Option 1:
*3.5.4.26)*  2,5-Diamino-6-(5-phospho-D-ribosylamino)pyrimidin-4(3H)-one + H2O <=> 5-Amino-6-(5'-phosphoribosylamino)uracil + Ammonia
*1.1.1.193)* 5-Amino-6-(5'-phosphoribosylamino)uracil + NADPH + H+ <=> 5-Amino-6-(5'-phospho-D-ribitylamino)uracil + NADP+

Option 2:
*1.1.1.302)* 2,5-Diamino-6-(5-phospho-D-ribosylamino)pyrimidin-4(3H)-one + NADPH + H+ <=> 2,5-Diamino-6-(5-phospho-D-ribitylamino)pyrimidin-4(3H)-one + NADP+
*5.4.99.28)* 2,5-Diamino-6-(5-phospho-D-ribitylamino)pyrimidin-4(3H)-one + H2O <=> 5-Amino-6-(5'-phospho-D-ribitylamino)uracil + Ammonia

*3.1.3.104)* 5-Amino-6-(5'-phospho-D-ribitylamino)uracil + H2O <=> 5-Amino-6-(1-D-ribitylamino)uracil + Orthophosphate
*2.5.1.78)*  5-Amino-6-(1-D-ribitylamino)uracil + L-3,4-Dihydroxybutan-2-one 4-phosphate <=> 6,7-Dimethyl-8-(D-ribityl)lumazine + 2 H2O + Orthophosphate

Aor =====================

*3.5.4.25)* GTP[c] => 2,5-diamino-6-hydroxy-4-(5'-phosphoribosylamino)-pyrimidine[c] + formate[c] + diphosphate[c]

Option 1:
*3.5.4.26)* 2,5-diamino-6-hydroxy-4-(5'-phosphoribosylamino)-pyrimidine[c] => *5-amino-6-(5-phosphoribosylamino)uracil[c]* + NH3[c]
*1.1.1.193)* 5-amino-6-(5-phosphoribosylamino)uracil[c] + NADPH[c] => 5-amino-6-(5'-phosphoribitylamino)uracil[c] + NADP(+)[c]

*3.1.3.-)* 5-amino-6-(5'-phosphoribitylamino)uracil[c] + H2O[c] => 5-amino-6-ribityluracil[c] + phosphate[c]	
*empty)* 5-amino-6-ribityluracil[c] + 2-hydroxy-3-oxobutyl phosphate[c] => 6,7-dimethyl-8-(1-D-ribityl)lumazine[c] + 2 H2O[c] + phosphate[c]

Ani =====================

*3.5.4.25)* GTP[c] + 3 Water[c] => Formate[c] + Pyrophosphate[c] + 2,5-diamino-6-(1-D-ribosylamino)pyrimidin-4(3H)-one 5'-phosphate[c]	

Option 2:
*1.1.1.302)* NADPH[c] + 2,5-diamino-6-(1-D-ribosylamino)pyrimidin-4(3H)-one 5'-phosphate[c] => NADP+[c] + 2,5-diamino-6-(5-phosphono)ribitylamino-4(3H)-pyrimidinone[c]	
*3.5.4.26)* Water[c] + 2,5-diamino-6-(5-phosphono)ribitylamino-4(3H)-pyrimidinone[c] => 2 Intracellular protons[c] + Ammonia[c] + *5-amino-6-(5-phospho-β-D-ribosylamino)uracil[c]*

*3.1.3.104)* 2 Intracellular protons[c] + Water[c] + 5-amino-6-(5-phospho-β-D-ribosylamino)uracil[c] => Orthophosphate[c] + 5-amino-6-(D-ribitylamino)uracil[c]
*2.5.1.78)* 2-hydroxy-3-oxobutyl phosphate[c] + 5-amino-6-(D-ribitylamino)uracil[c] => 2 Water[c] + Orthophosphate[c] + 6,7-dimethyl-8-(1-D-ribityl)lumazine[c]

So:
Strong suspicion that
1. In Aor, enzyme 3.1.3.- should be 3.1.3.104
2. In Ani, enzyme 3.5.4.26 should be 5.4.99.28
3. 5-amino-6-(5-phosphoribosylamino)uracil[c] and 5-amino-6-(5-phospho-β-D-ribosylamino)uracil[c] both have CHEBI:18337. However, the former is created by removing NH3 from the starting compound, while the later is created by reduction and removing NH3. In KEGG, they are two different

This leads me to:
1) Replace EC:3.5.4.26 with EC:5.4.99.28 in Ani_Draft
2) Remove metabolite 5-amino-6-(5-phosphoribosylamino)uracil[c] in Aor_draft. TODO: check if this automatically removes the reaction in RAVEN.
3) Consolidate: change 5-amino-6-(5'-phosphoribitylamino)uracil in Aor to 5-amino-6-(5-phospho-β-D-ribosylamino)uracil from Ani. Remove the CHEBI in both, and give both KEGG C00454.
4) Consolidate: change 5-amino-6-ribityluracil[c] in Aor to 5-amino-6-(D-ribitylamino)uracil[c], give both KEGG C04732
5) 



5-amino-6-(D-ribitylamino)uracil[c] == 5-amino-6-ribityluracil[c] == C04732
2-hydroxy-3-oxobutyl phosphate[c] == 2-hydroxy-3-oxobutyl phosphate[c] == C15556
6,7-dimethyl-8-(1-D-ribityl)lumazine[c] = 6,7-dimethyl-8-(1-D-ribityl)lumazine[c] == C04332

```{r}
rbind(target[which(target$ID == "6,7-dimethyl-8-(1-D-ribityl)lumazine[c]"),c("ID","CHEBI","KEGG")])
```

```{r}
which(Aor_reduced$METS$ID == "6,7-dimethyl-8-(1-D-ribityl)lumazine[c]")
Aor_draft$METS[grepl("oxobutyl", Aor_draft$METS$ID),]
```

```{r}
rbind(Aor_reduced$METS[which(Aor_reduced$METS$ID == "2-hydroxy-3-oxobutyl phosphate[c]"),c("ID","CHEBI","KEGG")])
```

```{r}
rbind(Ani_draft$METS[which(Ani_draft$METS$NAME == "2-hydroxy-3-oxobutyl phosphate"),c("ID","CHEBI","KEGG")])
```

```{r}
Ani_draft$METS[grepl("2,5-diamino-6-\\(5-phosphono\\)ribitylamino-4\\(3H\\)-pyrimidinone", Ani_draft$METS$NAME,ignore.case = TRUE),]
Ani_draft$METS[which(Ani_draft$METS$NAME == "5-amino-6-(5-phospho-β-D-ribosylamino)uracil"),]
```

# Issues

## zymosterol

For now I assumed the mechnaism are different, but they could just be synonyms 
or resolution differences. Asked Ed for help.

Ani ====================
Lanosterol[c] + 3 NADPH[c] + 3 Oxygen[c] => 4,4-Dimethyl-5alpha-cholesta-8,14,24-trien-3beta-ol[c] + Formate[c] + 4 Water[c] + 3 NADP+[c]
4,4-Dimethyl-5alpha-cholesta-8,14,24-trien-3beta-ol[c] + NADPH[c] => 14-Demethyllanosterol[c] + NADP+[c]
14-demethyllanosterol[c] + 3 NADPH[c] + 3 Oxygen[c] => 4-methylcholesta-8,24-diene-3-ol-4-carboxylate[c] + Intracellular protons[c] + 4 Water[c] + 3 NADP+[c]
4-methylcholesta-8,24-diene-3-ol-4-carboxylate[c] + Intracellular protons[c] => 4-methylcholesta-8,24-diene-3-ol[c] + Carbon dioxide[c]
4-methylcholesta-8,24-diene-3-ol[c] + 3 NADPH[c] + 3 Oxygen[c] => 3-dehydro-4-carboxyzymosterol[c] + 2 Intracellular protons[c] + 4 Water[c] + 3 NADP+[c]
3-dehydro-4-carboxyzymosterol[c] + 2 Intracellular protons[c] => Carbon dioxide[c] + Zymosterol[c]

Aor ====================
lanosterol[c] + oxygen[c] + FADH2[m] => 4,4-dimethylcholesta-8,14,24-trienol[c] + FAD[m]
4,4-dimethylcholesta-8,14,24-trienol[c] + NADPH[c] => 14-demethyllanosterol[c] + NADP(+)[c]
14-demethyllanosterol[c] + 3 oxygen[c] => intermediate methylzymosterol I[c]
intermediate methylzymosterol I[c] => CO2[c] + intermediate methylzymosterol II[c]
intermediate methylzymosterol II[c] + NADPH[c] => 4alpha-methylzymosterol[c] + NADP(+)[c]
4alpha-methylzymosterol[c] + 3 oxygen[c] => intermediate zymosterol I[c]
intermediate zymosterol I[c] => CO2[c] + intermediate zymosterol II[c]
intermediate zymosterol II[c] + NADPH[c] => zymosterol[c] + NADP(+)[c]

## cytochrome

Aor doesnt care about protons. But then these two reactions which should be
identical will be separate reactions in the drafts...

Aor ==============
NADPH[c] + 2 ferricytochrome c[m] => NADP(+)[c] + 2 ferrocytochrome c[m]

2 ferricytochrome c[m] + 4 H(+)[m] + ubiquinol[m] => 4 H(+) (energy metabolism)[c] + 2 ferrocytochrome c[m] + ubiquinone[m]

2 ferrocytochrome c[m] + 4 H(+)[m] + 0.5 oxygen[m] => 4 H(+) (energy metabolism)[c] + 2 ferricytochrome c[m] + H2O[m]


Ani =================
2 Ferricytochrome c[m] + NADPH[c] => 2 Ferrocytochrome c[m] + 2 Intracellular protons[c] + NADP+[c]

2 Ferricytochrome c[m] + 4 Proton (energy metabolism)[m] + Ubiquinol[m] => 2 Ferrocytochrome c[m] + 4 Proton (energy metabolism)[c] + Ubiquinone[m] + 2 Proton[m]

2 Ferrocytochrome c[m] + 4 Proton (energy metabolism)[m] + 0.5 Oxygen[m] + 2 Proton[m] => 2 Ferricytochrome c[m] + 4 Proton (energy metabolism)[c] + Water[m]

## Penicillin

N-[L-5-amino-5-carboxypentanoyl]-L-cysteinyl-D-valine[c] + oxygen[c] => 2 H2O[c] + isopenicillin N[c]
isopenicillin N[c] + phenylacetyl-CoA[c] => L-2-aminoadipate[c] + coenzyme A[c] + penicillins[c]
isopenicillin N[c] => penicillin N[c]
penicillins[c] => penicillanic acid[c]
penicillins[c] => penicillanic acid[c]
penicillins[c] => penicillanic acid[c]
penicillins[c] => penicillanic acid[c]
penicillins[c] => penicillanic acid[c]
penicillin G[c] => 

So essentially, 
1. penicillin G can be extrected but is never used/created anywhere.
2. penicillin N is a dead-end metabolite
3. penicillanic acid is a dead-end metabolite

## Chitosan

Chitoson is a huge polymer....

Ani ========
Chitosan[c] + Water[c] => D-Glucosamine[c]

Aor ========
chitosan[c] => D-glucosamine[c]

In kegg, R08715:
Chitosan(n+1) + H2O <=> Chitosan(n) + D-Glucosamine

## Pectin

Ani =========
Endo-Polygalacturonase	Water[e] + Polygalacturonate[e] => D-Galacturonate[e]	                          NA
Pectin methyl esterase	Water[e] + Methylated galacturonate[e] => Methanol[e] + Polygalacturonate[e]	  3.1.1.11


Aor ==========
Exo-polygalacturonase	    H2O[e] + pectate[e] => D-galacturonate[e]	    3.2.1.67
Endo-polygalacturonase	  pectate[e] => D-galacturonate[e]	            3.2.1.15
Endo-polygalacturonase	  pectate[e] => D-galacturonate[e]	            3.2.1.15
Endo-polygalacturonase	  pectate[e] => D-galacturonate[e]	            3.2.1.15

So, first off, I think these reactions should always involve H2O.
Second, it is strange that it is using "Methylated galacturonate" in stead of
"Methylated polygalacturonate".
Third, could we say pectate == Polygalacturonate?
Fourth, in KEGG, these reactions (e.g.) are always specified as
 	
Pectate + H2O <=> Pectate + D-Galacturonate

i.e., with the polymer on both sides. I suppose you cant do that in a GEM, 
because then the supply of pectate would be eternal.

## Riboflavin metabolism

https://www.kegg.jp/pathway/map00740+C04454

KEGG ====================
 	
*3.5.4.25)* GTP + 4 H2O <=> Formate + 2,5-Diamino-6-(5-phospho-D-ribosylamino)pyrimidin-4(3H)-one + 2 Orthophosphate

*3.5.4.26)* 2,5-Diamino-6-(5-phospho-D-ribosylamino)pyrimidin-4(3H)-one + H2O <=> 5-Amino-6-(5'-phosphoribosylamino)uracil + Ammonia
*1.1.1.193)* 5-Amino-6-(5'-phosphoribosylamino)uracil + NADPH + H+ <=> 5-Amino-6-(5'-phospho-D-ribitylamino)uracil + NADP+

*1.1.1.302)* 2,5-Diamino-6-(5-phospho-D-ribosylamino)pyrimidin-4(3H)-one + NADPH + H+ <=> 2,5-Diamino-6-(5-phospho-D-ribitylamino)pyrimidin-4(3H)-one + NADP+
*5.4.99.28)* 2,5-Diamino-6-(5-phospho-D-ribitylamino)pyrimidin-4(3H)-one + H2O <=> 5-Amino-6-(5'-phospho-D-ribitylamino)uracil + Ammonia

*3.1.3.104)* 5-Amino-6-(5'-phospho-D-ribitylamino)uracil + H2O <=> 5-Amino-6-(1-D-ribitylamino)uracil + Orthophosphate
*2.5.1.78)*  5-Amino-6-(1-D-ribitylamino)uracil + L-3,4-Dihydroxybutan-2-one 4-phosphate <=> 6,7-Dimethyl-8-(D-ribityl)lumazine + 2 H2O + Orthophosphate

Aor =====================

*3.5.4.25)* GTP[c] => 2,5-diamino-6-hydroxy-4-(5'-phosphoribosylamino)-pyrimidine[c] + formate[c] + diphosphate[c]
*3.5.4.26)* 2,5-diamino-6-hydroxy-4-(5'-phosphoribosylamino)-pyrimidine[c] => 5-amino-6-(5-phosphoribosylamino)uracil[c] + NH3[c]
*1.1.1.193)* 5-amino-6-(5-phosphoribosylamino)uracil[c] + NADPH[c] => 5-amino-6-(5'-phosphoribitylamino)uracil[c] + NADP(+)[c]
*3.1.3.-)* 5-amino-6-(5'-phosphoribitylamino)uracil[c] + H2O[c] => 5-amino-6-ribityluracil[c] + phosphate[c]	
*empty)* 5-amino-6-ribityluracil[c] + 2-hydroxy-3-oxobutyl phosphate[c] => 6,7-dimethyl-8-(1-D-ribityl)lumazine[c] + 2 H2O[c] + phosphate[c]

Ani =====================

*3.5.4.25)* GTP[c] + 3 Water[c] => Formate[c] + Pyrophosphate[c] + 2,5-diamino-6-(1-D-ribosylamino)pyrimidin-4(3H)-one 5'-phosphate[c]	
*1.1.1.302)* NADPH[c] + 2,5-diamino-6-(1-D-ribosylamino)pyrimidin-4(3H)-one 5'-phosphate[c] => NADP+[c] + 2,5-diamino-6-(5-phosphono)ribitylamino-4(3H)-pyrimidinone[c]	
*3.5.4.26)* Water[c] + 2,5-diamino-6-(5-phosphono)ribitylamino-4(3H)-pyrimidinone[c] => 2 Intracellular protons[c] + Ammonia[c] + 5-amino-6-(5-phospho-β-D-ribosylamino)uracil[c]
*3.1.3.104)* 2 Intracellular protons[c] + Water[c] + 5-amino-6-(5-phospho-β-D-ribosylamino)uracil[c] => Orthophosphate[c] + 5-amino-6-(D-ribitylamino)uracil[c]
*2.5.1.78)* 2-hydroxy-3-oxobutyl phosphate[c] + 5-amino-6-(D-ribitylamino)uracil[c] => 2 Water[c] + Orthophosphate[c] + 6,7-dimethyl-8-(1-D-ribityl)lumazine[c]

So:
Strong suspicion that
1. In Aor, enzyme 3.1.3.- should be 3.1.3.104
2. In Ani, enzyme 3.5.4.26 should be 5.4.99.28



---

# Revisit

What I do below can be very useful, but only when jointly looking to the other
GEMs. So, I will incorporate the code below in the other scripts later.

```
target <- niger_gem$METS[is.na(niger_gem$METS$KEGG)&is.na(niger_gem$METS$KEGG_v2)&is.na(niger_gem$METS$KEGG_v3)&is.na(niger_gem$METS$KEGG_v4),]
target <- target[!duplicated(target$NAME),]
```

```
View(target)
```

## Niger-specific metabolites

```
sum(grepl("Yanuthone",target$NAME,ignore.case = TRUE))
sum(grepl("nigrin",target$NAME,ignore.case = TRUE))
sum(grepl("fonsec",target$NAME,ignore.case = TRUE))
sum(grepl("Azanig",target$NAME,ignore.case = TRUE))
```

There are several classes of complex secondary metabolites described by Brandl
himself on the CHEBI website. We will never find a KEGG for these, and it is
very unlikely these will ever appear in the other GEMs.

- yanuthones: A class of meroterpenoids whose core structure consists of 5,6-epoxycyclohex-2-en-1-one which is substituted by varying side chains at positions 3 and 4, and by a sesquiterpene chain at position 6. The core structure may be derived from 6-methylsalicylic acid, which due to a decarboxylation results in a C7 scaffold (a 6-membered methylated ring); these are known as class I yanuthones. Class II yanuthones contain a C6 scaffold derived from an unknown precursor. -> search for yanuthone and epoxycyclo in Aor and Afu gave no hits

- Pyranonigrin: A member of the class of pyranopyrroles that is 2-[(1E)-prop-1-en-1-yl]-6,7-dihydropyrano[2,3-c]pyrrole-4,5-dione carrying two additional hydroxy substituents at positions 3 and 7. -> search for nigrin, pyrano and pyrrole in Aor and Afu gave no hits

- Fonsecinone: A dimeric naphthopyran with formula C32H28O11, originally isolated from Aspergillus niger. On pubchem, fonsecinone is 5-hydroxy-9-(5-hydroxy-6,8-dimethoxy-2-methyl-4-oxobenzo[g]chromen-10-yl)-8,10-dimethoxy-2-methylbenzo[h]chromen-4-one -> Search for fonsec, naphto and pyran in Aor and Afu gave no hits

-Azanigerone: An azaphilone that is the ester obtained by formal condensation of the carboxy group of 2,4-dimethylhexanoic acid with the tertiary hydroxy group of 7-hydroxy-7-methyl-6,8-dioxo-7,8-dihydro-6H-2-benzopyran-3-carboxylic acid. IUPAC: 7-[(2,4-dimethylhexanoyl)oxy]-7-methyl-6,8-dioxo-7,8-dihydro-6H-2-benzopyran-3-carboxylic acid
-> Search for aza and benzopyr in Aor and Afu gave no hits

```
target <- target[!grepl("Yanuthone",target$NAME,ignore.case = TRUE) &
                   !grepl("nigrin",target$NAME,ignore.case = TRUE) &
                   !grepl("fonsec",target$NAME,ignore.case = TRUE) &
                   !grepl("Azanig",target$NAME,ignore.case = TRUE)
                 ,]
```



## Fatty acids

```
sum(grepl("Undec",niger_gem$METS$NAME))
sum(grepl("Dodec",niger_gem$METS$NAME))
sum(grepl("Tridec",niger_gem$METS$NAME))
sum(grepl("Tetradec",niger_gem$METS$NAME))
sum(grepl("Pentadec",niger_gem$METS$NAME))
sum(grepl("Hexadec",niger_gem$METS$NAME))
sum(grepl("Heptadec",niger_gem$METS$NAME))
sum(grepl("Octadec",niger_gem$METS$NAME))
sum(grepl("Nonadec",niger_gem$METS$NAME))
sum(grepl("Eicos",niger_gem$METS$NAME))
```

In Aor and Afu, we use the regular names (stearoyl,oleoyl,myristoyl,...).
This translation can be done through KEGG. However, for many compounds
(e.g. those in target) we already dont have kegg. For some we could get
the kegg maybe, but for instance:

Octadecanoyl-CoA = stearoyl CoA = KEGG C00412 -> found
Octadecanoyl-carnitine = stearoyl carnitine -> not in kegg, but the name is in Aor


Try to change to these

```
target[5:89,]
```

Octadecanoyl-CoA = stearoyl CoA = KEGG C00412
Octadecenoyl-CoA = Oleoyl CoA = KEGG C00510
Tetradecanoyl-CoA = Myristoyl-CoA = C02593

Octadecanoate is stearate
Octadecanoic acid is Stearic acid =  	C01530

coenzyme A[p] + ATP[p] + stearate[p] <=> AMP[p] + diphosphate[p] + stearoyl-CoA[p]

Changes:

```
target[grepl("Octadecan",target$NAME,ignore.case = TRUE),]
```


- Octodecan to stear


## other


```
target$NAME[grepl("nigrin",target$NAME)]
```


```
sum(grepl("\\[acyl-carrier protein\\]",target$NAME))
sum(grepl("anoyl",target$NAME))

sum(grepl("Cerebrin",target$NAME))
sum(grepl("Aminoacyl-tRNA",target$NAME))
```
