---
title: "Matching GEMs and BPGA output V2"
author: "Jeroen Gilis"
date: "8 October 2024"
output: 
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

```{r, message=FALSE, warning=FALSE}
library(stringr)
library(openxlsx)
library("kableExtra")
```

```{r}
read_excel_allsheets <- function(filename) {
    sheets <- openxlsx::getSheetNames(filename)
    x <- lapply(sheets, function(X) openxlsx::read.xlsx(filename, sheet = X))
    names(x) <- sheets
    x
}
```

# Load gems

```{r}
# Note: excels should be closed when reading them in R
niger_gem <- read_excel_allsheets("../data/BaseGEMs/Niger_GEM.xlsx")
fumigatus_gem <- read_excel_allsheets("../data/BaseGEMs/Fumigatus_GEM.xlsx")
oryzae_gem <- read_excel_allsheets("../data/BaseGEMs/oryzae_GEM.xlsx")
```

# BPGA

```{r}
BPGA <- read.csv2("../data/genome/BPGA2ortho_results.csv", 
                  sep=",",
                  na.strings = "")
head(BPGA)
```

For some reason we have a trailin .0 in the niger gene names, let's just remove
those for now

```{r}
BPGA$niger <- gsub("\\..*","",BPGA$niger)
table(BPGA$cluster_type)
```

Huge amount of uniques, which makes sense, because we used a huge amount of
input only for niger (all proteins vs. CDS).

That doesn't matter now. Main question remains: did we lose paralogs.
Or maybe more accurately: for all genes in a GEM, did we (1) have them in the
BPGA input fasta, and (2) have them in the BPGA results

# Fumigatus

```{r}
length(fumigatus_gem$GENES$NAME[which(!fumigatus_gem$GENES$NAME %in% unique(BPGA$fumigatus[!is.na(BPGA$fumigatus)]))])
# there are 146 gene IDs in the GEM that are not in the BPGA output

hlp <- read.csv2("../../fasta/fumigatus_hlp.faa",header=FALSE)
hlp$V1 <- gsub(">","",hlp$V1)
fumigatus_missing <- fumigatus_gem$GENES$NAME[which(!fumigatus_gem$GENES$NAME %in% unique(BPGA$fumigatus[!is.na(BPGA$fumigatus)]))]
table(fumigatus_missing %in% hlp$V1)
# all of these genes were in the fBPGA input
# so, they were likely removed because they are paralogs
```

# Oryzae

```{r}
length(oryzae_gem$GENES$NAME[which(!oryzae_gem$GENES$NAME %in% unique(BPGA$oryzae[!is.na(BPGA$oryzae)]))])
# there are 274 gene IDs in the GEM that are not in the BPGA output

hlp <- read.csv2("../../fasta/oryzae_hlp.faa",header=FALSE)
hlp$V1 <- gsub(">","",hlp$V1)
oryzae_missing <- oryzae_gem$GENES$NAME[which(!oryzae_gem$GENES$NAME %in% unique(BPGA$oryzae[!is.na(BPGA$oryzae)]))]
table(oryzae_missing %in% hlp$V1)

oryzae_missing[which(!oryzae_missing %in% hlp$V1)]
# 67 of these genes were also not in the BPGA input!
# but the other 207 genes were likely removed because they are paralogs

# e.g. "AO090001000480" is not in the protein fasta file (also not the raw one)

```

```{r}
length(niger_gem$GENES$NAME[which(!niger_gem$GENES$NAME %in% unique(BPGA$niger[!is.na(BPGA$niger)]))])
# for niger, there are 1187 gene IDs in the GEM that are not in the BPGA output

hlp <- read.csv2("../../fasta/niger_hlp.faa",header=FALSE)
hlp$V1 <- gsub(">","",hlp$V1)
niger_missing <- niger_gem$GENES$NAME[which(!niger_gem$GENES$NAME %in% unique(BPGA$niger[!is.na(BPGA$niger)]))]
table(niger_missing %in% hlp$V1)

niger_missing[which(!niger_missing %in% hlp$V1)]
# 49 of these genes were also not in the BPGA input!
# but the other 1138 genes were likely removed because they are paralogs
```

So essentially, we have two problems on our hands:
- BPGA input does not have all IDs that we have in our starting GEMS
- BPGA removes IDs (likely paralogs)

However, maybe the latter is not true. Maybe they are there, but we are simply
not extracting them from the output! These data could be u_clusters.txt

But to be fair, output handling should not be left to the BPGA user. Perhaps
USEARCH does a better job?





