---
title: "Matching GEMs and BPGA output V2"
author: "Jeroen Gilis"
date: "8 October 2024"
output: 
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

```{r, message=FALSE, warning=FALSE}
library(stringr)
library(openxlsx)
library(kableExtra)
library(phyloseq)
library(metagMisc)
```

```{r}
read_excel_allsheets <- function(filename) {
    sheets <- openxlsx::getSheetNames(filename)
    x <- lapply(sheets, function(X) openxlsx::read.xlsx(filename, sheet = X))
    names(x) <- sheets
    x
}
```

# Load gems

```{r}
# Note: excels should be closed when reading them in R
niger_gem <- read_excel_allsheets("../data/BaseGEMs/Niger_GEM.xlsx")
fumigatus_gem <- read_excel_allsheets("../data/BaseGEMs/Fumigatus_GEM.xlsx")
oryzae_gem <- read_excel_allsheets("../data/BaseGEMs/oryzae_GEM.xlsx")
```

# BPGA

```{r}
BPGA <- read.csv2("../data/genome/BPGA_self.csv", 
                  sep=",",
                  na.strings = "")
head(BPGA)
```

Huge amount of uniques, which makes sense, because we used a huge amount of
input only for niger (all proteins vs. CDS).

That doesn't matter now. Main question remains: did we lose paralogs?
Or maybe more accurately: for all genes in a GEM, did we (1) have them in the
BPGA input fasta, and (2) have them in the BPGA results

# Fumigatus

```{r}
length(fumigatus_gem$GENES$NAME[which(!fumigatus_gem$GENES$NAME %in% unique(BPGA$fumigatus[!is.na(BPGA$fumigatus)]))])
# there are 146 gene IDs in the GEM that are not in the BPGA output

hlp <- read.csv2("../../fasta/fumigatus_hlp.faa",header=FALSE)
hlp$V1 <- gsub(">","",hlp$V1)
fumigatus_missing <- fumigatus_gem$GENES$NAME[which(!fumigatus_gem$GENES$NAME %in% unique(BPGA$fumigatus[!is.na(BPGA$fumigatus)]))]
table(fumigatus_missing %in% hlp$V1)
# all of these genes were in the fBPGA input
# so, they were likely removed because they are paralogs
```

# Oryzae

```{r}
length(oryzae_gem$GENES$NAME[which(!oryzae_gem$GENES$NAME %in% unique(BPGA$oryzae[!is.na(BPGA$oryzae)]))])
# there are 274 gene IDs in the GEM that are not in the BPGA output

hlp <- read.csv2("../../fasta/oryzae_hlp.faa",header=FALSE)
hlp$V1 <- gsub(">","",hlp$V1)
oryzae_missing <- oryzae_gem$GENES$NAME[which(!oryzae_gem$GENES$NAME %in% unique(BPGA$oryzae[!is.na(BPGA$oryzae)]))]
table(oryzae_missing %in% hlp$V1)

oryzae_missing[which(!oryzae_missing %in% hlp$V1)]
# 67 of these genes were also not in the BPGA input!
# but the other 207 genes were likely removed because they are paralogs

# e.g. "AO090001000480" is not in the protein fasta file (also not the raw one)

```

```{r}
length(niger_gem$GENES$NAME[which(!niger_gem$GENES$NAME %in% unique(BPGA$niger[!is.na(BPGA$niger)]))])
# for niger, there are 1187 gene IDs in the GEM that are not in the BPGA output

hlp <- read.csv2("../../fasta/niger_hlp.faa",header=FALSE)
hlp$V1 <- gsub(">","",hlp$V1)
niger_missing <- niger_gem$GENES$NAME[which(!niger_gem$GENES$NAME %in% unique(BPGA$niger[!is.na(BPGA$niger)]))]
table(niger_missing %in% hlp$V1)

niger_missing[which(!niger_missing %in% hlp$V1)]
# 49 of these genes were also not in the BPGA input!
# but the other 1138 genes were likely removed because they are paralogs
```

So essentially, we have two problems on our hands:
- BPGA input does not have all IDs that we have in our starting GEMS
- BPGA removes IDs (likely paralogs)

However, maybe the latter is not true. Maybe they are there, but we are simply
not extracting them from the output! These data could be u_clusters.txt

But to be fair, output handling should not be left to the BPGA user. Perhaps
USEARCH does a better job?

# USEARCH

USEARCH clustering generates a .uc file, with contents similar to those in
u_clusters.txt from BPGA. I think all info we need on homologs/paralog is here.
We just need to be able to extract this. Here, I search for existing methods.
If these don't work, we'd need to resort to another way to determine 
orthologs/paralogs, e.g., using the github Casper sent, as writing our own code
for this (like Casper did) will bring us too far.

Also, check clustering ouput format here: https://drive5.com/usearch/manual/opt_uc.html

```{r}
ucfile <- "../../clusters.uc"
clustering <- import_usearch_uc(ucfile,verbose = TRUE,readDelimiter = "\t")
clustering@.Data[1:3,1:3]
```

```{r}
ncol(clustering) # number of clusters?
```

```{r}
hlp <- clustering[,1]
hlp[hlp!=0]
```

Does this mean that fumigatus gene 1|Org1_Gene1004|fumigatus.faa# has 
- one paralog in fumigatus
- many orthologs in niger
- two orthologs in oryzae?

```{r}
table(unname(colSums(clustering@.Data)))
```

What is a one? 

```{r}
hlp <- clustering[,19]
hlp[hlp!=0]
```

So the import seems to only retain hits?  Query this gene in the original file.
-> can confirm, the .uc file has these two as only hit for each other. They are
orthologs without paralogs.


```{r}
hlp <- clustering[,"1|Org1_Gene1083|fumigatus.faa#"]
hlp[hlp!=0]
```

So now I need to wrangle this table to something more similar to what Casper 
has. 


```
# Save this function in case the github dies
metagMisc::parse_uc
function (x, map_only = FALSE, splitSeqID = TRUE, rm_dups = TRUE) 
{
    ii <- fread(file = x, header = FALSE, sep = "\t")
    ii <- ii[!V1 %in% "S"]
    ii[, `:=`(Query, tstrsplit(V9, ";", keep = 1))]
    ii[, `:=`(OTU, tstrsplit(V10, ";", keep = 1))]
    ii[V1 %in% "C", `:=`(OTU, Query)]
    if (nrow(ii[, .(Query, OTU)]) != nrow(unique(ii[, .(Query, 
        OTU)]))) {
        cat("Warning: duplicated rows found!\n")
        if (rm_dups == TRUE) {
            dups <- duplicated(ii[, .(Query, OTU)])
            cat("..", sum(dups), " duplicates removed\n")
            ii <- ii[!dups]
        }
    }
    setnames(x = ii, old = paste0("V", 1:10), new = c("recordType", 
        "clusterNumber", "seqLengthOrClusterSize", "percentIdentity", 
        "strand", "unusedField1", "unusedField2", "alignment", 
        "queryLabel", "targetLabel"), skip_absent = TRUE)
    ii[percentIdentity %in% "*", `:=`(percentIdentity, NA)]
    ii[, `:=`(percentIdentity, as.numeric(percentIdentity))]
    if (map_only == TRUE) {
        if (splitSeqID == TRUE) {
            ii <- ii[, .(Query, OTU)]
        }
        else {
            ii <- ii[, .(queryLabel, targetLabel)]
            setnames(x = ii, old = c("queryLabel", "targetLabel"), 
                new = c("Query", "OTU"))
        }
    }
    return(ii)
}
```

```{r}
hlp <- metagMisc::parse_uc(x=ucfile, map_only = FALSE, splitSeqID = TRUE, rm_dups = TRUE)

dim(hlp)
```

```{r}
table(hlp$recordType)
```

```{r}
hlp[1:10,]
```

```{r}
hlp[hlp$OTU == "1|Org1_Gene1004|fumigatus.faa#",]
```

```{r}
hlp[hlp$clusterNumber == 323,]
```

What I think we want is a dataframe, where each row is a cluster clusterNumber,
and we have column for each species, and their cells are a list of compatible 
IDs.

```{r}
res <- do.call(rbind,lapply(unique(hlp$clusterNumber), function(x){
   test <- hlp[hlp$clusterNumber == x,]
   data.frame(cluster = x,
              Afumigatus = toString(test$Query[grep("fumigatus",test$Query)]),
              Aoryzae = toString(test$Query[grep("oryzae",test$Query)]),
              Aniger = toString(test$Query[grep("niger",test$Query)]))
}))

head(res)
```

Maybe this is the most convenient structure for e.g. Casper to work with as you
could also look at this in excel, but honestly it would be much cleaner to work
with a named list where the new clusters are the names, and entries fumigatus,
niger and oryzae, and maybe some additional info like core/accessory/unique and
n_genomes.

TODO: map IDs (1|Org1_Gene9442|fumigatus.faa#) to original gene names.
TODO: ask Capser why there is a discrepancy between his cluster_type_manual and
BPGA's cluster_type_BPGA. I see how the former is computed, but not the latter.

