---
title: "Matching GEMs and BPGA output Jeroen"
author: "Jeroen Gilis"
date: "17 October 2024"
output: 
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

```{r, message=FALSE, warning=FALSE}
library(stringr)
library(openxlsx)
library(kableExtra)
library(phyloseq)
library(metagMisc)
```

```{r}
read_excel_allsheets <- function(filename) {
    sheets <- openxlsx::getSheetNames(filename)
    x <- lapply(sheets, function(X) openxlsx::read.xlsx(filename, sheet = X))
    names(x) <- sheets
    x
}
```

# Load gems

```{r}
# Note: excels should be closed when reading them in R
niger_gem <- read_excel_allsheets("../data/BaseGEMs/Niger_GEM.xlsx")
fumigatus_gem <- read_excel_allsheets("../data/BaseGEMs/Fumigatus_GEM.xlsx")
oryzae_gem <- read_excel_allsheets("../data/BaseGEMs/oryzae_GEM.xlsx")
```

# Load clustering output

In step 2 of their workflow, BPGA generates a file `u_clusters.txt` that 
contains the clustering information for each sequence in the BPGA input, i.e.,
containing both orthologs and paralogs. In USEARCH, the same file is also
generate, with extension `.uc`. The ouput format is described here:
https://drive5.com/usearch/manual/opt_uc.html

```
# Save this function in case the github dies
metagMisc::parse_uc
function (x, map_only = FALSE, splitSeqID = TRUE, rm_dups = TRUE) 
{
    ii <- fread(file = x, header = FALSE, sep = "\t")
    ii <- ii[!V1 %in% "S"]
    ii[, `:=`(Query, tstrsplit(V9, ";", keep = 1))]
    ii[, `:=`(OTU, tstrsplit(V10, ";", keep = 1))]
    ii[V1 %in% "C", `:=`(OTU, Query)]
    if (nrow(ii[, .(Query, OTU)]) != nrow(unique(ii[, .(Query, 
        OTU)]))) {
        cat("Warning: duplicated rows found!\n")
        if (rm_dups == TRUE) {
            dups <- duplicated(ii[, .(Query, OTU)])
            cat("..", sum(dups), " duplicates removed\n")
            ii <- ii[!dups]
        }
    }
    setnames(x = ii, old = paste0("V", 1:10), new = c("recordType", 
        "clusterNumber", "seqLengthOrClusterSize", "percentIdentity", 
        "strand", "unusedField1", "unusedField2", "alignment", 
        "queryLabel", "targetLabel"), skip_absent = TRUE)
    ii[percentIdentity %in% "*", `:=`(percentIdentity, NA)]
    ii[, `:=`(percentIdentity, as.numeric(percentIdentity))]
    if (map_only == TRUE) {
        if (splitSeqID == TRUE) {
            ii <- ii[, .(Query, OTU)]
        }
        else {
            ii <- ii[, .(queryLabel, targetLabel)]
            setnames(x = ii, old = c("queryLabel", "targetLabel"), 
                new = c("Query", "OTU"))
        }
    }
    return(ii)
}
```

```{r}
clustfile <- "../../BPGA/OutputJeroen/u_clusters.txt"
clustering <- metagMisc::parse_uc(x=clustfile, 
                                  map_only = FALSE, 
                                  splitSeqID = TRUE, 
                                  rm_dups = TRUE)
head(clustering)
```

# Translate IDs

TODO: map IDs (1|Org1_Gene9442|fumigatus.faa#) to original gene names.

BPGA stores this information in INPUT_all_original.fasta
Note that this step may not be required if you use USEARCH separately, and not
within BPGA. There, it just depends on whether you changed the names of the
FASTA headers.

```{r}
dict_fasta <- read.csv2("../../BPGA/OutputJeroen/INPUT_all_original.fasta", 
                        sep=">", header=FALSE)
dict_fasta$V1 <- NULL
colnames(dict_fasta) <- c("BPGA","original")

dict_fasta$BPGA <- sub("\\|$", "#", dict_fasta$BPGA)
dict_fasta$BPGA <- paste0(sub("\\|.*", "", dict_fasta$BPGA),
                          "|Org",
                          sub("\\|.*", "", dict_fasta$BPGA),
                          "_",
                          sub("^[^\\|]+\\|","",dict_fasta$BPGA))
rbind(head(dict_fasta,n=3),dict_fasta[15000:15002,],tail(dict_fasta,n=3))
```

But then for A. niger, we need an additional translation, from proper IDs to 
transcript numbers used for the GEM.

```{r}
awk_cmd <- sprintf("awk '$1 ~ /^>/' %s", "../../fasta/keep_Aspni7_all_transcripts_20131226.nt.fasta")
full_IDs <- readLines(pipe(awk_cmd))

tx_prot_names <- sub(".*\\|", "", full_IDs)
tx_num <- sub("^([^\\|]*\\|){2}([^\\|]+).*", "\\2", full_IDs)

dict_niger <- data.frame(tx_prot_names = tx_prot_names,
                         tx_num = tx_num)
head(dict_niger)
```

```{r}
target <- dict_fasta[sub(".*\\|(.*)", "\\1", dict_fasta$BPGA) == "niger.faa#",]
target$GEM <- dict_niger$tx_num[match(target$original, dict_niger$tx_prot_names)]
target[is.na(target$GEM),]

# So these are IDs that we have in the protein fasta, but that we can't link
# back to transcript numbers, and hence cant link back to the GEM.
lost <- target$BPGA[is.na(target$GEM)]

hlp[hlp$targetLabel %in% lost,]
hlp[hlp$Query %in% lost,]

# these IDs didnt receive many hits in the clustering either
```

```{r}
dict_fasta[sub(".*\\|(.*)", "\\1", dict_fasta$BPGA) == "niger.faa#","original"] <- target$GEM
dict_fasta$original[is.na(dict_fasta$original)] <- target$original[match(dict_fasta$BPGA[is.na(dict_fasta$original)],target$BPGA)]
rbind(head(dict_fasta,n=3),dict_fasta[15000:15002,],tail(dict_fasta,n=3))
```

Now translate all. Actually, this is much easier to do on the raw clustering
output, so lets do that. Actually, the only columns Im using from this output
are clusterNumber and query. So it suffices to translate query.

# Wrangle clustering output

```{r}
# eaier to store organism fasta name in separate column
hlp$fasta <- sub(".*\\|(.*)", "\\1", hlp$Query)
hlp$Query <- dict_fasta$original[match(hlp$Query, dict_fasta$BPGA)]

res <- do.call(rbind,lapply(unique(hlp$clusterNumber), function(x){
   test <- hlp[hlp$clusterNumber == x,]
   data.frame(cluster = x,
              Afumigatus = paste0(test$Query[grep("fumigatus",test$fasta)], collapse = ";"),
              Aoryzae = paste0(test$Query[grep("oryzae",test$fasta)], collapse = ";"),
              Aniger = paste0(test$Query[grep("niger",test$fasta)], collapse = ";"))
}))

res[res==""]<-NA
res$present_in_n_genomes <- rowSums(!is.na(res[,c("Afumigatus",
                                                  "Aoryzae",
                                                  "Aniger")]))
res$cluster_type_manual <- ""
res$cluster_type_manual[res$present_in_n_genomes == 1] <- "singleton"
res$cluster_type_manual[res$present_in_n_genomes == 2] <- "accessory"
res$cluster_type_manual[res$present_in_n_genomes == 3] <- "core"
res <- res[,c("cluster", "present_in_n_genomes", "cluster_type_manual",
              "Afumigatus","Aoryzae","Aniger")]
```

Now we only have the issue that some genes that are in the GEMs are not in this
table, because they were not in the BPGA input, because we don't have a protein
sequence for these features (and this can also not be readily obtained).

For Aniger this is one feature (Unknown5)
For Aoryzae this are 46 features (all ZY entries)

To facilitate downstream analyses, I will add these entries to res myself,
and assign them cluster_type_manual "artificial"

```{r}
unique(oryzae_gem$GENES$NAME)[!unique(oryzae_gem$GENES$NAME) %in% unique(unlist(strsplit(res$Aoryzae, split=";")))]
add_niger <- data.frame(cluster = max(res$cluster)+1,
                        present_in_n_genomes = 1,
                        cluster_type_manual = "artificial",
                        Afumigatus = NA,
                        Aoryzae = NA,
                        Aniger = "Unknown5")
res <- rbind(res, add_niger)

missing_oryzae <- unique(oryzae_gem$GENES$NAME)[!unique(oryzae_gem$GENES$NAME) %in% unique(unlist(strsplit(res$Aoryzae, split=";")))]

add_oryzae <- data.frame(cluster = max(res$cluster)+seq_along(missing_oryzae),
                         present_in_n_genomes = 1,
                        cluster_type_manual = "artificial",
                        Afumigatus = NA,
                        Aoryzae = missing_oryzae,
                        Aniger = NA)

res <- rbind(res,add_oryzae)
tail(res)
```

# Export final table

```{r}
res[is.na(res)]<-""
write.csv2(res, "../data/genome/BPGA2ortho_GEM_custom.csv")
```


# Appendix: missing Oryzae genes

```{r}
missing_oryzae
```

These IDs are present in the Oryzae GEM, but not in the FASTA input. These IDs
refer to ESTs, which are short cDNA sequences (RNA -> reverse ranscribed).
While I can retrieve a nucleotide sequence for the EST, it is not directly
possible to go to a protein sequence, as the link EST-gene-protein has not been
established. As such, I don't think there is much more we can do for these
IDs. However, below I have looked into each of the ZYs:

- 23 are isoenzy, we already have a properly characterized gene in Oryzae doing the same thing
- 15 are subunits of an enzyme complex, together with genes that are properly characterized
- 8 are truly unique, catalyzing a reaction no other gene is involved with

---

- Chr4_SC012_1076984__1077358: N-acetyl-y-glutamyl-phosphate reductase
Not in sup_table1, but same nomenclature a the ZYs, 
"Chromosome number with EST alignment" - "Contig name with EST alignment" - "EST position (Start-Stop)"

---

- ZY007107: subunit of Cytochrome c oxidase, together with
AO090010000481 and AO090026000651 and AO090026000421 and ZY007619 and ZY029203 and ZY029342
Looking at the other ZYs, they only appear here as subunit of this enzyme 
Looking at the AOs, they only appear here as subunit of this enzyme 

- ZY006639: subunit NADH dehydrogenase (ubiquinone, both mitochondrial and cytoplasm), together with
AO090003001115 and AO090011000782 and AO090001000553 and AO090026000229 and AO090001000661 and AO090011000578 and AO090003001313 and AO090102000645 and AO090012000402 and AO090011000502 and AO090103000199 and AO090672000005 and AO090001000459 and AO090001000300 and AO090003000400 and ZY006639 and ZY081278 and ZY083531 and ZY087692
Looking at the other ZYs, they only appear here as subunit of this enzyme 
Looking at the AOs, they also appear as subunit of NADH dehydrogenase (FAD-containing subunit)

- ZY007427: subunit of Ubiquinol-cytochrome c reductase, together with
AO090010000475 and AO090102000625, which also only appear as subunit of this enzyme

- ZY029275: subunit of Succinate dehydrogenase (flavoprotein subunit), together with
AO090020000415 and AO090020000596 and AO090010000505 
Looking at the AOs, they only appear here as subunit of this enzyme 

- ZY036719: subunit of F-type ATPase complex, together with
AO090005000617 and AO090005000749 and AO090005000604 and AO090026000372 and AO090010000482 and AO090701000168 and AO090206000115 and AO090010000482 and AO090003000607 and AO090023000599 

and subunit of V-type ATPase complex, together with
AO090102000349 and AO090038000334 and AO090026000283 and AO090120000313 and AO090120000091 and AO090003001090 and AO090012000797 and AO090026000517 and AO090038000579 and AO090038000352 and AO090012000274

- ZY077810: subunit of acyl-Coa dehydrogenase

For acyl %in% c(steaoryl, palmitoyl, myristoyl, lauroyl, decanoyl and octanoyl), other subunits are
AO090005000521 and AO090009000596 and AO090102000172 and AO090012000150 and AO090005000494 and AO090023000711 

For acyl %in% c(hexanoyl and butyrylyl), other subunits are
AO090103000106

- ZY083011: subunit of NADH dehydrogenase (FAD-containing subunit), together with
AO090003001115 and AO090011000782 and AO090001000553 and AO090026000229 and AO090001000661 and AO090011000578 and AO090003001313 and AO090102000645 and AO090012000402 and AO090011000502 and AO090103000199 and AO090672000005 and AO090001000459 and AO090001000300 and AO090003000400 and AO090001000657 and AO090011000358

- ZY098176: subunit of Acetyl-CoA carboxylase, together with 
AO090011000838 

- ZY098280: subunit of Acetolacetate synthase, together with 
AO090023000854 and AO090003000571 and AO090166000076

---

Unique, no other proteins catalyzing this reaction: 

- ZY006430: Phosphatidylinositol-3,4,5-trisphosphate 3-phosphatase

- ZY080692: P-hydroxybenzoate hydroxylase

- ZY088142: Di-trans,poly-cis-decaprenylcistransferase

- ZY097848: Methylglyoxal synthase

- ZY104986: Polyphosphate kinase

- ZY110993: Alanine racemase

- ZY111007: Scytalone dehydratase

- ZY140875: Selenocysteine synthase

---

! Im pretty sure the enzyme for r84 should be AO090001000642 in stead of AO090001000484

```{r}
isoenzymes <- list("ZY006677" = "AO090026000021",
                   "ZY007112", "AO090026000800",
                   "ZY080376" = c("AO090011000353", "AO090701000542", "AO090012000766", 
                                  "AO090102000588", "AO090005000837", "AO090026000395", 
                                  "AO090120000164", "AO090003001432", "AO090103000036",
                                  "AO090001000143", "AO090701000644"),
                   "ZY111292" = c("AO090011000353", "AO090701000542", "AO090012000766", 
                                  "AO090102000588", "AO090005000837", "AO090026000395", 
                                  "AO090120000164", "AO090003001432", "AO090103000036",
                                  "AO090001000143", "AO090701000644"),
                   "ZY029856" = "AO090005001098",
                   "ZY007726" = c("AO090020000670", "AO090009000711", "AO090001000641",
                                  "AO090005001098"),
                   "ZY083916" = c("AO090020000670", "AO090009000711","AO090020000668",
                                  "AO090001000641"),
                   "ZY029184" = c("AO090005001370","AO090011000449","AO090012000084",
                                  "AO090206000079","AO090701000589","AO090005001370",
                                  "AO090005001371","AO090113000128","AO090005000579",
                                  "AO090026000212","AO090026000321","AO090026000323"),
                   "ZY077820" = c("AO090005001300","AO090120000145","AO090011000152"),
                   "ZY085753" = c("AO090005000045", "AO090001000484","AO090001000642",
                                  "AO090003000415"),
                   "ZY111420" = c("AO090026000498","AO090026000149"), 
                   "ZY086694" = c("AO090026000498","AO090026000149"),
                   "ZY087055" = c("AO090001000162","AO090023000467","AO090009000417"),
                   "ZY087096" = c("AO090003000631","AO090005000973","AO090103000149",
                                  "AO090103000485","AO090102000005","AO090003001039",
                                  "AO090103000134","AO090102000478","AO090012000378",
                                  "AO090012000112","AO090023000353"),
                   "ZY087286" = "AO090012000998",
                   "ZY087560" = c("AO090038000057","AO090038000010"),
                   "ZY093903" = c("AO090038000057","AO090038000010"),
                   "ZY087836" = "AO090011000561",
                   "ZY097910" = c("AO090005001216","AO090026000084"),
                   "ZY110723" = c("AO090005001124","AO090012000733"),
                   "ZY111104" = c("AO090005000433","AO090102000204","AO090020000034"),
                   "ZY135516" = "AO090701000187")
```

Note: the oryzae GEM has no "or" gene associations, they are cataloged as
separate reactions.




TODO: ask Capser why there is a discrepancy between his cluster_type_manual and
BPGA's cluster_type_BPGA. I see how the former is computed, but not the latter.
