---
title: "Matching GEMs and BPGA output V2"
author: "Jeroen Gilis"
date: "8 October 2024"
output: 
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

```{r, message=FALSE, warning=FALSE}
library(stringr)
library(openxlsx)
library(kableExtra)
library(phyloseq)
library(metagMisc)
```

```{r}
read_excel_allsheets <- function(filename) {
    sheets <- openxlsx::getSheetNames(filename)
    x <- lapply(sheets, function(X) openxlsx::read.xlsx(filename, sheet = X))
    names(x) <- sheets
    x
}
```

# Load gems

```{r}
# Note: excels should be closed when reading them in R
niger_gem <- read_excel_allsheets("../data/BaseGEMs/Niger_GEM.xlsx")
fumigatus_gem <- read_excel_allsheets("../data/BaseGEMs/Fumigatus_GEM.xlsx")
oryzae_gem <- read_excel_allsheets("../data/BaseGEMs/oryzae_GEM.xlsx")
```

# BPGA

```{r}
BPGA <- read.csv2("../data/genome/BPGA_self.csv", 
                  sep=",",
                  na.strings = "")

BPGA$niger <- gsub("ANIG_","",BPGA$niger)

head(BPGA)
```

1	0	unique	AFUA_5G12730	NA	NA
2	1	unique	NA	1186152	NA
3	2	unique	NA	NA	AO090011000043
4	3	core	AFUA_1G10380	1113533	AO090038000390
5	4	unique	NA	1163248	NA
6	5	unique	NA	1211107	NA

Huge amount of uniques, which makes sense, because we used a huge amount of
input only for niger (all proteins vs. CDS).

That doesn't matter now. Main question remains: did we lose paralogs?
Or maybe more accurately: for all genes in a GEM, did we (1) have them in the
BPGA input fasta, and (2) have them in the BPGA results

# Fumigatus

Im am very confused.
Where are genes AFUA_1G15440 and AFUA_3G00910?? -> removed by BPGA. They are 
still in `u_clusters.txt`, but not in the downstream sequence files

```{r}
length(fumigatus_gem$GENES$NAME[which(!fumigatus_gem$GENES$NAME %in% unique(BPGA$fumigatus[!is.na(BPGA$fumigatus)]))])
# there are 146 gene IDs in the GEM that are not in the BPGA output

hlp <- read.csv2("../../fasta/fumigatus_hlp.faa",header=FALSE)
hlp$V1 <- gsub(">","",hlp$V1)
fumigatus_missing <- fumigatus_gem$GENES$NAME[which(!fumigatus_gem$GENES$NAME %in% unique(BPGA$fumigatus[!is.na(BPGA$fumigatus)]))]
table(fumigatus_missing %in% hlp$V1)
# all of these genes were in the BPGA input
# so, they were likely removed because they are paralogs
```

```{r}
table(unique(fumigatus_gem$GENES$NAME) %in% unique(hlp$V1))
```

# Oryzae

```{r}
length(oryzae_gem$GENES$NAME[which(!oryzae_gem$GENES$NAME %in% unique(BPGA$oryzae[!is.na(BPGA$oryzae)]))])
# there are 256 gene IDs in the GEM that are not in the BPGA output

hlp <- read.csv2("../../fasta/oryzae_hlp.faa",header=FALSE)
hlp$V1 <- gsub(">","",hlp$V1)
oryzae_missing <- oryzae_gem$GENES$NAME[which(!oryzae_gem$GENES$NAME %in% unique(BPGA$oryzae[!is.na(BPGA$oryzae)]))]
table(oryzae_missing %in% hlp$V1)

oryzae_missing[which(!oryzae_missing %in% hlp$V1)]
# 46 of these genes were also not in the BPGA input!
# but the other 210 genes were likely removed because they are paralogs
```

---

- Chr4_SC012_1076984__1077358: N-acetyl-y-glutamyl-phosphate reductase
Not in sup_table1, but same nomenclature a the ZYs, 
"Chromosome number with EST alignment" - "Contig name with EST alignment" - "EST position (Start-Stop)"

- ZY007107: subunit of Cytochrome c oxidase, together with
AO090010000481 and AO090026000651 and AO090026000421 and ZY007107 and ZY007619 and ZY029203 and ZY029342
Looking at the other ZYs, they only appear here as subunit of this enzyme 
Looking at the AOs, they only appear here as subunit of this enzyme 

- ZY006639: subunit NADH dehydrogenase (ubiquinone, both mitochondrial and cytoplasm), together with
AO090003001115 and AO090011000782 and AO090001000553 and AO090026000229 and AO090001000661 and AO090011000578 and AO090003001313 and AO090102000645 and AO090012000402 and AO090011000502 and AO090103000199 and AO090672000005 and AO090001000459 and AO090001000300 and AO090003000400 and ZY006639 and ZY081278 and ZY083531 and ZY087692
Looking at the other ZYs, they only appear here as subunit of this enzyme 
Looking at the AOs, they also appear as subunit of NADH dehydrogenase (FAD-containing subunit)

- ZY007427: subunit of Ubiquinol-cytochrome c reductase, together with
AO090010000475 and AO090102000625, which also only appear as subunit of this enzyme

- ZY029275: subunit of Succinate dehydrogenase (flavoprotein subunit), together with
AO090020000415 and AO090020000596 and AO090010000505 
Looking at the AOs, they only appear here as subunit of this enzyme 

- ZY036719: subunit of F-type ATPase complex, together with
AO090005000617 and AO090005000749 and AO090005000604 and AO090026000372 and AO090010000482 and AO090701000168 and AO090206000115 and AO090010000482 and AO090003000607 and AO090023000599 

and subunit of V-type ATPase complex, together with
AO090102000349 and AO090038000334 and AO090026000283 and AO090120000313 and AO090120000091 and AO090003001090 and AO090012000797 and AO090026000517 and AO090038000579 and AO090038000352 and AO090012000274

- ZY077810: subunit of acyl-Coa dehydrogenase

For acyl %in% c(steaoryl, palmitoyl, myristoyl, lauroyl, decanoyl and octanoyl), other subunits are
AO090005000521 and AO090009000596 and AO090102000172 and AO090012000150 and AO090005000494 and AO090023000711 

For acyl %in% c(hexanoyl and butyrylyl), other subunits are
AO090103000106

- ZY083011: subunit of NADH dehydrogenase (FAD-containing subunit), together with
AO090003001115 and AO090011000782 and AO090001000553 and AO090026000229 and AO090001000661 and AO090011000578 and AO090003001313 and AO090102000645 and AO090012000402 and AO090011000502 and AO090103000199 and AO090672000005 and AO090001000459 and AO090001000300 and AO090003000400 and AO090001000657 and AO090011000358

- ZY098176: subunit of Acetyl-CoA carboxylase, together with 
AO090011000838 

- ZY098280: subunit of Acetolacetate synthase, together with 
AO090023000854 and AO090003000571 and AO090166000076

---

Unique, no other proteins catalyzing this reaction: 

- ZY006430: Phosphatidylinositol-3,4,5-trisphosphate 3-phosphatase

- ZY080692: P-hydroxybenzoate hydroxylase

- ZY088142: Di-trans,poly-cis-decaprenylcistransferase

- ZY097848: Methylglyoxal synthase

- ZY104986: Polyphosphate kinase

- ZY110993: Alanine racemase

- ZY111007: Scytalone dehydratase

- ZY140875: Selenocysteine synthase

---

! Im pretty sure the enzyme for r84 should be AO090001000642 in stead of AO090001000484

```{r}
isoenzymes <- list("ZY006677" = "AO090026000021",
                   "ZY007112", "AO090026000800",
                   "ZY080376" = c("AO090011000353", "AO090701000542", "AO090012000766", 
                                  "AO090102000588", "AO090005000837", "AO090026000395", 
                                  "AO090120000164", "AO090003001432", "AO090103000036",
                                  "AO090001000143", "AO090701000644"),
                   "ZY111292" = c("AO090011000353", "AO090701000542", "AO090012000766", 
                                  "AO090102000588", "AO090005000837", "AO090026000395", 
                                  "AO090120000164", "AO090003001432", "AO090103000036",
                                  "AO090001000143", "AO090701000644"),
                   "ZY029856" = "AO090005001098",
                   "ZY007726" = c("AO090020000670", "AO090009000711", "AO090001000641",
                                  "AO090005001098"),
                   "ZY083916" = c("AO090020000670", "AO090009000711","AO090020000668",
                                  "AO090001000641"),
                   "ZY029184" = c("AO090005001370","AO090011000449","AO090012000084",
                                  "AO090206000079","AO090701000589","AO090005001370",
                                  "AO090005001371","AO090113000128","AO090005000579",
                                  "AO090026000212","AO090026000321","AO090026000323"),
                   "ZY077820" = c("AO090005001300","AO090120000145","AO090011000152"),
                   "ZY085753" = c("AO090005000045", "AO090001000484","AO090001000642",
                                  "AO090003000415"),
                   "ZY111420" = c("AO090026000498","AO090026000149"), 
                   "ZY086694" = c("AO090026000498","AO090026000149"),
                   "ZY087055" = c("AO090001000162","AO090023000467","AO090009000417"),
                   "ZY087096" = c("AO090003000631","AO090005000973","AO090103000149",
                                  "AO090103000485","AO090102000005","AO090003001039",
                                  "AO090103000134","AO090102000478","AO090012000378",
                                  "AO090012000112","AO090023000353"),
                   "ZY087286" = "AO090012000998",
                   "ZY087560" = c("AO090038000057","AO090038000010"),
                   "ZY093903" = c("AO090038000057","AO090038000010"),
                   "ZY087836" = "AO090011000561",
                   "ZY097910" = c("AO090005001216","AO090026000084"),
                   "ZY110723" = c("AO090005001124","AO090012000733"),
                   "ZY111104" = c("AO090005000433","AO090102000204","AO090020000034"),
                   "ZY135516" = "AO090701000187")
```

Note: the oryzae GEM has no "or" gene associations, they are cataloged as
separate reactions.

# niger

For niger, we used inputs like "estExt_Genemark1.C_chr_5030003"
However, the GEM IDs are transcript IDs, which are 6-digit numbers.
Here, I generate a translation table.

```{r}
awk_cmd <- sprintf("awk '$1 ~ /^>/' %s", "../../fasta/keep_Aspni7_all_transcripts_20131226.nt.fasta")
full_IDs <- readLines(pipe(awk_cmd))

tx_prot_names <- sub(".*\\|", "", full_IDs)
tx_IDs <- sub("^([^\\|]*\\|){2}([^\\|]+).*", "\\2", full_IDs)

dict_niger <- data.frame(tx_prot_names = tx_prot_names,
                         tx_IDs = tx_IDs)
head(dict_niger)
```

```{r}
table(unique(BPGA$niger[!is.na(BPGA$niger)]) %in% dict_niger$tx_prot_names)
```

```{r}
target <- unique(BPGA$niger[!is.na(BPGA$niger)])
target[!target %in% dict_niger$tx_prot_names]
```

```{r}
lost <- target[!target %in% dict_niger$tx_prot_names]
BPGA[BPGA$niger %in% lost,]
```

I think it's fair to say that it doesn't matter too much we loose these
clusters.

```{r}
BPGA <- BPGA[!BPGA$niger %in% lost,] # remove A. niger clusters for which we
# wouldnt be able to link to the GEM anyway.

BPGA$niger_txID <- dict_niger$tx_IDs[match(BPGA$niger, dict_niger$tx_prot_names)]
BPGA
```

```{r}
table(niger_gem$GENES$NAME %in% unique(BPGA$niger_txID[!is.na(BPGA$niger_txID)]))
# for niger, there are 1215 gene IDs in the GEM that are not in the BPGA output

hlp <- read.csv2("../../fasta/niger_hlp.faa",header=FALSE)
hlp$V1 <- gsub(">","",hlp$V1)
hlp$V2 <- dict_niger$tx_IDs[match(hlp$V1,dict_niger$tx_prot_names)]


niger_missing <- niger_gem$GENES$NAME[which(!niger_gem$GENES$NAME %in% unique(BPGA$niger[!is.na(BPGA$niger)]))]
table(niger_missing %in% hlp$V2)

niger_missing[which(!niger_missing %in% hlp$V2)]
# 1 of these genes were also not in the BPGA input!
# but the other 1214 genes were likely removed because they are paralogs
```

Nothing we can do about "unknown5", it's probably a placeholder for an enzyme
that has to exist but that has not been identified yet.

It is this enzyme in Ecoli https://www.uniprot.org/uniprotkb/Q46857/entry

## old worries

All the worries below no longer apply.
But I keep the code and text just be sure for now.

Transcripts

All these IDs are found back in the "transcripts" from jgi, but I dont know what "MIX" refers to.
But note that many MIX and estExt are present in the protein fasta, just not these 49.

1218960: >jgi|Aspni7|1218960|MIX18557_1_4
1189342: >jgi|Aspni7|1189342|estExt_Genemark1.C_chr_802_t20252
1208460: >jgi|Aspni7|1208460|MIX8057_104_71
1202623: >jgi|Aspni7|1202623|MIX2220_5_45

Isozymes:


```
target <- niger_missing[which(!niger_missing %in% hlp$V1)]
c("1148603","1175567","1218960","1181034") %in% target
# while 1218960 isnt in the input, its potential isozymes are in the input
```

Do for all

```
has_isozyme <- unlist(lapply(target, function(id){
  hlp <- niger_gem$RXNS$GENE.ASSOCIATION[grepl(id, niger_gem$RXNS$GENE.ASSOCIATION)]
  hlp <- gsub(" ", "",hlp)
  hlp <- str_split(hlp, "or")
  if(all(lengths(hlp) > 1)){
    return(TRUE)
  } else{
    return(FALSE)
  }
}))
table(has_isozyme)
target[!has_isozyme]
```

```
do.call(rbind,lapply(target[!has_isozyme], function(id){
  return(niger_gem$RXNS[grepl(id, niger_gem$RXNS$GENE.ASSOCIATION),c("NAME","EQUATION","GENE.ASSOCIATION")])
}))
```


- 1208460: is Acetyl-CoA synthase (mitochondria)
Acetyl-CoA synthase for cytosol is present, ID = 1146906. But different sequence.

https://www.uniprot.org/uniprotkb?query=%28taxonomy_id%3A380704%29+AND+%22Acetyl-Coenzyme+A+synthetase%22
There are four in uniprot, but how to link to my "ugly" transcript ID :/

If I throw the transcript sequence: 

>jgi|Aspni7|1208460|MIX8057_104_71
ATGTCACTTCACCCGCAGCAGTTGGTCCACAACACCTCCCTGCGCGACCCAGAAGGCTTCTGGTCTCGCC
AGGCCGAACAACTGCACTGGCACAAAAAACCATCGCGTGCCATCGACCAGCAATTGAAAACAACCCCCAG
TGGCTCCAGCTATACTCATTGGAGCTGGTTCCCGGATGGGGAGATCTCGACGACCTACAACTGTGTCGAT
CGGCATGTCCATTCAGGCAAGGGAGATAATGTTGCTATCATCTGGGAGTCCCCGGTGACAGGAACGAAGG
AGAAATATACCTATAAACAGCTATTGGACGAAGTCGAGGTTCTAGCAGGCGTGCTAAGAGAAGAAGGTGT
GCGAAAGGGCGACGTGGTTATTATTTATATGCCTATGATTCCTGCGGCACTTATTGCTGCTTTGGCAATT
GCACGCCTGGGTGCTATTCATTCCGCAGTCTTTGGCGGGTTCGCAGCCAAGTCGCTCGCACAACGCATTG
AAGCGGCCAAGCCACGCGTCCTCATGACAGCGTCGTGCGGTATCGAAGGCTCAAAAGGTCCCATCGCCTA
TCGACCGTTAGTTGAGGGTGCTATCGAGCTGAGCCACTTCAAGCCAGAGAAAGTGATCGTCTGGCAGAGG
GATCAACTGCGGTGGAATAACCCCGATAAGTTGGCTGGACAACGTAACTGGCAACGACTGGTGAAAAGCG
CACGCATGCGTGGCATAAGAGCCGAGCCGGTACCCGTGAAGAGCACGGATGGACTATATATCATCTATAC
TAGCGGCACTACGGGCTCGCCCAAAGGTGTGTTTCGAGAGGCCGGTGGACACGCAGTGGGACTCCAGCTG
TCGATGAAACACCTATTTGATATACAAGGGCCCGGGGACGTCATCTTCTGCGCCTCTGACATTGGATGGG
TGGTCGGTCATTCGTACATCTTGTATGGGCCACTCCTGGTTGGAGCAACGACAATCCTTTTCGAAGGAAA
GCCGGTGGGGACACCGGACGCAGGCACCTTCTGGCGGATTATCGAGGAGCATAAGGCAAATGTGTTGTTC
ACGGCGCCCACGGCCATGCGGGCCATCCGCAAGGATGATCCCAACAATCGCTTCTTTGAAGAGGTGGCTC
GGCGTGGCGGCCTCCAACATTTCCGCGCACTGTTTCTTGCGGGAGAAAGGAGCGAACCCAGTATTGTGCA
AGCCTACCAAGAACTTCTGAGCCAATACGCTGCTCCCGGTGCAGTGGTTGTGGACAATTGGTGGTCGTCA
GAATCAGGGTCTCCCATATCTGGGCTGGCACTTAGAAGCGCTGCGGCAATGGAGTCGTCCTCGCGTCAGA
ATGATGCAGTGAAGCCATTGACCGTTCGACCAGGATCAGCAGGTCTGCCCATGCCTGGATTTGACGTGCG
CGTGGTCGACGACGAGGGAAAGGCACTGCCTCCGGGCACGATGGGGAACATTGTCTTAGGCATTCCTTTG
GCGCCCACTGCATTTACCCGTCTCTTCAACGATGACGAGCGGTTCTACCGGAGTTATCTGCGTAGATTTA
ATGGACAGTGGATTGACACCGGAGACGCGGGGTTAATCGATAAAGACGGGTATATCCATATCATGGCTCG
ATCCGACGACATCATCAATGTCGCGGCGCATCGCTTCAGTACGGGCGCCATTGAGCAAGCAATCCTCTTG
CACCCACAAATTGGCGAAGCCAGCGTGGTGGGCGTTCCGGACCCGCTGAAGGGACACCTCCCGTTTGCTT
TCATTCAGCCACGAACCGCGACCGGGCCCATACCCGCGAAACCTACAGCAGAACTGTTTCAGGAGGTGAA
TGGGCTGGTGCGCGAGCAGATCGGGGCAATCGCGTCCCTAGGGGGCATGATACAAGGCCGGGGCATGATC
CCCAAGACGCGCAGTGGGAAGACGCTGCGCCGAGTGCTGCGAGAGCTGGTGGAGCATGCAGTGCGCGGGG
AGTACGAGGCGCCGGTGAATGTGCCGCCGACGGTGGAAGACGCCGACGTCGTGGATGAGGCACGACGACG
GGTGCGTGAGTACTTTGATCGACAGGGACGGGCGAAGCTGTAATCGCTTGTTGGATGCAGTTGGATGCAG
TTGGATGCAGTTGGATGCAGTTGGATGCAGTTGGATGCAGTTGGATGCAGTTGGATGCAGTTGGATGCAG
TTGGATGCAGTTGGATGCAGTTGGATGCAGTTGGATGCAGTTGGAGCTGTACAGGTGACATTAAATGACA
TTTATCGTTGTACTTACCGAGTTGATTACCATTTGTGGGAAGTAACAGTGTTGATGATCGATGCGTATCG
TATTCGTCAGTGCACGAATCAGAATTAGTATGTATCTATTGAATGGATATACGCAAGTATTTTAGGCAAA
CATTGTATGTATCTGTATCTGAAGCTGCTTCAGAGAC

We match the the fourth entry of uniprot, https://www.uniprot.org/uniprotkb/G3XV63/entry

But if we'd then do the same for the cytosolic one (just as a test)

>jgi|Aspni7|1146906|fgenesh1_kg.chr_602_#_355_#_590_1_CATW_CATX_CATY_CATZ
GGGCATTGGGTTCAATAAATAACTGTCTCTCTTCCCCATCCTGGGCACCAGTTGCTTTCTTTTCCCCCCT
CTCTTCCTCATTTCTTCCTAATCTTTCTATTCTCTTTCCTCTCTACTCCTCTCTCTCTCCCTCCTCTCTT
ACTCCCTCCTACACTTCTCCCTCTCTTCCCCCCCCTCGCGAAAGTTTGCCTGAACCCCACCAACCAGCAC
TGCTGCAATTCTTTTGACAATTCAAACACACAGTTTCGTTGGGACACAACAAAACATGGCCGACGGCAGT
GTTATGCCCCCCAAGTCCTCCGTGGTGGTCGAGGCCCACGAAGTCGACACCTTCCATGTCCCCGAAGCAT
TCCACCAGAAGCACCCGACAGGTACCCACTTGAAAGACCTTGACGAGTACAAGAAGCTCTATGAGGAGTC
CATCCGCAGCCCAGACACTTTTTGGGCTCGCATGGCCCGCGAACTTCTTCACTTCGACCAAGACTTCCAG
ACCACCCACAGTGGGTCGCTGGAGAATGGCGACAATGCCTGGTTTGTCGAGGGTCGTCTGAACGCCTCCT
ACAACTGCGTCGATCGCCATGCCTTCAAGAACCCCGACAAGGTCGCCATTATCTACGAGGCCGATGAGCC
CAGCGAGGGCCGCACAATCACCTACGGTGAGCTCCTGCGCGAGGTGTCTCGTGTGGCCTGGGTTCTCCGC
CAGCAGGGTGTGAAGAAGGGTGACACCGTTGCTATCTACCTTCCCATGATTCCGGAGGCCATTGTCGCTT
TCTTGGCTTGCTCTCGTATCGGTGCCGTCCACTCCGTTGTTTTCGCCGGTTTCTCGTCCGACTCTCTCCG
GGACCGTGTCCTGGACGCTGGCTCCAAGGTCGTCATCACGACCGATGAGGGCAAGCGTGGAGGAAAGGTC
ATTGGCACCAAGCGCATTGTCGATGAAGCCCTGAAGCAGTGCCCCGATGTCACTAGCGTCCTGGTCTACA
AGCGCACCGGCGCGGAGGTCCCCTGGACCAAGGGCCGTGATGTCTGGTGGCACGAGGAGGTCGAGAAGTA
CCCCAACTACCTTGCTCCTGAGCCGGTGAACTCGGAAGATCCCCTCTTCCTGCTCTACACCTCTGGTTCT
ACCGGAAAGCCCAAGGGTGTCATGCACACCACGGCTGGTTACCTTTTGGGTGCTGCCATGACCGGCAAGT
ATGTCTTTGATATCCACGACGATGACCGCTACTTCTGTGGTGGTGATGTTGGCTGGATCACCGGCCACAC
GTACGTCGTATATGCACCCTTGCTGCTCGGATGCTCTACCGTGGTCTTCGAGAGTACCCCGGCTTACCCC
AACTTCTCCCGTTACTGGGACGTGATCGAGAAGCACAAGGTGACCCAATTCTACGTCGCCCCCACTGCCC
TGCGACTTCTGAAGCGGGCTGGCGACGAGCACATTCACCACAAGATGGCACACCTGCGTGTCCTGGGCTC
TGTCGGTGAGCCTATTGCTGCCGAGGTCTGGAAATGGTACTTCGAGGTGGTCGGAAAGGAAGAAGCCCAC
ATCTGCGATACCTACTGGCAGACTGAGACGGGTTCGAATGTCATCACTCCCCTGGGTGGTATTACCCCGA
CGAAACCCGGAAGCGCGTCGCTTCCCTTCTTTGGCATTGAGCCCGCCATCATTGACCCGGTCTCTGGTGA
GGAGATCACTGGCAACGATGTGGAGGGTGTGCTTGCCTTCAAGCAACCCTGGCCGAGCATGGCCCGTACC
GTGTGGGGCGCCCACAAGCGCTACATGGATACCTACTTGAACGTTTACAAGGGCTACTACTTCACTGGAG
ATGGTGCAGGCCGCGATCACGACGGATACTACTGGATCCGGGGTCGTGTTGACGATGTGGTCAACGTTTC
CGGACACCGTCTGTCCACGGCTGAGATTGAAGCCGCCCTCATTGAGCACCACATGGTGGCTGAGGCTGCT
GTCGTTGGTATTGCTGACGAGCTTACCGGTCAGGCCGTCAACGCATTCGTGGCCCTGAAAGAGGGCAACG
AGACCAGCGACCAGGTCCGCAAGGATCTGATCATGCAGGTTCGTCGGTCCATCGGACCTTTCGCTGCCCC
CAAGGCTGTCTACGTTGTCGAGGATCTCCCCAAGACCCGTAGCGGTAAGATCATGCGCCGTATTCTCCGG
AAGATCCTGAGCGGTGAGGAGGACAGCCTGGGTGACACTTCTACTCTGTCGGACCCTTCGGTTGTTGACA
AGATCATTGCCACGGTGCACGCTGCCCGTGGCAAATAAGATCTTCAAATGAACCGAATCCAAGACGAGCG
ATATTCACAGCCATGAAATGAAAGCGATTAATCAGGATGTTATGATGAGTTGATTAGCATTATTGGGTGG
ACACAATTGTCCGGAGGCGCAGTCATATTCCTATAAACCCCGTTTCGTGATTCCCAACGTGTTTGAGAGA
TGAACATTATGGGTGGTCGCTGATCACCTTTTCAATATACATAAGTTCTATTCTTCATAGTAATGTTCTT
TTGTTCTTAAATGTAGTATGGGCCGTTCCTTTACATGACAGCGGTTGTCGGGG

This fails. Which is worrying. Perhaps I cant assume that the tx IDs and
and protein IDs for Aspni7 are 1-1 matches. This would be a problem...

- 1202754: Phosphomevalonate kinase. I do find a sequence for it, e.g.
https://www.ncbi.nlm.nih.gov/protein/EHA27845.1
https://www.uniprot.org/uniprotkb/G3XN19/entry
>tr|G3XN19|G3XN19_ASPNA Phosphomevalonate kinase OS=Aspergillus niger (strain ATCC 1015 / CBS 113.46 / FGSC A1144 / LSHB Ac4 / NCTC 3858a / NRRL 328 / USDA 3528.7) OX=380704 GN=ASPNIDRAFT_211137 PE=3 SV=1
MSYKAPTVTAVSAPGKVLLAGGYLVLDRDYTGTVFALDARIHVVVQQLRRSHRITEDEAA
DEQASEDVIVVRSPQFVDAVWEYGIQRCDNGGGVKVVQKGDGRANPFVETSLNYALTYIS
YVAASKDFGSLSITILADNDYYSETAFSKISSLQSPGRFVNFGVPLHEAHKTGLGSSAAL
VTAFVSALVIHRTLQPEDLGAAREKLHNLAQAAHCAAQGKVGSGFDVAAAIYGSCLYRRF
SPSILESVGDAGSPGFEERLFSIVEDADPNHTWDTECLDFGMKLPRGMQMVLCDVECGSQ
TPSMVRKVLEWRKQNPEESLMLWQALQSNNERLCLELKQLAQSPDVEAPNSFDDTRRLIE
RSRDLIRTMTQKAGVPIEPKVQTELLDAITSIEGVVGGVVPGAGGYDALAVLIEDKAEVI
QRLNELFETWESKLEDDFGGKIGNVRLLGVRHGSAGVQNESLDQYAGWF

So essentially, we have two problems on our hands:
- BPGA input does not have all IDs that we have in our starting GEMS
- BPGA removes IDs (likely paralogs)

However, maybe the latter is not true. Maybe they are there, but we are simply
not extracting them from the output! These data could be u_clusters.txt

But to be fair, output handling should not be left to the BPGA user. Perhaps
USEARCH does a better job?

# USEARCH

USEARCH clustering generates a .uc file, with contents similar to those in
u_clusters.txt from BPGA. I think all info we need on homologs/paralog is here.
We just need to be able to extract this. Here, I search for existing methods.
If these don't work, we'd need to resort to another way to determine 
orthologs/paralogs, e.g., using the github Casper sent, as writing our own code
for this (like Casper did) will bring us too far.

Also, check clustering ouput format here: https://drive5.com/usearch/manual/opt_uc.html

```{r}
ucfile <- "../../clusters.uc"
clustering <- import_usearch_uc(ucfile,verbose = TRUE,readDelimiter = "\t")
clustering@.Data[1:3,1:3]
```

```{r}
ncol(clustering) # number of clusters?
```

```{r}
hlp <- clustering[,1]
hlp[hlp!=0]
```

Does this mean that fumigatus gene 1|Org1_Gene1004|fumigatus.faa# has 
- one paralog in fumigatus
- many orthologs in niger
- two orthologs in oryzae?

```{r}
table(unname(colSums(clustering@.Data)))
```

What is a one? 

```{r}
hlp <- clustering[,19]
hlp[hlp!=0]
```

So the import seems to only retain hits?  Query this gene in the original file.
-> can confirm, the .uc file has these two as only hit for each other. They are
orthologs without paralogs.


```{r}
hlp <- clustering[,"1|Org1_Gene1083|fumigatus.faa#"]
hlp[hlp!=0]
```

So now I need to wrangle this table to something more similar to what Casper 
has. 


```
# Save this function in case the github dies
metagMisc::parse_uc
function (x, map_only = FALSE, splitSeqID = TRUE, rm_dups = TRUE) 
{
    ii <- fread(file = x, header = FALSE, sep = "\t")
    ii <- ii[!V1 %in% "S"]
    ii[, `:=`(Query, tstrsplit(V9, ";", keep = 1))]
    ii[, `:=`(OTU, tstrsplit(V10, ";", keep = 1))]
    ii[V1 %in% "C", `:=`(OTU, Query)]
    if (nrow(ii[, .(Query, OTU)]) != nrow(unique(ii[, .(Query, 
        OTU)]))) {
        cat("Warning: duplicated rows found!\n")
        if (rm_dups == TRUE) {
            dups <- duplicated(ii[, .(Query, OTU)])
            cat("..", sum(dups), " duplicates removed\n")
            ii <- ii[!dups]
        }
    }
    setnames(x = ii, old = paste0("V", 1:10), new = c("recordType", 
        "clusterNumber", "seqLengthOrClusterSize", "percentIdentity", 
        "strand", "unusedField1", "unusedField2", "alignment", 
        "queryLabel", "targetLabel"), skip_absent = TRUE)
    ii[percentIdentity %in% "*", `:=`(percentIdentity, NA)]
    ii[, `:=`(percentIdentity, as.numeric(percentIdentity))]
    if (map_only == TRUE) {
        if (splitSeqID == TRUE) {
            ii <- ii[, .(Query, OTU)]
        }
        else {
            ii <- ii[, .(queryLabel, targetLabel)]
            setnames(x = ii, old = c("queryLabel", "targetLabel"), 
                new = c("Query", "OTU"))
        }
    }
    return(ii)
}
```

```{r}
hlp <- metagMisc::parse_uc(x=ucfile, map_only = FALSE, splitSeqID = TRUE, rm_dups = TRUE)

dim(hlp)
```

```{r}
table(hlp$recordType)
```

```{r}
hlp[1:10,]
```

```{r}
hlp[hlp$OTU == "1|Org1_Gene1004|fumigatus.faa#",]
```

```{r}
hlp[hlp$clusterNumber == 323,]
```

What I think we want is a dataframe, where each row is a cluster clusterNumber,
and we have column for each species, and their cells are a list of compatible 
IDs.

```{r}
res <- do.call(rbind,lapply(unique(hlp$clusterNumber), function(x){
   test <- hlp[hlp$clusterNumber == x,]
   data.frame(cluster = x,
              Afumigatus = toString(test$Query[grep("fumigatus",test$Query)]),
              Aoryzae = toString(test$Query[grep("oryzae",test$Query)]),
              Aniger = toString(test$Query[grep("niger",test$Query)]))
}))

head(res)
```

Maybe this is the most convenient structure for e.g. Casper to work with as you
could also look at this in excel, but honestly it would be much cleaner to work
with a named list where the new clusters are the names, and entries fumigatus,
niger and oryzae, and maybe some additional info like core/accessory/unique and
n_genomes.

TODO: map IDs (1|Org1_Gene9442|fumigatus.faa#) to original gene names.
TODO: ask Capser why there is a discrepancy between his cluster_type_manual and
BPGA's cluster_type_BPGA. I see how the former is computed, but not the latter.

