---
title: "Matching GEMs and BPGA output"
author: "Jeroen Gilis"
date: "6 January 2025"
output: 
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

```{r, message=FALSE, warning=FALSE}
library(metagMisc)
```

```{r}
source("./Rutilities.R")
```

# Load clustering output

In step 2 of their workflow, BPGA generates a file `u_clusters.txt` that 
contains the clustering information for each sequence in the BPGA input, i.e.,
containing both orthologs and paralogs. In USEARCH, the same file is also
generate, with extension `.uc`. The ouput format is described here:
http://drive5.com/usearch/manual/opt_uc.html

```{r}
clustfile <- "../data/genome/u_clusters.txt"
clustering <- metagMisc::parse_uc(x=clustfile, 
                                  map_only = FALSE, 
                                  splitSeqID = TRUE, 
                                  rm_dups = TRUE)
head(clustering)
```

```
# Save this function in case the github dies
metagMisc::parse_uc
function (x, map_only = FALSE, splitSeqID = TRUE, rm_dups = TRUE) 
{
    ii <- fread(file = x, header = FALSE, sep = "\t")
    ii <- ii[!V1 %in% "S"]
    ii[, `:=`(Query, tstrsplit(V9, ";", keep = 1))]
    ii[, `:=`(OTU, tstrsplit(V10, ";", keep = 1))]
    ii[V1 %in% "C", `:=`(OTU, Query)]
    if (nrow(ii[, .(Query, OTU)]) != nrow(unique(ii[, .(Query, 
        OTU)]))) {
        cat("Warning: duplicated rows found!\n")
        if (rm_dups == TRUE) {
            dups <- duplicated(ii[, .(Query, OTU)])
            cat("..", sum(dups), " duplicates removed\n")
            ii <- ii[!dups]
        }
    }
    setnames(x = ii, old = paste0("V", 1:10), new = c("recordType", 
        "clusterNumber", "seqLengthOrClusterSize", "percentIdentity", 
        "strand", "unusedField1", "unusedField2", "alignment", 
        "queryLabel", "targetLabel"), skip_absent = TRUE)
    ii[percentIdentity %in% "*", `:=`(percentIdentity, NA)]
    ii[, `:=`(percentIdentity, as.numeric(percentIdentity))]
    if (map_only == TRUE) {
        if (splitSeqID == TRUE) {
            ii <- ii[, .(Query, OTU)]
        }
        else {
            ii <- ii[, .(queryLabel, targetLabel)]
            setnames(x = ii, old = c("queryLabel", "targetLabel"), 
                new = c("Query", "OTU"))
        }
    }
    return(ii)
}
```

# Translate IDs

The goal of this section is to establish a link between the sequence identifiers
that are used in the BPGA output, and the sequence identifiers that are used in
the three GEMs that we will use to construct a draft panGEM fro Aspergillus.

The queryLabel and targetLabel columns from the BPGA output contain labels for
each fasta sequence that was used as input to BPGA. The link between these 
labels and the original fasta sequence labels is stored in the BPGA output file
`INPUT_all_original.fasta`. 

Note that these steps may not be required if you use USEARCH separately, and not
within BPGA. Indeed, in USEARCH it just depends on whether you manually changed 
the names of the FASTA headers yourself or not.

```{r}
dict_fasta <- read.csv2("../data/genome/INPUT_all_original.fasta", 
                        sep=">", header=FALSE)
dict_fasta$V1 <- NULL
colnames(dict_fasta) <- c("BPGA","original")
dict_fasta$BPGA <- sub("\\|$", "#", dict_fasta$BPGA)
dict_fasta$BPGA <- paste0(sub("\\|.*", "", dict_fasta$BPGA),
                          "|Org",
                          sub("\\|.*", "", dict_fasta$BPGA),
                          "_",
                          sub("^[^\\|]+\\|","",dict_fasta$BPGA))
head(dict_fasta)
```

To link the BPGA sequence identifiers to the sequence identifiers from the 
niger GEM, we need an additional translation step. Indeed, we used protein
sequences as input for BPGA, but the sequence identifiers in the niger GEM
are transcript identifiers. As such, we need to establish a link between the 
protein and the transcript sequence identifiers of the niger GEM.

```{r}
awk_cmd <- sprintf("awk '$1 ~ /^>/' %s", "../data/genome/keep_Aspni7_all_transcripts_20131226.nt.fasta")
full_IDs <- readLines(pipe(awk_cmd))
tx_prot_names <- sub(".*\\|", "", full_IDs)
tx_num <- sub("^([^\\|]*\\|){2}([^\\|]+).*", "\\2", full_IDs)
dict_niger <- data.frame(tx_prot_names = tx_prot_names,
                         tx_num = tx_num)
target <- dict_fasta[sub(".*\\|(.*)", "\\1", dict_fasta$BPGA) == "niger.faa#",]
target$GEM <- dict_niger$tx_num[match(target$original, dict_niger$tx_prot_names)]
```

There are 16 niger protein sequences that we used as input to BPGA for which
we cannot establish a link with any transcript identifiers.

```{r}
target[is.na(target$GEM),]

# So these are IDs that we have in the protein fasta, but that we can't link
# back to transcript numbers, and hence cant link back to the GEM.
lost <- target[is.na(target$GEM),]

clustering[clustering$targetLabel %in% lost$BPGA,]
clustering[clustering$Query %in% lost$BPGA,]
```

Note that these IDs didnt receive many hits in the clustering either. I will
thus just leave them as NA. As such, they will count for establishing in how 
many genomes a cluster was observed, they will also be present in the final
BPGA table , but we know in advance they wont be matched to IDs in the GEM.

```{r}
dict_fasta[sub(".*\\|(.*)", "\\1", dict_fasta$BPGA) == "niger.faa#","original"] <- target$GEM
dict_fasta$original[is.na(dict_fasta$original)] <- target$original[match(dict_fasta$BPGA[is.na(dict_fasta$original)],target$BPGA)]
```

# Wrangle clustering output

```{r}
clustering$fasta <- sub(".*\\|(.*)", "\\1", clustering$Query)
all_organisms <- unique(clustering$fasta) # easier to store organism fasta name in separate column
clustering$Query <- dict_fasta$original[match(clustering$Query, dict_fasta$BPGA)]

res <- do.call(rbind,pbapply::pblapply(unique(clustering$clusterNumber), function(x){
    test <- clustering[clustering$clusterNumber == x,]
    hlp <- as.data.frame(rbind(unlist(lapply(all_organisms, function(organism){
    paste0(test$Query[grep(organism,test$fasta)], collapse = ";")
        }))))
    colnames(hlp) <- all_organisms
    df_row <- cbind(cluster = x, hlp)
    return(df_row)
}))
```

I think I have a 1 more column than Casper. So he probably sent me sequences
from an isolate that he didnt include in his BPGA run.

```{r}
res[res==""] <- NA
colnames(res) <- gsub(".faa#","",colnames(res))

res$present_in_n_genomes <- rowSums(!is.na(res[,-1]))
res$cluster_type_manual <- ""

res$cluster_type_manual[res$present_in_n_genomes == 1] <- "singleton"
res$cluster_type_manual[res$present_in_n_genomes > 1] <- "accessory"
res$cluster_type_manual[res$present_in_n_genomes > 150] <- "core"

putfirst <- c("cluster", "present_in_n_genomes", "cluster_type_manual","niger",
              "fumigatus","oryzae")
res <- res[,c(match(putfirst,colnames(res)),
       which(!colnames(res) %in% putfirst))]
```

```
res[grep("PEPCE", res$niger),]
```

Now we only have the issue that some genes that are in the GEMs are not in this
table, because they were not in the BPGA input, because we don't have a protein
sequence for these features (and these can also not be readily obtained).

For A. niger this is one feature (Unknown5).
For A. oryzae this are 46 features (all ZY entries).

To facilitate downstream analyses, I will add these entries to the BPGA output
table myself, and assign them cluster_type_manual "artificial".

```{r}
add_niger <- as.data.frame(matrix(data=NA,nrow=1,ncol=162,dimnames = list(NULL,colnames(res))))
add_niger$cluster <- max(res$cluster)+1
add_niger$present_in_n_genomes <- 1
add_niger$cluster_type_manual <- "artificial"
add_niger$niger <- "Unknown5"
res <- rbind(res, add_niger)
```

```{r}
oryzae_gem <- read_excel_allsheets("../model/templates/iWV1314.xlsx")
missing_oryzae <- unique(oryzae_gem$GENES$NAME)[!unique(oryzae_gem$GENES$NAME) %in% unique(unlist(strsplit(res$oryzae, split=";")))]

add_oryzae <- as.data.frame(matrix(data=NA,nrow=length(missing_oryzae),
                                  ncol=162,dimnames = list(NULL,colnames(res))))
add_oryzae$cluster <- max(res$cluster)+seq_along(missing_oryzae)
add_oryzae$present_in_n_genomes <- 1
add_oryzae$cluster_type_manual <- "artificial"
add_oryzae$oryzae <- missing_oryzae
res <- rbind(res, add_oryzae)
```

# Export final table

```{r}
res[is.na(res)]<-""
res$cluster <- paste0("cluster_",res$cluster)
write.csv2(res, "../data/genome/BPGA2ortho_GEM_custom.csv",row.names=FALSE)
```

# Appendix: missing Oryzae genes

```{r}
missing_oryzae
```

These IDs are present in the Oryzae GEM, but not in the FASTA input. These IDs
refer to ESTs, which are short cDNA sequences (RNA -> reverse ranscribed).
While I can retrieve a nucleotide sequence for the EST, it is not directly
possible to go to a protein sequence, as the link EST-gene-protein has not been
established. As such, I don't think there is much more we can do for these
IDs. However, below I have looked into each of the ZYs:

- 23 are isoenzymes, we already have a properly characterized gene in Oryzae doing the same thing
- 15 are subunits of an enzyme complex, together with genes that are properly characterized
- 8 are truly unique, catalyzing a reaction no other gene is involved with

## Chr4

- Chr4_SC012_1076984__1077358: N-acetyl-y-glutamyl-phosphate reductase
Not in sup_table1, but same nomenclature a the ZYs, 
"Chromosome number with EST alignment" - "Contig name with EST alignment" - "EST position (Start-Stop)"

## Isoenzymes

- ZY007107: subunit of Cytochrome c oxidase, together with
AO090010000481 and AO090026000651 and AO090026000421 and ZY007619 and ZY029203 and ZY029342
Looking at the other ZYs, they only appear here as subunit of this enzyme 
Looking at the AOs, they only appear here as subunit of this enzyme 

- ZY006639: subunit NADH dehydrogenase (ubiquinone, both mitochondrial and cytoplasm), together with
AO090003001115 and AO090011000782 and AO090001000553 and AO090026000229 and AO090001000661 and AO090011000578 and AO090003001313 and AO090102000645 and AO090012000402 and AO090011000502 and AO090103000199 and AO090672000005 and AO090001000459 and AO090001000300 and AO090003000400 and ZY006639 and ZY081278 and ZY083531 and ZY087692
Looking at the other ZYs, they only appear here as subunit of this enzyme 
Looking at the AOs, they also appear as subunit of NADH dehydrogenase (FAD-containing subunit)

- ZY007427: subunit of Ubiquinol-cytochrome c reductase, together with
AO090010000475 and AO090102000625, which also only appear as subunit of this enzyme

- ZY029275: subunit of Succinate dehydrogenase (flavoprotein subunit), together with
AO090020000415 and AO090020000596 and AO090010000505 
Looking at the AOs, they only appear here as subunit of this enzyme 

- ZY036719: subunit of F-type ATPase complex, together with
AO090005000617 and AO090005000749 and AO090005000604 and AO090026000372 and AO090010000482 and AO090701000168 and AO090206000115 and AO090010000482 and AO090003000607 and AO090023000599 

and subunit of V-type ATPase complex, together with
AO090102000349 and AO090038000334 and AO090026000283 and AO090120000313 and AO090120000091 and AO090003001090 and AO090012000797 and AO090026000517 and AO090038000579 and AO090038000352 and AO090012000274

- ZY077810: subunit of acyl-Coa dehydrogenase

For acyl %in% c(steaoryl, palmitoyl, myristoyl, lauroyl, decanoyl and octanoyl), other subunits are
AO090005000521 and AO090009000596 and AO090102000172 and AO090012000150 and AO090005000494 and AO090023000711 

For acyl %in% c(hexanoyl and butyrylyl), other subunits are
AO090103000106

- ZY083011: subunit of NADH dehydrogenase (FAD-containing subunit), together with
AO090003001115 and AO090011000782 and AO090001000553 and AO090026000229 and AO090001000661 and AO090011000578 and AO090003001313 and AO090102000645 and AO090012000402 and AO090011000502 and AO090103000199 and AO090672000005 and AO090001000459 and AO090001000300 and AO090003000400 and AO090001000657 and AO090011000358

- ZY098176: subunit of Acetyl-CoA carboxylase, together with 
AO090011000838 

- ZY098280: subunit of Acetolacetate synthase, together with 
AO090023000854 and AO090003000571 and AO090166000076

## Unique enzymes

Unique, no other proteins catalyzing this reaction: 

- ZY006430: Phosphatidylinositol-3,4,5-trisphosphate 3-phosphatase

- ZY080692: P-hydroxybenzoate hydroxylase

- ZY088142: Di-trans,poly-cis-decaprenylcistransferase

- ZY097848: Methylglyoxal synthase

- ZY104986: Polyphosphate kinase

- ZY110993: Alanine racemase

- ZY111007: Scytalone dehydratase

- ZY140875: Selenocysteine synthase

! Im pretty sure the enzyme for r84 should be AO090001000642 in stead of AO090001000484

```{r}
isoenzymes <- list("ZY006677" = "AO090026000021",
                   "ZY007112" = "AO090026000800",
                   "ZY080376" = c("AO090011000353", "AO090701000542", "AO090012000766", 
                                  "AO090102000588", "AO090005000837", "AO090026000395", 
                                  "AO090120000164", "AO090003001432", "AO090103000036",
                                  "AO090001000143", "AO090701000644"),
                   "ZY111292" = c("AO090011000353", "AO090701000542", "AO090012000766", 
                                  "AO090102000588", "AO090005000837", "AO090026000395", 
                                  "AO090120000164", "AO090003001432", "AO090103000036",
                                  "AO090001000143", "AO090701000644"),
                   "ZY029856" = "AO090005001098",
                   "ZY007726" = c("AO090020000670", "AO090009000711", "AO090001000641",
                                  "AO090005001098"),
                   "ZY083916" = c("AO090020000670", "AO090009000711","AO090020000668",
                                  "AO090001000641"),
                   "ZY029184" = c("AO090005001370","AO090011000449","AO090012000084",
                                  "AO090206000079","AO090701000589","AO090005001370",
                                  "AO090005001371","AO090113000128","AO090005000579",
                                  "AO090026000212","AO090026000321","AO090026000323"),
                   "ZY077820" = c("AO090005001300","AO090120000145","AO090011000152"),
                   "ZY085753" = c("AO090005000045", "AO090001000484","AO090001000642",
                                  "AO090003000415"),
                   "ZY111420" = c("AO090026000498","AO090026000149"), 
                   "ZY086694" = c("AO090026000498","AO090026000149"),
                   "ZY087055" = c("AO090001000162","AO090023000467","AO090009000417"),
                   "ZY087096" = c("AO090003000631","AO090005000973","AO090103000149",
                                  "AO090103000485","AO090102000005","AO090003001039",
                                  "AO090103000134","AO090102000478","AO090012000378",
                                  "AO090012000112","AO090023000353"),
                   "ZY087286" = "AO090012000998",
                   "ZY087560" = c("AO090038000057","AO090038000010"),
                   "ZY093903" = c("AO090038000057","AO090038000010"),
                   "ZY087836" = "AO090011000561",
                   "ZY097910" = c("AO090005001216","AO090026000084"),
                   "ZY110723" = c("AO090005001124","AO090012000733"),
                   "ZY111104" = c("AO090005000433","AO090102000204","AO090020000034"),
                   "ZY135516" = "AO090701000187")
```

Note: the oryzae GEM has no "or" gene associations, they are cataloged as
separate reactions.

```{r}
res <- read.csv2("../data/genome/BPGA2ortho_GEM_custom.csv")
res[res==""]<-NA
res[which(res$oryzae == "AO090038000057"),]

res[which(res$cluster == "cluster_12964"),which(!is.na(res[which(res$cluster == "cluster_12964"),]))]

res_sub <- res[,c('cluster', 'oryzae', 'Aspergillus_oryzae_RIB40_CAoGD', 
                  'Aspergillus_oryzae_NRRL_2217','Aspergillus_oryzae_NRRL_3483',
                  'Aspergillus_oryzae_NRRL_5589', 'Aspergillus_oryzae_NRRL_3488',
                  'Aspergillus_oryzae_NRRL_5592', 'Aspergillus_oryzae_NRRL_471',
                  'Aspergillus_oryzae_NRRL_35890')]

# hlp <- apply(res_sub, 1, is.na)
res_sub <- res_sub[-which(rowSums(is.na(res_sub[,2:10])) == 9),]
rownames(res_sub) <- res_sub$cluster
res_sub$cluster <- NULL

result <- res_sub[is.na(res_sub[[9]]) & !apply(res_sub[, -9], 1, function(x) any(is.na(x))), ]
rownames(result)

source("../code/Rutilities.R")
panAsp_v2 <- read_excel_allsheets("../model/panAsp_v2.xlsx")

rownames(result)[rownames(result) %in% panAsp_v2$GENES$NAME]
```

