---
title: "Inspecting BPGA2ortho_GEM_custom.txt"
author: "Jeroen Gilis"
date: "2 October 2024"
output: 
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

```{r, message=FALSE, warning=FALSE}
library(stringr)
library(openxlsx)
library("kableExtra")
```

```{r}
read_excel_allsheets <- function(filename) {
    sheets <- openxlsx::getSheetNames(filename)
    x <- lapply(sheets, function(X) openxlsx::read.xlsx(filename, sheet = X))
    names(x) <- sheets
    x
}
```


# BPGA

```{r}
BPGA <- read.csv2("../panAsp-GEM/data/genome/BPGA2ortho_GEM_custom.txt", 
                  sep="\t",
                  na.strings = "")
BPGA[1:5,1:14] %>%
  kbl() %>%
  kable_styling(full_width = F)
```

```{r}
table(str_extract(colnames(BPGA)[9:ncol(BPGA)], "[^_]*_[^_]*"))
```

<span style="color: red;">**Q: How to interpret the existence of 156 Oryzae genomes?**</span>
<span style="color: red;">**Answer: These are 156 different isolates. Note that the main goal of this project is to improve the Oryzae GEM, and that we will use Fumigatus and Niger to help constructing a pan-genome that helps with that.**</span>

```{r}
table(BPGA$cluster_type_BPGA, BPGA$cluster_type_manual)
```

```{r}
table(BPGA$present_in_n_genomes, BPGA$cluster_type_manual)
```
```{r}
table(BPGA$present_in_n_genomes, BPGA$cluster_type_BPGA)
```

# Aspergillus niger GEM

This file was obtained by

1. Downloading the Aspergillus niger GEM from Brandl *et al.* (XML)
2. Opening the XML in MatLab using RAVEN toolbox.
3. Saving as Excel

```{r}
niger_gem <- read_excel_allsheets("../GEMs_orig/AspNiger/40694_2018_60_MOESM2_ESM.xlsx")
```

```{r}
head(niger_gem$RXNS$NAME,n=10)
# "Aldose 1-epimerase" vs. "aldose-1-epimerase" -> might not be clean!
```

<span style="color: red;">**Q: The names are not consistent (Aldose 1-epimerase vs. aldose-1-epimerase), is that a problem?**</span>
<span style="color: red;">**No, the names are not used in the model, everything is based on the unambiguous identifiers (not sure if it's using EC number, gene association, or both).**</span>

```{r}
head(niger_gem$GENES,n=10)
```

```{r}
sum(grepl("or",niger_gem$RXNS$GENE.ASSOCIATION))
sum(grepl("and",niger_gem$RXNS$GENE.ASSOCIATION))
```

# Aspergillus fumigatus GEM

```{r}
fumigatus_gem <- read_excel_allsheets("../GEMs_orig/AspFumigatus/Pan_Aspergillus_fumigatus.xlsx")
```

# Aspergillus oryzae GEM

```{r}
oryzae_gem <- read_excel_allsheets("../GEMs_orig/AspOryzae/iWV1314.xlsx")
```

# Establishing a link between BPGA and GEM gene IDs

## Aspergillus fumigatus

```{r}
length(fumigatus_gem$GENES$NAME)
length(unique(fumigatus_gem$GENES$NAME)) # 1444 unique genes in GEM
```

```{r}
length(BPGA$Afu_ID)
length(unique(BPGA$Afu_ID[!is.na(BPGA$Afu_ID)])) # 8377 unique genes in BPGA output
```

```{r}
table(fumigatus_gem$GENES$NAME %in% unique(BPGA$Afu_ID[!is.na(BPGA$Afu_ID)]))
# out of 1444 genes in GEM, only 1214 are in BPGA, 230 are not
```

```{r}
fumigatus_input <- read.csv2(file="../BPGA/fumigatus_table.tsv",
                             sep="\t")

length(unique(fumigatus_input$X.Format.Gene.ID))
length(unique(BPGA$Afu_ID))
length(Fumigatus_GENES$NAME)

# all genes of the GEM are in the BPGA input
table(Fumigatus_GENES$NAME %in% unique(BPGA$Afu_ID))
table(Fumigatus_GENES$NAME %in% unique(fumigatus_input$X.Format.Gene.ID))
```

```{r}
# linking clusters to reactions (based on only this):
fumigatus_link_table <- data.frame(Afu_ID = BPGA$Afu_ID,
                                   cluster = paste0("cluster_",BPGA$cluster_number))

fumigatus_link_table <- fumigatus_link_table[fumigatus_link_table$Afu_ID %in% Fumigatus_GENES$NAME,]
rownames(fumigatus_link_table) <- NULL
head(fumigatus_link_table)

write.csv2(fumigatus_link_table, file='../GEMs_orig/AspFumigatus/orthologPairs_Fumigatus_test.csv', 
            quote=FALSE, row.names=FALSE)

# So this is table linking the ortholog cluster numbers to the gene IDs in the GEM,
# As these are all in the GEM, these are all enzymes.

# Note that we'd expect to have a cluster number for each gene in the GEM. 
# However, at the moment, 230 genes are missing, because they were in the GEM but
# not in the BPGA ouptut.

# In addition, we also have many Afu_ids that are not in the Afu GEM. Eduard suggested
# that for these remaining genes (which may still include enzymes), we could e.g. run
# getKEGGModelForOrganism (?) to link genes to reactions.
```

```{r}
# inspect these genes
hlp <- BPGA[!BPGA$Afu_ID %in% fumigatus_link_table$Afu_ID,]
hlp <- hlp[!is.na(hlp$Afu_ID),]
hlp
```

## Aspergillus oryzae

```{r}
length(oryzae_gem$GENES$NAME)
length(unique(oryzae_gem$GENES$NAME)) # 1346 unique genes in GEM
```

```{r}
length(BPGA$CAoGD_ID)
length(unique(BPGA$CAoGD_ID[!is.na(BPGA$CAoGD_ID)])) # 9356 genes in BPGA
```

```{r}
table(oryzae_gem$GENES$NAME %in% unique(BPGA$CAoGD_ID[!is.na(BPGA$CAoGD_ID)]))
```

```{r}
oryzae_input <- read.csv2(file="../BPGA/oryzae_table.tsv",
                          sep="\t") # BPGA input

length(unique(oryzae_input$X.Format.Gene.ID))
length(unique(BPGA$CAoGD_ID))
length(oryzae_gem$GENES$NAME)

table(oryzae_gem$GENES$NAME %in% unique(BPGA$CAoGD_ID))
table(oryzae_gem$GENES$NAME %in% unique(oryzae_input$X.Format.Gene.ID))
# unfortunately, not all of them are present :(
```

## Aspergillus niger

TLDR; Not sure yet how to link these

```{r}
length(unique(BPGA$Ani_ID))
head(unique(BPGA$Ani_ID))
```

```{r}
head(niger_gem$GENES$NAME,n=10) # different format
```

```{r}
length(unique(Niger_GENES$NAME))
```

```{r}
which(BPGA$Ani_ID == "ASPNIDRAFT_225574")
# link ASPNIDRAFT_225574 = 1162446 = An01g06930 = fum1 (found from 
# https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6590338/, who worked with 
# fungiDB, which is temporarily offline)

which(Niger_RXNS$GENE.ASSOCIATION == "1162446")
```

```{r}
c("1109007","1178899") %in% Niger_GENES$NAME # Brandl transcriptIDs
c("1129557","1126350","1081533") %in% Niger_GENES$NAME # Brandl proteinID
```

For Brandl, Gene IDs supposedly correspond to ATCC1015 Aspni7 transcriptIds. 
I can find all of them back in the transcript/proteins FASTAs, but not in CDS FASTA

For BPGA, the "EHA_." are 
https://www.ebi.ac.uk/ena/browser/view/EHA17890.1?show=publications

ASPNIDRAFT_...: the ... are locus tags

## biomaRt

```{r}
library(biomaRt)
ensembl_fungi <- useMart(host="https://fungi.ensembl.org", 
                         biomart="fungi_mart")
searchDatasets(mart = ensembl_fungi, pattern = "niger")
searchDatasets(mart = ensembl_fungi, pattern = "fumigatus")
searchDatasets(mart = ensembl_fungi, pattern = "oryzae")
```

```{r}
ensembl_fum <- useEnsemblGenomes(biomart = "fungi_mart", dataset = "afumigatus_eg_gene")
ensembl_or <- useEnsemblGenomes(biomart = "fungi_mart", dataset = "aoryzae_eg_gene")
# ensembl <- useEnsemblGenomes(biomart = "fungi_mart", dataset = "aniger_eg_gene")
# unfortunately, not the right one for A. niger
```

```{r}
test <- getBM(attributes = c("ensembl_gene_id","protein_id","ensembl_transcript_id"), 
              mart = ensembl_fum)
head(test)
```

# Next steps

- Make panModel: assign all ortholog clusters from the pan-genome data table to reactions <span style="color: red;">**(if they are enzymes of course)**</span>.
    - The panModel will then have the ortholog cluster numbers (ranging from 0 to 18974) as gene identifiers. Good to prefix these with “cluster_”?

```{r}
# Am I understanding correctly that all is required for (3) is:
head(cbind(paste0("cluster_",BPGA$cluster_number),
           BPGA$Ani_ID)) # for only the enzymes
```

<span style="color: red;">**Q: What do we do with the info from `cluster_type`??**</span>
<span style="color: red;">**Answer: we will not used this information.**</span>

- Order of assign ortholog clusters to reactions:
    - Extract the ortholog cluster numbers and Ani_ID genes, and feed this 2-column table to getModelFromHomology, with the A. niger model as template. The output (let’s call it Ani-draft) is the first draft panModel.

`getModelFromHomology` works as follows:

function [draftModel, hitGenes]=getModelFromHomology(models,blastStructure,...
    getModelFor,preferredOrder,strictness,onlyGenesInModels,maxE,...
    minLen,minIde,mapNewGenesToOld)
    
I've already seen it used as:

model = `getModelFromHomology`({modelSce, modelRhto}, blast, 'hanpo',{'sce', 'rhto'});

<span style="color: red;">**Q: would we then use the existing model Ani as `model`, and the Pan as `getModelFor`? Would the Pan model then be a duplicate of the Ani model, but with update gene names (cluster_) that allow for linking with the other two Aspergillus?**</span>
<span style="color: red;">**Answer: Exactly!**</span>

# biomaRt

## Fumigatus

```{r}
BPGA[which(BPGA$CAoGD_ID == "AO090003001500"),] # linked to AFUA_2G11270 (ortholog)
# from https://nribf21.nrib.go.jp/CAoGD/gen.cgi?prj=01909&gnm=aor0-5&gen=AO090003001500,
# the orthologs in fumigatus are: c("AFUA_2G11270","AFUA_3G00910","AFUA_1G15440")
```

```{r}
c("AFUA_2G11270","AFUA_3G00910","AFUA_1G15440") %in% fumigatus_input$X.Format.Gene.ID # all in BPGA input
c("AFUA_2G11270","AFUA_3G00910","AFUA_1G15440") %in% fumigatus_gem$GENES$NAME # all in GEM
c("AFUA_2G11270","AFUA_3G00910","AFUA_1G15440") %in% BPGA$Afu_ID # but only one of them in BPGA
```

-> BPGA likely removed the paralogs -> we must restore them

```{r}
target<-fumigatus_gem$GENES$NAME[which(!fumigatus_gem$GENES$NAME %in% unique(BPGA$Afu_ID[!is.na(BPGA$Afu_ID)]))]
fumigatus_input[fumigatus_input$X.Format.Gene.ID%in%target,]
```

```{r}
listAttributes(ensembl_fum)$name[grepl("para", listAttributes(ensembl_fum)$name)]
```

```{r}
library(biomaRt)
ensembl_fum <- useEnsemblGenomes(biomart = "fungi_mart", dataset = "afumigatus_eg_gene")
paralog_table <- getBM(attributes = c("ensembl_gene_id",
                             "afumigatus_eg_paralog_ensembl_gene",
                             "afumigatus_eg_paralog_orthology_type",
                             "afumigatus_eg_paralog_perc_id_r1"), 
              mart = ensembl_fum,
              filters = "ensembl_gene_id",
              values=target)

paralog_table[paralog_table==""]<-NA
paralog_table
```

```{r}
paralog_table[is.na(paralog_table$afumigatus_eg_paralog_ensembl_gene),]
length(unique(paralog_table$ensembl_gene_id[is.na(paralog_table$afumigatus_eg_paralog_ensembl_gene)]))

# for 48 targets, not homologs were found!

# This leads me to believe that BPGA has a rule based on which they determine 
# which genes to remove from the output (e.g., paralogs based on BLAST but not in the database, or even something else).
```

```{r}
# we might need to manually look for paralogs for these genes
# e.g. AFUA_3G08660 (test also AFUA_2G11270 as positive control)
# Input https://blast.ncbi.nlm.nih.gov/Blast.cgi
# Input for https://orthomcl.org/orthomcl/app/search/sequence/ByBlast
```




Assign each target to the paralog that is in Afu_id, and that has the highest
afumigatus_eg_paralog_perc_id_r1.

```{r}
test <- paralog_table[paralog_table$afumigatus_eg_paralog_ensembl_gene%in%BPGA$Afu_ID,] #TODO:check!
table(target%in%test$ensembl_gene_id)
```

```{r}
target[!target%in%test$ensembl_gene_id]
```

```{r}
paralog_table[paralog_table$ensembl_gene_id == "AFUA_3G08660",]

sum(is.na(paralog_table$afumigatus_eg_paralog_ensembl_gene))

paralog_table[24,2]
```


So these are genes that are in the GEM, not in BPGA (so no cluster_id yet),
but that also has no paralogs. From CAoGD, I can confirm this gene has
no orthologs.




```{r}
# only retain within-species paralogs
test <- paralog_table[paralog_table$afumigatus_eg_paralog_orthology_type=="within_species_paralog",]
table(target%in%test$ensembl_gene_id) # for some of the targets, we did not identify a within-specie paralog. Inspect:
```



```{r}
"AFUA_2G17500" %in% fumigatus_gem$GENES$NAME
"AFUA_2G17500" %in% BPGA$Afu_ID
paralog_table[paralog_table$ensembl_gene_id == "AFUA_2G17500",]
```





Goal: for each target, find a paralog that is already in BPGA

```{r}
test <- test[test$afumigatus_eg_paralog_ensembl_gene%in%BPGA$Afu_ID,] #TODO:check!
```

```{r}
match(test$ensembl_gene_id,target)
```

```{r}
target[1]
table(target%in%test)
```




```{r}
test[test$afumigatus_eg_paralog_ensembl_gene=="AFUA_3G00910",]
c("AFUA_2G11270","AFUA_3G00910","AFUA_1G15440")
```

```{r}
test <- test[test$afumigatus_eg_paralog_ensembl_gene%in%BPGA$Afu_ID,] #TODO:check!
test
```


```{r}
"AFUA_1G00510" %in% BPGA$Afu_ID
"AFUA_1G00510" %in% Fumigatus_GENES$NAME
"AFUA_1G00510" %in% fumigatus_input$X.Format.Gene.ID

test[test$ensembl_gene_id == "AFUA_1G00510",]
```

```{r}
BPGA[BPGA$Afu_ID%in%test$afumigatus_eg_paralog_ensembl_gene[test$ensembl_gene_id == "AFUA_1G00510"],]
```


For each of the genes in GEM but not in BPGA, I need to add them to BPGA, and 
assign them with the same cluster label as their paralogs

```{r}
# df to add
df <- data.frame(Afu_id = target,)
```





```{r}
test
```




## Oryzae

```{r}
ensembl_or <- useEnsemblGenomes(biomart = "fungi_mart", dataset = "aoryzae_eg_gene")
test <- getBM(attributes = c("ensembl_gene_id","aoryzae_eg_paralog_ensembl_gene"), 
              mart = ensembl_or,
              filters = "ensembl_gene_id",
              values="AO090026000523")
test
```

```{r}
ensembl_nig <- useEnsemblGenomes(biomart = "fungi_mart", dataset = "aniger_eg_gene")  # wrong strain, but let's see
test <- getBM(attributes = c("ensembl_gene_id","aniger_eg_paralog_ensembl_gene"), 
              mart = ensembl_nig,
              filters = "ensembl_gene_id",
              values="AO090026000523")
test
```



