---
title: "Consolidate metabolites for A. niger"
author: "Jeroen Gilis"
date: "7 October 2024"
output: 
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

```{r, message=FALSE, warning=FALSE}
library(stringr)
library(openxlsx)
library(RCurl)
library(kableExtra)
library(data.table)
```

## Helpers

```{r}
source("./Rutilities.R")
```

# Obtain GEM

```{r}
# Note: excels should be closed when reading them in R
niger_gem <- read_excel_allsheets("../data/BaseGEMs/Niger_GEM.xlsx")
```

# Inspect

```{r}
head(niger_gem$METS)
```

```{r}
table(grepl("kegg.compound",niger_gem$METS$MIRIAM)) # KEGG for half
table(grepl("chebi/CHEBI",niger_gem$METS$MIRIAM)) # Chebi for most
table(is.na(niger_gem$METS$InChI)) # inchi for half
```

- kegg.compound and Chebi for many (see also REPLACEMENT.ID)
- Inchi for half
- for some, no formalized, only names

```{r}
table(grepl("kegg",niger_gem$METS$MIRIAM),
      grepl("chebi",niger_gem$METS$MIRIAM))

table(is.na(niger_gem$METS[!grepl("kegg",niger_gem$METS$MIRIAM) &
                 !grepl("chebi",niger_gem$METS$MIRIAM),"InChI"]))
# if no KEGG nor CHEBI, inchi can help a bit
```

# Wrangle

```{r}
# Get KEGG column
niger_gem$METS$KEGG <- niger_gem$METS$MIRIAM
niger_gem$METS$KEGG[!grepl("kegg",niger_gem$METS$KEGG)] <- NA
niger_gem$METS$KEGG <- gsub("\\;.*","",gsub(".*kegg.compound/(.+)", "\\1", niger_gem$METS$KEGG))

# Get chebi column
niger_gem$METS$CHEBI <- niger_gem$METS$MIRIAM
niger_gem$METS$CHEBI[!grepl("chebi",niger_gem$METS$CHEBI)] <- NA
niger_gem$METS$CHEBI <- gsub("\\;.*","",gsub(".*chebi/(.+)", "\\1", niger_gem$METS$CHEBI))
```


# Get chebi based on KEGG

Get chebi based on KEGG. Perhaps best to do this for all KEGG IDs, because they 
can be linked to  multiple CHEBIs, and we currently only have 1 per KEGG.

## Pubchem compound

```{r, eval=FALSE}
kegg_query <- niger_gem$METS$KEGG[which(!is.na(niger_gem$METS$KEGG))]
kegg_query <- unique(kegg_query) # only search each query once
system.time({
CHEBI_comp <- linkMetaboliteIdentifiers(query = kegg_query, 
                                       target = "^CHEBI", 
                                       reader = "read.csv",
                                       database = "pubchem_compound",
                                       verbose = TRUE)
}) 
# 1244.88s for 1584 queries (0.79s/query)

table(!is.na(CHEBI_comp$target)) # 2/3 found back
CHEBI_comp <- CHEBI_comp[!is.na(CHEBI_comp$target),]
saveRDS(CHEBI_comp,"../data/intermediate/niger_KEGGtoCHEBI_pubcomp.rds")
```

```{r}
CHEBI_comp <- readRDS("../data/intermediate/niger_KEGGtoCHEBI_pubcomp.rds")
```

TODO
1. Some of the new CHEBIs differ from the original CHEBIs
2. Line 22 is CHEBI:15490; (Acyl-CoA); [M+H]+;

```{r}
niger_gem$METS <- consolidateIDs(data = niger_gem$METS,
                                     queryID = "KEGG",
                                     originalID = "CHEBI",
                                     link = CHEBI_comp,
                                     strategy = "union",
                                     keepOriginal = TRUE)
```

```{r}
niger_gem$METS[c(1,8,16,22), c("KEGG","CHEBI_original","CHEBI_new","CHEBI_consolidated")]
```

```{r}
table(!is.na(niger_gem$METS$KEGG),!is.na(niger_gem$METS$CHEBI_original))
table(!is.na(niger_gem$METS$KEGG),!is.na(niger_gem$METS$CHEBI_consolidated))
```

```{r}
niger_gem$METS$CHEBI <- niger_gem$METS$CHEBI_consolidated
niger_gem$METS$CHEBI_original <- NULL
niger_gem$METS$CHEBI_new <- NULL
niger_gem$METS$CHEBI_consolidated <- NULL
```

## Pubchem substance

Here we attempted to search all KEGG IDs in pubchem_substance to identify
links with CHEBI. However, this resulted in only 9 succesful queries, and all of
these were already obtained from pubchem_compound

```{r}
kegg_query <- niger_gem$METS$KEGG[which(!is.na(niger_gem$METS$KEGG))]
kegg_query <- unique(kegg_query) # only search each query once
system.time({
CHEBI_sub <- linkMetaboliteIdentifiers(query = kegg_query, 
                                       target = "^CHEBI", 
                                       reader = "read.csv",
                                       database = "pubchem_substance",
                                       verbose = TRUE)
}) 
saveRDS(CHEBI_sub,"../data/intermediate/niger_KEGGtoCHEBI_pubsub.rds")
```

```{r}
CHEBI_sub <- readRDS("../data/intermediate/niger_KEGGtoCHEBI_pubsub.rds")
(CHEBI_sub <- CHEBI_sub[!is.na(CHEBI_sub$target),])
# We are only able to add 9 CHEBIs based on this
```

```
niger_gem$METS <- consolidateIDs(data = niger_gem$METS,
                                     queryID = "KEGG",
                                     originalID = "CHEBI",
                                     link = CHEBI_sub,
                                     strategy = "union",
                                     keepOriginal = TRUE)

niger_gem$METS[!is.na(niger_gem$METS$CHEBI_new),
                   c("KEGG","CHEBI_original","CHEBI_new","CHEBI_consolidated")]
```

# Get KEGG based on chebi

## pubchem compound

Now, try to add KEGGs for the IDs for which we have CHEBI.

Previously, I searched all KEGGs, because they can be associated with multiple 
CHEBIs. As each CHEBI should only have one KEGG, I here only search for the 
missing ones. Searching for all could be relevant if we need to find mistakes in
the original annotation.

```{r}
chebi_query <- niger_gem$METS$CHEBI[
  which(is.na(niger_gem$METS$KEGG) &
          !is.na(niger_gem$METS$CHEBI))]
chebi_query <- unique(unlist(str_split(chebi_query,pattern = "/")))

KEGG_comp <- linkMetaboliteIdentifiers(query = chebi_query, 
                                       target = "^C(\\d{5})\\b", 
                                       reader = "read.csv",
                                       database = "pubchem_compound",
                                       verbose = TRUE)

saveRDS(KEGG_comp,"../data/intermediate/niger_CHEBItoKEGG_pubcomp.rds")
```

```
niger_gem$METS <- consolidateIDs(data = niger_gem$METS,
                                     queryID = "CHEBI",
                                     originalID = "KEGG",
                                     link = KEGG_comp,
                                     strategy = "union",
                                     keepOriginal = TRUE)

niger_gem$METS[!is.na(niger_gem$METS$KEGG_new),
                   c("CHEBI","KEGG_original","KEGG_new","KEGG_consolidated")]
```

```
niger_gem$METS$KEGG <- niger_gem$METS$KEGG_consolidated
niger_gem$METS$KEGG_original <- NULL
niger_gem$METS$KEGG_new <- NULL
niger_gem$METS$KEGG_consolidated <- NULL
```

```
table(grepl("kegg",niger_gem$METS$MIRIAM),
      grepl("chebi",niger_gem$METS$MIRIAM))

table(!is.na(niger_gem$METS$KEGG),
      !is.na(niger_gem$METS$CHEBI)) # Added 24 links
```

## pubchem substance

Provides no additional hits!

```{r}
chebi_query <- niger_gem$METS$CHEBI[
  which(is.na(niger_gem$METS$KEGG) &
          !is.na(niger_gem$METS$CHEBI))]
chebi_query <- unique(unlist(str_split(chebi_query,pattern = "/")))

KEGG_sub <- linkMetaboliteIdentifiers(query = chebi_query, 
                                      target = "^C(\\d{5})\\b", 
                                      reader = "read.csv",
                                      database = "pubchem_substance",
                                      verbose = TRUE)

saveRDS(KEGG_sub,"../data/intermediate/niger_KEGGtoCHEBI_pubcomp.rds")
```













