---
title: "Consolidating metabolites across GEMs"
author: "Jeroen Gilis"
date: "4 October 2024"
output: 
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

```{r, message=FALSE, warning=FALSE}
library(stringr)
library(openxlsx)
library(RCurl)
library("kableExtra")
```

## Helpers

```{r}
read_excel_allsheets <- function(filename) {
    sheets <- openxlsx::getSheetNames(filename)
    x <- lapply(sheets, function(X) openxlsx::read.xlsx(filename, sheet = X))
    names(x) <- sheets
    x
}
```

```{r}
#TODO: if necessary, allow timeout option. Currently, we are probably working
# under global options timeout, getOption('timeout'), but not sure how the
# error will look like in the case of timeout (i.e., if it will be properly
# caught by try-error)

#TODO: perhaps useful to add a "bidirectional option": for each query, find each
# target, and for each target, find each of type query. This could also be done
# in subsequent runs, but having it in one function could be handy
# Do different chebi's ever return multiple KEGG? Probably not.
# But the other way around this can happen.

#TODO: at the moment, the function has placeholders for (1) including multiple
# types of targets and (2) using different R functions to read the data from the 
# web

linkMetaboliteIdentifiers <- function(query, 
                                      target = c("CHEBI"), 
                                      reader = c("csv"),
                                      database = c("pubchem_compound",
                                                   "pubchem_substance"),
                                      bidirectional = c(FALSE, TRUE),
                                      verbose = c(TRUE, FALSE)){
  
  res <- unlist(lapply(query, function(q){
      
      if(database == "pubchem_compound"){
        url <- paste0("https://pubchem.ncbi.nlm.nih.gov/rest/pug/compound/name/",
                      q,
                      "/synonyms/TXT")
      } else if(database == "pubchem_substance"){
        url <- paste0("https://pubchem.ncbi.nlm.nih.gov/rest/pug/substance/name/",
                      q,
                      "/synonyms/TXT")
      }
      
      if(reader == "csv"){
        data <- tryCatch({
          suppressWarnings(read.csv(url,header = FALSE))
          },error = function(e) e)
      }
      
      if (inherits(data, "error")) {
        return("Unknown query")
      }
      
      if(sum(grepl(target,data$V1)) > 0){ # see if faster option exists
        return(paste0(data$V1[grep(target,data$V1)],collapse = "/")) # works also if multiple chebi's
      } else{
        return(paste0("No ", target))
      }

  }))
  
  if(verbose){
      warning("Number of unknown queries: ", sum(res == "Unknown query"), "\n")
      warning("Number of queries without ", target,  ": ", sum(res == paste0("No ", target)))
  }
  
  error <- rep(NA,length=length(res))
  error[res == "Unknown query"] <- "Unknown query"
  error[res == paste0("No ", target)] <- paste0("No ", target)
  res[which(res %in% c("Unknown query",
                       paste0("No ", target)))] <- NA

  return(data.frame(query = query,
                    target = res,
                    error = error))
}
```

```
test <- linkMetaboliteIdentifiers(query = kegg_query[120:130], 
                                  target = "CHEBI",
                                  reader = "csv",
                                  database = "pubchem",
                                  verbose = TRUE)
test
```

```{r}
consolidateIDs <- function(data,
                           queryID,
                           originalID,
                           link,
                           strategy = c("union","intersection","keep_original","keep_new"),
                           keepOriginal = c(TRUE, FALSE)){
  
    stopifnot("queryID not in colnames(data)" = queryID %in% colnames(data))
    #TODO make less cryptic
    stopifnot("Mismatch between link and data" = all(!is.na(match(link$query, data[[queryID]]))))
    
    newLink <- link$target[match(data[[queryID]],link$query)]
    
    consolidated <- unlist(sapply(seq_along(data[[originalID]]), 
                                                   function(x){
      if(is.na(data[[originalID]][x])){
        return(newLink[x])
      } else if(is.na(newLink[x])){
        return(data[[originalID]][x])
      } else{
        
          original_x <- unlist(str_split(data[[originalID]][x], pattern = "/"))
          new_x <- unlist(str_split(newLink[x], pattern = "/"))
          
          if(strategy == "union"){
            combined <- union(original_x,new_x)
          } else if(strategy == "intersect") {
            combined <- intersect(original_x,new_x)
          } else if(strategy == "original_only"){
            combined <- original_x
          } else if(strategy == "new_only"){
            combined <- new_x
          }
          
          return(paste0(combined,collapse = "/"))
      }
    }))
    
    if(keepOriginal){
      data[[paste0(originalID,"_new")]] <- newLink
      data[[paste0(originalID,"_consolidated")]] <- consolidated
      colnames(data)[which(colnames(data) == originalID)] <- paste0(originalID,"_original")
      return(data)
    } else {
      data[[originalID]] <- consolidated
    }
}
```

# Obtain gems

```{r}
# Note: excels should be closed when reading them in R
niger_gem <- read_excel_allsheets("../data/BaseGEMs/Niger_GEM.xlsx")
fumigatus_gem <- read_excel_allsheets("../data/BaseGEMs/Fumigatus_GEM.xlsx")
oryzae_gem <- read_excel_allsheets("../data/BaseGEMs/Niger_GEM.xlsx")
```

# Fumigatus

## Inspect

```{r}
colnames(fumigatus_gem$METS)
```

```{r}
sum(is.na(fumigatus_gem$METS$REPLACEMENT.ID)) # they used this in their analysis
# If available, they used the KEGG+compartment
# If not, then they used the CHEBI+compartment
# If that is also not available, they use "Corecocluster"+compartment
# Finally, they have a replacement ID for e.g. biomass, lipid, ...
```

```{r}
table(grepl("C(\\d{5})\\[",fumigatus_gem$METS$REPLACEMENT.ID)) # KEGG
table(grepl("CHEBI",fumigatus_gem$METS$REPLACEMENT.ID)) # CHEBI
table(grepl("Cluster",fumigatus_gem$METS$REPLACEMENT.ID)) # coreco cluster
table(grepl("C(\\d{3})\\[",fumigatus_gem$METS$REPLACEMENT.ID)) # CoreCo/C[digits]
```

```{r}
fumigatus_gem$METS[!grepl("C(\\d{5})\\[",fumigatus_gem$METS$REPLACEMENT.ID)&
                                    !grepl("CHEBI",fumigatus_gem$METS$REPLACEMENT.ID) &
                                    !grepl("Cluster",fumigatus_gem$METS$REPLACEMENT.ID) &
                                    !grepl("C(\\d{3})\\[",fumigatus_gem$METS$REPLACEMENT.ID),
                   c("NAME","MIRIAM","REPLACEMENT.ID")]
# these last 305 compounds are mainly tagged CoreCo or A.fumigatus.central.metabolism 
# in miriam. For instance, A.fumigatus.central.metabolism/CERB1A seems to refer to
# cerberin1 comp. A (not sure if comp. stands for compound, component, or compartiment).
```

- kegg.compound for many (see also REPLACEMENT.ID), except biomass/lipid/...
- InChi, bigg, biocyc, chebi, hmdb, metanet, pubchem, and seed.compound for many


```{r}
table(grepl("kegg.compound",fumigatus_gem$METS$MIRIAM))
table(grepl("chebi/CHEBI",fumigatus_gem$METS$MIRIAM))

table(grepl("kegg",fumigatus_gem$METS$MIRIAM),
      grepl("chebi",fumigatus_gem$METS$MIRIAM))
```

## Wrangle

```{r}
# Get KEGG column
fumigatus_gem$METS$KEGG <- fumigatus_gem$METS$MIRIAM
fumigatus_gem$METS$KEGG[!grepl("kegg",fumigatus_gem$METS$KEGG)] <- NA
fumigatus_gem$METS$KEGG <- gsub("\\;.*","",gsub(".*kegg.compound/(.+)", "\\1", fumigatus_gem$METS$KEGG))

# Get chebi column
fumigatus_gem$METS$CHEBI <- fumigatus_gem$METS$MIRIAM
fumigatus_gem$METS$CHEBI[!grepl("chebi",fumigatus_gem$METS$CHEBI)] <- NA
fumigatus_gem$METS$CHEBI <- gsub("\\;.*","",gsub(".*chebi/(.+)", "\\1", fumigatus_gem$METS$CHEBI))

table(!is.na(fumigatus_gem$METS$KEGG),
      !is.na(fumigatus_gem$METS$CHEBI))
```

## Get chebi based on KEGG

Get chebi based on KEGG. Perhaps best to do this for all KEGG IDs, because they 
can be linked to  multiple CHEBIs, and we currently only have 1 per KEGG.

```{r}
kegg_query <- fumigatus_gem$METS$KEGG[which(!is.na(fumigatus_gem$METS$KEGG))]
kegg_query <- unique(kegg_query) # only search each query once
system.time({
CHEBI_res <- linkMetaboliteIdentifiers(query = kegg_query, 
                                       target = "CHEBI", 
                                       reader = "csv",
                                       database = "pubchem_compound",
                                       verbose = TRUE)
}) 
# 1244.88s for 1584 queries (0.79s/query)
```

Some of the new CHEBIs differ from the original CHEBIs:

```{r}
fumigatus_gem$METS <- consolidateIDs(data = fumigatus_gem$METS,
                                     queryID = "KEGG",
                                     originalID = "CHEBI",
                                     link = CHEBI_res,
                                     strategy = "union",
                                     keepOriginal = TRUE)
```

```{r}
table(grepl("kegg",fumigatus_gem$METS$MIRIAM),
      grepl("chebi",fumigatus_gem$METS$MIRIAM))

table(grepl("kegg",fumigatus_gem$METS$MIRIAM),
      !is.na(fumigatus_gem$METS$CHEBI_consolidated)) # added 745 links between KEGG and CHEBI!
```

## Get KEGG based on chebi

Now, try to add KEGGs for the IDs for which we have CHEBI.
Previously, I searched all KEGGs, because they can be associated with multiple CHEBIs.
As each CHEBI should only have one KEGG, I here only search for the missing ones.
Searching for all could be relevant if we need to find mistakes in the original
annotation.

```{r}
chebi_query <- fumigatus_gem$METS$CHEBI_consolidated[
  which(is.na(fumigatus_gem$METS$KEGG) &
          !is.na(fumigatus_gem$METS$CHEBI_consolidated))]

chebi_query <- unique(unlist(str_split(chebi_query,pattern = "/")))

KEGG_res <- linkMetaboliteIdentifiers(query = chebi_query, 
                                      target = "\\bC(\\d{5})\\b", 
                                      reader = "csv",
                                      database = "pubchem_compound",
                                      verbose = TRUE)
```

```{r}
fumigatus_gem$METS <- consolidateIDs(data = fumigatus_gem$METS,
                                     queryID = "CHEBI_consolidated",
                                     originalID = "KEGG",
                                     link = KEGG_res,
                                     strategy = "union",
                                     keepOriginal = TRUE)
```

```{r}
table(grepl("kegg",fumigatus_gem$METS$MIRIAM),
      grepl("chebi",fumigatus_gem$METS$MIRIAM))

table(!is.na(fumigatus_gem$METS$KEGG_consolidated),
      !is.na(fumigatus_gem$METS$CHEBI_consolidated)) # Added 12 more links. 
# This is very limited. But interestingly, 152/165 CHEBIs without KEGG are
# CHEBI:52320

chebi_query <- fumigatus_gem$METS$CHEBI_consolidated[which(is.na(fumigatus_gem$METS$KEGG_consolidated) &
                                                     !is.na(fumigatus_gem$METS$CHEBI_consolidated))]
table(chebi_query)
```
Curate later! This could be done based either on replacement.ID or name.

## Get CHEBI based on name

```{r}
query_name <- fumigatus_gem$METS$NAME[is.na(fumigatus_gem$METS$CHEBI_both)]
length(unique(query_name))

# trim localization
query_name <- gsub(pattern=" \\[.*","",query_name)
length(unique(query_name))

# decapitalize
query_name <- tolower(query_name)
length(unique(query_name))

# only uniques
query_name <- unique(query_name)
```

```{r}
res_comp <- linkMetaboliteIdentifiers(query = query_name[100:199], 
                                 target = "CHEBI",
                                 reader = "csv",
                                 database = "pubchem_compound",
                                 verbose = TRUE)
# 65 fails: 57 unknown, 8 without chebi
```
```{r}
res_sub <- linkMetaboliteIdentifiers(query = query_name[100:199], 
                                 target = "CHEBI",
                                 reader = "csv",
                                 database = "pubchem_substance",
                                 verbose = TRUE)
# 53 fails: 37 unknown, 16 without chebi
```

```{r}
res <- data.frame(query = res_comp$query,
                  compound=res_comp$target,
                  compound_error=res_comp$error,
                  substance=res_sub$target, 
                  substance_error=res_sub$error)


res$both <- unlist(sapply(seq_along(res$substance),function(x){
  if(is.na(res$substance[x])){
    return(res$compound[x])
  } else if(is.na(res$compound[x])){
    return(res$substance[x])
  } else{
    combined <- union(unlist(str_split(res$substance[x], 
                                       pattern = "/")),
                      unlist(str_split(res$compound[x], 
                                       pattern = "/")))
    return(paste0(combined,collapse = "/"))
  }
}))
table(is.na(res$both))
```

Approximately 50% of the compounds could be annotated already, not bad.
Of course, we are not 100% sure if all of these are all correct.
But let's first focus on the ones we didnt find:

Why are the following not found

```{r}
table(res$compound_error,res$substance_error) # mostly, string not retrieved
```




Many are in the form of: 

- (glcnac)2 (man)8 (pp-dol)1
- protein asparagine
- 1r, (1r,2s)
- trailin (in dna)

```{r}
query_name <- res$query[is.na(res$both)]
query_name
```

```{r}
q <- "(glcn)1 (ino(acyl)-p)1 (man)3 (etn)1 (p)1"
q <- "(GlcN)1 (Ino(acyl)-P)1 (Man)2 (EtN)1 (P)1"
q <- "G00148"
url <- paste0("https://pubchem.ncbi.nlm.nih.gov/rest/pug/substance/name/",
              q,
              "/synonyms/TXT")
read.csv(url,header = FALSE)
```

maybe I shouldnt be looking for name, but for ... formula?


```{r}
q <- "galactosylceramide"
q <- "CHEBI:36498"
url <- paste0("https://pubchem.ncbi.nlm.nih.gov/rest/pug/substance/name/",
              q,
              "/synonyms/TXT")
data <- tryCatch({
  suppressWarnings(read.csv(url,header = FALSE))
  },error = function(e) e)
data
```


# Oryzae

## Inspect


```{r}
head(oryzae_gem$METS)
```

- Problematic: no formalized identifiers, only ID, NAME and REPLACEMENT.ID

# Niger

## Inspect

```{r}
head(niger_gem$METS)
```

```{r}
niger_gem$METS[which(is.na(niger_gem$METS$MIRIAM)),]
```

```{r}
table(is.na(niger_gem$METS$InChI))
```


- kegg.compound and Chebi for many (see also REPLACEMENT.ID)
- Inchi for half
- for some, no formalized identifiers

It seems best to go for both KEGG and Chebi.

# Scrap

--- 3-cyanopyridine

```{r}
niger_gem$METS[which(is.na(niger_gem$METS$MIRIAM)),][1,]
```

For this one, I have the inchi.
I can search the inchi using InChI:InChI=1S/C6H4N2/c7-4-6-2-1-3-8-5-6/h1-3,5H
in https://metacyc.org/metabolite-translation-service.shtml.

This gives me several identifiers, but not kegg.

I then look for 3-cyanopyridine and 3-Cyanopyridine in 
https://www.genome.jp/kegg/compound/ -> no hit.

I then take the molecular mass from the metacyc search, and feed that to kegg:
no hit, but kegg does have 2-Cyanopyridine and 4-Cyanopyridine, just no
3-Cyanopyridine.

But, from metacyc we also get Chebi (86556). If we query that in 
https://www.ebi.ac.uk/chebi/, we get 3-pyridinecarbonitrile, which turns out to 
be synonym for 3-Cyanopyridine.

However, again, searching this in kegg does not give a hit.

This compound seems really not exist in KEGG. So what's next? Not use KEGG at all?
But is there even a database in which all required compounds exist?

--- C00659

```{r}
fumigatus_gem$METS[1:2,]
```
  
We have the kegg compound for this, let's try to find chebi

Via KEGG, we can manually click to chebi (C00659 -> 982)

But when trying to do this more automatically in metacyc, it fails.
Kegg:C00659   ChEBI:982   InChI:InChI=1S/C6H10O4/c1-3-6(10,4(2)7)5(8)9/h10H,3H2,1-2H3,(H,8,9)   InChIkey:InChIKey=VUQLHQFKACOHNZ-UHFFFAOYSA-N   CF:C6H10O4

This is probably inevitable; I strongly suspect that the people that constructed
this GEM in the first place used Metabolite Translation Service for MetaCyc.
As such, I would return as much information as they did, not more.

I could also input C000659 in
https://www.metaboanalyst.ca/faces/upload/ConvertView.xhtml
this succeeds in returning Chebi

One more possibility would be to do this in R using MetaFetcheR. However, this
is a non-maintained github page that failed to install out of the box and that
requires installing additional things; as such, this seems far from ideal.

One other possibility is to use the pypi library bioservices
https://bioservices.readthedocs.io/en/latest/convertor_tutorial.html

This seems to work well! I could actually just obtain a Kegg-Chebi dictionary
from there and import it here!

```{r}
kegg_chebi_dict <- read.csv2("../data/intermediate//kegg_chebi_dict.csv", 
                  sep=",",
                  na.strings = "",
                  header = FALSE)
colnames(kegg_chebi_dict) <- c("KEGG","CHEBI")
kegg_chebi_dict$KEGG <- gsub(".*cpd:(.+)", "\\1", kegg_chebi_dict$KEGG)
kegg_chebi_dict$CHEBI <- gsub(".*chebi:(.+)", "\\1", kegg_chebi_dict$CHEBI)

head(kegg_chebi_dict)
```

```{r}
# update using our KEGG/CHEBI dictionary
table(fumigatus_gem$METS$KEGG %in% kegg_chebi_dict$KEGG)
table(fumigatus_gem$METS$CHEBI %in% kegg_chebi_dict$CHEBI)

# THE DICTIONARY IS INCOMPLETE...
# For instance
"C04253" %in% fumigatus_gem$METS$KEGG
"C04253" %in% kegg_chebi_dict$KEGG 
# indeed, no chebi for this kegg, also not on kegg website (could link to pubchem though)
# also no kegg-chebi link through metacyc (could link to inchi though)
# This makes sense, on KEGG the compound is called  	
# Electron-transferring flavoprotein, which is  generic term, not a real compound


# Conclusion, not all KEGGs have a chebi
# Let's see if we can still get something from it
```

```{r}
fumigatus_gem$METS$CHEBI_dic <- kegg_chebi_dict$CHEBI[match(fumigatus_gem$METS$KEGG,kegg_chebi_dict$KEGG)]
```

```{r}
head(fumigatus_gem$METS[,c("KEGG","CHEBI","CHEBI_dic")])
```

```{r}
# notice that there are discrepancies between CHEBI and CHEBI_dic:
fumigatus_gem$METS[which(fumigatus_gem$METS$KEGG=="C00009"),c("KEGG","CHEBI","CHEBI_dic")]
# these are the protonated vs the non-protonated forms of the compound!
```

```{r}
fumigatus_gem$METS[fumigatus_gem$METS$KEGG=="C00659",c("KEGG","CHEBI","CHEBI_dic")]
```


```{r}
fumigatus_gem$METS$KEGG[which(!is.na(fumigatus_gem$METS$KEGG)&
                                is.na(fumigatus_gem$METS$CHEBI))]
```


```{r}
# then we also have compounds with chebi but no kegg
fumigatus_gem$METS[c(59,60),c("KEGG","CHEBI","CHEBI_dic")]

# if the CHEBI is in kegg_chebi_dict, add it:
target <- fumigatus_gem$METS$CHEBI[which(!is.na(fumigatus_gem$METS$CHEBI)&
                           is.na(fumigatus_gem$METS$CHEBI_dic))]
table(target) # mostly chebi compound 52320 (ergosteryl ester??)
```


```{r}
fumigatus_gem$METS[which(fumigatus_gem$METS$CHEBI=="52320"),]
```

Clearly indicates that, at least for these compounds, we'd better work
with the replacement ID column, which has the proper CHEBIs!







