---
title: "Link oryzae to niger metabolites"
author: "Jeroen Gilis"
date: "5 November 2024"
output: 
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

# Libraries

```{r, message=FALSE, warning=FALSE}
library(stringr)
library(openxlsx)
library(RCurl)
library(kableExtra)
```

# Helpers

```{r}
source("./Rutilities.R")
```

# Load GEMs

```{r}
# Note: excels should be closed when reading them in R
Aor_draft <- read_excel_allsheets("../model/Aor_draft_comp.xlsx")

# Aor_reduced <- read_excel_allsheets("../scrap/Aor_reduced.xlsx")
Aor_reduced <- read_excel_allsheets("../data/intermediate/Aor_reduced.xlsx")

#Ani_draft <- read_excel_allsheets("../model/Ani_draft_consolidated.xlsx")
Ani_draft <- read_excel_allsheets("../data/intermediate/Ani_draft_comp_mets.xlsx")
```

# Wrangle

```{r}
Aor_reduced$METS$CHEBI <- Aor_reduced$METS$MIRIAM
Aor_reduced$METS$CHEBI[!grepl("chebi",Aor_reduced$METS$CHEBI)] <- NA
Aor_reduced$METS$CHEBI <- gsub("chebi:","",Aor_reduced$METS$CHEBI)
Aor_reduced$METS$CHEBI <- gsub("/",":",Aor_reduced$METS$CHEBI)

Aor_reduced$METS$KEGG <- Aor_reduced$METS$MIRIAM
Aor_reduced$METS$KEGG[!grepl("kegg",Aor_reduced$METS$KEGG)] <- NA
Aor_reduced$METS$KEGG <- gsub("kegg.compound/","",Aor_reduced$METS$KEGG)
Aor_reduced$METS$KEGG[!is.na(Aor_reduced$METS$KEGG)]
```

The replacement IDs for both oryzae niger can be improved. If the
compartment is not cytosol, a trailing compartment name was added. Fix this:

```{r}
Aor_draft$METS[Aor_draft$METS$NAME %in% c("3-oxoacyl-[acp]","acetoacetyl-CoA"),c(2,3,9)]
Aor_reduced$METS[Aor_reduced$METS$NAME %in% c("3-oxoacyl-[acp]","acetoacetyl-CoA"),c(2,3,9)]
```

```{r}
Aor_reduced$METS$`REPLACEMENT ID` <- gsub("m\\[m\\]", "\\[m\\]", Aor_reduced$METS$`REPLACEMENT ID`)
Aor_reduced$METS$`REPLACEMENT ID` <- gsub("e\\[e\\]", "\\[e\\]", Aor_reduced$METS$`REPLACEMENT ID`)
Aor_reduced$METS$`REPLACEMENT ID` <- gsub("p\\[p\\]", "\\[p\\]", Aor_reduced$METS$`REPLACEMENT ID`)
```

```{r}
length(Aor_reduced$METS$`REPLACEMENT ID`)
length(unique(Aor_reduced$METS$`REPLACEMENT ID`))
```

# Consolidate metabolites

```{r}
sum(!is.na(Aor_reduced$METS$KEGG)) # very few keggs in Aor
sum(Aor_reduced$METS$KEGG %in% Ani_draft$METS$KEGG[!is.na(Ani_draft$METS$KEGG)])
sum(Aor_reduced$METS$KEGG %in% Ani_draft$METS$KEGG_self[!is.na(Ani_draft$METS$KEGG_self)])

sum(!is.na(Aor_reduced$METS$CHEBI))
sum(Aor_reduced$METS$CHEBI %in% Ani_draft$METS$CHEBI[!is.na(Ani_draft$METS$CHEBI)])
# so the 54 you lose here are by definition unique to Aor, because while the 
# name is existing in Ani, it is not existing in that compartment

sum(tolower(Aor_reduced$METS$NAME) %in% tolower(Ani_draft$METS$NAME))
sum(tolower(Aor_reduced$METS$ID) %in% tolower(Ani_draft$METS$ID)) 
# so the 54 you lose here are by definition unique to Aor, because while the 
# name is existing in Ani, it is not existing in that compartment

sum(Aor_reduced$METS$`REPLACEMENT ID` %in% Ani_draft$METS$`REPLACEMENT ID`)
sum(tolower(Aor_reduced$METS$`REPLACEMENT ID`) %in% tolower(Ani_draft$METS$`REPLACEMENT ID`))
# the main question is: do we trust matches based on replacement id?
# -> match on proper IDs first, then manually inspect if we can trust the
# remaining matches based on replacement_id.
```

## CHEBI

```{r}
Aor_reduced$METS$CHEBI_ID <- paste0(Aor_reduced$METS$CHEBI,"[", Aor_reduced$METS$COMPARTMENT, "]")
Aor_reduced$METS$CHEBI_ID[!grepl("CHEBI",Aor_reduced$METS$CHEBI_ID)] <- NA

Ani_draft$METS$CHEBI_ID <- paste0(Ani_draft$METS$CHEBI,"[", Ani_draft$METS$COMPARTMENT, "]")
Ani_draft$METS$CHEBI_ID[!grepl("CHEBI",Ani_draft$METS$CHEBI_ID)] <- NA
```

```{r}
sum(Aor_reduced$METS$CHEBI %in% Ani_draft$METS$CHEBI[!is.na(Ani_draft$METS$CHEBI)])
sum(Aor_reduced$METS$CHEBI_ID %in% Ani_draft$METS$CHEBI_ID[!is.na(Ani_draft$METS$CHEBI_ID)])
# so the 45 you lose here are by definition unique to Aor, because while the 
# chebi is existing in Ani, it is not existing in that compartment
```

```{r}
CHEBI_both <- Aor_reduced$METS$CHEBI[Aor_reduced$METS$CHEBI %in% Ani_draft$METS$CHEBI[!is.na(Ani_draft$METS$CHEBI)]]
CHEBI_ID_both <- Aor_reduced$METS$CHEBI_ID[Aor_reduced$METS$CHEBI_ID %in% Ani_draft$METS$CHEBI_ID[!is.na(Ani_draft$METS$CHEBI_ID)]]
```

We probably will never be able to map these metabolites between Aor and Ani,
i.e., they exist in both Aor and Ani, but only in Aor it (also) exists a 
different compartment. Save these for later:

```{r}
CHEBI_altcomp <- Aor_reduced$METS$CHEBI_ID[Aor_reduced$METS$CHEBI %in% CHEBI_both &
                            !Aor_reduced$METS$CHEBI_ID %in% CHEBI_ID_both]
```

Map oryzae metabolites to niger:

```{r}
table(Aor_reduced$METS$`REPLACEMENT ID` %in% Ani_draft$METS$`REPLACEMENT ID`)
Aor_reduced$METS$`REPLACEMENT ID`[match(CHEBI_ID_both, Aor_reduced$METS$CHEBI_ID)] <- Ani_draft$METS$`REPLACEMENT ID`[match(CHEBI_ID_both, Ani_draft$METS$CHEBI_ID)]
table(Aor_reduced$METS$`REPLACEMENT ID` %in% Ani_draft$METS$`REPLACEMENT ID`)
```

```{r}
length(Aor_reduced$METS$`REPLACEMENT ID`)
length(unique(Aor_reduced$METS$`REPLACEMENT ID`))
```

```{r}
hlp <- Aor_reduced$METS[Aor_reduced$METS$`REPLACEMENT ID` %in% 
                   Aor_reduced$METS$`REPLACEMENT ID`[duplicated(Aor_reduced$METS$`REPLACEMENT ID`)],c(2,3,5,9)]
hlp

Aor_before <- read_excel_allsheets("../data/intermediate/Aor_reduced.xlsx")
Aor_before$METS[match(hlp$ID, Aor_before$METS$ID),c(2,3,5,9)]
```


```{r}
Ani_draft$METS[which(Ani_draft$METS$CHEBI == "CHEBI:24636"),c(2,5,9)]
```

Two different situations:

- H[c]: in Aor_draft, they made a distinction between protons in the cytoplasm 
and protons in the cytoplasm participating in energy metabolism. In Ani draft,
hydrogen does not exist, only H+. But when we link through chebi, hydrogen and
H+ get the same replacement_id in Aor_draft.

- For the fatty acid, probably there was a mistake in the original annotation.
linolenoyl-CoA should really be C183COA and not C182COA, as it has 3 double 
bonds. The one with 2 double bonds is linoleoyl-CoA, so easy to make a mistake.

For linolenoyl-CoA[c], leave the change as is. This should be fine, we just have
two different names linked to a single ID, but that should be fine and should 
happen often anyway when mapping between species.

For H_PO, it is probably best to keep the original ID. Make sure that in the end
of the script this is still true.

```{r}
Aor_reduced$METS[Aor_reduced$METS$ID == "H(+) (energy metabolism)[c]","REPLACEMENT ID"] <- "H_PO[c]"
```

Continue with the unlinked metabolites:

```{r}
target <- Aor_reduced$METS[!Aor_reduced$METS$CHEBI_ID %in% CHEBI_ID_both,]
```

## Name

```{r}
sum(tolower(target$NAME) %in% tolower(Ani_draft$METS$NAME))
sum(tolower(target$ID) %in% tolower(Ani_draft$METS$ID)) 
# so the 44 you lose here are by definition unique to Aor, because while the 
# name is existing in Ani, it is not existing in that compartment
```

```{r}
NAME_both <- tolower(target$NAME)[tolower(target$NAME) %in% tolower(Ani_draft$METS$NAME)]
ID_both <-  tolower(target$ID)[tolower(target$ID) %in% tolower(Ani_draft$METS$ID)]
```

We probably will never be able to map these metabolites between Aor and Ani,
i.e., they exist in both Aor and Ani, but only in Aor it (also) exists a 
different compartment.

```{r}
ID_altcomp <- tolower(target$ID)[tolower(target$NAME) %in% NAME_both &
                            !tolower(target$ID) %in% ID_both]
```

```{r}
table(Aor_reduced$METS$`REPLACEMENT ID` %in% Ani_draft$METS$`REPLACEMENT ID`)
Aor_reduced$METS$`REPLACEMENT ID`[match(ID_both, tolower(Aor_reduced$METS$ID))] <- Ani_draft$METS$`REPLACEMENT ID`[match(ID_both, tolower(Ani_draft$METS$ID))]
table(Aor_reduced$METS$`REPLACEMENT ID` %in% Ani_draft$METS$`REPLACEMENT ID`)
```

```{r}
length(Aor_reduced$METS$`REPLACEMENT ID`)
length(unique(Aor_reduced$METS$`REPLACEMENT ID`)) # linolenoyl-CoA, not a problem
```

Continue with the unlinked metabolites:

```{r}
target <- target[!tolower(target$ID) %in% ID_both,]
```

## Replacement ID

```{r}
sum(target$`REPLACEMENT ID` %in% Ani_draft$METS$`REPLACEMENT ID`) # these are okay
```

```{r}
Ani_draft$METS$`REPLACEMENT NAME` <- gsub("\\[.*","",Ani_draft$METS$`REPLACEMENT ID`)
target$`REPLACEMENT NAME` <- gsub("\\[.*","",target$`REPLACEMENT ID`)

replacement_NAME_both <- target$`REPLACEMENT NAME`[target$`REPLACEMENT NAME` %in% Ani_draft$METS$`REPLACEMENT NAME`]
replacement_ID_both <- target$`REPLACEMENT ID`[target$`REPLACEMENT ID` %in% Ani_draft$METS$`REPLACEMENT ID`]
```

```{r}
replacement_altcomp <- target$`REPLACEMENT ID`[target$`REPLACEMENT NAME` %in% replacement_NAME_both & 
                                   !target$`REPLACEMENT ID` %in% replacement_ID_both]
```

Map oryzae metabolites to niger:

```{r}
table(Aor_reduced$METS$`REPLACEMENT ID` %in% Ani_draft$METS$`REPLACEMENT ID`)
Aor_reduced$METS$`REPLACEMENT ID`[match(replacement_ID_both, Aor_reduced$METS$`REPLACEMENT ID`)] <- Ani_draft$METS$`REPLACEMENT ID`[match(replacement_ID_both, Ani_draft$METS$`REPLACEMENT ID`)]
table(Aor_reduced$METS$`REPLACEMENT ID` %in% Ani_draft$METS$`REPLACEMENT ID`)
```

```{r}
length(Aor_reduced$METS$`REPLACEMENT ID`)
length(unique(Aor_reduced$METS$`REPLACEMENT ID`)) # linolenoyl-CoA, not a problem
```

Continue with the unlinked metabolites:

```{r}
target <- target[!target$`REPLACEMENT ID` %in% replacement_ID_both,]
```

## KEGG

### Get kegg based on Chebi

The file kegg_chebi_dict.csv was created by running the script 
`map_kegg_chebi.ipynb`. This file contains curated links between CHEBI and KEGG
metabolite identifiers, and was constructed for the Bioservices Python library.

```{r}
kegg_chebi_dict <- read.csv2("../data/intermediate/kegg_chebi_dict.csv", header=FALSE, sep=",")

# wrangle
colnames(kegg_chebi_dict) <- c("KEGG","CHEBI")
kegg_chebi_dict$KEGG <- gsub("cpd:","",kegg_chebi_dict$KEGG)
kegg_chebi_dict$CHEBI <- gsub("chebi:","CHEBI:",kegg_chebi_dict$CHEBI)

target$KEGG_v2 <- kegg_chebi_dict$KEGG[match(target$CHEBI,kegg_chebi_dict$CHEBI)]
```

```{r}
# 2-Deoxy-D-ribose 5-phosphate should get C00673
# succinate semialdehyde should get C00232 
# (2S)-2-isopropyl-3-oxosuccinate should get C04236
Aor_before$METS[Aor_before$METS$ID %in% c("2-deoxy-D-ribose 5-phosphate[c]",
                                          "(2S)-2-isopropyl-3-oxosuccinate[c]",
                                          "succinate semialdehyde[c]",
                                          "D-ribose 5-phosphate[m]",
                                          "succinate[p]"),c(2,9)]


rbind(target[target$ID == "2-deoxy-D-ribose 5-phosphate[c]",c(2,12,15)],
      target[target$ID == "(2S)-2-isopropyl-3-oxosuccinate[c]",c(2,12,15)],
      target[target$ID == "succinate semialdehyde[c]",c(2,12,15)])
```

### Get kegg based on name

```{r}
name_query <- tolower(unique(target$NAME))
namesLink <- match(tolower(target$NAME),name_query)
name_query <- gsub(",", " ", name_query) # it works to replace commas for spaces
```

```{r,eval=FALSE}
addKEGG <- lapply(name_query, function(x){
  #print(x)
  query <- keggFind(database = "compound", query = x)
  return(cbind(x,
               paste0(gsub("cpd:", "", names(query)),collapse="/"),
               paste0(unname(query),collapse="/")))
})

saveRDS(addKEGG,"../data/intermediate/oryzae_NAMEtoKEGG_keggrest.rds")
```

```{r}
addKEGG <- readRDS("../data/intermediate/oryzae_NAMEtoKEGG_keggrest.rds")

target$KEGG_v3 <- sapply(addKEGG, function(x){x[2]})[namesLink]
target$KEGG_v3[target$KEGG_v3==""] <- NA

target$AllNames <- sapply(addKEGG, function(x){x[3]})[namesLink]
target$AllNames[target$AllNames==""] <- NA
```

```{r}
table(!is.na(target$KEGG))
table(!is.na(target$KEGG)|!is.na(target$KEGG_v2))
table(!is.na(target$KEGG)|!is.na(target$KEGG_v2)|!is.na(target$KEGG_v3))
```

```{r}
rbind(target[target$ID == "2-deoxy-D-ribose 5-phosphate[c]",c(2,12,15,16)],
      target[target$ID == "(2S)-2-isopropyl-3-oxosuccinate[c]",c(2,12,15,16)],
      target[target$ID == "succinate semialdehyde[c]",c(2,12,15,16)])
```

TODO: my acquired KEGGs are correct, these three originals for sure are wrong.
My only mistake is that I wanted to for sure keep v2, but then I overwrite with
v1 for some. But in addition, use v3 for these in stead of v1.


### Consolidate KEGGs

I will assume that Kegg_v2 is always right, as it builds on a link between
CHEBI and KEGG

```{r}
target$KEGG_final <- target$KEGG_v2
target[is.na(target$KEGG_final) &
         ! is.na(target$KEGG),c("KEGG","KEGG_v3")] # all original KEGGs also look good
index <- which(is.na(target$KEGG_final) & !is.na(target$KEGG))
target$KEGG_final[index] <- target$KEGG[index]

# Add kegg_v3 if it is a unique hit
index <- which(is.na(target$KEGG_final) &
                 !is.na(target$KEGG_v3) &
                 lengths(strsplit(target$KEGG_v3,"/")) == 1)
target$KEGG_final[index] <- target$KEGG_v3[index]


hlp <- target[is.na(target$KEGG_final) &
         !is.na(target$KEGG_v3),c("NAME","KEGG_v3")]
# for these 41 remaining it is tricky: we matched multiple keggs to the name.
# It is usually the first hit, but not always. We need to manually inspect this
```

```{r}
# "N-carbamoyl-?-alanine" should be N-Carbamoyl-beta-alanine, which is the same as  	
# 3-Ureidopropionate
# target$NAME[grepl("N-carbamoyl-\\?-alanine", target$NAME)] <- "3-Ureidopropionate"

curated <- list(c("4-hydroxybenzoate", "C00156"),
                c("pimelate", "C02656"),
                c("8-amino-7-oxononanoate", "C01092"),
                c("averantin", "C20499"),
                c("carbamate", "C01563"),
                c("chitobiose", "C01674"),
                c("cinnamate","C00423"), # this is actually trans-cinnamate, nut thats whats used in the GEM
                c("citramalate", "C00815"),
                c("D-aspartate", "C00402"),
                c("dihydrosterigmatocystin", "C20445"),
                c("geranylgeranyl diphosphate", "C00353"),
                c("HX", "C00462"),
                c("indole-3-acetate","C00954"),
                c("itaconate", "C00490"),
                c("L-ribulose", "C00508"),
                c("mesaconate", "C01732"),
                c("maltotriose", "C01835"),
                c("1D-myo-inositol 1,4-bisphosphate", "C01220"),
                c("4alpha-methylzymosterol", "C05103"),
                c("naringenin", "C00509"),
                c("norsolorinic acid","C20452"),
                c("3,4-dihydroxybenzoate", "C00230"),
                c("RX", "C01322"),
                c("tagatose 6-phosphate","C01097"), # this is D-tagatose-6P, for L- its C06312
                c("tagatose", "C00795"), # This is for D-Tagatose, for L- its C21523
                c("thiamin diphosphate", "C00068"),
                c("N-carbamoyl-\\?-alanine", "C02642"),
                c("urocanate", "C00785"),
                c("versiconal", "C20507"),
                c("chitobiose", "C01674"),
                c("itaconate", "C00490"),
                c("L-ribulose", "C00508"),
                c("CMP", "C00055"),
                c("heme a", "C15670"),
                c("nicotinate", "C00253"),
                c("orotate", "C00295"),
                c("D-ribose 5-phosphate", "C00117"),
                c("glycolate", "C00160"),
                c("succinate", "C00042"),
                c("xanthine", "C00385"))
```

```{r}
for (element in curated) {
  target$KEGG_final[grep(element[1],target$NAME)] <- element[2]
}

target$KEGG_v2 <- target$KEGG_v3 <- NULL
sum(is.na(target$KEGG_final)) # So only 40 metabolites without kegg left.

target$KEGG <- target$KEGG_final
target$KEGG_final <- NULL
```

### Map KEGG

```{r}
target$KEGG_ID <- paste0(target$KEGG, "[", target$COMPARTMENT, "]")
target$KEGG_ID[grepl("NA\\[",target$KEGG_ID)] <- NA
```

```{r}
Ani_draft$METS$KEGG_ID <- paste0(Ani_draft$METS$KEGG, "[", Ani_draft$METS$COMPARTMENT, "]")
Ani_draft$METS$KEGG_ID[grepl("NA\\[",Ani_draft$METS$KEGG_ID)] <- NA

# Get KEGG_self_ID columns
Ani_draft$METS$KEGG_self_ID <- NA
Ani_draft$METS$KEGG_self_ID[which(!is.na(Ani_draft$METS$KEGG_self))] <- unlist(lapply(which(!is.na(Ani_draft$METS$KEGG_self)),function(index){
  hlp <- unlist(strsplit(Ani_draft$METS$KEGG_self[index],"/"))
  paste0(hlp,"[",Ani_draft$METS$COMPARTMENT[index],"]",collapse = "/")
}))
```

```{r}
table(target$KEGG %in% Ani_draft$METS$KEGG[!is.na(Ani_draft$METS$KEGG)])
table(target$KEGG_ID %in% Ani_draft$METS$KEGG_ID[!is.na(Ani_draft$METS$KEGG_ID)])

# Does not take properly handle metabolites with multiple KEGGs...
table(target$KEGG %in% Ani_draft$METS$KEGG_self[!is.na(Ani_draft$METS$KEGG_self)])
table(target$KEGG_ID %in% Ani_draft$METS$KEGG_self_ID[!is.na(Ani_draft$METS$KEGG_self_ID)])

# But since this gives no extra hits, we can safely use the above.
table(target$KEGG[!is.na(target$KEGG)] %in% unlist(strsplit(Ani_draft$METS$KEGG_self,"/")))
table(target$KEGG_ID[!is.na(target$KEGG_ID)] %in% unlist(strsplit(Ani_draft$METS$KEGG_self_ID,"/")))
```

```{r}
KEGG_both <- target$KEGG[target$KEGG %in% Ani_draft$METS$KEGG_self[!is.na(Ani_draft$METS$KEGG_self)]]
KEGG_ID_both <- target$KEGG_ID[target$KEGG_ID %in% Ani_draft$METS$KEGG_self_ID[!is.na(Ani_draft$METS$KEGG_self_ID)]]
```

```{r}
KEGG_altcomp <- target$KEGG_ID[target$KEGG %in% KEGG_both & 
                                 !target$KEGG_ID %in% KEGG_ID_both]
KEGG_altcomp
```
Map oryzae metabolites to niger:

```{r}
Aor_reduced$METS$KEGG[match(target$ID, Aor_reduced$METS$ID)] <- target$KEGG
Aor_reduced$METS$KEGG_ID[match(target$ID, Aor_reduced$METS$ID)] <- target$KEGG_ID

table(Aor_reduced$METS$`REPLACEMENT ID` %in% Ani_draft$METS$`REPLACEMENT ID`)
Aor_reduced$METS$`REPLACEMENT ID`[match(KEGG_ID_both, Aor_reduced$METS$KEGG_ID)] <- Ani_draft$METS$`REPLACEMENT ID`[match(KEGG_ID_both, Ani_draft$METS$KEGG_self_ID)]
table(Aor_reduced$METS$`REPLACEMENT ID` %in% Ani_draft$METS$`REPLACEMENT ID`)
```

```{r}
length(Aor_reduced$METS$`REPLACEMENT ID`)
length(unique(Aor_reduced$METS$`REPLACEMENT ID`))
```

```{r}
hlp <- Aor_reduced$METS[Aor_reduced$METS$`REPLACEMENT ID` %in% 
                   Aor_reduced$METS$`REPLACEMENT ID`[duplicated(Aor_reduced$METS$`REPLACEMENT ID`)],c(2,3,5,9)]
hlp
```

Possible issue: KEGGs C00117 and C00042

```{r}
Aor_reduced$METS[Aor_reduced$METS$KEGG %in% c("C00117","C00042"),c(2,9,11,12)]
Aor_before$METS[Aor_before$METS$ID %in% c("2-deoxy-D-ribose 5-phosphate[c]",
                                          "(2S)-2-isopropyl-3-oxosuccinate[c]",
                                          "succinate semialdehyde[c]",
                                          "D-ribose 5-phosphate[m]",
                                          "succinate[p]"),c(2,9)]
Ani_draft$METS[Ani_draft$METS$KEGG %in% c("C00117","C00042"),c(2,9,12,11)]
```

Indeed a mistake in our automatic KEGG annotation for:

```{r}
# 2-Deoxy-D-ribose 5-phosphate should get C00673
# succinate semialdehyde should get C00232 
# (2S)-2-isopropyl-3-oxosuccinate should get C04236


```

An important quest is: where do these mistakes come from? 
Which KEGG acquirement strategy introduced these mistakes?

---> good news, not a mistake of mine ;) all three are mistakes in the original
annotation!


Continue with the unlinked metabolites:

```{r}
target <- target[!target$KEGG_ID %in% KEGG_ID_both,]
```

# Altcomp

```{r}
altcomp <- unique(c(which(target$`REPLACEMENT ID` %in% replacement_altcomp),
                    which(target$ID %in% ID_altcomp),
                    which(target$CHEBI_ID %in% CHEBI_altcomp),
                    which(target$KEGG_ID %in% KEGG_altcomp)))
length(altcomp) # I expect up front that these metabolites do not exist in the
# same compartment in Aor as they do in Ani. The remaining 217 could potentially 
# be saved, or just be metabolites unique to oryzae compared to niger
```

```{r}
index_Aor <- which(Aor_reduced$METS$KEGG_ID %in% KEGG_altcomp)

# link
Aor_reduced$METS$`REPLACEMENT ID`[index_Aor] <- Ani_draft$METS$`REPLACEMENT ID`[match(gsub("\\[.*","",KEGG_altcomp),Ani_draft$METS$KEGG_self)]

# fix compartments
Aor_reduced$METS$`REPLACEMENT ID`[index_Aor] <- gsub("\\[.\\]", "",
     Aor_reduced$METS$`REPLACEMENT ID`[index_Aor])

Aor_reduced$METS$`REPLACEMENT ID`[index_Aor] <- paste0(Aor_reduced$METS$`REPLACEMENT ID`[index_Aor],
                                                  "[",
                                                  Aor_reduced$METS$COMPARTMENT[index_Aor],
                                                  "]")
```

```{r}
altcomp_todo <- setdiff(altcomp,which(target$KEGG_ID %in% KEGG_altcomp))
all(altcomp_todo %in% which(target$`REPLACEMENT ID` %in% replacement_altcomp))
all(altcomp_todo %in% which(target$ID %in% ID_altcomp))
all(altcomp_todo %in% which(target$CHEBI_ID %in% CHEBI_altcomp))

# all remaining altcomps can be fixed through replacement ID

# these are the replacement IDs in Aor
target$`REPLACEMENT ID`[setdiff(altcomp,which(target$KEGG_ID %in% KEGG_altcomp))]

# these are the replacement IDs in Ani
index_Aor <- which(Aor_reduced$METS$`REPLACEMENT ID` %in% target$`REPLACEMENT ID`[setdiff(altcomp,which(target$KEGG_ID %in% KEGG_altcomp))])
hlp <- Aor_reduced$METS$`REPLACEMENT ID`[index_Aor]
Ani_draft$METS$`REPLACEMENT ID`[match(gsub("\\[.*","",hlp),Ani_draft$METS$`REPLACEMENT NAME`)]

# which essentially means we dont need to do anything. The metabolites are
# properly matched through replacement ID, and are just in different 
# compartments
```

```
Aor_reduced_consolidated <- Aor_reduced
write.xlsx(Aor_reduced_consolidated, "../model/Aor_reduced_consolidated.xlsx",sep.names = " ")
Aor_reduced_consolidated <- read_excel_allsheets("../model/Aor_reduced_consolidated.xlsx")
```

```{r}
Aor_reduced_consolidated <- Aor_reduced
write.xlsx(Aor_reduced_consolidated$METS[,c("NAME", "COMPARTMENT","REPLACEMENT ID")], "../data/intermediate/Aor_reduced_metabolite_table.xlsx")
```


# Final rescue

```{r}
table(tolower(Aor_reduced_consolidated$METS$`REPLACEMENT ID`) %in% tolower(Ani_draft$METS$`REPLACEMENT ID`))
# of which 65 for sure cannot be fixed
```

```{r}
target <- target[-altcomp,]
```

# Final rescue - devel

We have a lot of info on these, but still no match with Ani (where we also have
quite a lot of info). So, these really might be unique metabolites to Aor 
(or at least our Aor draft). But how can we be sure. Because most of these 
metabolites dont look that exotic.

```{r}
table(!is.na(target$KEGG))
table(!is.na(target$CHEBI))
table(!is.na(target$CHEBI)|!is.na(target$KEGG))
```

## curation

TODO: do this for Aor_reduced

```{r}
# manually add/curate some info
target[grep("trans-4-coumarate", target$NAME),c("ID", "NAME", "KEGG","KEGG_ID")] <- c(
  "4-coumarate[c]","4-coumarate","C00811","C00811[c]")

target[grep("1,8-DHN", target$NAME),c("ID", "NAME", "CHEBI","CHEBI_ID")] <- c(
  "naphthalene-1,8-diol[c]","naphthalene-1,8-diol","CHEBI:167604","CHEBI:167604[c]")

target[which(target$ID == "N-carbamoyl-?-alanine[c]"),"ID"] <- "3-Ureidopropionate[c]"

# could change 2',4,4',6'-tetrahydroxychalcone[c] to Naringenin chalcone[c]

target[which(target$ID == "2,5-diamino-6-hydroxy-4-(5'-phosphoribosylamino)-pyrimidine[c]"),c("ID","NAME","KEGG","KEGG_ID")] <-
  c("2,5-diamino-6-(1-D-ribosylamino)pyrimidin-4(3H)-one 5'-phosphate[c]",
    "2,5-diamino-6-(1-D-ribosylamino)pyrimidin-4(3H)-one 5'-phosphate",
    "C01304","C01304[c]")

# Remove from Aor because it makes consolidation of Aor and Ani diifficult and
# the same reaction is already present (but with different intermediate).
5-amino-6-(5-phosphoribosylamino)uracil[c]

```

## "Surely" unique for oryzae

We never know for sure, but at least we can be pretty conficent here:

```{r}
Aor_unique <- c("aflatoxin B1[c]","aflatoxin B2[c]","aflatoxin G1[c]",
                "aflatoxin G2[c]", "intermediate methylzymosterol II[c]",
                "intermediate zymosterol II[c]", 
                "intermediate methylzymosterol I[c]", 
                "intermediate zymosterol I[c]", "4alpha-methylzymosterol[c]",
                "heme a[m]", "heme o[m]", "apocytochrome C[m]",
                "cytochrome c[m]","isopenicillin N[c]","penicillins[c]",
                "penicillanic acid[c]","penicillin G[c]",
                "L-selenocysteinyl-tRNA(Sec)[c]", "selenophosphate[c]",
                "L-cysteinylglycine[c]",
                "N-[L-5-amino-5-carboxypentanoyl]-L-cysteinyl-D-valine[c]",
                "tagatose 6-phosphate[c]",
                "D-tagatofuranose 1,6-bisphosphate[c]","tagatose[c]",
                "tagatose[e]","averufin[c]","averantin[c]",
                "5-hydroxyaverantin[c]","versiconal hemiacetal acetate[c]",
                "versicolorin B[c]","versiconal[c]","versicolorin A[c]",
                "dihydrodemethylsterigmatocystin[c]",
                "dihydro-O-methylsterigmatocystin[c]",
                "dihydrosterigmatocystin[c]","6-demethylsterigmatocystin[c]",
                "7-O-methylsterigmatocystin[c]","sterigmatocystin[c]",
                "vermelone[c]","scytalone[c]","naphthalene-1,3,8-triol[c]",
                "1,8-DHN[c]","melanins[c]", "sarcosine[c]", "sarcosine[m]", 
                "N,N-dimethylglycine[c]", "N,N-dimethylglycine[m]",
                "betaine aldehyde[c]", "glycine betaine[c]", "chitobiose[c]",
                "chitobiose[e]","norsolorinic acid[c]","FMN[c]","FMN[e]",
                "FMN[m]","vestine[c]","ferreirin[c]","eriodictyol[c]",
                "2',4,4',6'-tetrahydroxychalcone[c]","4-coumaroyl-CoA[c]",
                "tryptamine[c]","3-Ureidopropionate[c]","5,6-dihydrouracil[c]")

target <- target[!target$ID %in% Aor_unique,]
View(target[!duplicated(target$NAME),c("ID","CHEBI","KEGG","AllNames"),])
```

Riboflavin metabolism

KEGG ====================
 	
*3.5.4.25)* GTP + 4 H2O <=> Formate + 2,5-Diamino-6-(5-phospho-D-ribosylamino)pyrimidin-4(3H)-one + 2 Orthophosphate

Option 1:
*3.5.4.26)*  2,5-Diamino-6-(5-phospho-D-ribosylamino)pyrimidin-4(3H)-one + H2O <=> 5-Amino-6-(5'-phosphoribosylamino)uracil + Ammonia
*1.1.1.193)* 5-Amino-6-(5'-phosphoribosylamino)uracil + NADPH + H+ <=> 5-Amino-6-(5'-phospho-D-ribitylamino)uracil + NADP+

Option 2:
*1.1.1.302)* 2,5-Diamino-6-(5-phospho-D-ribosylamino)pyrimidin-4(3H)-one + NADPH + H+ <=> 2,5-Diamino-6-(5-phospho-D-ribitylamino)pyrimidin-4(3H)-one + NADP+
*5.4.99.28)* 2,5-Diamino-6-(5-phospho-D-ribitylamino)pyrimidin-4(3H)-one + H2O <=> 5-Amino-6-(5'-phospho-D-ribitylamino)uracil + Ammonia

*3.1.3.104)* 5-Amino-6-(5'-phospho-D-ribitylamino)uracil + H2O <=> 5-Amino-6-(1-D-ribitylamino)uracil + Orthophosphate
*2.5.1.78)*  5-Amino-6-(1-D-ribitylamino)uracil + L-3,4-Dihydroxybutan-2-one 4-phosphate <=> 6,7-Dimethyl-8-(D-ribityl)lumazine + 2 H2O + Orthophosphate

Aor =====================

*3.5.4.25)* GTP[c] => 2,5-diamino-6-hydroxy-4-(5'-phosphoribosylamino)-pyrimidine[c] + formate[c] + diphosphate[c]

Option 1:
*3.5.4.26)* 2,5-diamino-6-hydroxy-4-(5'-phosphoribosylamino)-pyrimidine[c] => *5-amino-6-(5-phosphoribosylamino)uracil[c]* + NH3[c]
*1.1.1.193)* 5-amino-6-(5-phosphoribosylamino)uracil[c] + NADPH[c] => 5-amino-6-(5'-phosphoribitylamino)uracil[c] + NADP(+)[c]

*3.1.3.-)* 5-amino-6-(5'-phosphoribitylamino)uracil[c] + H2O[c] => 5-amino-6-ribityluracil[c] + phosphate[c]	
*empty)* 5-amino-6-ribityluracil[c] + 2-hydroxy-3-oxobutyl phosphate[c] => 6,7-dimethyl-8-(1-D-ribityl)lumazine[c] + 2 H2O[c] + phosphate[c]

Ani =====================

*3.5.4.25)* GTP[c] + 3 Water[c] => Formate[c] + Pyrophosphate[c] + 2,5-diamino-6-(1-D-ribosylamino)pyrimidin-4(3H)-one 5'-phosphate[c]	

Option 2:
*1.1.1.302)* NADPH[c] + 2,5-diamino-6-(1-D-ribosylamino)pyrimidin-4(3H)-one 5'-phosphate[c] => NADP+[c] + 2,5-diamino-6-(5-phosphono)ribitylamino-4(3H)-pyrimidinone[c]	
*3.5.4.26)* Water[c] + 2,5-diamino-6-(5-phosphono)ribitylamino-4(3H)-pyrimidinone[c] => 2 Intracellular protons[c] + Ammonia[c] + *5-amino-6-(5-phospho-β-D-ribosylamino)uracil[c]*

*3.1.3.104)* 2 Intracellular protons[c] + Water[c] + 5-amino-6-(5-phospho-β-D-ribosylamino)uracil[c] => Orthophosphate[c] + 5-amino-6-(D-ribitylamino)uracil[c]
*2.5.1.78)* 2-hydroxy-3-oxobutyl phosphate[c] + 5-amino-6-(D-ribitylamino)uracil[c] => 2 Water[c] + Orthophosphate[c] + 6,7-dimethyl-8-(1-D-ribityl)lumazine[c]

So:
Strong suspicion that
1. In Aor, enzyme 3.1.3.- should be 3.1.3.104
2. In Ani, enzyme 3.5.4.26 should be 5.4.99.28
3. 5-amino-6-(5-phosphoribosylamino)uracil[c] and 5-amino-6-(5-phospho-β-D-ribosylamino)uracil[c] both have CHEBI:18337. However, the former is created by removing NH3 from the starting compound, while the later is created by reduction and removing NH3. In KEGG, they are two different

This leads me to:
1) Replace EC:3.5.4.26 with EC:5.4.99.28 in Ani_Draft
2) Remove metabolite 5-amino-6-(5-phosphoribosylamino)uracil[c] in Aor_draft. TODO: check if this automatically removes the reaction in RAVEN.
3) Consolidate: change 5-amino-6-(5'-phosphoribitylamino)uracil in Aor to 5-amino-6-(5-phospho-β-D-ribosylamino)uracil from Ani. Remove the CHEBI in both, and give both KEGG C00454.
4) Consolidate: change 5-amino-6-ribityluracil[c] in Aor to 5-amino-6-(D-ribitylamino)uracil[c], give both KEGG C04732
5) 



5-amino-6-(D-ribitylamino)uracil[c] == 5-amino-6-ribityluracil[c] == C04732
2-hydroxy-3-oxobutyl phosphate[c] == 2-hydroxy-3-oxobutyl phosphate[c] == C15556
6,7-dimethyl-8-(1-D-ribityl)lumazine[c] = 6,7-dimethyl-8-(1-D-ribityl)lumazine[c] == C04332

```{r}
rbind(target[which(target$ID == "6,7-dimethyl-8-(1-D-ribityl)lumazine[c]"),c("ID","CHEBI","KEGG")])
```

```{r}
which(Aor_reduced$METS$ID == "6,7-dimethyl-8-(1-D-ribityl)lumazine[c]")
Aor_draft$METS[grepl("oxobutyl", Aor_draft$METS$ID),]
```

```{r}
rbind(Aor_reduced$METS[which(Aor_reduced$METS$ID == "2-hydroxy-3-oxobutyl phosphate[c]"),c("ID","CHEBI","KEGG")])
```

```{r}
rbind(Ani_draft$METS[which(Ani_draft$METS$NAME == "2-hydroxy-3-oxobutyl phosphate"),c("ID","CHEBI","KEGG")])
```

```{r}
Ani_draft$METS[grepl("2,5-diamino-6-\\(5-phosphono\\)ribitylamino-4\\(3H\\)-pyrimidinone", Ani_draft$METS$NAME,ignore.case = TRUE),]
Ani_draft$METS[which(Ani_draft$METS$NAME == "5-amino-6-(5-phospho-β-D-ribosylamino)uracil"),]
```

# Issues

## zymosterol

For now I assumed the mechnaism are different, but they could just be synonyms 
or resolution differences. Asked Ed for help.

Ani ====================
Lanosterol[c] + 3 NADPH[c] + 3 Oxygen[c] => 4,4-Dimethyl-5alpha-cholesta-8,14,24-trien-3beta-ol[c] + Formate[c] + 4 Water[c] + 3 NADP+[c]
4,4-Dimethyl-5alpha-cholesta-8,14,24-trien-3beta-ol[c] + NADPH[c] => 14-Demethyllanosterol[c] + NADP+[c]
14-demethyllanosterol[c] + 3 NADPH[c] + 3 Oxygen[c] => 4-methylcholesta-8,24-diene-3-ol-4-carboxylate[c] + Intracellular protons[c] + 4 Water[c] + 3 NADP+[c]
4-methylcholesta-8,24-diene-3-ol-4-carboxylate[c] + Intracellular protons[c] => 4-methylcholesta-8,24-diene-3-ol[c] + Carbon dioxide[c]
4-methylcholesta-8,24-diene-3-ol[c] + 3 NADPH[c] + 3 Oxygen[c] => 3-dehydro-4-carboxyzymosterol[c] + 2 Intracellular protons[c] + 4 Water[c] + 3 NADP+[c]
3-dehydro-4-carboxyzymosterol[c] + 2 Intracellular protons[c] => Carbon dioxide[c] + Zymosterol[c]

Aor ====================
lanosterol[c] + oxygen[c] + FADH2[m] => 4,4-dimethylcholesta-8,14,24-trienol[c] + FAD[m]
4,4-dimethylcholesta-8,14,24-trienol[c] + NADPH[c] => 14-demethyllanosterol[c] + NADP(+)[c]
14-demethyllanosterol[c] + 3 oxygen[c] => intermediate methylzymosterol I[c]
intermediate methylzymosterol I[c] => CO2[c] + intermediate methylzymosterol II[c]
intermediate methylzymosterol II[c] + NADPH[c] => 4alpha-methylzymosterol[c] + NADP(+)[c]
4alpha-methylzymosterol[c] + 3 oxygen[c] => intermediate zymosterol I[c]
intermediate zymosterol I[c] => CO2[c] + intermediate zymosterol II[c]
intermediate zymosterol II[c] + NADPH[c] => zymosterol[c] + NADP(+)[c]

## cytochrome

Aor doesnt care about protons. But then these two reactions which should be
identical will be separate reactions in the drafts...

Aor ==============
NADPH[c] + 2 ferricytochrome c[m] => NADP(+)[c] + 2 ferrocytochrome c[m]

2 ferricytochrome c[m] + 4 H(+)[m] + ubiquinol[m] => 4 H(+) (energy metabolism)[c] + 2 ferrocytochrome c[m] + ubiquinone[m]

2 ferrocytochrome c[m] + 4 H(+)[m] + 0.5 oxygen[m] => 4 H(+) (energy metabolism)[c] + 2 ferricytochrome c[m] + H2O[m]


Ani =================
2 Ferricytochrome c[m] + NADPH[c] => 2 Ferrocytochrome c[m] + 2 Intracellular protons[c] + NADP+[c]

2 Ferricytochrome c[m] + 4 Proton (energy metabolism)[m] + Ubiquinol[m] => 2 Ferrocytochrome c[m] + 4 Proton (energy metabolism)[c] + Ubiquinone[m] + 2 Proton[m]

2 Ferrocytochrome c[m] + 4 Proton (energy metabolism)[m] + 0.5 Oxygen[m] + 2 Proton[m] => 2 Ferricytochrome c[m] + 4 Proton (energy metabolism)[c] + Water[m]

## Penicillin

N-[L-5-amino-5-carboxypentanoyl]-L-cysteinyl-D-valine[c] + oxygen[c] => 2 H2O[c] + isopenicillin N[c]
isopenicillin N[c] + phenylacetyl-CoA[c] => L-2-aminoadipate[c] + coenzyme A[c] + penicillins[c]
isopenicillin N[c] => penicillin N[c]
penicillins[c] => penicillanic acid[c]
penicillins[c] => penicillanic acid[c]
penicillins[c] => penicillanic acid[c]
penicillins[c] => penicillanic acid[c]
penicillins[c] => penicillanic acid[c]
penicillin G[c] => 

So essentially, 
1. penicillin G can be extrected but is never used/created anywhere.
2. penicillin N is a dead-end metabolite
3. penicillanic acid is a dead-end metabolite

## Chitosan

Chitoson is a huge polymer....

Ani ========
Chitosan[c] + Water[c] => D-Glucosamine[c]

Aor ========
chitosan[c] => D-glucosamine[c]

In kegg, R08715:
Chitosan(n+1) + H2O <=> Chitosan(n) + D-Glucosamine

## Pectin

Ani =========
Endo-Polygalacturonase	Water[e] + Polygalacturonate[e] => D-Galacturonate[e]	                          NA
Pectin methyl esterase	Water[e] + Methylated galacturonate[e] => Methanol[e] + Polygalacturonate[e]	  3.1.1.11


Aor ==========
Exo-polygalacturonase	    H2O[e] + pectate[e] => D-galacturonate[e]	    3.2.1.67
Endo-polygalacturonase	  pectate[e] => D-galacturonate[e]	            3.2.1.15
Endo-polygalacturonase	  pectate[e] => D-galacturonate[e]	            3.2.1.15
Endo-polygalacturonase	  pectate[e] => D-galacturonate[e]	            3.2.1.15

So, first off, I think these reactions should always involve H2O.
Second, it is strange that it is using "Methylated galacturonate" in stead of
"Methylated polygalacturonate".
Third, could we say pectate == Polygalacturonate?
Fourth, in KEGG, these reactions (e.g.) are always specified as
 	
Pectate + H2O <=> Pectate + D-Galacturonate

i.e., with the polymer on both sides. I suppose you cant do that in a GEM, 
because then the supply of pectate would be eternal.

## Riboflavin metabolism

https://www.kegg.jp/pathway/map00740+C04454

KEGG ====================
 	
*3.5.4.25)* GTP + 4 H2O <=> Formate + 2,5-Diamino-6-(5-phospho-D-ribosylamino)pyrimidin-4(3H)-one + 2 Orthophosphate

*3.5.4.26)* 2,5-Diamino-6-(5-phospho-D-ribosylamino)pyrimidin-4(3H)-one + H2O <=> 5-Amino-6-(5'-phosphoribosylamino)uracil + Ammonia
*1.1.1.193)* 5-Amino-6-(5'-phosphoribosylamino)uracil + NADPH + H+ <=> 5-Amino-6-(5'-phospho-D-ribitylamino)uracil + NADP+

*1.1.1.302)* 2,5-Diamino-6-(5-phospho-D-ribosylamino)pyrimidin-4(3H)-one + NADPH + H+ <=> 2,5-Diamino-6-(5-phospho-D-ribitylamino)pyrimidin-4(3H)-one + NADP+
*5.4.99.28)* 2,5-Diamino-6-(5-phospho-D-ribitylamino)pyrimidin-4(3H)-one + H2O <=> 5-Amino-6-(5'-phospho-D-ribitylamino)uracil + Ammonia

*3.1.3.104)* 5-Amino-6-(5'-phospho-D-ribitylamino)uracil + H2O <=> 5-Amino-6-(1-D-ribitylamino)uracil + Orthophosphate
*2.5.1.78)*  5-Amino-6-(1-D-ribitylamino)uracil + L-3,4-Dihydroxybutan-2-one 4-phosphate <=> 6,7-Dimethyl-8-(D-ribityl)lumazine + 2 H2O + Orthophosphate

Aor =====================

*3.5.4.25)* GTP[c] => 2,5-diamino-6-hydroxy-4-(5'-phosphoribosylamino)-pyrimidine[c] + formate[c] + diphosphate[c]
*3.5.4.26)* 2,5-diamino-6-hydroxy-4-(5'-phosphoribosylamino)-pyrimidine[c] => 5-amino-6-(5-phosphoribosylamino)uracil[c] + NH3[c]
*1.1.1.193)* 5-amino-6-(5-phosphoribosylamino)uracil[c] + NADPH[c] => 5-amino-6-(5'-phosphoribitylamino)uracil[c] + NADP(+)[c]
*3.1.3.-)* 5-amino-6-(5'-phosphoribitylamino)uracil[c] + H2O[c] => 5-amino-6-ribityluracil[c] + phosphate[c]	
*empty)* 5-amino-6-ribityluracil[c] + 2-hydroxy-3-oxobutyl phosphate[c] => 6,7-dimethyl-8-(1-D-ribityl)lumazine[c] + 2 H2O[c] + phosphate[c]

Ani =====================

*3.5.4.25)* GTP[c] + 3 Water[c] => Formate[c] + Pyrophosphate[c] + 2,5-diamino-6-(1-D-ribosylamino)pyrimidin-4(3H)-one 5'-phosphate[c]	
*1.1.1.302)* NADPH[c] + 2,5-diamino-6-(1-D-ribosylamino)pyrimidin-4(3H)-one 5'-phosphate[c] => NADP+[c] + 2,5-diamino-6-(5-phosphono)ribitylamino-4(3H)-pyrimidinone[c]	
*3.5.4.26)* Water[c] + 2,5-diamino-6-(5-phosphono)ribitylamino-4(3H)-pyrimidinone[c] => 2 Intracellular protons[c] + Ammonia[c] + 5-amino-6-(5-phospho-β-D-ribosylamino)uracil[c]
*3.1.3.104)* 2 Intracellular protons[c] + Water[c] + 5-amino-6-(5-phospho-β-D-ribosylamino)uracil[c] => Orthophosphate[c] + 5-amino-6-(D-ribitylamino)uracil[c]
*2.5.1.78)* 2-hydroxy-3-oxobutyl phosphate[c] + 5-amino-6-(D-ribitylamino)uracil[c] => 2 Water[c] + Orthophosphate[c] + 6,7-dimethyl-8-(1-D-ribityl)lumazine[c]

So:
Strong suspicion that
1. In Aor, enzyme 3.1.3.- should be 3.1.3.104
2. In Ani, enzyme 3.5.4.26 should be 5.4.99.28
