---
title: "Consolidate metabolites for A. niger and A. fumigatus"
author: "Jeroen Gilis"
date: "6 January 2025"
output: 
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

# Libraries

```{r, message=FALSE, warning=FALSE}
library(stringr)
library(openxlsx)
library(RCurl)
library(KEGGREST)
```

# Helpers

```{r}
source("./Rutilities.R")
```

# Load GEMs

```{r}
# Note: excels should be closed when reading them in R
Aor_draft <- read_excel_allsheets("../data/intermediate/Aor_draft_comp.xlsx")
Aor_kegg <- read.xlsx("../data/intermediate/oryzae_kegg.xlsx")
Ani_draft <- read_excel_allsheets("../data/intermediate/Ani_draft_comp_mets.xlsx")
Afu_draft <- read_excel_allsheets("../data/intermediate/Afu_draft_comp.xlsx")
Ani_reduced <- read_excel_allsheets("../scrap/Ani_reduced.xlsx")
Afu_reduced <- read_excel_allsheets("../scrap/Afu_reduced.xlsx")
OtherAsp <- read_excel_allsheets("../scrap/OtherAsp_10_50_loc.xlsx");
```

# Wrangle input

## Oryzae

```{r}
# Get CHEBI column
Aor_draft$METS$CHEBI <- Aor_draft$METS$MIRIAM
Aor_draft$METS$CHEBI[!grepl("chebi",Aor_draft$METS$CHEBI)] <- NA
Aor_draft$METS$CHEBI <- gsub("chebi:","",Aor_draft$METS$CHEBI)
Aor_draft$METS$CHEBI <- gsub("/",":",Aor_draft$METS$CHEBI)

# Get KEGG column
Aor_draft$METS$KEGG <- Aor_kegg$KEGG
```

## Niger

```{r}
# Get KEGG column
Ani_reduced$METS$KEGG <- Ani_reduced$METS$MIRIAM
Ani_reduced$METS$KEGG[!grepl("kegg",Ani_reduced$METS$KEGG)] <- NA
Ani_reduced$METS$KEGG <- gsub("\\;.*","",gsub(".*kegg.compound/(.+)", "\\1", Ani_reduced$METS$KEGG))

# Get chebi column
Ani_reduced$METS$CHEBI <- Ani_reduced$METS$MIRIAM
Ani_reduced$METS$CHEBI[!grepl("chebi",Ani_reduced$METS$CHEBI)] <- NA
Ani_reduced$METS$CHEBI <- gsub("\\;.*","",gsub(".*chebi/(.+)", "\\1", Ani_reduced$METS$CHEBI))
```

## Fumigatus

```{r}
# Get KEGG column
Afu_reduced$METS$KEGG <- Afu_reduced$METS$MIRIAM
Afu_reduced$METS$KEGG[!grepl("kegg",Afu_reduced$METS$KEGG)] <- NA
Afu_reduced$METS$KEGG <- gsub("\\;.*","",gsub(".*kegg.compound/(.+)", "\\1", Afu_reduced$METS$KEGG))

# Get chebi column
Afu_reduced$METS$CHEBI <- Afu_reduced$METS$MIRIAM
Afu_reduced$METS$CHEBI[!grepl("chebi",Afu_reduced$METS$CHEBI)] <- NA
Afu_reduced$METS$CHEBI <- gsub("\\;.*","",gsub(".*chebi/(.+)", "\\1", Afu_reduced$METS$CHEBI))
```

# Niger

## KEGG available

For 170/252 compounds we already have a KEGG ID available.

```{r}
table(!is.na(Ani_reduced$METS$KEGG))
table(!is.na(Ani_reduced$METS$CHEBI))
```

```{r}
target <- Ani_reduced$METS[is.na(Ani_reduced$METS$KEGG),]
```

## Obtain KEGG directly thorugh KEGG-CHEBI dictionary

```{r}
kegg_chebi_dict <- read.csv2("../data/intermediate/kegg_chebi_dict.csv", header=FALSE, sep=",")

# wrangle
colnames(kegg_chebi_dict) <- c("KEGG","CHEBI")
kegg_chebi_dict$KEGG <- gsub("cpd:","",kegg_chebi_dict$KEGG)
kegg_chebi_dict$CHEBI <- gsub("chebi:","CHEBI:",kegg_chebi_dict$CHEBI)

target$KEGG <- kegg_chebi_dict$KEGG[match(target$CHEBI,kegg_chebi_dict$CHEBI)]

Ani_reduced$METS$KEGG[match(target$`REPLACEMENT ID`,Ani_reduced$METS$`REPLACEMENT ID`)] <- target$KEGG
target <- Ani_reduced$METS[is.na(Ani_reduced$METS$KEGG),]
```

## Based on KEGGREST

At this point, almost all compounds that are left are fatty acids bound to CoA
or ACP with a oxo- or hydroxy- group.

```{r}
target
```

```{r}
name_query <- tolower(target$NAME)
name_query <- gsub(",", " ", name_query) # it works to replace commas for spaces
```

```{r}
addKEGG <- lapply(name_query, function(x){
  #print(x)
  query <- keggFind(database = "compound", query = x)
  return(cbind(x,
               paste0(gsub("cpd:", "", names(query)),collapse="/"),
               paste0(unname(query),collapse="/")))
})

target$KEGG <- sapply(addKEGG, function(x){x[2]})
target$KEGG[target$KEGG==""] <- NA
```

```{r}
# metabolites that we could assign a unique KEGG id based on its name
# -> use directly
index <- which(!is.na(target$KEGG) &
                 lengths(strsplit(target$KEGG,"/")) == 1)
target$KEGG_final <- NA
target$KEGG_final[index] <- target$KEGG[index]

# metabolites that we could assign a KEGG id, but not unambiguously
# -> try to identify the correct ID if possible
target$KEGG_final[target$NAME == "Palmitoyl-[acyl-carrier protein]"] <- "C05764"
target$KEGG_final[target$NAME == "Ubiquinol"] <- "C00390"
```

```{r}
target[is.na(target$KEGG_final),]
# The first 43/55 compounds are fatty acids with an additional group.
# These are not in KEGG (not for uneven numbers and >20C anyway).
# We can only dd one of these:
target$KEGG_final[target$NAME == "3-oxooctadecanoyl-[acyl-carrier protein]"] <- "C16219"

# The rest of the compounds are also not described in KEGG.
```

Add the compounds we found.

```{r}
Ani_reduced$METS$KEGG[match(target$`REPLACEMENT ID`,Ani_reduced$METS$`REPLACEMENT ID`)] <- target$KEGG_final
target_Ani <- Ani_reduced$METS[is.na(Ani_reduced$METS$KEGG),]
```

## Compare to oryzae GEM IDs

For the remaining 54 compounds, we will need to check if we already have them in
the oryzae GEM. If so, we need to take the same identifier as in the oryzae GEM,
even if this is not a KEGG ID.

The fatty acid derivatives are described in the replacement ID using the same
nomenclature/structure in both GEMs. As such, we can be quite certain that the
remaining metabolites in this category are unique to niger compred to oryzae.
Except for hexadecadienoyl-[acp] and octadecatrienoyl-[acp], but these already
have the same ID, so nothing needs to be done there either.

For the remaining few compounds, it looks like we do not have them in 
`Ani_reduced`. As such, we keep the original identifiers for now.

```{r}
target_Ani[gsub("\\[.\\]","",target_Ani$`REPLACEMENT ID`) %in% 
         gsub("\\[.\\]","",Aor_draft$METS$`REPLACEMENT ID`),]
```

# Fumigatus

```{r}
table(!is.na(Afu_reduced$METS$KEGG))
table(!is.na(Afu_reduced$METS$CHEBI))
```

## KEGG available

For 258/289 compounds we already have a KEGG ID available.

```{r}
table(!is.na(Afu_reduced$METS$KEGG))
table(!is.na(Afu_reduced$METS$CHEBI))
```

```{r}
target <- Afu_reduced$METS[is.na(Afu_reduced$METS$KEGG),]
```

## Manually inspect the few remaining metabolites

CHEBI available, but note that we have many CHEBI:52320, which is 
"ergosteryl ester"; a generic term. Interestingly, the REPLACEMENT.ID seems a
more accurate CHEBI of these compounds. It seems that we can safely replace the 
CHEBI column with the REPLACEMENT.ID column for these instances:

```{r}
hlp <- target$`REPLACEMENT ID`[which(target$CHEBI == "CHEBI:52320")]
hlp <- gsub("CHEBI","CHEBI:",hlp)
hlp <- gsub("\\[.\\]","",hlp)
target$CHEBI[which(target$CHEBI == "CHEBI:52320")] <- hlp
```

```{r}
target[!duplicated(target$NAME),] # only 16 compounds
```

```{r}
target$KEGG[target$NAME == "(R)-2,3-Dihydroxy-3-methylpentanoate"] <- "C06007"
target$KEGG[target$NAME == "chitin"] <- "C00461"
target$KEGG[target$NAME == "(R)-3-methyl-2-oxobutanoate"] <- "C00141"
target$KEGG[target$NAME == "3-hydroxymyristoyl-[acp]"] <- "C04688"
```

```{r}
Afu_reduced$METS$KEGG[match(target$`REPLACEMENT ID`,Afu_reduced$METS$`REPLACEMENT ID`)] <- target$KEGG
```

```{r}
target[is.na(target$KEGG) & 
         !duplicated(target$NAME),]
```

# Write output

```{r}
write.xlsx(Ani_reduced$METS[,c("NAME","KEGG")], "../data/intermediate/niger_kegg.xlsx")
write.xlsx(Afu_reduced$METS[,c("NAME","KEGG")], "../data/intermediate/fumigatus_kegg.xlsx")
```

# Appendix: Anticipate downstream issues

There are a few compounds that are probably truly unique to the fumigatus GEM,
however, the reactions they are used in are not. Exmples:

## B00001 and B00002

The compounds B00001 and B00002 are interesting. They seem to encode something
like "a polymer that is one building block longer/shorter than before".

Unfortunately, this is encoded differently in the oryzae GEM. Hence, we need
to manually curate these reactions.

```{r}
Afu_reduced$RXNS[grep("B0000", Afu_reduced$RXNS$EQUATION),]
```

```{r}
Afu_draft$RXNS[grep("B0000", Afu_draft$RXNS$EQUATION),]
```

E.g. 3 H2O[cytoplasm] + alpha-L-arabinan[cytoplasm] <=> 2 B00001[cytoplasm] + 3 L-arabinose[cytoplasm]
is reaction with EC 3.2.1.55.
It probably wants to encode the same reaction as:

```{r}
Aor_draft$RXNS[which(Aor_draft$RXNS$`EC-NUMBER` == "3.2.1.55"),]
```

## PO ration

Another tricky case is the "compound" PO ratio, which stands for phosphor/oxygen
ratio.

```{r}
Afu_reduced$RXNS[grep("PO ratio", Afu_reduced$RXNS$EQUATION),]
```

```{r}
Afu_draft$RXNS[grep("PO ratio", Afu_draft$RXNS$EQUATION),]
```

```{r}
Aor_draft$RXNS[grep("1.9.3.1", Aor_draft$RXNS$`EC-NUMBER`),]
```













