---
title: "Link OtherAsp to panAsp_v1 metabolites"
author: "Jeroen Gilis"
date: "21 November 2024"
output: 
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

# Libraries

```{r, message=FALSE, warning=FALSE}
library(openxlsx)
```

# Helpers

```{r}
source("../code/Rutilities.R")
```

# Load data

```{r}
Asp_kegg <- read_excel_allsheets("../scrap/Asp_kegg_loc.xlsx")
panAsp_v1 <- read_excel_allsheets("../model/panAsp_v1.xlsx")
Ani_idtokegg <- read.xlsx("../data/intermediate/niger_metabolite_table.xlsx")
```

# Wrangle data

## panAsp_v1

```{r}
# Get RepName column
panAsp_v1$METS$RepName <- gsub("\\[.\\]","",panAsp_v1$METS$`REPLACEMENT ID`)

# Get KEGG column
panAsp_v1$METS$KEGG <- panAsp_v1$METS$MIRIAM
panAsp_v1$METS$KEGG[!grepl("kegg",panAsp_v1$METS$KEGG)] <- NA
panAsp_v1$METS$KEGG <- gsub("\\;.*","",gsub(".*kegg.compound/(.+)", "\\1", panAsp_v1$METS$KEGG))

# Get chebi column
panAsp_v1$METS$CHEBI <- panAsp_v1$METS$MIRIAM
panAsp_v1$METS$CHEBI[!grepl("chebi",panAsp_v1$METS$CHEBI)] <- NA
panAsp_v1$METS$CHEBI <- gsub("\\;.*","",gsub(".*chebi/(.+)", "\\1", panAsp_v1$METS$CHEBI))
panAsp_v1$METS$CHEBI <- gsub("chebi:CHEBI/", "CHEBI:", panAsp_v1$METS$CHEBI)

# Get KEGG_self column
panAsp_v1$METS$KEGG_self <- Ani_idtokegg$KEGG_self[match(panAsp_v1$METS$NAME,Ani_idtokegg$NAME)]
```

## Asp_kegg

```{r}
# Get RepName column
Asp_kegg$METS$RepName <- NA

# Get KEGG column
Asp_kegg$METS$KEGG <- Asp_kegg$METS$MIRIAM
Asp_kegg$METS$KEGG[!grepl("kegg",Asp_kegg$METS$KEGG)] <- NA
Asp_kegg$METS$KEGG <- gsub("kegg.compound/","",Asp_kegg$METS$KEGG)
Asp_kegg$METS$KEGG <- gsub("kegg.glycan/","",Asp_kegg$METS$KEGG)
Asp_kegg$METS$KEGG <- gsub(";.*","",Asp_kegg$METS$KEGG)

# Get CHEBI column
Asp_kegg$METS$CHEBI <- Asp_kegg$METS$MIRIAM
Asp_kegg$METS$CHEBI[!grepl("chebi",Asp_kegg$METS$CHEBI)] <- NA
Asp_kegg$METS$CHEBI <- gsub("\\;.*","",gsub(".*chebi/(.+)", "\\1", Asp_kegg$METS$CHEBI))
```

# Update replacementID

## Based on KEGG_self

```{r}
table(Asp_kegg$METS$KEGG %in% panAsp_v1$METS$KEGG_self)
```

Note that there are 10 instances where a KEGG ID is ambiguously assigned a 
replacement ID in the pan genome. As such, we will need to decide (manually) 
which replacement ID these KEGG IDs should get.

But still be mindful of these metabolites! For instance, C00080 in the pan 
genome either refers to protons, or protons involved in energy metabolism. If 
we here set C00080 to just protons, that means that none of the protons in these
newly added reactions are involved in energy metabolism. Same goes for e.g.
CDP-diacylglycerol versus Cdp-diacylglycerol used for cardiolipin, 
Cdp-diacylglycerol used for phosphatidylcholine, etc.

Also difficult, the pan genome only has 1-acyl-sn-glycerol-3-phosphate used for 
cardiolipin/phosphatidylcholine/phosphatidylethanolamine/phosphatidylserine, but
not just 1-acyl-sn-glycerol-3-phosphate. Same for 1-acyl-sn-glycerone phosphate.

For Glucocerebroside vs Glucocerebroside2 it is just a difference between
double bond or not.

```{r}
panAsp_v1$METS[which(panAsp_v1$METS$KEGG_self == "C00026"),c("ID","REPLACEMENT ID")]
panAsp_v1$METS[which(panAsp_v1$METS$KEGG_self == "C00080"),c("ID","REPLACEMENT ID")]
panAsp_v1$METS[which(panAsp_v1$METS$KEGG_self == "C00221"),c("ID","REPLACEMENT ID")]
panAsp_v1$METS[which(panAsp_v1$METS$KEGG_self == "C00269"),c("ID","REPLACEMENT ID")]
panAsp_v1$METS[which(panAsp_v1$METS$KEGG_self == "C00334"),c("ID","REPLACEMENT ID")]
panAsp_v1$METS[which(panAsp_v1$METS$KEGG_self == "C00350"),c("ID","REPLACEMENT ID")]
panAsp_v1$METS[which(panAsp_v1$METS$KEGG_self == "C00681"),c("ID","REPLACEMENT ID")]
panAsp_v1$METS[which(panAsp_v1$METS$KEGG_self == "C01190"),c("ID", "COMPOSITION", "REPLACEMENT ID")]
panAsp_v1$METS[which(panAsp_v1$METS$KEGG_self == "C03372"),c("ID","REPLACEMENT ID")]
panAsp_v1$METS[which(panAsp_v1$METS$KEGG_self == "C05764"),c("ID","REPLACEMENT ID")]
```

```{r}
overwrite <- list("C00026"="AKG",
     "C00080"="H",
     "C00221"="bDGLC",
     "C00269"="CDPDG",
     "C00334"="GABA",
     "C00350"="PE",
     "C00681"="AGL3P",
     "C01190"="GLUCER",
     "C03372"="AT3P2",
     "C05764"="C160ACP")
```

```{r}
hlp <- Asp_kegg$METS$KEGG[Asp_kegg$METS$KEGG %in% panAsp_v1$METS$KEGG_self]
hlp <- unique(hlp[!is.na(hlp)])
for (element in hlp) {
    newrepname <- unique(panAsp_v1$METS$RepName[which(panAsp_v1$METS$KEGG_self == element)])
    target <- which(Asp_kegg$METS$KEGG == element)
    Asp_kegg$METS$RepName[target] <- rep(newrepname,length(target))
}
Asp_kegg$METS[which(Asp_kegg$METS$KEGG %in% names(overwrite)),c("KEGG","RepName")]

for (i in seq_along(overwrite)) { # overwrite ambiguous kegg-replacement links
    newrepname <- unlist(unname(overwrite)[i])
    target <- which(Asp_kegg$METS$KEGG == names(overwrite)[i])
    Asp_kegg$METS$RepName[target] <- rep(newrepname,length(target))
}
Asp_kegg$METS[which(Asp_kegg$METS$KEGG %in% names(overwrite)),c("KEGG","RepName")]
```

```{r}
table(!is.na(Asp_kegg$METS$RepName))
```

## Based on KEGG

Again, there is 1 instance where a KEGG ID is ambiguously assigned a 
replacement ID in the pan genome. As such, we will need to decide (manually) 
which replacement ID this KEGG ID should get.

```{r}
panAsp_v1$METS[which(panAsp_v1$METS$KEGG == "C02737"),c("ID","REPLACEMENT ID")]
overwrite <- list("C02737"="PS")
```

```{r}
target <- Asp_kegg$METS[is.na(Asp_kegg$METS$RepName),]
hlp <- target$KEGG[target$KEGG %in% panAsp_v1$METS$KEGG]
hlp <- unique(hlp[!is.na(hlp)])

for (element in hlp) {
    newrepname <- unique(panAsp_v1$METS$RepName[which(panAsp_v1$METS$KEGG == element)])
    print(element);print(newrepname)
    target <- which(Asp_kegg$METS$KEGG == element)
    Asp_kegg$METS$RepName[target] <- rep(newrepname,length(target))
}
Asp_kegg$METS$`REPLACEMENT ID`[which(Asp_kegg$METS$KEGG == "C02737")]<-overwrite$C02737
```

## Based on CHEBI

```{r}
target <- Asp_kegg$METS[is.na(Asp_kegg$METS$RepName),]
hlp <- target$CHEBI[target$CHEBI %in% panAsp_v1$METS$CHEBI]
hlp <- unique(hlp[!is.na(hlp)])

for (element in hlp) {
    newrepname <- unique(panAsp_v1$METS$RepName[which(panAsp_v1$METS$CHEBI == element)])
    #print(element);print(newrepname)
    target <- which(Asp_kegg$METS$CHEBI == element)
    Asp_kegg$METS$RepName[target] <- rep(newrepname,length(target))
}
```

```{r}
table(!is.na(Asp_kegg$METS$RepName))
```

## Based on NAME

```{r}
target <- Asp_kegg$METS[is.na(Asp_kegg$METS$RepName),]
hlp <- tolower(target$NAME)[tolower(target$NAME) %in% tolower(panAsp_v1$METS$NAME)]
hlp <- unique(hlp[!is.na(hlp)])

for (element in hlp) {
    newrepname <- unique(panAsp_v1$METS$RepName[which(tolower(panAsp_v1$METS$NAME) == element)])
    #print(element);print(newrepname)
    target <- which(tolower(Asp_kegg$METS$NAME) == element)
    Asp_kegg$METS$RepName[target] <- rep(newrepname,length(target))
}
```

```{r}
table(!is.na(Asp_kegg$METS$RepName))
```

```{r}
Asp_kegg$METS$RepIdNew <- Asp_kegg$METS$`REPLACEMENT ID`
Asp_kegg$METS$RepIdNew[!is.na(Asp_kegg$METS$RepName)] <- paste0(Asp_kegg$METS$RepName[!is.na(Asp_kegg$METS$RepName)],
                            "[",
                            Asp_kegg$METS$COMPARTMENT[!is.na(Asp_kegg$METS$RepName)],
                            "]")
Asp_kegg$METS$RepIdNew[is.na(Asp_kegg$METS$RepName)] <- paste0("Added_", Asp_kegg$METS$RepIdNew[is.na(Asp_kegg$METS$RepName)])

output <- data.frame(RepOriginal = Asp_kegg$METS$`REPLACEMENT ID`,
                     RepConsolidated = Asp_kegg$METS$RepIdNew)
```

# Write output

```{r}
write.xlsx(output, "../data/intermediate/OtherAsp_metabolite_table.xlsx")
```

# Downstream

```
table(is.na(Asp_kegg$METS$RepName))
```

So then we are still left with 427 metabolites, and suggestion is that these are
all unique. To some extent, this does not surprise me, as the fumigatus GEM had 
a lot of metabolites that were unobserved in the oryzae and niger GEMs.

```
View(Asp_kegg$METS[is.na(Asp_kegg$METS$RepName),])
```

But if we inspect these metabolites, I am quite sceptical that some of these 
would not exist in panAspGEM_v1 yet.

- Glutamate: glutamate in hlp, L-glutamate in pan
- Fatty acids: always problematic, e.g. Arachidonate is C20, in hlp we also have
a dump term "Fatty acid", ...
- Fructose: again stereochemistry differences, perhaps pointing to different
resolution
- Glucose: same as fructose, e.g., pan has alpha-D-Glucose 6-phosphate and 
beta-D-Glucose 6-phosphate, while hlp has D-Glucose 6-phosphate

```
panAsp_v1$METS[grep("manni", panAsp_v1$METS$NAME, ignore.case = TRUE),
               c("NAME","KEGG","CHEBI")]
```

```
hlp[grep("glucose", hlp$NAME, ignore.case = TRUE),
    c("NAME","KEGG","CHEBI")]
```

So, I suppose I connect these 796-427=369 metabolites to panAsp, and then 
consider the rest as unique, but then curate downstream.

Perhaps it could be useful to keep track of where the unique metabolites in the
panGEM came from.




