---
title: "Metabolites"
output: html_document
date: "2024-10-03"
---

```{r, message=FALSE, warning=FALSE}
library(stringr)
library(openxlsx)
library("kableExtra")
```

```{r}
read_excel_allsheets <- function(filename) {
    sheets <- openxlsx::getSheetNames(filename)
    x <- lapply(sheets, function(X) openxlsx::read.xlsx(filename, sheet = X))
    names(x) <- sheets
    x
}
```

# Obtain gems

```{r}
niger_gem <- read_excel_allsheets("../GEMs_orig/AspNiger/40694_2018_60_MOESM2_ESM.xlsx")
fumigatus_gem <- read_excel_allsheets("../GEMs_orig/AspFumigatus/Pan_Aspergillus_fumigatus.xlsx")
oryzae_gem <- read_excel_allsheets("../GEMs_orig/AspOryzae/iWV1314.xlsx")
```

# Establishing a link between metabolites

```{r}
head(fumigatus_gem$METS)
```


- kegg.compound for many (see also REPLACEMENT.ID), except biomass/lipid/...
- InChi, bigg, biocyc, chebi, hmdb, metanet, pubchem, and seed.compound for many

```{r}
head(oryzae_gem$METS)
```

- Problematic: no formalized identifiers, only ID, NAME and REPLACEMENT.ID

```{r}
head(niger_gem$METS)
```

```{r}
niger_gem$METS[which(is.na(niger_gem$METS$MIRIAM)),]
```

```{r}
table(is.na(niger_gem$METS$InChI))
```


- kegg.compound and Chebi for many (see also REPLACEMENT.ID)
- Inchi for half
- for some, no formalized identifiers

## TRY

--- 3-cyanopyridine

```{r}
niger_gem$METS[which(is.na(niger_gem$METS$MIRIAM)),][1,]
```

For this one, I have the inchi.
I can search the inchi using InChI:InChI=1S/C6H4N2/c7-4-6-2-1-3-8-5-6/h1-3,5H
in https://metacyc.org/metabolite-translation-service.shtml.

This gives me several identifiers, but not kegg.

I then look for 3-cyanopyridine and 3-Cyanopyridine in 
https://www.genome.jp/kegg/compound/ -> no hit.

I then take the molecular mass from the metacyc search, and feed that to kegg:
no hit, but kegg does have 2-Cyanopyridine and 4-Cyanopyridine, just no
3-Cyanopyridine.

But, from metacyc we also get Chebi (86556). If we query that in 
https://www.ebi.ac.uk/chebi/, we get 3-pyridinecarbonitrile, which turns out to 
be synonym for 3-Cyanopyridine.

However, again, searching this in kegg does not give a hit.

This compound seems really not exist in KEGG. So what's next? Not use KEGG at all?
But is there even a database in which all required compounds exist?

--- C00659

```{r}
fumigatus_gem$METS[1:2,]
```
  
We have the kegg compound for this, let's try to find chebi

Via KEGG, we can manually click to chebi (C00659 -> 982)

But when trying to do this more automatically in metacyc, it fails.
Kegg:C00659   ChEBI:982   InChI:InChI=1S/C6H10O4/c1-3-6(10,4(2)7)5(8)9/h10H,3H2,1-2H3,(H,8,9)   InChIkey:InChIKey=VUQLHQFKACOHNZ-UHFFFAOYSA-N   CF:C6H10O4

This is probably inevitable; I strongly suspect that the people that constructed
this GEM in the first place used Metabolite Translation Service for MetaCyc.
As such, I would return as much information as they did, not more.

I could also input C000659 in
https://www.metaboanalyst.ca/faces/upload/ConvertView.xhtml
this succeeds in returning Chebi

One more possibility would be to do this in R using MetaFetcheR. However, this
is a non-maintained github page that failed to install out of the box and that
requires installing additional things; as such, this seems far from ideal.

One other possibility is to use the pypi library bioservices
https://bioservices.readthedocs.io/en/latest/convertor_tutorial.html

This seems to work well! I could actually just obtain a Kegg-Chebi dictionary
from there and import it here!

```{r}
kegg_chebi_dict <- read.csv2("../pythonAnalyses/kegg_chebi_dict.csv", 
                  sep=",",
                  na.strings = "",
                  header = FALSE)
colnames(kegg_chebi_dict) <- c("KEGG","CHEBI")
kegg_chebi_dict$KEGG <- gsub(".*cpd:(.+)", "\\1", kegg_chebi_dict$KEGG)
kegg_chebi_dict$CHEBI <- gsub(".*chebi:(.+)", "\\1", kegg_chebi_dict$CHEBI)

head(kegg_chebi_dict)
```


# Consolidate

## Make KEGG and Chebi for fumigatus

```{r}
head(fumigatus_gem$METS$MIRIAM)
```

```{r}
head(fumigatus_gem$METS$REPLACEMENT.ID)
# this turns out to be a promising column.
# If available, they used the KEGG+compartment
# If not, then they used the CHEBI+compartment
# If that is also not available, they use "Corecocluster"+compartment

table(grepl("C(\\d{5})\\[",fumigatus_gem$METS$REPLACEMENT.ID)) # KEGG
table(grepl("CHEBI",fumigatus_gem$METS$REPLACEMENT.ID)) # CHEBI
table(grepl("Cluster",fumigatus_gem$METS$REPLACEMENT.ID)) # coreco cluster
table(grepl("C(\\d{3})\\[",fumigatus_gem$METS$REPLACEMENT.ID)) # CoreCo/C[digits]


fumigatus_gem$METS[!grepl("C(\\d{5})\\[",fumigatus_gem$METS$REPLACEMENT.ID)&
                                    !grepl("CHEBI",fumigatus_gem$METS$REPLACEMENT.ID) &
                                    !grepl("Cluster",fumigatus_gem$METS$REPLACEMENT.ID) &
                                    !grepl("C(\\d{3})\\[",fumigatus_gem$METS$REPLACEMENT.ID),
                   c("NAME","MIRIAM","REPLACEMENT.ID")]
# these last 305 compounds are mainly tagged CoreCo or A.fumigatus.central.metabolism 
# in miriam. For instance, A.fumigatus.central.metabolism/CERB1A seems to refer to
# cerberin1 comp. A (not sure if comp. stands for compound, component, or compartiment).
```

```{r}
fumigatus_gem$METS[605,]
```


```{r}
table(grepl("kegg.compound",fumigatus_gem$METS$MIRIAM))
table(grepl("chebi/CHEBI",fumigatus_gem$METS$MIRIAM))

table(grepl("kegg",fumigatus_gem$METS$MIRIAM),
      grepl("chebi",fumigatus_gem$METS$MIRIAM))
```

```{r}
# Get KEGG column
fumigatus_gem$METS$KEGG <- fumigatus_gem$METS$MIRIAM
fumigatus_gem$METS$KEGG[!grepl("kegg",fumigatus_gem$METS$KEGG)] <- NA
fumigatus_gem$METS$KEGG <- gsub("\\;.*","",gsub(".*kegg.compound/(.+)", "\\1", fumigatus_gem$METS$KEGG))
```


```{r}
# Get chebi column
fumigatus_gem$METS$CHEBI <- fumigatus_gem$METS$MIRIAM
fumigatus_gem$METS$CHEBI[!grepl("chebi",fumigatus_gem$METS$CHEBI)] <- NA
fumigatus_gem$METS$CHEBI <- gsub("\\;.*","",gsub(".*CHEBI:(.+)", "\\1", fumigatus_gem$METS$CHEBI))
```

```{r}
sum(is.na(fumigatus_gem$METS[,c("KEGG","CHEBI")]))
```

```{r}
# update using our KEGG/CHEBI dictionary
table(fumigatus_gem$METS$KEGG %in% kegg_chebi_dict$KEGG)
table(fumigatus_gem$METS$CHEBI %in% kegg_chebi_dict$CHEBI)

# THE DICTIONARY IS INCOMPLETE...
# For instance
"C04253" %in% fumigatus_gem$METS$KEGG
"C04253" %in% kegg_chebi_dict$KEGG 
# indeed, no chebi for this kegg, also not on kegg website (could link to pubchem though)
# also no kegg-chebi link through metacyc (could link to inchi though)
# This makes sense, on KEGG the compound is called  	
# Electron-transferring flavoprotein, which is  generic term, not a real compound


# Conclusion, not all KEGGs have a chebi
# Let's see if we can still get something from it
```

```{r}
fumigatus_gem$METS$CHEBI_dic <- kegg_chebi_dict$CHEBI[match(fumigatus_gem$METS$KEGG,kegg_chebi_dict$KEGG)]
```

```{r}
head(fumigatus_gem$METS[,c("KEGG","CHEBI","CHEBI_dic")])
```

```{r}
# notice that there are discrepancies between CHEBI and CHEBI_dic:
fumigatus_gem$METS[which(fumigatus_gem$METS$KEGG=="C00009"),c("KEGG","CHEBI","CHEBI_dic")]
# these are the protonated vs the non-protonated forms of the compound!
```

```{r}
# then we also have compounds with chebi but no kegg
fumigatus_gem$METS[c(59,60),c("KEGG","CHEBI","CHEBI_dic")]

# if the CHEBI is in kegg_chebi_dict, add it:
target <- fumigatus_gem$METS$CHEBI[which(!is.na(fumigatus_gem$METS$CHEBI)&
                           is.na(fumigatus_gem$METS$CHEBI_dic))]
table(target) # mostly chebi compound 52320 (ergosteryl ester??)
```


```{r}
fumigatus_gem$METS[which(fumigatus_gem$METS$CHEBI=="52320"),]
```

Clearly indicates that, at least for these compounds, we'd better work
with the replacement ID column, which has the proper CHEBIs!


Hi Ed,

Yesterday, I am have been looking into
(1) consolidating the gene IDs between the BPGA output and the invidual GEMs,
(2) mapping the metabolites across the GEMs, and
(3) consolidating the compartments across the GEMs.

While I made progress in (1-3), I also identified several obstacles in each.
