---
title: "Consolidate metabolites for A. niger"
author: "Jeroen Gilis"
date: "12 November 2024"
output: 
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

```{r, message=FALSE, warning=FALSE}
library(stringr)
library(openxlsx)
library(RCurl)
library(data.table)
library(KEGGREST)
```

## Helpers

```{r}
source("./Rutilities.R")
```

# Obtain GEM

```{r}
# Note: excels should be closed when reading them in R
#niger_gem <- read_excel_allsheets("../model/Ani_draft_comp.xlsx")
niger_gem <- read_excel_allsheets("../data/intermediate/Ani_draft_comp.xlsx")
```

# Inspect

```{r}
head(niger_gem$METS)
```

```{r}
table(grepl("kegg.compound",niger_gem$METS$MIRIAM)) # KEGG for half
table(grepl("chebi/CHEBI",niger_gem$METS$MIRIAM)) # Chebi for most
table(!is.na(niger_gem$METS$InChI)) # inchi for half
```

- kegg.compound and Chebi for many (see also REPLACEMENT.ID)
- Inchi for half
- for some, nothing formalized, only names

```{r}
table(grepl("kegg",niger_gem$METS$MIRIAM),
      grepl("chebi",niger_gem$METS$MIRIAM))

table(is.na(niger_gem$METS[!grepl("kegg",niger_gem$METS$MIRIAM) &
                 !grepl("chebi",niger_gem$METS$MIRIAM),"InChI"]))
# if no KEGG nor CHEBI, inchi can help a bit
```

# Wrangle

```{r}
# Get KEGG column
niger_gem$METS$KEGG <- niger_gem$METS$MIRIAM
niger_gem$METS$KEGG[!grepl("kegg",niger_gem$METS$KEGG)] <- NA
niger_gem$METS$KEGG <- gsub("\\;.*","",gsub(".*kegg.compound/(.+)", "\\1", niger_gem$METS$KEGG))

# Get chebi column
niger_gem$METS$CHEBI <- niger_gem$METS$MIRIAM
niger_gem$METS$CHEBI[!grepl("chebi",niger_gem$METS$CHEBI)] <- NA
niger_gem$METS$CHEBI <- gsub("\\;.*","",gsub(".*chebi/(.+)", "\\1", niger_gem$METS$CHEBI))
```

# Filter

The code in this section is not required anymore.

```
# By moving metabolites from a compartment to cytoplasm, we have created
# duplicate IDs.
niger_gem$METS[which(duplicated(niger_gem$METS$`REPLACEMENT ID`)),c(2,3,9)]
```

```
#Inspect the reactions that were moved
Ani_draft <- read_excel_allsheets("../model/Ani_draft.xlsx")
Ani_draft$RXNS[grepl("\\[v\\]",Ani_draft$RXNS$EQUATION),c(3,4)]
Ani_draft$RXNS[grepl("\\[g\\]",Ani_draft$RXNS$EQUATION),c(3,4)]
Ani_draft$RXNS[grepl("\\[n\\]",Ani_draft$RXNS$EQUATION),"EQUATION"]
```

```
# I think that the only problematic reaction is the V-ATPase. It does not make
# sense to keep this reaction, and as such, we can also remove the 
# vacuolar protons entry.
which(duplicated(niger_gem$METS$`REPLACEMENT ID`))
niger_gem$METS[niger_gem$METS$`REPLACEMENT ID`=="H[c]",]

niger_gem$RXNS[grep("UDP-galactose \\(furanose\\)\\[c\\]",niger_gem$RXNS$EQUATION),][2,c(3,4)]

# niger_gem$METS <- niger_gem$METS[-grep("Vacuolar protons",niger_gem$METS$ID),]
# niger_gem$RXNS <- niger_gem$RXNS[-grep("Vacuolar protons",niger_gem$RXNS$EQUATION),]

# remove the others too -> no longer do this, causes problems to port to matlab
#niger_gem$METS <- niger_gem$METS[-which(duplicated(niger_gem$METS$`REPLACEMENT ID`)),]
```

# Get KEGG based on chebi

## pubchem compound

Now, try to add KEGGs for the IDs for which we have CHEBI.

Since each CHEBI should only have one KEGG, I here only search for the 
missing ones. Searching for all could be relevant if we need to find mistakes in
the original annotation.

```{r}
chebi_query <- niger_gem$METS$CHEBI[
  which(is.na(niger_gem$METS$KEGG) &
          !is.na(niger_gem$METS$CHEBI))]
chebi_query <- unique(unlist(str_split(chebi_query,pattern = "/")))
```

```{r,eval=FALSE}
CHEBItoKEGG_comp <- linkMetaboliteIdentifiers(query = chebi_query, 
                                       target = "^C(\\d{5})\\b", 
                                       reader = "read.csv",
                                       database = "pubchem_compound_name",
                                       verbose = TRUE)
saveRDS(CHEBItoKEGG_comp,"../data/intermediate/niger_CHEBItoKEGG_comp.rds")
```

```{r}
CHEBItoKEGG_comp <- readRDS("../data/intermediate/niger_CHEBItoKEGG_comp.rds")

niger_gem$METS$KEGG_v2 <- CHEBItoKEGG_comp$target[match(niger_gem$METS$CHEBI, CHEBItoKEGG_comp$query)]
niger_gem$METS[!is.na(niger_gem$METS$KEGG_v2),c("CHEBI", "KEGG", "KEGG_v2")][11:20,]
# while we add some truly new things, we also add things the original GEM filled
# in in come spaces but not all
```

## pubchem substance

Did not provide additional hits

## bioservices dictionary

The file kegg_chebi_dict.csv was created by running the script 
`map_kegg_chebi.ipynb`. This file contains curated links between CHEBI and KEGG
metabolite identifiers, and was constructed for the Bioservices Python library.

```{r}
kegg_chebi_dict <- read.csv2("../data/intermediate/kegg_chebi_dict.csv", header=FALSE, sep=",")

# wrangle
colnames(kegg_chebi_dict) <- c("KEGG","CHEBI")
kegg_chebi_dict$KEGG <- gsub("cpd:","",kegg_chebi_dict$KEGG)
kegg_chebi_dict$CHEBI <- gsub("chebi:","CHEBI:",kegg_chebi_dict$CHEBI)

niger_gem$METS$KEGG_v3 <- kegg_chebi_dict$KEGG[match(niger_gem$METS$CHEBI,kegg_chebi_dict$CHEBI)]
```

# Get KEGG based on name

```{r}
name_query <- niger_gem$METS$NAME[
  which(is.na(niger_gem$METS$KEGG) &
          is.na(niger_gem$METS$KEGG_v2) &
          is.na(niger_gem$METS$KEGG_v3) &
          !is.na(niger_gem$METS$NAME))]
name_query <- tolower(unique(name_query))
namesLink <- match(tolower(niger_gem$METS$NAME),name_query)
```

```{r,eval=FALSE}
name_query <- gsub(":"," ",name_query)
name_query <- gsub(","," ",name_query)
name_query <- gsub("\\{","\\[",name_query)
name_query <- gsub("\\}","\\]",name_query)
name_query[grepl("\\(\\+\\)\\-",name_query)] <- "kotanin"
name_query[grepl("\\(\\−\\)\\-",name_query)] <- "gallocatechin"
name_query <- gsub("β","beta",name_query)
name_query <- gsub("γ","gamma",name_query)
# probably wont find these either, but lets try

addKEGG <- lapply(name_query, function(x){
  #print(x)
  query <- keggFind(database = "compound", query = x)
  return(cbind(x,
               paste0(gsub("cpd:", "", names(query)),collapse="/"),
               paste0(unname(query),collapse="/")))
})

saveRDS(addKEGG,"../data/intermediate/niger_NAMEtoKEGG_keggrest.rds")
```

```{r}
addKEGG <- readRDS("../data/intermediate/niger_NAMEtoKEGG_keggrest.rds")

niger_gem$METS$KEGG_v4 <- sapply(addKEGG, function(x){x[2]})[namesLink]
niger_gem$METS$KEGG_v4[niger_gem$METS$KEGG_v4==""] <- NA

niger_gem$METS$AllNames <- sapply(addKEGG, function(x){x[3]})[namesLink]
niger_gem$METS$AllNames[niger_gem$METS$AllNames==""] <- NA
```

```{r}
table(!is.na(niger_gem$METS$KEGG))
table(!is.na(niger_gem$METS$KEGG)|!is.na(niger_gem$METS$KEGG_v2))
table(!is.na(niger_gem$METS$KEGG)|!is.na(niger_gem$METS$KEGG_v2)|!is.na(niger_gem$METS$KEGG_v3))
table(!is.na(niger_gem$METS$KEGG)|!is.na(niger_gem$METS$KEGG_v2)|!is.na(niger_gem$METS$KEGG_v3)|!is.na(niger_gem$METS$KEGG_v4))
```

```{r}
hlp <- niger_gem$METS[is.na(niger_gem$METS$KEGG)&is.na(niger_gem$METS$KEGG_v2)&is.na(niger_gem$METS$KEGG_v3)&is.na(niger_gem$METS$KEGG_v4),]
head(hlp[!duplicated(hlp$NAME),])
```

# Classes that require curation

## tRNA reactions

```{r}
sum(grepl("Aminoacyl-tRNA",niger_gem$METS$NAME))
```

```{r}
index <- grep("Aminoacyl-tRNA",niger_gem$METS$NAME)
niger_gem$METS$ID[index]
```

```{r}
kegg_query <- gsub("Aminoacyl-tRNA ", "tRNA", niger_gem$METS$NAME)
kegg_query[grepl("\\) loaded",kegg_query)] <- c("L-Alanyl-tRNA(Ala)",
                                                "L-Seryl-tRNA(Ser)",
                                                "L-Tryptophanyl-tRNA(Trp)",
                                                "L-Valyl-tRNA(Val)",
                                                "L-Glutamyl-tRNA(Glu)",
                                                "Glycyl-tRNA(Gly)",
                                                "L-Lysyl-tRNA(Lys)",
                                                "L-Phenylalanyl-tRNA(Phe)",
                                                "L-Leucyl-tRNA(Leu)",
                                                "L-Arginyl-tRNA(Arg)",
                                                "L-Isoleucyl-tRNA(Ile)",
                                                "L-Asparaginyl-tRNA(asn)",
                                                "L-Methionyl-tRNA(Met)",
                                                "L-Glutaminyl-tRNA(Gln)",
                                                "L-Tyrosyl-tRNA(Tyr)",
                                                "L-Prolyl-tRNA(Pro)",
                                                "L-Aspartyl-tRNA(Asp)",
                                                "L-Histidyl-tRNA(His)",
                                                "L-Cysteinyl-tRNA(Cys)",
                                                "L-Threonyl-tRNA(Thr)")
```

```{r}
addKEGG <- lapply(kegg_query[index], function(x){
  query <- keggFind(database = "compound", query = x)
  return(cbind(x,
               paste0(gsub("cpd:", "", names(query)),collapse="/"),
               paste0(unname(query),collapse="/")))
})
```

```{r}
hlp <- as.data.frame(cbind(sapply(addKEGG, function(x){x[1]}),
                           sapply(addKEGG, function(x){x[2]})))
hlp$V2[hlp$V2=="C00886/C01635/C20458/C20459"] <- "C01635"
hlp$V2 <- gsub("/.*", "", hlp$V2)
head(hlp)
```

```{r}
niger_gem$METS$KEGG_v5 <- ""
niger_gem$METS$KEGG_v5[index] <- hlp$V2
```

# Consolidate keggs for different sources

```{r}
niger_gem$METS$KEGG_v2[is.na(niger_gem$METS$KEGG_v2)] <- ""
niger_gem$METS$KEGG_v3[is.na(niger_gem$METS$KEGG_v3)] <- ""
niger_gem$METS$KEGG_v4[is.na(niger_gem$METS$KEGG_v4)] <- ""

niger_gem$METS$KEGG_v2[niger_gem$METS$KEGG_v2 == ""] <- NA
niger_gem$METS$KEGG_v3[niger_gem$METS$KEGG_v3 == ""] <- NA
niger_gem$METS$KEGG_v4[niger_gem$METS$KEGG_v4 == ""] <- NA
niger_gem$METS$KEGG_v5[niger_gem$METS$KEGG_v5 == ""] <- NA

niger_gem$METS$KEGG_self <- unlist(lapply(1:nrow(niger_gem$METS), function(x){
  hlp <- niger_gem$METS[x,c(13,14,15,17)]
  hlp <- paste0(hlp[,!is.na(hlp)],collapse = "/")
  paste0(unique(strsplit(hlp, "/")[[1]]),collapse = "/")
}))
niger_gem$METS$KEGG_self[niger_gem$METS$KEGG_self==""] <- NA
niger_gem$METS$KEGG_v2 <- niger_gem$METS$KEGG_v3 <- niger_gem$METS$KEGG_v4 <- niger_gem$METS$KEGG_v5 <- NULL
```

```{r}
multikegg <- niger_gem$METS[lengths(strsplit(niger_gem$METS$KEGG_self, "/")) > 1,]
head(multikegg[!duplicated(multikegg$NAME),c("NAME", "KEGG_self")])
```

```{r}
curated <- list(c("Benzaldehyde","C00261"),
                c("Tetradecenoyl-CoA", "C05273/C06736/C06737"), # location of double bond...
                c("Palmitoyl-[acyl-carrier protein]", "C05764"),
                c("Hexadecenoyl-CoA", "C05272/C21072/C21943"), # location of double bond...
                c("Hexadecenoate", "C13792/C13946"), # location of double bond...
                c("Octadecenoate", "C13787/C14641/C14832/C14834/C20702"), # location of double bond...
                c("Octadecadienoate; Linoleic acid", "C01595"),
                c("Octadecadienoate", "C01595/C04056/C13800/C14827/C15441/C19937"),
                c("But-2-enoyl-[acyl-carrier protein]", "C04246"),
                c("But-2-enoyl-CoA", "C00877"),
                c("Pent-2-enoyl-CoA", "C02451"),
                c("Heptanoyl-CoA", "C11407/C11421/C11947/C11948"),
                c("Octanoate", "C06423"),
                c("Cerebroside 1",NA),
                c("Cerebroside 2",NA),
                c("Lysophosphatidylcholine","C04230/C04233"), # 1-lyso and 2-lyso
                c("Lysophosphatidylethanolamine","C04438/C05973"), # 1-lyso and 2-lyso
                c("Pentanal",NA),
                c("Pseudonigeran (1,3-alpha-glucan)","C02616"),
                c("Ubiquinol", "C00390"),
                c("Glutathione", "C00051"),
                c("Sulfur", "C00087"),
                c("Mevalon", NA), # couldnt find
                c("Rubrofusarin B",NA),
                c("Protoporphyrin IX","C02191"),
                c("Ethanesulfonic acid",NA),
                c("Cellulose", "C00760"),
                c("Carboxylic acid ", "C00060"),
                c("Gallocatechin gallate",NA), # Epigallocatechin gallate is C09731
                c("(−)-gallocatechin",NA), #
                c("trans-caffeic acid","C01197"), # note that in KEGG, caffeic acid and trans-caffeic acid are synonyms
                c("Styrene","C07083"),
                c("Naringenin","C00509"),
                c("Putrescine","C00134"),
                c("Methylamine","C00128"),
                c("hexanal",NA),  # isohexanal is C02373
                c("propylamine", NA),
                c("Cyanate", "C01417"),
                c("Methionine S-oxide", "C02989") ,# Actually for L-Methionine S-oxide 
                c("beta-1,4-Mannan", "C02492"), # 1,4-beta-D-Mannan in KEGG
                c("beta-mannobiose", "C20861"), # Mannobiose, beta-1,4-Mannobiose in KEGG
                c("alpha 1,3-glucan", "C02616"), # 1,3-alpha-D-Glucan
                c("Rhamnogalacturonan", NA),
                c("Dolichol","C00381"),
                c("5'-S-methyl-5'-thioadenosine", "C00170"), #5'-Methylthioadenosine
                c("4-aminobutanal", "C00555"),
                c("Precursor Z", "C18239"),
                c("Molybdenum cofactor","C18237"),
                c("Ethylenediamine",NA),
                c("Hydroquinone","C00530"),
                c("Melezitose","C08243"),
                c("Methyl pyruvate",NA),
                c("2-aminopentanoic acid", "C01799/C01826"), # for D-2-Aminopentanoic acid and L-2-Aminopentanoic acid
                c("Maltodextrin", "C01935"))

for (element in curated) {
  niger_gem$METS$KEGG_self[which(niger_gem$METS$NAME %in% element[1])] <- element[2]
}
```

I dont think its worth it to replace these names here. If we'd really want to do
this (i.e., making the names prettier), we should probably do it in matlab.

```
#  curate some names
# in METS
niger_gem$METS[which(niger_gem$METS$ID == "Octadecadienoate; Linoleic acid[c]"),c("ID","NAME")] <- c(
  "Linoleic acid[c]","Linoleic acid")
niger_gem$METS[which(niger_gem$METS$NAME == "Octadecatrienoate; α-linolenic acid"),c("ID","NAME")] <- c(
  "alpha-Linolenic acid[c]","alpha-Linolenic acid")
niger_gem$METS[which(niger_gem$METS$NAME == "Pseudonigeran (1,3-alpha-glucan)"),c("ID","NAME")] <- c(
  "1,3-alpha-D-Glucan[c]","1,3-alpha-D-Glucan")

# in RXNS
# note that for these components gsub should be fine, but this is not generally
niger_gem$METS[grepl("Octadecadienoate; Linoleic acid", niger_gem$METS$NAME, ignore.case = TRUE),]
niger_gem$METS[grepl("Octadecatrienoate; α-linolenic acid", niger_gem$METS$NAME, ignore.case = TRUE),]
niger_gem$METS[grepl("Pseudonigeran \\(1,3-alpha-glucan\\)", niger_gem$METS$NAME, ignore.case = TRUE),]
# true!

niger_gem$RXNS$EQUATION <- gsub("Octadecadienoate; Linoleic acid[c]",
                                "Linoleic acid[c]",
                                niger_gem$RXNS$EQUATION)
niger_gem$RXNS$EQUATION <- gsub("Octadecatrienoate; α-linolenic acid[c]",
                                "alpha-Linolenic acid[c]",
                                niger_gem$RXNS$EQUATION)
niger_gem$RXNS$EQUATION <- gsub("Pseudonigeran (1,3-alpha-glucan)",
                                "Linoleic acid[c]",
                                niger_gem$RXNS$EQUATION)
```

# Export output

```{r}
# needed for metabolites_niger_oryzae.Rmd
write.xlsx(niger_gem, "../data/intermediate/Ani_draft_comp_mets.xlsx",sep.names = " ")

# Save table to import in matlab (not strictly necessary to have, can remove later)
write.xlsx(niger_gem$METS[,c("NAME","KEGG_self")], "../data/intermediate/niger_metabolite_table.xlsx")
```


# Fix mistakes in original Ani_draft

Fix mistakes that became apparent in metabolites_niger_oryzae.rmd or 
metabolites_niger_fumigatus.rmd.

```
niger_gem <- read_excel_allsheets("../model/Ani_draft_consolidated.xlsx")
niger_gem <- read_excel_allsheets("../scrap/Ani_draft_v2.xlsx")
```

```
# the chebi of Acetoacetyl [acyl-carrier protein][c] is wrong
niger_gem$METS[which(niger_gem$METS$CHEBI == "CHEBI:15345"),c(2,5,9)]
niger_gem$METS[which(niger_gem$METS$ID == "Acetoacetyl [acyl-carrier protein][c]"),
               c("MIRIAM","CHEBI","CHEBI_ID")] <- c("kegg.compound/C05744;chebi/CHEBI:2393",
                                                    "CHEBI:2393","CHEBI:2393[c]")
```

```
write.xlsx(niger_gem, "../model/Ani_draft_consolidated.xlsx",sep.names = " ")
```

---

# Revisit

What I do below can be very useful, but only when jointly looking to the other
GEMs. So, I will incorporate the code below in the other scripts later.

```
target <- niger_gem$METS[is.na(niger_gem$METS$KEGG)&is.na(niger_gem$METS$KEGG_v2)&is.na(niger_gem$METS$KEGG_v3)&is.na(niger_gem$METS$KEGG_v4),]
target <- target[!duplicated(target$NAME),]
```

```
View(target)
```

## Niger-specific metabolites

```
sum(grepl("Yanuthone",target$NAME,ignore.case = TRUE))
sum(grepl("nigrin",target$NAME,ignore.case = TRUE))
sum(grepl("fonsec",target$NAME,ignore.case = TRUE))
sum(grepl("Azanig",target$NAME,ignore.case = TRUE))
```

There are several classes of complex secondary metabolites described by Brandl
himself on the CHEBI website. We will never find a KEGG for these, and it is
very unlikely these will ever appear in the other GEMs.

- yanuthones: A class of meroterpenoids whose core structure consists of 5,6-epoxycyclohex-2-en-1-one which is substituted by varying side chains at positions 3 and 4, and by a sesquiterpene chain at position 6. The core structure may be derived from 6-methylsalicylic acid, which due to a decarboxylation results in a C7 scaffold (a 6-membered methylated ring); these are known as class I yanuthones. Class II yanuthones contain a C6 scaffold derived from an unknown precursor. -> search for yanuthone and epoxycyclo in Aor and Afu gave no hits

- Pyranonigrin: A member of the class of pyranopyrroles that is 2-[(1E)-prop-1-en-1-yl]-6,7-dihydropyrano[2,3-c]pyrrole-4,5-dione carrying two additional hydroxy substituents at positions 3 and 7. -> search for nigrin, pyrano and pyrrole in Aor and Afu gave no hits

- Fonsecinone: A dimeric naphthopyran with formula C32H28O11, originally isolated from Aspergillus niger. On pubchem, fonsecinone is 5-hydroxy-9-(5-hydroxy-6,8-dimethoxy-2-methyl-4-oxobenzo[g]chromen-10-yl)-8,10-dimethoxy-2-methylbenzo[h]chromen-4-one -> Search for fonsec, naphto and pyran in Aor and Afu gave no hits

-Azanigerone: An azaphilone that is the ester obtained by formal condensation of the carboxy group of 2,4-dimethylhexanoic acid with the tertiary hydroxy group of 7-hydroxy-7-methyl-6,8-dioxo-7,8-dihydro-6H-2-benzopyran-3-carboxylic acid. IUPAC: 7-[(2,4-dimethylhexanoyl)oxy]-7-methyl-6,8-dioxo-7,8-dihydro-6H-2-benzopyran-3-carboxylic acid
-> Search for aza and benzopyr in Aor and Afu gave no hits

```
target <- target[!grepl("Yanuthone",target$NAME,ignore.case = TRUE) &
                   !grepl("nigrin",target$NAME,ignore.case = TRUE) &
                   !grepl("fonsec",target$NAME,ignore.case = TRUE) &
                   !grepl("Azanig",target$NAME,ignore.case = TRUE)
                 ,]
```

## Cerebrin

In Ani
1. Octadecanoyl-CoA[c] + Phytosphingosine(C18)[c] => Cerebrin 1 (C18 phytosphingosin and C18:0 fatty acid)[c] + Coenzyme A[c]
2. Octadecenoyl-CoA[c] + Phytosphingosine(C18)[c] => Cerebrin 1 (C18 phytosphingosin and C18:1 fatty acid)[c] + Coenzyme A[c]
3. Octadecanoyl-CoA[c] + Phytosphingosine(C20)[c] => Cerebrin 1 (C20 phytosphingosin and C18:0 fatty acid)[c] + Coenzyme A[c]
4. Octadecenoyl-CoA[c] + Phytosphingosine(C20)[c] => Cerebrin 1 (C20 phytosphingosin and C18:1 fatty acid)[c] + Coenzyme A[c]

5. Octadecanoyl-CoA[c] + Sphinganine(C18)[c] => Cerebrin 2 (C18 sphinganine and C18:0 fatty acid)[c] + Coenzyme A[c]
6. Octadecenoyl-CoA[c] + Sphinganine(C18)[c] => Cerebrin 2 (C18 sphinganine and C18:1 fatty acid)[c] + Coenzyme A[c]
7. Octadecanoyl-CoA[c] + Sphinganine(C20)[c] => Cerebrin 2 (C20 sphinganine and C18:0 fatty acid)[c] + Coenzyme A[c]
8. Octadecenoyl-CoA[c] + Sphinganine(C20)[c] => Cerebrin 2 (C20 sphinganine and C18:1 fatty acid)[c] + Coenzyme A[c]

9. Octadecanoyl-CoA[c] + Sphinganine(C19:1)[c] => Cerebroside 1[c] + Coenzyme A[c]

In Afu ==>
1. stearoyl-CoA[c] +	phytosphingosine[c] => Cerebrin 1[c] + coenzyme A[c]
2. oleoyl-CoA[c] + phytosphingosine[c] => Cerebrin 1[c] + coenzyme A[c]
3. stearoyl-CoA[c] + Phytosphingosine(C20)[c] => Cerebrin 1[c] + coenzyme A[c]
4. oleoyl-CoA[c] + Phytosphingosine(C20)[c] => Cerebrin 1[c] + coenzyme A[c]

5. stearoyl-CoA[c] + sphinganine[c] => Cerebrin 2[c] + coenzyme A[c]
6. oleoyl-CoA[c] + sphinganine[c] => Cerebrin 2[c] + coenzyme A[c]
7. stearoyl-CoA[c] + sphinganine (C18)[c] => Cerebrin 2[c] + coenzyme A[c]
8. oleoyl-CoA[c] + sphinganine (C18)[c] => Cerebrin 2[c] + coenzyme A[c]

9. stearoyl-CoA[c] + Sphinganine[c]  => Cerebroside 1[c] + coenzyme A[c]

################################################################################

################################################################################


Sphinganine(C18), Sphinganine(C19:1) and Sphinganine(C20) -> Sphinganine
Phytosphingosine(C18) and Phytosphingosine(C20) -> Phytosphingosine
UDP-galactose (pyranose) -> UDP-galactose

---  Cerebrin 1 after Octadecanoyl or stearoyl -> Cerebrin 1a
Cerebrin 1 (C18 phytosphingosin and C18:0 fatty acid) and Cerebrin 1 (C20 phytosphingosin and C18:0 fatty acid) -> Cerebrin 1a

---  Cerebrin 1 after Octadecenoyl or oleoyl -> Cerebrin 1b
Cerebrin 1 (C18 phytosphingosin and C18:1 fatty acid) and Cerebrin 1 (C20 phytosphingosin and C18:1 fatty acid) -> Cerebrin 1b

Cerebrin 1 (C18 phytosphingosin and C14:0 fatty acid) and Cerebrin 1 (C20 phytosphingosin and C14:0 fatty acid) -> Cerebrin 1c


Cerebrin 2 (C18 sphinganine and C18:0 fatty acid) and Cerebrin 2 (C20 sphinganine and C18:0 fatty acid) -> Cerebrin 2a
Cerebrin 2 (C18 sphinganine and C18:1 fatty acid) and Cerebrin 2 (C20 sphinganine and C18:1 fatty acid) -> Cerebrin 2b
Cerebrin 2 (C18 sphinganine and C14:0 fatty acid) and Cerebrin 2 (C20 sphinganine and C14:0 fatty acid) -> Cerebrin 2c
Cerebrin 2 (C18 sphinganine and a 2-hydroxy C18:1 fatty acid) -> Cerebrin 2d

Cerebrin 3 (C18 sphingosine and C18:0 fatty acid) and Cerebrin 3 (C20 sphingosine and C18:0 fatty acid) -> Cerebrin 3a
Cerebrin 3 (C18 sphingosine and C18:1 fatty acid) and Cerebrin 3 (C20 sphingosine and C18:1 fatty acid) -> Cerebrin 3b
Cerebrin 3 (C18 sphingosine and C14:0 fatty acid) and Cerebrin 3 (C20 sphingosine and C14:0 fatty acid) -> Cerebrin 3c
Cerebrin 3 (C18 sphingosine and a 2-hydroxy C18:1 fatty acid) -> Cerebrin 3d

Cerebrin 1 (Mixture of C18 and C20 phytosphingosin) -> Cerebrin 1ab
Cerebrin 2 (Mixture of C18 and C20 sphingosin) -> Cerebrin 2ab

Galactosylceramide(Comp. A) -> Galactosylceramide A
Galactosylceramide(Comp. B) -> Galactosylceramide B
Galactosylceramide (Mixture between GALCERA and GALCERB) -> Galactosylceramide AB

Galactosylceramide (C18 sphinganine and 2-hydroxy C18:1 fatty acid) -> Galactosylceramide C
Galactosylceramide (C18 sphingosine and 2-hydroxy C18:1 fatty acid) -> Galactosylceramide D

################################################################################

Octadecanoyl-CoA[c] + Phytosphingosine[c] => Cerebrin 1a[c] + Coenzyme A[c]
Octadecenoyl-CoA[c] + Phytosphingosine[c] => Cerebrin 1b[c] + Coenzyme A[c]

0.25 Cerebrin 1a[c] + 0.25 Cerebrin 1b[c] + 0.25 Cerebrin 1a[c] + 0.25 Cerebrin 1b[c] => Cerebrin 1ab[c]

Octadecanoyl-CoA[c] + Sphinganine[c] => Cerebrin 2a[c] + Coenzyme A[c]
Octadecenoyl-CoA[c] + Sphinganine[c] => Cerebrin 2b[c] + Coenzyme A[c]

0.25 Cerebrin 2a[c] + 0.25 Cerebrin 2b[c] + 0.25 Cerebrin 2a[c] + 0.25 Cerebrin 2b[c] => Cerebrin 2ab[c]

Cerebrin 2a[c] + UDP-galactose[c] => Galactosylceramide A[c] + UDP[c]
Cerebrin 2b[c] + UDP-galactose[c] => Galactosylceramide B[c] + UDP[c]

Octadecanoyl-CoA[c] + Sphinganine[c] => Cerebroside 1[c] + Coenzyme A[c]
Cerebroside 1[c] + UDPglucose[c] => Glucocerebroside[c] + UDP[c]
Octadecenoyl-CoA[c] + Sphinganine[c] => Cerebroside 2[c] + Coenzyme A[c]
Cerebroside 2[c] + UDPglucose[c] => Glucocerebroside 2[c] + UDP[c]

Cerebrin 2ab[c] + 1-Phosphatidyl-D-myo-inositol[c] => Water[c] + Inositol phosphorylceramide[c]

0.5 Galactosylceramide A[c] + 0.5 Galactosylceramide B[c] => Galactosylceramide AB[c]

0.000223 Myristic acid[c] + 0.00148 Palmitate[c] + 0.000245 Stearate[c] + 0.001678 Octadecenoate; Oleic acid[c] + 0.001664 Octadecadienoate; Linoleic acid[c] + 4.7e-05 Octadecatrienoate; α-linolenic acid[c] + 3e-05 Cerebrin 1ab[c] + 3.1e-05 Cerebrin 2ab[c] + 0.001746 Cardiolipin[m] + 0.001009 Diacylglycerol[c] + 0.007859 Digalactosyl-diacylglycerol[c] + 0.010038 Ergosterolester[c] + 0.034062 Ergosterol[c] + 2.5e-05 Galactosylceramide AB[c] + 2.4e-05 Glucocerebroside[c] + 2.4e-05 Glucocerebroside 2[c] + 0.008912 Monoacylglycerol[c] + 0.012374 Monoglucosyloxyoctadecenoic acid[c] + 0.030053 Monogalactosyl-diacylglycerol[c] + 0.015312 Phosphatidylcholine[c] + 0.034807 Phosphatidylethanolamine[c] + 0.000359 Phosphatidylserine[c] + 0.010083 Triacylglycerol[c] + 5e-06 Trigalactosyldimannosylinositol-P-ceramide[c] => Artificial metabolite. The metabolites found in the biomass. Measured in grams[c]

Tetradecanoyl-CoA[c] + Phytosphingosine[c] => Coenzyme A[c] + Cerebrin 1c[c]
Tetradecanoyl-CoA[c] + Sphinganine[c] => Coenzyme A[c] + Cerebrin 2c[c]

UDP-galactose[c] + Cerebrin 2d[c] <=> UDP[c] + Galactosylceramide C[c]
Sphinganine[c] + 2-hydroxy octadecenoyl-CoA[c] <=> Coenzyme A[c] + Cerebrin 2d[c]

Cerebrin 2a[c] + 2 Intracellular protons[c] + Oxygen[c] => 2 Water[c] + Cerebrin 3a[c]
Cerebrin 2b[c] + 2 Intracellular protons[c] + Oxygen[c] => 2 Water[c] + Cerebrin 3b[c]
2 Intracellular protons[c] + Oxygen[c] + Cerebrin 2c[c] => 2 Water[c] + Cerebrin 3c[c]
2 Intracellular protons[c] + Oxygen[c] + Cerebrin 2d[c] => 2 Water[c] + Cerebrin 3d[c]

UDP-galactose[c] + Cerebrin 3d[c] <=> UDP[c] + Galactosylceramide D[c]

Cerebrin 1a[c] + Water[c] => Stearate[c] + Phytosphingosine[c]
Sphinganine[c] + NADPH[c] + Oxygen[c] => Phytosphingosine[c] + Water[c] + NADP+[c]


################################## Afu

sphinganine[c] + stearoyl-CoA[c] => coenzyme A[c] + Cerebrin 2[c]
phytosphingosine[c] + oleoyl-CoA[c] => coenzyme A[c] + Cerebrin 1[c]
stearoyl-CoA[c] + sphinganine (C18)[c] => coenzyme A[c] + Cerebrin 2[c]
oleoyl-CoA[c] + Phytosphingosine(C20)[c] => coenzyme A[c] + Cerebrin 1[c]
Sphinganine[c] + stearoyl-CoA[c] => coenzyme A[c] + Cerebroside 1[c]
phytosphingosine[c] + stearoyl-CoA[c] => coenzyme A[c] + Cerebrin 1[c]
sphinganine[c] + oleoyl-CoA[c] => coenzyme A[c] + Cerebrin 2[c]
stearoyl-CoA[c] + Phytosphingosine(C20)[c] => coenzyme A[c] + Cerebrin 1[c]
oleoyl-CoA[c] + sphinganine (C18)[c] => coenzyme A[c] + Cerebrin 2[c]


################################


```
target[grepl("Cerebrin",target$NAME,ignore.case = TRUE),]
```

```
target[grepl("sphing",target$NAME,ignore.case = TRUE),]
```

```
niger_gem$RXNS[grepl("ceramid",niger_gem$RXNS$NAME,ignore.case = TRUE),]

target[grepl("sphing",target$NAME,ignore.case = TRUE),]
```

## Fatty acids

```
sum(grepl("Undec",niger_gem$METS$NAME))
sum(grepl("Dodec",niger_gem$METS$NAME))
sum(grepl("Tridec",niger_gem$METS$NAME))
sum(grepl("Tetradec",niger_gem$METS$NAME))
sum(grepl("Pentadec",niger_gem$METS$NAME))
sum(grepl("Hexadec",niger_gem$METS$NAME))
sum(grepl("Heptadec",niger_gem$METS$NAME))
sum(grepl("Octadec",niger_gem$METS$NAME))
sum(grepl("Nonadec",niger_gem$METS$NAME))
sum(grepl("Eicos",niger_gem$METS$NAME))
```

In Aor and Afu, we use the regular names (stearoyl,oleoyl,myristoyl,...).
This translation can be done through KEGG. However, for many compounds
(e.g. those in target) we already dont have kegg. For some we could get
the kegg maybe, but for instance:

Octadecanoyl-CoA = stearoyl CoA = KEGG C00412 -> found
Octadecanoyl-carnitine = stearoyl carnitine -> not in kegg, but the name is in Aor


Try to change to these

```
target[5:89,]
```

Octadecanoyl-CoA = stearoyl CoA = KEGG C00412
Octadecenoyl-CoA = Oleoyl CoA = KEGG C00510
Tetradecanoyl-CoA = Myristoyl-CoA = C02593

Octadecanoate is stearate
Octadecanoic acid is Stearic acid =  	C01530

coenzyme A[p] + ATP[p] + stearate[p] <=> AMP[p] + diphosphate[p] + stearoyl-CoA[p]

Changes:

```
target[grepl("Octadecan",target$NAME,ignore.case = TRUE),]
```


- Octodecan to stear


## other


```
target$NAME[grepl("nigrin",target$NAME)]
```


```
sum(grepl("\\[acyl-carrier protein\\]",target$NAME))
sum(grepl("anoyl",target$NAME))

sum(grepl("Cerebrin",target$NAME))
sum(grepl("Aminoacyl-tRNA",target$NAME))
```
