---
title: "Consolidate metabolites for A. fumigatus"
author: "Jeroen Gilis"
date: "7 October 2024"
output: 
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

```{r, message=FALSE, warning=FALSE}
library(stringr)
library(openxlsx)
library(RCurl)
library(kableExtra)
```

## Helpers

```{r}
source("./Rutilities.R")
```

# Obtain GEM

```{r}
# Note: excels should be closed when reading them in R
fumigatus_gem <- read_excel_allsheets("../data/BaseGEMs/Fumigatus_GEM.xlsx")
```

# Inspect

```{r}
colnames(fumigatus_gem$METS)
```

```{r}
sum(is.na(fumigatus_gem$METS$REPLACEMENT.ID)) # they used this in their analysis
# If available, they used the KEGG+compartment
# If not, then they used the CHEBI+compartment
# If that is also not available, they use "Corecocluster"+compartment
# Finally, they have a replacement ID for e.g. biomass, lipid, ...
```

```{r}
table(grepl("C(\\d{5})\\[",fumigatus_gem$METS$REPLACEMENT.ID)) # KEGG
table(grepl("CHEBI",fumigatus_gem$METS$REPLACEMENT.ID)) # CHEBI
table(grepl("Cluster",fumigatus_gem$METS$REPLACEMENT.ID)) # coreco cluster
table(grepl("C(\\d{3})\\[",fumigatus_gem$METS$REPLACEMENT.ID)) # CoreCo/C[digits]
```

```{r}
fumigatus_gem$METS[!grepl("C(\\d{5})\\[",fumigatus_gem$METS$REPLACEMENT.ID)&
                                    !grepl("CHEBI",fumigatus_gem$METS$REPLACEMENT.ID) &
                                    !grepl("Cluster",fumigatus_gem$METS$REPLACEMENT.ID) &
                                    !grepl("C(\\d{3})\\[",fumigatus_gem$METS$REPLACEMENT.ID),
                   c("NAME","MIRIAM","REPLACEMENT.ID")]
# these last 305 compounds are mainly tagged CoreCo or A.fumigatus.central.metabolism 
# in miriam. For instance, A.fumigatus.central.metabolism/CERB1A seems to refer to
# cerberin1 comp. A (not sure if comp. stands for compound, component, or compartiment).
```

- kegg.compound for many (see also REPLACEMENT.ID), except biomass/lipid/...
- InChi, bigg, biocyc, chebi, hmdb, metanet, pubchem, and seed.compound for many


```{r}
table(grepl("kegg.compound",fumigatus_gem$METS$MIRIAM))
table(grepl("chebi/CHEBI",fumigatus_gem$METS$MIRIAM))

table(grepl("kegg",fumigatus_gem$METS$MIRIAM),
      grepl("chebi",fumigatus_gem$METS$MIRIAM))
```

# Wrangle

```{r}
# Get KEGG column
fumigatus_gem$METS$KEGG <- fumigatus_gem$METS$MIRIAM
fumigatus_gem$METS$KEGG[!grepl("kegg",fumigatus_gem$METS$KEGG)] <- NA
fumigatus_gem$METS$KEGG <- gsub("\\;.*","",gsub(".*kegg.compound/(.+)", "\\1", fumigatus_gem$METS$KEGG))

# Get chebi column
fumigatus_gem$METS$CHEBI <- fumigatus_gem$METS$MIRIAM
fumigatus_gem$METS$CHEBI[!grepl("chebi",fumigatus_gem$METS$CHEBI)] <- NA
fumigatus_gem$METS$CHEBI <- gsub("\\;.*","",gsub(".*chebi/(.+)", "\\1", fumigatus_gem$METS$CHEBI))

table(!is.na(fumigatus_gem$METS$KEGG),
      !is.na(fumigatus_gem$METS$CHEBI))
```

# Get chebi based on KEGG

Get chebi based on KEGG. Perhaps best to do this for all KEGG IDs, because they 
can be linked to  multiple CHEBIs, and we currently only have 1 per KEGG.

## Pubchem compound

```{r}
kegg_query <- fumigatus_gem$METS$KEGG[which(!is.na(fumigatus_gem$METS$KEGG))]
kegg_query <- unique(kegg_query) # only search each query once
system.time({
CHEBI_comp <- linkMetaboliteIdentifiers(query = kegg_query, 
                                       target = "^CHEBI", 
                                       reader = "read.csv",
                                       database = "pubchem_compound",
                                       verbose = TRUE)
}) 
# 1244.88s for 1584 queries (0.79s/query)

table(!is.na(CHEBI_comp$target)) # 2/3 found back
CHEBI_comp <- CHEBI_comp[!is.na(CHEBI_comp$target),]
saveRDS(CHEBI_comp,"../data/intermediate/fumigatus_KEGG_CHEBI_pubchemcomp.rds")
```

Some of the new CHEBIs differ from the original CHEBIs:

```{r}
CHEBI_comp <- readRDS("../data/intermediate/fumigatus_KEGG_CHEBI_pubchemcomp.rds")
fumigatus_gem$METS <- consolidateIDs(data = fumigatus_gem$METS,
                                     queryID = "KEGG",
                                     originalID = "CHEBI",
                                     link = CHEBI_comp,
                                     strategy = "union",
                                     keepOriginal = TRUE)
```

```{r}
fumigatus_gem$METS[c(1,2,15,63), 
                   c("KEGG","CHEBI_original","CHEBI_new","CHEBI_consolidated")]
```

```{r}
table(grepl("kegg",fumigatus_gem$METS$MIRIAM),
      grepl("chebi",fumigatus_gem$METS$MIRIAM))

table(grepl("kegg",fumigatus_gem$METS$MIRIAM),
      !is.na(fumigatus_gem$METS$CHEBI_consolidated)) 
# added 745 links between KEGG and CHEBI!
# But, still failed to add 809 CHEBI for which we have the KEGG.
# ! these might be substances, not compounds -> next step
```

```{r}
fumigatus_gem$METS$CHEBI <- fumigatus_gem$METS$CHEBI_consolidated
fumigatus_gem$METS$CHEBI_original <- NULL
fumigatus_gem$METS$CHEBI_new <- NULL
fumigatus_gem$METS$CHEBI_consolidated <- NULL
```

## Pubchem substance

Here we attempted to search all KEGG IDs in pubchem_substance to identify
links with CHEBI. However, this resulted in only 9 succesful queries, and all of
these were already obtained from pubchem_compound

```
kegg_query <- fumigatus_gem$METS$KEGG[which(!is.na(fumigatus_gem$METS$KEGG))]
kegg_query <- unique(kegg_query) # only search each query once
system.time({
CHEBI_sub <- linkMetaboliteIdentifiers(query = kegg_query, 
                                       target = "^CHEBI", 
                                       reader = "read.csv",
                                       database = "pubchem_substance",
                                       verbose = TRUE)
}) 
# 1244.88s for 1584 queries (0.79s/query)
```

```
(CHEBI_sub <- CHEBI_sub[!is.na(CHEBI_sub$target),])
# We are only able to add 9 CHEBIs based on this
```

```
fumigatus_gem$METS <- consolidateIDs(data = fumigatus_gem$METS,
                                     queryID = "KEGG",
                                     originalID = "CHEBI",
                                     link = CHEBI_sub,
                                     strategy = "union",
                                     keepOriginal = TRUE)

fumigatus_gem$METS[!is.na(fumigatus_gem$METS$CHEBI_new),
                   c("KEGG","CHEBI_original","CHEBI_new","CHEBI_consolidated")]
```

# Add CHEBI based on replacement ID

Nearly all instances for which we have a CHEBI but not KEGG are CHEBI:52320

```{r}
table(!is.na(fumigatus_gem$METS$KEGG),
      !is.na(fumigatus_gem$METS$CHEBI))

table(fumigatus_gem$METS$CHEBI[which(is.na(fumigatus_gem$METS$KEGG) &
                                       !is.na(fumigatus_gem$METS$CHEBI))])
```

CHEBI:52320 is "ergosteryl ester". Interestingly, the REPLACEMENT.ID seems a
more accurate CHEBI of these compounds. It seems that we can safely replace the 
CHEBI column with the REPLACEMENT.ID column for these instances:

```{r}
head(fumigatus_gem$METS[which(fumigatus_gem$METS$CHEBI == "CHEBI:52320"),
                        c("NAME", "REPLACEMENT.ID")])
```

```{r}
target <- which(fumigatus_gem$METS$CHEBI == "CHEBI:52320")
hlp <- fumigatus_gem$METS$REPLACEMENT.ID[which(fumigatus_gem$METS$CHEBI == "CHEBI:52320")]

hlp <- gsub(pattern="\\[.*","",hlp)
hlp <- gsub(pattern="CHEBI","CHEBI:",hlp)
fumigatus_gem$METS$CHEBI[target] <- hlp

fumigatus_gem$METS[target[1:5],c("CHEBI","REPLACEMENT.ID")]
```

# Get KEGG based on chebi

## pubchem compound

Now, try to add KEGGs for the IDs for which we have CHEBI.

Previously, I searched all KEGGs, because they can be associated with multiple 
CHEBIs. As each CHEBI should only have one KEGG, I here only search for the 
missing ones. Searching for all could be relevant if we need to find mistakes in
the original annotation.

```{r}
chebi_query <- fumigatus_gem$METS$CHEBI[
  which(is.na(fumigatus_gem$METS$KEGG) &
          !is.na(fumigatus_gem$METS$CHEBI))]
chebi_query <- unique(unlist(str_split(chebi_query,pattern = "/")))

KEGG_comp <- linkMetaboliteIdentifiers(query = chebi_query, 
                                       target = "^C(\\d{5})\\b", 
                                       reader = "read.csv",
                                       database = "pubchem_compound",
                                       verbose = TRUE)
```

```{r}
fumigatus_gem$METS <- consolidateIDs(data = fumigatus_gem$METS,
                                     queryID = "CHEBI",
                                     originalID = "KEGG",
                                     link = KEGG_comp,
                                     strategy = "union",
                                     keepOriginal = TRUE)

fumigatus_gem$METS[!is.na(fumigatus_gem$METS$KEGG_new),
                   c("CHEBI","KEGG_original","KEGG_new","KEGG_consolidated")]
```

```{r}
fumigatus_gem$METS$KEGG <- fumigatus_gem$METS$KEGG_consolidated
fumigatus_gem$METS$KEGG_original <- NULL
fumigatus_gem$METS$KEGG_new <- NULL
fumigatus_gem$METS$KEGG_consolidated <- NULL
```

```{r}
table(grepl("kegg",fumigatus_gem$METS$MIRIAM),
      grepl("chebi",fumigatus_gem$METS$MIRIAM))

table(!is.na(fumigatus_gem$METS$KEGG),
      !is.na(fumigatus_gem$METS$CHEBI)) # Added 24 links
```

## pubchem substance

Provides no additional hits!

```{r}
chebi_query <- fumigatus_gem$METS$CHEBI[
  which(is.na(fumigatus_gem$METS$KEGG) &
          !is.na(fumigatus_gem$METS$CHEBI))]
chebi_query <- unique(unlist(str_split(chebi_query,pattern = "/")))

KEGG_sub <- linkMetaboliteIdentifiers(query = chebi_query, 
                                      target = "^C(\\d{5})\\b", 
                                      reader = "read.csv",
                                      database = "pubchem_substance",
                                      verbose = TRUE)
```

# Curate names

We've done what we could linking KEGG to CHEBI and vice versa.
Now, we will need to rely on the names.
The names are messy, and we will need to wrangle them to make sense.
In addition, we may also want to exclude some names from the GEM altogether, 
i.e, because they are purely artificial or dead-end metabolites.

Here, we aim to define some rule to mask those compounds.

```{r}
query_name <- fumigatus_gem$METS$NAME[is.na(fumigatus_gem$METS$KEGG)|
                                        is.na(fumigatus_gem$METS$CHEBI)]
length(query_name)
length(unique(query_name))

# trim localization
query_name <- gsub(pattern=" \\[.*","",query_name)
length(unique(query_name))

# decapitalize
query_name <- tolower(query_name)
length(unique(query_name))

# only uniques
query_name <- unique(query_name)
```

There are 743 metabolites for which we don't have KEGG or CHEBI, or neither.
We will not care about the follwoing, because they are artifcial metabolites 
only used for GEM simulations:

```{r}
mask_GEM <- c("carbohydrate","lipid","protein","deoxyribonucleic acids",
              "ribonucleic acids", "cofactor", "biomass")

query_name <- query_name[!query_name %in% mask_GEM]
length(query_name)
```

Next, we notice that many of these metabolites involve ACP 
(acyl-carrier protein). We wsh to keep these metabolites, because acp is a 
cofactor that is being recycled. For now, though, we will set these aside.

Comment Eduard: Maybe you can map the other models also to KEGG and that way 
map them? If needed anyway, good chance that beta-oxidation is well/similarly 
described in all of these models, so that no new clusters are introduced with 
the second and third template model.

```{r}
mask_acp <- grepl("\\[acp\\]", query_name) | grepl("\\[acyl-carrier protein\\]", query_name)
head(query_name[mask_acp])

query_name <- query_name[!mask_acp]
length(query_name)
```

Then, we have many entries with (glcn) or (glcnac),
where (glcn) is glucosamine and glcnac acetylglucosamine

```{r}
mask_glcn_glcnac <- grepl("(glcn)", query_name) | grepl("(glcnac)", query_name)
head(query_name[mask_glcn_glcnac])

query_name <- query_name[!mask_glcn_glcnac]
length(query_name)
```

We have several entries involving the unspecific term "protein":

```{r}
mask_protein <- grepl("protein", query_name)
head(query_name[mask_protein])

query_name <- query_name[!mask_protein]
length(query_name)
```

So more masking. But we should discuss this again.

```{r}
query_name[grep("aromatic",query_name)]
query_name[grep("rna",query_name)]
query_name[grep("dna",query_name)]


mask_more <- grepl("aromatic", query_name) | 
             grepl("rna", query_name) |
             grepl("dna", query_name)
query_name <- query_name[!mask_more]
length(query_name)
```

# Get KEGG based on name

Everything failed! ote that even though there were 50% unknown queries
(which may indicate we can improve name curation), the other 50% simply did
not have a KEGG in pubchem. This may indicate that we are using the same 
annotation strategy as the original authors.

```
NAMEtoKEGG_pubcomp <- linkMetaboliteIdentifiers(query = query_name, 
                                 target = "KEGG",
                                 reader = "read.csv",
                                 database = "pubchem_compound",
                                 verbose = TRUE)
# 579/579 fails: 282 unknown, 297 without KEGG
# -> might indicate that the authors of the GEM tried exactly this!

NAMEtoKEGG_pubsub <- linkMetaboliteIdentifiers(query = query_name, 
                                 target = "KEGG",
                                 reader = "read.csv",
                                 database = "pubchem_substance",
                                 verbose = TRUE)
# 579/579 fails: 202 unknown, 377 without KEGG
```

# Get CHEBI based on name

```{r}
NAMEtoCHEBI_pubcomp <- linkMetaboliteIdentifiers(query = query_name, 
                                 target = "CHEBI",
                                 reader = "read.csv",
                                 database = "pubchem_compound",
                                 verbose = TRUE)
# 342/579 fails: 282 unknown, 60 without chebi
saveRDS(NAMEtoCHEBI_pubcomp,"../data/intermediate/fumigatus_NAME_CHEBI_pubchemcomp.rds")

NAMEtoCHEBI_pubsub <- linkMetaboliteIdentifiers(query = query_name, 
                                 target = "CHEBI",
                                 reader = "read.csv",
                                 database = "pubchem_substance",
                                 verbose = TRUE)
# 328/579 fails: 202 unknown, 126 without chebi
saveRDS(NAMEtoCHEBI_pubsub,"../data/intermediate/fumigatus_NAME_CHEBI_pubchemsub.rds")
```

```{r}
NAMEtoCHEBI_pubcomp <- readRDS("../data/intermediate/fumigatus_NAME_CHEBI_pubchemcomp.rds")
NAMEtoCHEBI_pubsub <- readRDS("../data/intermediate/fumigatus_NAME_CHEBI_pubchemsub.rds")
```

```{r}
res <- data.frame(query = NAMEtoCHEBI_pubcomp$query,
                  compound=NAMEtoCHEBI_pubcomp$target,
                  compound_error=NAMEtoCHEBI_pubcomp$error,
                  substance=NAMEtoCHEBI_pubsub$target, 
                  substance_error=NAMEtoCHEBI_pubsub$error)

res$both <- unlist(sapply(seq_along(res$substance),function(x){
  if(is.na(res$substance[x])){
    return(res$compound[x])
  } else if(is.na(res$compound[x])){
    return(res$substance[x])
  } else{
    combined <- union(unlist(str_split(res$substance[x], 
                                       pattern = "/")),
                      unlist(str_split(res$compound[x], 
                                       pattern = "/")))
    return(paste0(combined,collapse = "/"))
  }
}))
table(!is.na(res$both))
```

```{r}
View(res[!is.na(res$both),c(1,2,4,6)])
```

We were able to annotate 300/668 metabolites based on name. Not too bad.
Of course, we are not 100% sure if all of these are all correct. For instance,
"hx" and "aldehyde" are bad, chebi for compound is correct, but for substance it 
isn't!!

1. Try to categorize the names we don't find. Maybe there are still categories that we won't even care about.For others, we can try to tailor the names so that we can find them back.
2. In addition, based on the new chebi, get their kegg.
3. Would it be useful to look at the funigatus GEM and see which are dead-end metabolites? My only concern is the same metabolite may not be dead-end in another GEM, so the we'd still need the annotation right? But then maybe we can more easily get it there...

zymosterol intermediate 1c should be on pubchem and chebi: why not found?
chlorophyllide a should be on pubchem and KEGG: why not found?

```{r}
res[is.na(res$both),]
```

Many are in the form of: 

- (glcnac)2 (man)8 (pp-dol)1
- protein asparagine
- 1r, (1r,2s)
- trailin (in dna)

```{r}
query_name <- res$query[is.na(res$both)]
query_name
```

```
knitr::knit_exit()
```

```{r}
q <- "(glcn)1 (ino(acyl)-p)1 (man)3 (etn)1 (p)1"
q <- "(GlcN)1 (Ino(acyl)-P)1 (Man)2 (EtN)1 (P)1"
q <- "G00148"
url <- paste0("https://pubchem.ncbi.nlm.nih.gov/rest/pug/substance/name/",
              q,
              "/synonyms/TXT")
read.csv(url,header = FALSE)
```

# Scrap

--- 3-cyanopyridine

```{r}
niger_gem$METS[which(is.na(niger_gem$METS$MIRIAM)),][1,]
```

For this one, I have the inchi.
I can search the inchi using InChI:InChI=1S/C6H4N2/c7-4-6-2-1-3-8-5-6/h1-3,5H
in https://metacyc.org/metabolite-translation-service.shtml.

This gives me several identifiers, but not kegg.

I then look for 3-cyanopyridine and 3-Cyanopyridine in 
https://www.genome.jp/kegg/compound/ -> no hit.

I then take the molecular mass from the metacyc search, and feed that to kegg:
no hit, but kegg does have 2-Cyanopyridine and 4-Cyanopyridine, just no
3-Cyanopyridine.

But, from metacyc we also get Chebi (86556). If we query that in 
https://www.ebi.ac.uk/chebi/, we get 3-pyridinecarbonitrile, which turns out to 
be synonym for 3-Cyanopyridine.

However, again, searching this in kegg does not give a hit.

This compound seems really not exist in KEGG. So what's next? Not use KEGG at all?
But is there even a database in which all required compounds exist?

--- C00659

```{r}
fumigatus_gem$METS[1:2,]
```
  
We have the kegg compound for this, let's try to find chebi

Via KEGG, we can manually click to chebi (C00659 -> 982)

But when trying to do this more automatically in metacyc, it fails.
Kegg:C00659   ChEBI:982   InChI:InChI=1S/C6H10O4/c1-3-6(10,4(2)7)5(8)9/h10H,3H2,1-2H3,(H,8,9)   InChIkey:InChIKey=VUQLHQFKACOHNZ-UHFFFAOYSA-N   CF:C6H10O4

This is probably inevitable; I strongly suspect that the people that constructed
this GEM in the first place used Metabolite Translation Service for MetaCyc.
As such, I would return as much information as they did, not more.

I could also input C000659 in
https://www.metaboanalyst.ca/faces/upload/ConvertView.xhtml
this succeeds in returning Chebi

One more possibility would be to do this in R using MetaFetcheR. However, this
is a non-maintained github page that failed to install out of the box and that
requires installing additional things; as such, this seems far from ideal.

One other possibility is to use the pypi library bioservices
https://bioservices.readthedocs.io/en/latest/convertor_tutorial.html

This seems to work well! I could actually just obtain a Kegg-Chebi dictionary
from there and import it here!

```{r}
kegg_chebi_dict <- read.csv2("../data/intermediate//kegg_chebi_dict.csv", 
                  sep=",",
                  na.strings = "",
                  header = FALSE)
colnames(kegg_chebi_dict) <- c("KEGG","CHEBI")
kegg_chebi_dict$KEGG <- gsub(".*cpd:(.+)", "\\1", kegg_chebi_dict$KEGG)
kegg_chebi_dict$CHEBI <- gsub(".*chebi:(.+)", "\\1", kegg_chebi_dict$CHEBI)

head(kegg_chebi_dict)
```

```{r}
# update using our KEGG/CHEBI dictionary
table(fumigatus_gem$METS$KEGG %in% kegg_chebi_dict$KEGG)
table(fumigatus_gem$METS$CHEBI %in% kegg_chebi_dict$CHEBI)

# THE DICTIONARY IS INCOMPLETE...
# For instance
"C04253" %in% fumigatus_gem$METS$KEGG
"C04253" %in% kegg_chebi_dict$KEGG 
# indeed, no chebi for this kegg, also not on kegg website (could link to pubchem though)
# also no kegg-chebi link through metacyc (could link to inchi though)
# This makes sense, on KEGG the compound is called  	
# Electron-transferring flavoprotein, which is  generic term, not a real compound


# Conclusion, not all KEGGs have a chebi
# Let's see if we can still get something from it
```









