---
title: "Consolidate metabolites for A. oryzae"
author: "Jeroen Gilis"
date: "7 October 2024"
output: 
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

```{r, message=FALSE, warning=FALSE}
library(stringr)
library(openxlsx)
library(RCurl)
library(kableExtra)
```

## Helpers

```{r}
source("./Rutilities.R")
```

# Obtain GEM

```{r}
# Note: excels should be closed when reading them in R
oryzae_gem <- read_excel_allsheets("../data/BaseGEMs/Oryzae_GEM.xlsx")
```

# Inspect

```{r}
View(oryzae_gem$METS)
```

- Problematic: no formalized identifiers, only ID, NAME and REPLACEMENT.ID

Also, the NAMEs themselves are messy. E.g.:

```{r}
oryzae_gem$METS$NAME[c(1,3)] # trailing composition and trailing underscores
```

I think it's easiest to take the IDs, everything before the last underscore is
the NAME, everything behind it is eiter the compasition and the compartment or
only the compartment

```{r}
# fix names
oryzae_gem$METS$NAME <- gsub("_[^_]*$", "", oryzae_gem$METS$ID)

# extract composition
composition <- gsub("\\[.*", "", gsub('.*_', '', oryzae_gem$METS$ID))
composition[composition==""]<-NA
oryzae_gem$METS$COMPOSITION <- composition
```

# Mask names

We may want to exclude some names from the GEM altogether, 
i.e, because they are purely artificial or dead-end metabolites.

Here, we aim to define some rules to mask those compounds.

! BUT LETS FIRST NOT CARE ABOUT MASKS AND QUERY EVERYTHING AND SEE WHAT HAPPENS

```{r}
query_name <- oryzae_gem$METS$NAME
length(query_name)
length(unique(query_name))

# decapitalize
#query_name <- tolower(query_name)
#length(unique(query_name))

# only uniques
query_name <- unique(query_name)
```

```{r}
NAMEtoKEGG_pubcomp <- linkMetaboliteIdentifiers(query = query_name, 
                                 target = "^C(\\d{5})\\b",
                                 reader = "read.csv",
                                 database = "pubchem_compound",
                                 verbose = TRUE)
# 579/579 fails: 214 unknown, 297 without KEGG
saveRDS(NAMEtoKEGG_pubcomp,"../data/intermediate/oryzae_NAME_KEGG_pubcomp.rds")

NAMEtoKEGG_pubsub <- linkMetaboliteIdentifiers(query = query_name, 
                                 target = "^C(\\d{5})\\b",
                                 reader = "read.csv",
                                 database = "pubchem_substance",
                                 verbose = TRUE)
# 579/579 fails: 202 unknown, 377 without KEGG
saveRDS(NAMEtoKEGG_pubsub,"../data/intermediate/oryzae_NAME_KEGG_pubsub.rds")
```

```{r}
NAMEtoCHEBI_pubcomp <- linkMetaboliteIdentifiers(query = query_name, 
                                 target = "CHEBI",
                                 reader = "read.csv",
                                 database = "pubchem_compound",
                                 verbose = TRUE)
# 342/579 fails: 282 unknown, 60 without chebi
saveRDS(NAMEtoCHEBI_pubcomp,"../data/intermediate/oryzae_NAME_CHEBI_pubcomp.rds")

NAMEtoCHEBI_pubsub <- linkMetaboliteIdentifiers(query = query_name, 
                                 target = "CHEBI",
                                 reader = "read.csv",
                                 database = "pubchem_substance",
                                 verbose = TRUE)
# 328/579 fails: 202 unknown, 126 without chebi
saveRDS(NAMEtoCHEBI_pubsub,"../data/intermediate/oryzae_NAME_CHEBI_pubsub.rds")
```

```{r}
NAMEtoKEGG_pubcomp <- readRDS("../data/intermediate/oryzae_NAME_KEGG_pubcomp.rds")
NAMEtoKEGG_pubsub <- readRDS("../data/intermediate/oryzae_NAME_KEGG_pubsub.rds")
NAMEtoCHEBI_pubcomp <- readRDS("../data/intermediate/oryzae_NAME_CHEBI_pubcomp.rds")
NAMEtoCHEBI_pubsub <- readRDS("../data/intermediate/oryzae_NAME_CHEBI_pubsub.rds")
```

```{r}
table(NAMEtoKEGG_pubcomp$error) #why no chebi? not no KEGG or no //bC...
table(NAMEtoKEGG_pubsub$error)
table(NAMEtoCHEBI_pubcomp$error)
table(NAMEtoCHEBI_pubsub$error)
```






There are 743 metabolites for which we don't have KEGG or CHEBI, or neither.
We will not care about the follwoing, because they are artifcial metabolites 
only used for GEM simulations:

```
mask_GEM <- c("carbohydrate","lipid","protein","deoxyribonucleic acids",
              "ribonucleic acids", "cofactor", "biomass")

query_name <- query_name[!query_name %in% mask_GEM]
length(query_name)
```

Next, we notice that many of these metabolites involve ACP 
(acyl-carrier protein). We wsh to keep these metabolites, because acp is a 
cofactor that is being recycled. For now, though, we will set these aside.

Comment Eduard: Maybe you can map the other models also to KEGG and that way 
map them? If needed anyway, good chance that beta-oxidation is well/similarly 
described in all of these models, so that no new clusters are introduced with 
the second and third template model.

```
mask_acp <- grepl("\\[acp\\]", query_name) | grepl("\\[acyl-carrier protein\\]", query_name)
head(query_name[mask_acp])

query_name <- query_name[!mask_acp]
length(query_name)
```

Then, we have many entries with (glcn) or (glcnac),
where (glcn) is glucosamine and glcnac acetylglucosamine

```
mask_glcn_glcnac <- grepl("(glcn)", query_name) | grepl("(glcnac)", query_name)
head(query_name[mask_glcn_glcnac])

query_name <- query_name[!mask_glcn_glcnac]
length(query_name)
```

We have several entries involving the unspecific term "protein":

```
mask_protein <- grepl("protein", query_name)
head(query_name[mask_protein])

query_name <- query_name[!mask_protein]
length(query_name)
```

So more masking. But we should discuss this again.

```
query_name[grep("aromatic",query_name)]
query_name[grep("rna",query_name)]
query_name[grep("dna",query_name)]


mask_more <- grepl("aromatic", query_name) | 
             grepl("rna", query_name) |
             grepl("dna", query_name)
query_name <- query_name[!mask_more]
length(query_name)
```

