---
title: "Consolidate metabolites for A. oryzae"
author: "Jeroen Gilis"
date: "6 January 2025"
output: 
    html_document:
      code_download: true
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

This script aims to obtain as many KEGG metabolite identifiers for the 
metabolites in the oryzae GEM. KEGG identifiers are obtained from

1. The information in the oryzae GEM itself
2. A KEGG-CHEBI dictionary obtained from the bioservices python database
3. Links present in the niger GEM
4. Links present in the fumigatus GEM
5. Manually

The result of this script is a .xlsx file with links between the original 
metabolite identifiers and corresponding KEGG identifiers.

# Libraries

```{r, message=FALSE, warning=FALSE}
library(stringr)
library(openxlsx)
library(RCurl)
library(KEGGREST)
```

# Helpers

```{r}
source("./helperFunctions/Rutilities.R")
```

# Load GEMs

```{r}
# Note: excel files should be closed when reading them in R
Aor_draft <- read_excel_allsheets("../data/intermediate/Aor_draft_comp.xlsx")
Ani_draft <- read_excel_allsheets("../data/intermediate/Ani_draft.xlsx")
```

# Wrangle

```{r}
# Get RepName column
Ani_draft$METS$RepName <- gsub("\\[.\\]","",Ani_draft$METS$`REPLACEMENT ID`)
```

```{r}
# Get RepName column
Aor_draft$METS$RepName <- NA

# Get CHEBI column
Aor_draft$METS$CHEBI <- Aor_draft$METS$MIRIAM
Aor_draft$METS$CHEBI[!grepl("chebi",Aor_draft$METS$CHEBI)] <- NA
Aor_draft$METS$CHEBI <- gsub("chebi:","",Aor_draft$METS$CHEBI)
Aor_draft$METS$CHEBI <- gsub("/",":",Aor_draft$METS$CHEBI)

# Get KEGG column
Aor_draft$METS$KEGG <- Aor_draft$METS$MIRIAM
Aor_draft$METS$KEGG[!grepl("kegg",Aor_draft$METS$KEGG)] <- NA
Aor_draft$METS$KEGG <- gsub("kegg.compound/","",Aor_draft$METS$KEGG)
```

The replacement IDs for both oryzae and niger can be improved. If the
compartment is not cytosol, a trailing compartment name was added. Fix this:

```{r}
Aor_draft$METS$`REPLACEMENT ID` <- gsub("m\\[m\\]", "\\[m\\]", Aor_draft$METS$`REPLACEMENT ID`)
Aor_draft$METS$`REPLACEMENT ID` <- gsub("e\\[e\\]", "\\[e\\]", Aor_draft$METS$`REPLACEMENT ID`)
Aor_draft$METS$`REPLACEMENT ID` <- gsub("p\\[p\\]", "\\[p\\]", Aor_draft$METS$`REPLACEMENT ID`)
```

```{r}
sum(!is.na(Aor_draft$METS$KEGG)) # very few keggs in Aor
sum(Aor_draft$METS$KEGG %in% Ani_draft$METS$KEGG[!is.na(Ani_draft$METS$KEGG)])
sum(Aor_draft$METS$KEGG %in% Ani_draft$METS$KEGG_self[!is.na(Ani_draft$METS$KEGG_self)])

sum(!is.na(Aor_draft$METS$CHEBI))
sum(Aor_draft$METS$CHEBI %in% Ani_draft$METS$CHEBI[!is.na(Ani_draft$METS$CHEBI)])
# so the 54 you lose here are by definition unique to Aor, because while the 
# name is existing in Ani, it is not existing in that compartment

sum(tolower(Aor_draft$METS$NAME) %in% tolower(Ani_draft$METS$NAME))
sum(tolower(Aor_draft$METS$ID) %in% tolower(Ani_draft$METS$ID)) 
# so the 54 you lose here are by definition unique to Aor, because while the 
# name is existing in Ani, it is not existing in that compartment

sum(Aor_draft$METS$`REPLACEMENT ID` %in% Ani_draft$METS$`REPLACEMENT ID`)
sum(tolower(Aor_draft$METS$`REPLACEMENT ID`) %in% tolower(Ani_draft$METS$`REPLACEMENT ID`))
# the main question is: do we trust matches based on replacement id?
# -> match on proper IDs first, then manually inspect if we can trust the
# remaining matches based on replacement_id.
```

# KEGG available

For a few metabolites, we already have a KEGG identifier available. Keep those.

```{r}
target <- Aor_draft$METS[is.na(Aor_draft$METS$KEGG),]
```

# Obtain KEGG directly thorugh KEGG-CHEBI dictionary

```{r}
kegg_chebi_dict <- read.csv2("../data/intermediate/kegg_chebi_dict.csv", header=FALSE, sep=",")

# wrangle
colnames(kegg_chebi_dict) <- c("KEGG","CHEBI")
kegg_chebi_dict$KEGG <- gsub("cpd:","",kegg_chebi_dict$KEGG)
kegg_chebi_dict$CHEBI <- gsub("chebi:","CHEBI:",kegg_chebi_dict$CHEBI)

target$KEGG <- kegg_chebi_dict$KEGG[match(target$CHEBI,kegg_chebi_dict$CHEBI)]

Aor_draft$METS$KEGG[match(target$`REPLACEMENT ID`,Aor_draft$METS$`REPLACEMENT ID`)] <- target$KEGG

target <- Aor_draft$METS[is.na(Aor_draft$METS$KEGG),]
```

# Based on niger GEM

```{r}
Ani_mets <- Ani_draft$METS

# Get KEGG column
Ani_mets$KEGG <- Ani_mets$MIRIAM
Ani_mets$KEGG[!grepl("kegg",Ani_mets$KEGG)] <- NA
Ani_mets$KEGG <- gsub("\\;.*","",gsub(".*kegg.compound/(.+)", "\\1", Ani_mets$KEGG))

# Get chebi column
Ani_mets$CHEBI <- Ani_mets$MIRIAM
Ani_mets$CHEBI[!grepl("chebi",Ani_mets$CHEBI)] <- NA
Ani_mets$CHEBI <- gsub("\\;.*","",gsub(".*chebi/(.+)", "\\1", Ani_mets$CHEBI))

# Get RepName column
Ani_mets$RepName <- gsub("\\[.\\]","",Ani_mets$`REPLACEMENT ID`)
```

```{r}
# Get RepName column
target$RepName <- gsub("\\[.\\]","",target$`REPLACEMENT ID`)

# Through CHEBI
target$KEGG <- Ani_mets$KEGG[match(target$CHEBI,Ani_mets$CHEBI)]
Aor_draft$METS$KEGG[match(target$`REPLACEMENT ID`,Aor_draft$METS$`REPLACEMENT ID`)] <- target$KEGG
target <- Aor_draft$METS[is.na(Aor_draft$METS$KEGG),]

# Through name
target$KEGG <- Ani_mets$KEGG[match(tolower(target$NAME),
                                   tolower(Ani_mets$NAME))]
Aor_draft$METS$KEGG[match(target$`REPLACEMENT ID`,Aor_draft$METS$`REPLACEMENT ID`)] <- target$KEGG
target <- Aor_draft$METS[is.na(Aor_draft$METS$KEGG),]

# Through replacement ID (without compartment identfier, which we called RepName)
target$KEGG <- Ani_mets$KEGG[match(target$RepName, Ani_mets$RepName)]
Aor_draft$METS$KEGG[match(target$`REPLACEMENT ID`,Aor_draft$METS$`REPLACEMENT ID`)] <- target$KEGG
target <- Aor_draft$METS[is.na(Aor_draft$METS$KEGG),]
```

# Based on KEGGREST

```{r}
name_query <- tolower(target$NAME)
name_query <- gsub(",", " ", name_query) # it works to replace commas for spaces
```

```{r}
addKEGG <- lapply(name_query, function(x){
  #print(x)
  query <- keggFind(database = "compound", query = x)
  return(cbind(x,
               paste0(gsub("cpd:", "", names(query)),collapse="/"),
               paste0(unname(query),collapse="/")))
})

saveRDS(addKEGG,"../data/intermediate/oryzae_NAMEtoKEGG_keggrest.rds")

# Comment above and and uncomment below to save time
# addKEGG <- readRDS("../data/intermediate/oryzae_NAMEtoKEGG_keggrest.rds")
```

```{r}
target$KEGG <- sapply(addKEGG, function(x){x[2]})
target$KEGG[target$KEGG==""] <- NA

target$AllNames <- sapply(addKEGG, function(x){x[3]})
target$AllNames[target$AllNames==""] <- NA

table(!is.na(target$KEGG))
table(!is.na(target$KEGG) &
                 lengths(strsplit(target$KEGG,"/")) == 1) # unambiguous
table(!is.na(target$KEGG) &
                 lengths(strsplit(target$KEGG,"/")) != 1) # ambiguous
```

```{r}
# metabolites that we could assign a unique KEGG id based on its name
# -> use directly
index <- which(!is.na(target$KEGG) &
                 lengths(strsplit(target$KEGG,"/")) == 1)
target$KEGG_final <- NA
target$KEGG_final[index] <- target$KEGG[index]
```

```{r}
# metabolites that we could assign a KEGG id, but not unambiguously
# -> try to identify the correct ID if possible
index <- which(!is.na(target$KEGG) &
                 lengths(strsplit(target$KEGG,"/")) != 1)
hlp <- target[index,]
hlp <- hlp[!duplicated(hlp$NAME),]

curated <- list(c("4-hydroxybenzoate", "C00156"),
                c("pimelate", "C02656"),
                c("8-amino-7-oxononanoate", "C01092"),
                c("averantin", "C20499"),
                c("carbamate", "C01563"),
                c("chitobiose", "C01674"),
                c("cinnamate","C00423"), # this is actually trans-cinnamate, but thats whats used in the GEM
                c("citramalate", "C00815"),
                c("D-aspartate", "C00402"),
                c("dihydrosterigmatocystin", "C20445"),
                c("geranylgeranyl diphosphate", "C00353"),
                c("HX", "C00462"),
                c("indole-3-acetate","C00954"),
                c("itaconate", "C00490"),
                c("L-ribulose", "C00508"),
                c("mesaconate", "C01732"),
                c("maltotriose", "C01835"),
                c("1D-myo-inositol 1,4-bisphosphate", "C01220"),
                c("4alpha-methylzymosterol", "C05103"),
                c("naringenin", "C00509"),
                c("norsolorinic acid","C20452"),
                c("3,4-dihydroxybenzoate", "C00230"),
                c("RX", "C01322"),
                c("tagatose 6-phosphate","C01097"), # this is D-tagatose-6P, for L- its C06312
                c("tagatose", "C00795"), # This is for D-Tagatose, for L- its C21523
                c("thiamin diphosphate", "C00068"),
                c("urocanate", "C00785"),
                c("versiconal", "C20507"),
                c("chitobiose", "C01674"),
                c("itaconate", "C00490"),
                c("L-ribulose", "C00508"),
                c("heme a", "C15670"),
                c("L-2-aminoadipate", "C00918"),
                c("myristate", "C06424"),
                c("hexadecenoate", "C13792/C13946"), # depending on location double bond
                c("hexadecenoyl-CoA", "C05272/C21072/C21943"), # depending on location double bond
                c("linoleate", "C01595"),
                c("octadecatrienoic acid", "C06426/C06427/C08315/C08364"), # depending on location double bond
                c("butyrate", "C00246"),
                c("octanoate", "C06423"),
                c("cyanate", "C01417"),
                c("hydrogen", "C00080"), # When inspecting the reactions involving hydrogen, it seems like H+ is intended
                c("(R)-lactate", "C00256"),
                c("lysophosphatidylcholine", "C04230/C04233"), # 1- vs 2-lyso
                c("phosphate", "C00009"),
                c("diphosphate", "C00013"),
                c("propionate", "C00163"),
                c("N-carbamoyl-?-alanine", "C02642"), # \\? is beta, a synonym is 3-Ureidopropionate
                c("cellulose", "C00760"),
                c("N-acetyl-L-glutamate", "C00624"),
                c("urate", "C00366"))

hlp[!hlp$NAME %in% sapply(curated, function(element){element[1]}),c("NAME","KEGG")] # got all of them
```

```{r}
for (element in curated) {
  target$KEGG_final[which(target$NAME==element[1])] <- element[2]
}

target$KEGG <- NULL
table(!is.na(target$KEGG_final)) # So only 79 metabolites without KEGG left
```

```{r}
Aor_draft$METS$KEGG[match(target$`REPLACEMENT ID`,Aor_draft$METS$`REPLACEMENT ID`)] <- target$KEGG_final
target <- Aor_draft$METS[is.na(Aor_draft$METS$KEGG),]
```

# Manually

```{r}
target <- target[is.na(target$KEGG),]
target[!duplicated(target$NAME),]
```

```{r}
curated <- list(c("1,8-DHN", NA), # dihydroxynapthalene, KEGG only knows 1,2/1,4/2,3 (C03012/C14785/C16213)
                c("trans-4-coumarate", "C00811"), # kegg: 4-coumarate/trans-4-hydroxycinnamate
                c("5'-S-methyl-5'-thioadenosine", "C00170"), # kegg: 5'methylthioadenosine/S-methyl-5-thioadenosine
                c("5-amino-6-ribityluracil", "C04732"), # doublecheck, could be 5-amino-6-(D-ribitylamino)uracil
                c("acyl CoAs", "C00040"), # kegg: Acyl-CoA
                c("1-aminocyclopropanecarboxylate", "C01234"), # kegg: 1-aminocyclopropane-1-carboxylate
                c("L-gamma-glutamyl-L-alanine", NA),
                c("alpha-amylase", NA),
                c("acylglycerone phosphates", NA),
                c("biomass", NA),
                c("laurate", "C02679"),
                c("lauroyl-[acp]", "C05223"),
                c("dodecanoyl-carnitine", NA), # C02301 is O-acylcarnitine, C02571 is O-acetylcarnitine
                c("myristoyl-carnitine", NA), # C02301 is O-acylcarnitine, C02571 is O-acetylcarnitine
                c("palmitoyl-carnitine", "C02990"), #	L-Palmitoylcarnitine
                c("hexadecadienoate", "C21980"), # (6Z,9Z)-Hexadecadienoic acid
                c("hexadecadienoyl-[acp]", NA),
                c("stearoyl-carnitine", NA),
                c("octadecatrienoyl-[acp]", "C16161"),
                c("trans-but-2-enoyl-[acp]", "C04246"),
                c("2,5-diamino-6-hydroxy-4-(5'-phosphoribosylamino)-pyrimidine", NA),
                c("digalactosyl diglyceride", NA),
                c("limited dextrin-[debranched glycogen phosphorylase]", NA),
                c("diacylglycerol pyrophosphate", NA),
                c("2-deoxyribose 1-phosphate", "C00672"), # 2-Deoxy-D-ribose 1-phosphate
                c("free fatty acids", "C00162"), # kegg for Fatty acid
                c("galactocyl ceremide", NA),
                c("glycoaldehyde", "C00266"),
                c("beta-D-glucosiduronic acids", NA),
                c("limited dextrin-[glycogen phosphorylase]", NA),
                c("5-hydroxyaverantin", "C20501"),
                c("1-(indol-3-yl)glycerol 3-phosphate", "C03506"),
                c("intermediate methylzymosterol II", NA),
                c("intermediate zymosterol II", NA),
                c("(S)-3-(5-oxo-4,5-dihydro-3H-imidazol-4-yl)propanoate", "C03680"), # assuming this is 4,5-Dihydro-4-oxo-5-imidazolepropanoate
                c("intermediate methylzymosterol I", NA),
                c("inositol-mannose-P-inositol-P-D-ceramide", NA),
                c("mannose-inositol-P-ceramide", NA),
                c("oxidonitrate", "C00533"), # Nitric oxide, NO^1-
                c("(R)-3-methyl-2-oxobutanoate", "C00141"), # 3-Methyl-2-oxobutanoate
                c("2-keto-3-methyl-valerate", "C06008"), # (R)-2-Oxo-3-methylpentanoate
                c("phosphatidylamine", NA),
                c("phosphatidylcholines", NA),
                c("phytosphingosine 1-phosphate", NA),
                c("precorrins", NA), # there are many in kegg
                c("protein for biomass and amylase", NA),
                c("ribonucleic acids", NA),
                c("ceramide phosphocholines", NA),
                c("galactosylceramide sulfates", NA),
                c("threo-3-methyl-L-aspartate", "C03618"), # L-threo-3-Methylaspartate 	
                c("vestine", NA),
                c("2-methyl-2-hydroxybutyryl-CoA", NA),
                c("inositol phosphorylceramide", NA),
                c("intermediate zymosterol I", NA),
                c("3-demethylubiquinone-8", "C05815"), # 5-hydroxy-6-methoxy-3-methyl-2-octaprenyl-1,4-benzoquinone
                c("acetohydroxybutyrate", "C00659"), # 2-Aceto-2-hydroxybutanoate
                c("trans-dodec-2-enoyl-CoA", "C03221"), # 2-trans-Dodecenoyl-CoA/(2E)-Dodec-2-enoyl-CoA
                c("(R)-3-hydroxy-3-methyl-2-oxobutanoate", "C04181"), # 3-Hydroxy-3-methyl-2-oxobutanoate
                c("2-(3-amino-3-carboxypropyl)-L-histidine", "C04441"),
                c("2-keto-myo-inositol", "C00691/C17577"),
                c("lysophosphatidylamine",NA),
                c("lysophosphatidyl-N-dimethylethanolamine", NA),
                c("lysophosphatidylglycerol", NA),
                c("lysophosphatidyl-N-methylethanolamine", NA),
                c("lysophosphatidylserine", NA),
                c("melanins", NA),
                c("cysteine-[enzyme]", NA),
                c("S-sulfanylcysteine-[enzyme]", NA),
                c("acyl-[acp] (for phospholipids)", NA))

target[!target$NAME %in% sapply(curated, function(element){element[1]}),c("NAME","KEGG")]
```

```{r}
for (element in curated) {
  target$KEGG_final[which(target$NAME==element[1])] <- element[2]
}

table(!is.na(target$KEGG_final)) # So only 45 metabolites without KEGG left
```

```{r}
Aor_draft$METS$KEGG[match(target$`REPLACEMENT ID`,Aor_draft$METS$`REPLACEMENT ID`)] <- target$KEGG_final
target <- Aor_draft$METS[is.na(Aor_draft$METS$KEGG),]
```

```{r}
table(!is.na(Aor_draft$METS$KEGG))
```
   
## Handle hydrogen/protons

```{r}
Aor_draft$METS[grepl("hydrogen\\[", Aor_draft$METS$ID) |
                 grepl("H\\(", Aor_draft$METS$ID),c("ID", "NAME","KEGG")]
```

If we continue this conversion process, there will no longer be a distinction between
hydrogen[c] and H(+) (energy metabolism)[c], which is bad. As such, I will update
the KEGG_ID for H(+) (energy metabolism)[c] from C00080 (which will become C00080[c])
to energy_C00080.

```{r}
Aor_draft$METS[which(Aor_draft$METS$ID == "H(+) (energy metabolism)[c]"),"KEGG"] <- "energy_C00080"
```

# Write output

```{r}
write.xlsx(Aor_draft$METS[,c("NAME","KEGG")], "../data/intermediate/oryzae_kegg_187_50_1385.xlsx")
```

# Appendix

```
list(c("CMP", "C00055"),
    c("palmitate", "C00249"), 
    c("N-acetyl-D-glucosamine", "C00140"),
    c("oxaloacetate", "C00036"),
    c("2-oxobutanoate", "C00109"),
    c("oxalate", "C00209"),
    c("4-aminobenzoate", "C00568"),
    c("phenylacetate", "C07086"),
    c("prephenate", "C00254"),
    c("sulfite", "C00094"),
    c("nitrite", "C00088"),
    c("nitrate", "C00244"),
    c("isocitrate", "C00311"),
    c("L-glutamate", "C00025"),
    c("D-glucuronate", "C00191"),
    c("glyoxylate", "C00048"),
    c("hexanoate", "C01585"),
    c("carnitine", "C00487/C00318/C15025"), # carnitine/L-carnitine/D-carnitine
    c("chorismate", "C00251"),
    c("citrate", "C00158"),
    c("nicotinate", "C00253"),
    c("stearate", "C01530"),
    c("linolenoyl-CoA", "C03035/C16162"),
    c("octadecatrienoyl-CoA", "C03035/C16162"),
    c("D-gluconate", "C00257"),
    c("orotate", "C00295"),
    c("D-ribose 5-phosphate", "C00117"),
    c("glycolate", "C00160"),
    c("succinate", "C00042"),
    c("xanthine", "C00385"),
    c("acetate", "C00033"),
    c("acetoacetate", "C00164"),
    c("acetyl phosphate", "C00227"),
    c("5-amino-1-(5-phospho-D-ribosyl)imidazole", "C03373"),
    c("2-oxoglutarate", "C00026"),
    c("6-phospho-D-gluconate", "C00345"),
    c("D-glucose", "C00031"), # note, alpha-D-glucose is C00267, beta-D-glucose is C00221
    c("3-dehydroshikimate", "C02637"),
    c("D-erythrose 4-phosphate", "C00279"),
    c("episterol", "C15777"),
    c("formate", "C00058"),
    c("D-fructose", "C00095"), # beta-D-fructose is C02336
    c("fumarate", "C00122"),
    c("D-glycerate", "C00258"),
    c("D-galactonate", "C00880"),
    c("D-galacturonate", "C00333"),
    c("D-galactose", "C00124"), # note, alpha-D-galactose is C00962, beta-D-galactose is C00984
    c("D-glucosamine", "C00352"),
    c("anthranilate", "C00108"),
    c("D-arabinose", "C00216"),
    c("L-aspartate", "C00049"),
    c("decanoate", "C01571"),
    c("raffinose", "C00492"),
    c("acyl-carrier protein", "C00229"),
    c("cis-aconitate", "C00417"),
    c("D-erythrose", "C01796"),
    c("UDP-N-acetyl-D-galactosamine", "C00203"),
    c("protein", "C00017"),
    c("pyruvate", "C00022"),
    c("quinate", "C00296"),
    c("D-ribose 1-phosphate", "C00620"),
    c("D-ribose", "C00121"),
    c("sulfur", "C00087"),
    c("sulfate", "C00059"),
    c("shikimate", "C00493"),
    c("L-sorbose", "C00247"),
    c("tartrate", "C00898"), # D-tartrate would be C02107
    c("tetrahydrofolate", "C00101"),
    c("L-arabinose", "C00259"),
    c("L-xylulose", "C00312"),
    c("(S)-malate", "C00149"),
    c("D-mannose", "C00159"),
    c("(R)-mevalonate", "C00418"),
    c("2-formamido-N(1)-(5-phospho-D-ribosyl)acetamidine", "C04640"),
    c("(2-amino-4-hydroxy-7,8-dihydropteridin-6-yl)methyl trihydrogen diphosphate", "C04807"),
    c("	2-amino-6-(hydroxymethyl)-7,8-dihydropteridin-4-ol", "C01300"))
```





