---
title: "Prepare BioLog experiment output for FBA analysis"
output: html_document
date: "2025-01-14"
---

In this script, we prepare the outputs of the BioLog phenotyping experiment
to serve as input for the FBA analyses performed in *simulations.ipynb*. 
In addition, we define a set of core metabolites based on the BioLog 
experimental data. More specifically, we identify for each metabolite how many 
strains have an observed metabolic activity that is (a) 100 or (b) 150 units
higher than the baseline in at least 2 out of 3 replicates. In 
*simulations.ipynb*, we will define a cutoff on these values, and metabolites 
above these cutoffs are the ones that we will focus on there. The goal of that 
script is to gapfill the GEMs of the strains such that the growth predicted 
through FBA aligns with the observations from the BioLog experiment.

# Load libraries

```{r, message=FALSE, warning=FALSE}
library(purrr)
library(dplyr)
library(ComplexHeatmap)
library(circlize)
library(matrixStats)
library(tidyr)
```

# Helpers

```{r}
source("./helperFunctions/Rutilities.R")
```

# Add information to the Biolog results file

## Load Biolog results

```{r}
# Load BioLog assay results (for each plate)
PM_01_wide <- read.csv("../data/biolog/Results_PM01.csv")
PM_02_wide <- read.csv("../data/biolog/Results_PM02.csv")
PM_03_wide <- read.csv("../data/biolog/Results_PM03.csv")
```

## Wrangle data

```{r}
# update column names
colnames(PM_01_wide) <- sapply(colnames(PM_01_wide), update_colnames)
colnames(PM_02_wide) <- sapply(colnames(PM_02_wide), update_colnames)
colnames(PM_03_wide) <- sapply(colnames(PM_03_wide), update_colnames)

PM_01_long <- wrangleResults(PM_01_wide)
PM_02_long <- wrangleResults(PM_02_wide)
PM_03_long <- wrangleResults(PM_03_wide)

PM_01_long$PlateID <- "PM_01"
PM_02_long$PlateID <- "PM_02"
PM_03_long$PlateID <- "PM_03"

PM_all <- rbind(PM_01_long, PM_02_long, PM_03_long)
colnames(PM_all)[c(3,4)] <- c("Medium", "Observed")
```

## Manually obtain KEGG IDs for Biolog metabolites

```{r}
# Link biolog media to oryzae metabolites
media_link = list("2__Deoxy_Adenosine" = "C05198[c]",
                  "L_Tryptophan" = "C00078[e]",
                  "Ammonia" = "C00014[e]",
                  "Capric_Acid" = "C01571[c]", # called decanoate in GEM, same according to kegg
                  "Dulcitol" = "C01697[e]", # called galactitol in GEM, same according to kegg
                  "Glyoxylic_Acid" = "C00048[c]",
                  "L_Glutamic_Acid" = "C00025[e]",
                  "Itaconic_Acid" = "C00490[e]",
                  "Acetoacetic_Acid" = "C00164[c]",
                  "L_Citrulline" = "C00327[c]",
                  "L_Homoserine" = "C00263[c]",
                  "Propionic_Acid" = "C00163[e]",
                  "L_Ornithine" = "C00077[c]",
                  "Fumaric_Acid" = "C00122[e]", 
                  "D_Mannitol" = "C00392[c]",
                  "Malonic_Acid" = "C00383[c]",
                  "D_Tagatose" = "C00795[e]",
                  "D_Raffinose" = "C00492[e]", # raffinose in GEM, not D-
                  "D_Alanine" = "C00133[c]",
                  "Histamine" = "C00388[c]",
                  "Stachyose" = "C01613[c]",
                  "Allantoin" = "C01551[c]",
                  "Xanthosine" = "C01762[c]",
                  "Xanthine" = "C00385[c]",
                  "Thymine" = "C00178[c]",
                  "Guanosine" = "C00387[c]",
                  "Cytidine" = "C00475[c]",
                  "Formamide" = "C00488[c]",
                  "Methylamine" = "C00218[c]",
                  "Biuret" = "C06555[c]",
                  "Amygdalin" = "C08325[c]",
                  "Ethanolamine" = "C00189[c]",
                  "Tyramine" = "C00483[c]",
                  "Inosine" = "C00294[c]", 
                  "Adenosine" = "C00212[c]",
                  "Maltotriose" = "C01835[c]",
                  "Uridine" = "C00299[c]",
                  "Thymidine" = "C00214[c]",
                  "D_Glucose" = "C00031[e]",
                  "D_Galactose" = "C00124[e]",
                  "D_Mannose" = "C00159[e]",
                  "L_Sorbose" = "C00247[e]",
                  "D_Arabinose" = "C00216[e]",
                  "L_Arabinose" = "C00259[e]",
                  "D_Fructose" = "C00095[e]",
                  "D_Ribose" = "C00121[e]",
                  "D_Xylose" = "C00181[e]",
                  "D_Maltose" = "C00208[e]",  # in GEM: Maltose
                  "Pectin" = "C00470[e]", # in GEM: Pectate
                  "D_Cellobiose" = "C00185[e]", # in GEM: Cellobiose
                  "Xylitol" = "C00379[e]",
                  "Citric_Acid" = "C00158[e]", # in GEM: Citrate
                  "Acetic_Acid" = "C00033[e]", # in GEM: Acetate
                  "L_Alanine" = "C00041[e]",
                  "L_Arginine" = "C00062[e]",
                  "L_Asparagine" = "C00152[e]",
                  "L_Cysteine" = "C00097[e]",
                  "L_Glutamine" = "C00064[e]",
                  "L_Histidine" = "C00135[e]",
                  "L_Isoleucine" = "C00407[e]",
                  "L_Leucine" = "C00123[e]",
                  "L_Lysine" = "C00047[e]",
                  "L_Methionine" = "C00073[e]",
                  "L_Phenylalanine" = "C00079[e]",
                  "L_Proline" = "C00148[e]",
                  "L_Serine" = "C00065[e]",
                  "L_Threonine" = "C00188[e]",
                  "L_Tyrosine" = "C00082[e]",
                  "L_Valine" = "C00183[e]",
                  "D_Trehalose" = "C01083[e]", # in GEM: alpha,alpha-Trehalose
                  "D_Sorbitol" = "C00794[e]",
                  "Glycogen" = "C00182[e]",
                  "Oxalic_Acid" = "C00209[e]", # in GEM: oxalate
                  "Adenine" = "C00147[e]",
                  "Cytosine" = "C00380[e]",
                  "Guanine" = "C00242[e]",
                  "Uracil" = "C00106[e]",
                  "Glycerol" = "C00116[e]",
                  "Glycine" = "C00037[e]",
                  "Putrescine" = "C00134[e]",
                  "Pyruvic_Acid" = "C00022[e]",
                  "Mannan" = "C00464[e]", # in GEM: Mannans
                  "Sucrose" = "C00089[e]",
                  "D_Melibiose" = "C05402[e]", # in GEM: melibiose
                  "Succinic_Acid" = "C00042[e]",
                  "Urea" = "C00086[e]",
                  "Agmatine" = "C00179[e]",
                  "Glycolic_Acid" = "C00160[c]",
                  "D_Glucosamine" = "C00329[e]",
                  "Dihydroxy_Acetone" = "C00184[c]")

PM_all$Metabolite <- unname(unlist(media_link))[match(PM_all$Medium,names(media_link))]
```

## Export data

```{r}
write.csv2(PM_all, "../data/biolog/PM_all.csv", row.names = FALSE)
```

# Define a core set of metabolites

For each metabolite, identify how many isolates have an observed metabolic 
activity that is (a) 100 or (b) 150 units higher than the baseline in at least
2 replicates. Then rank based on those numbers. Then walk down that ranked list.
Currently, I identify a core metabolite as a metabolite where there are at least
two isolates that have two replicates with a metabolic activity that is at least 
150 units higher than their baselines, or where where there are at least
four isolates that have two replicates with a metabolic activity that is at least 
100 units higher than their baselines

## Read and wrangle Biolog raw data

```{r}
csv_Files <- list.files(path = "../data/biolog/Raw", 
                        pattern = "\\.csv$",
                        full.names = TRUE)

raw_data <- lapply(csv_Files, function(x){
    data = read.csv(x)
    tail(data, n=1) # select the last time point
})
raw_data <- do.call(rbind, raw_data)

# Remove file location
raw_data$Plate.File <- NULL

# Update colname
colnames(raw_data)[which(colnames(raw_data) == "Plate.Type")] <- "Plate"

# Clean isolate names
raw_data$Field.2 <- gsub(" ", "_", raw_data$Field.2)

# Establish a baseline
raw_data$baseline <- NA
cols_to_consider <- grep("^([A-H][0-9]{2})$", names(raw_data), value = TRUE)
for (i in 1:nrow(raw_data)) {
    raw_data$baseline[i] <- mean(sort(unlist(raw_data[i,cols_to_consider]))[1:10])
}
```

## Assign wells to compounds

```{r}
wellDf1 <- data.frame(Medium = unname(sapply(colnames(PM_01_wide), update_colnames)),
                      Plate = "PM01",
                      Well = substr(colnames(PM_01_wide),1,3))
wellDf2 <- data.frame(Medium = unname(sapply(colnames(PM_02_wide), update_colnames)),
                      Plate = "PM02A",
                      Well = substr(colnames(PM_02_wide),1,3))
wellDf3 <- data.frame(Medium = unname(sapply(colnames(PM_03_wide), update_colnames)),
                      Plate = "PM03B",
                      Well = substr(colnames(PM_03_wide),1,3))
wellDfall <- rbind(wellDf1,wellDf2,wellDf3)
wellDfall <- wellDfall[-which(wellDfall$Well == "X"),]
```

## Create a dataframe ranking metabolites according to metabolic activity

```{r}
metaboliteDf <- data.frame(Plate = rep(c("PM01", "PM02A", "PM03B"), each = 96),
                           Well = rep(grep("^([A-H][0-9]{2})$", names(raw_data), value = TRUE),3),
                           Medium = wellDfall$Medium)
metaboliteDf$Condition <- paste0(metaboliteDf$Medium, "_", metaboliteDf$Plate)
metaboliteDf$basePlus150 <- metaboliteDf$basePlus100 <- 0

for (i in seq(nrow(metaboliteDf))) {
    
    data_i <- data.frame(isolates = raw_data$Field.2[which(raw_data$Plate == metaboliteDf$Plate[i])],
                         baselineSignal = raw_data$baseline[which(raw_data$Plate == metaboliteDf$Plate[i])],
                         metaboliteSignal = raw_data[which(raw_data$Plate == metaboliteDf$Plate[i]),
                                                      which(colnames(raw_data) == metaboliteDf$Well[i])])
    data_i$normalizedSignal <- data_i$metaboliteSignal - data_i$baselineSignal
    
    for (isolate in unique(data_i$isolates)) {
        index <- which(metaboliteDf$Plate == metaboliteDf$Plate[i] & metaboliteDf$Well == metaboliteDf$Well[i])
        if(sum(data_i$normalizedSignal[which(data_i$isolates == isolate)] > 150) >= 2){
          metaboliteDf$basePlus150[index] <- metaboliteDf$basePlus150[index] + 1
          metaboliteDf$basePlus100[index] <- metaboliteDf$basePlus100[index] + 1
        } else if(sum(data_i$normalizedSignal[which(data_i$isolates == isolate)] > 100) >= 2) {
          metaboliteDf$basePlus100[index] <- metaboliteDf$basePlus100[index] + 1
        }
    }
}
head(metaboliteDf)
```

```{r}
KEGG <- unname(media_link)[match(metaboliteDf$Medium, names(media_link))]
KEGG[sapply(KEGG, is.null)] <- NA
metaboliteDf$KEGG <- unlist(KEGG)
```

```{r}
write.csv2(metaboliteDf, "../data/biolog/metaboliteDf.csv", row.names = FALSE)
```



